### 2022-04-09 manuscript/manuscript.Rmd
Shell of the manuscript

---
title: Estimating Uncertainty for Tipping Points with Deep Learning
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  Lorem Ipsum whatever

keywords:
  - Time Series
  - Forecasting
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |

header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
#  - \usepackage{lineno}
#  - \usepackage{endfloat}
#  - \linenumbers

output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---
# Introduction

# Materials and Methods

# Results

# Discussion

\pagebreak

# References

### 2022-04-11 darts/train_ensemble.py
More up to date now; didn't add bash scripts to avoid cluttering

@@ -1,0 +1,140 @@
+import pandas as pd
+import numpy as np
+import os
+import sys
+sys.path.append("../")
+import matplotlib
+matplotlib.use("Agg")
+import matplotlib.pyplot as plt
+from darts import TimeSeries
+import torch
+from darts.models import RNNModel, BlockRNNModel
+from darts.utils.likelihood_models import LaplaceLikelihood
+from utils import preprocessed_t_series, truth_dist, import_hypers
+import argparse
+import yaml
+from functools import reduce
+
+parser = argparse.ArgumentParser()
+parser.add_argument(
+    "-m",
+    "--model",
+    default="lstm",
+    type=str,
+    help="model to train with (lstm or block_rnn)",
+)
+parser.add_argument(
+    "-s",
+    "--n_samples",
+    default=1000,
+    type=int,
+    help="# of samples to train on",
+)
+parser.add_argument(
+    "-o",
+    "--output_file_name",
+    default="trash",
+    type=str,
+    help="File name of plots",
+)
+parser.add_argument(
+    "-p",
+    "--plot",
+    action="store_true",
+    help="Flag whether to plot",
+)
+parser.add_argument(
+    "--seed",
+    default=42,
+    type=int,
+    help="Seed selection",
+)
+parser.add_argument(
+    "-t",
+    "--tp_model",
+    default="stochastic",
+    type=str,
+    help="Select which model to use for the tipping point (stochastic/saddle)",
+)
+parser.add_argument(
+    "--hyper_file",
+    default="lstm",
+    help="Select which hyperparameter file to load hypers from",
+)
+parser.add_argument(
+    "-c",
+    "--case",
+    default="none",
+    type=str,
+    help="Special cases to consider (none, tipped, non_tipped, both)",
+)
+args = parser.parse_args()
+
+
+if args.hyper_file == "stochastic_lstm":
+    from train_hyperparams.stochastic_lstm import hyperparameters
+elif args.hyper_file == "saddle_lstm":
+    from train_hyperparams.saddle_lstm import hyperparameters
+elif args.hyper_file == "stochastic_block":
+    from train_hyperparams.stochastic_block import hyperparameters
+elif args.hyper_file == "saddle_block":
+    from train_hyperparams.saddle_block import hyperparameters
+
+model = {"lstm" : RNNModel, "block_rnn" : BlockRNNModel}[args.model.lower()]
+
+models = []
+for i in range(42, 62):
+    hyperparameters["random_state"] = i
+    np.random.seed(i)
+    
[…]

### 2022-04-11 darts/utils.py

@@ -10,14 +10,9 @@ models = {"stochastic" : stochastic_tp,
           }
 
 # Need to change this so that it doesn't intake args
-def preprocessed_t_series(model, n_samples, random_alpha=False):
+def preprocessed_t_series(model, n_samples):
     _data = models[model.lower()]()
-    
-    # Generating time series
-    if random_alpha:
-        training_data = _data.collect_samples(n_samples, random_alpha)
-    else:
-        training_data = _data.collect_samples(n_samples)
+    training_data = _data.collect_samples(n_samples)
 
     # Preprocessing time series to
     _ts = [[] for i in range(training_data.shape[1])]
@@ -28,7 +23,7 @@ def preprocessed_t_series(model, n_samples, random_alpha=False):
     vals = np.array(_ts).reshape(training_data.shape[1], 1, n_samples)
     return TimeSeries.from_times_and_values(RangeIndex(250),vals), vals.reshape(250, n_samples).T
 
-def truth_dist(model, t_series, input_len, output_len, n_samples=100, random_alpha=False):
+def truth_dist(model, t_series, input_len, output_len, n_samples=100):
     N_init = t_series[-1].values()[0][0]
     t_max = 250-input_len
         
@@ -41,10 +36,7 @@ def truth_dist(model, t_series, input_len, output_len, n_samples=100, random_alp
         _data.h_init = _data.h_init + len(t_series) * _data.alpha
 
     for i in range(n_samples):
-      if random_alpha:
-          training_data = _data.collect_samples(n_samples, random_alpha)
-      else:
-          training_data = _data.collect_samples(n_samples)
+      training_data = _data.collect_samples(n_samples)
       
       # Preprocessing time series to
       _ts = [[] for i in range(training_data.shape[1])]
@@ -55,7 +47,7 @@ def truth_dist(model, t_series, input_len, output_len, n_samples=100, random_alp
       vals = np.array(_ts).reshape(training_data.shape[1], 1, n_samples)
     return TimeSeries.from_times_and_values(RangeIndex(start=input_len, stop=250), vals)
   
-def count_tipped(vals, model):
+def count_tipped(vals):
     n_vals = len(vals)
     count = 0
     for i in range(n_vals):
@@ -63,4 +55,22 @@ def count_tipped(vals, model):
             count += 1
     return count / n_vals
 
-#
+def import_hypers(hyper_file_name):
+    if hyper_file_name == "lstm":
+        from train_hyperparams.lstm import hyperparameters
[…]

### 2022-04-11 darts/train_ensemble.py
Model training stuff almost a wrap

@@ -7,10 +7,11 @@ import matplotlib
 matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from darts import TimeSeries
+from pandas import DataFrame
 import torch
 from darts.models import RNNModel, BlockRNNModel
 from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist, import_hypers
+from utils import preprocessed_t_series, truth_dist, import_hypers, make_df
 import argparse
 import yaml
 from functools import reduce
@@ -38,10 +39,10 @@ parser.add_argument(
     help="File name of plots",
 )
 parser.add_argument(
-    "-p",
-    "--plot",
+    "-e",
+    "--evaluate",
     action="store_true",
-    help="Flag whether to plot",
+    help="Flag whether to plot and save a csv of samples",
 )
 parser.add_argument(
     "--seed",
@@ -83,7 +84,7 @@ elif args.hyper_file == "saddle_block":
 model = {"lstm" : RNNModel, "block_rnn" : BlockRNNModel}[args.model.lower()]
 
 models = []
-for i in range(42, 62):
+for i in range(42, 47):
     hyperparameters["random_state"] = i
     np.random.seed(i)
     
@@ -95,10 +96,11 @@ for i in range(42, 62):
     elif args.case.lower() == "both" and args.tp_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
         train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-        train_series = train_series.stack(train_series_)
+        train_series = train_series.concatenate(train_series_, axis="sample")
     elif args.case.lower() == "none":
         # Generating time series
         train_series, _ = preprocessed_t_series(args.tp_model, args.n_samples)
+        
 
     my_model = model(
         likelihood=LaplaceLikelihood(),
@@ -114,13 +116,14 @@ for i in range(42, 62):
     del my_model
 
 # Generating time series
-if args.plot:
+if args.evaluate:
     if not os.path.exists("plots/"):
         os.makedirs("plots/")
     input_len = hyperparameters["input_chunk_length"]
     output_len = hyperparameters["output_chunk_length"]
+    final_df = DataFrame()
   
-    for i in range(3):
+    for i in range(5):
         train_series, _ = preprocessed_t_series(args.tp_model, 1)
[…]

### 2022-04-11 darts/utils.py

@@ -3,7 +3,7 @@ sys.path.append("../")
 from model import stochastic_tp, saddle_node_tp
 import numpy as np
 from darts import TimeSeries
-from pandas import RangeIndex
+from pandas import RangeIndex, DataFrame
 
 models = {"stochastic" : stochastic_tp, 
           "saddle" : saddle_node_tp,
@@ -74,3 +74,34 @@ def import_hypers(hyper_file_name):
         from train_hyperparams.saddle_100 import hyperparameters
     elif hyper_file_name == "saddle_1000":
         from train_hyperparams.saddle_1000 import hyperparameters
+
+def make_df(prediction_series, truth_series, tp_model, ml_model, case, n_samples):
+    predictions_array = prediction_series.all_values()[:, 0, :]
+    truth_array = truth_series.all_values()[:, 0, :]
+    
+    predictions_df = DataFrame(predictions_array)
+    truth_df = DataFrame(truth_array)
+    
+    predictions_df["time"] = prediction_series.time_index
+    truth_df["time"] = truth_series.time_index
+    
+    predictions_df = predictions_df.melt(id_vars="time", var_name="ensemble", value_name="value")
+    truth_df = truth_df.melt(id_vars="time", var_name="ensemble", value_name="value")
+    
+    
+    predictions_df["tp_model"] = tp_model
+    predictions_df["ml_model"] = ml_model
+    predictions_df["case"] = n_samples if case == "none" else case
+    predictions_df["true_model"] = False
+    truth_df["tp_model"] = tp_model
+    truth_df["ml_model"] = ml_model
+    truth_df["case"] = n_samples if case == "none" else case
+    truth_df["true_model"] = True
+    df = predictions_df.append(truth_df, ignore_index=True)
+    
+    
+    return df
+    
+  
+    
+    

### 2022-04-12 darts/train_ensemble.py
Keeping up to date

@@ -91,7 +91,7 @@ for i in range(42, 47):
     # Selecting case to pick training set
     if args.case.lower() == "tipped" and args.tp_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-    elif args.case.lower() == "non_tipped" and args.tp_model == "stochastic":
+    elif args.case.lower() == "nontipped" and args.tp_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
     elif args.case.lower() == "both" and args.tp_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
@@ -119,6 +119,10 @@ for i in range(42, 47):
 if args.evaluate:
     if not os.path.exists("plots/"):
         os.makedirs("plots/")
+    if not os.path.exists(f"plots/ensemble_{args.output_file_name}/"):
+        os.makedirs(f"plots/ensemble_{args.output_file_name}/")
+    if not os.path.exists("forecasts/"):
+            os.makedirs("forecasts/")
     input_len = hyperparameters["input_chunk_length"]
     output_len = hyperparameters["output_chunk_length"]
     final_df = DataFrame()
@@ -141,7 +145,7 @@ if args.evaluate:
         df = make_df(ensemble_series, t_dist, args.tp_model, args.model.lower(), args.case, args.n_samples)
         df["iter"] = i
         final_df = final_df.append(df, ignore_index=True)
-        
-        plt.savefig(f"plots/ensemble_{args.output_file_name}_{i}")
+        plt.savefig(f"plots/ensemble_{args.output_file_name}/{i}")
         plt.clf()
-    final_df.to_csv(f"{args.output_file_name}.csv.gz")
+        
+    final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz")

### 2022-04-12 greta/mcmc.R
experiment

@@ -1,0 +1,59 @@
+library(tidyverse)
+set.seed(1234567)
+
+np.clip <- function(x, a, b) {
+  if(x < a) return(a)
+  if(x > b) return(b)
+  x
+}
+
+# View the bistable curve: 
+# curve( (0.75 - x)**2 * (0.25 - x), 0, 1)
+
+step <- function(N, eta,  a = 0.75, b = 0.25) {
+  N <- N + (a - N)**2 * (b - N) + eta
+  N <- np.clip(N, 0, 1)
+  N
+}
+
+# simulate
+simulate <- function(N_init = 0.75,
+                     t_max = 250L,
+                     mu =0, # -5e-4,
+                     sigma = 1e-2,
+                     a = 0.75, 
+                     b = 0.25) {
+  eta <- rnorm(t_max, mu, sigma)
+  N   <- numeric(t_max)
+  N[1] <- N_init
+  for (t in 1:(t_max-1)) {
+    N[t+1] <- step(N[t], eta[t], a = a, b = b)
+  }
+  tibble::tibble(t = 1:t_max, N = N)
+}
+
+sims <- purrr::map_dfr(1:100, \(i) simulate(), .id = "i")
+sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
+
+sims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) | filter(t<max(t))
+
+x_t <- sims$N
+x_t1 <- sims$xt1
+
+library(greta)
+a <- uniform(0, 2)
+b <- uniform(0, 2)
+sigma <- uniform(0, 1)
+
+# Model   (mean <-  may(x_t, p))
+mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
+distribution(x_t1) <- normal(mean, sigma )
+
+m <- model(a, b, sigma)
+
+mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))
+
+bench::bench_time({                 
+  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
+})
+

### 2022-04-12 greta/mcmc.R
typo

@@ -35,7 +35,7 @@ simulate <- function(N_init = 0.75,
 sims <- purrr::map_dfr(1:100, \(i) simulate(), .id = "i")
 sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
 
-sims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) | filter(t<max(t))
+sims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
 
 x_t <- sims$N
 x_t1 <- sims$xt1

### 2022-04-13 greta/mcmc.Rmd
mcmc comparison

---
output: github_document
---

```{r}
library(tidyverse)
set.seed(1234567)

np.clip <- function(x, a, b) {
  if(x < a) return(a)
  if(x > b) return(b)
  x
}

# View the bistable curve: 
# curve( (0.75 - x)**2 * (0.25 - x), 0, 1)

step <- function(N, eta,  a = 0.75, b = 0.25) {
  N <- N + (a - N)**2 * (b - N) + eta
  N <- np.clip(N, 0, 1)
  N
}

# simulate
simulate <- function(N_init = 0.75,
                     t_max = 250L,
                     mu =0, # -5e-4,
                     sigma = 1e-2,
                     a = 0.75, 
                     b = 0.25) {
  eta <- rnorm(t_max, mu, sigma)
  N   <- numeric(t_max)
  N[1] <- N_init
  for (t in 1:(t_max-1)) {
    N[t+1] <- step(N[t], eta[t], a = a, b = b)
  }
  tibble::tibble(t = 1:t_max, N = N)
}

sims <- purrr::map_dfr(1:100, \(i) simulate(), .id = "i")
sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
```

```{r}
gsims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))

x_t <- gsims$N
x_t1 <- gsims$xt1

library(greta)
a <- uniform(0, 10)
b <- uniform(0, 10)
sigma <- uniform(0, 10)
mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
distribution(x_t1) <- normal(mean, sigma )
m <- model(a, b, sigma)
```

```{r}
mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))

bench::bench_time({                 
  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
})
```

```{r}
bayesplot::mcmc_trace(draws)
```

```{r}
posterior_samples <- 
  bind_rows(map(draws, as_tibble)) %>% 
  sample_n(100)
posterior_sims <- posterior_samples %>%
  purrr::transpose() %>%
  map_dfr(function(q) simulate() ,.id = "i")

bind_rows(sims, posterior_sims, .id = "model")|> 
  ggplot(aes(t, N, col=model, group=interaction(model,i))) +
           geom_line(alpha=0.3)

```

# Scoring

```{r}

library(scoringRules)

observed <- sims %>% filter(i == 1) %>% pull(N)
posterior_sims
dat <- posterior_sims |> pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> select(-t) |> as.matrix()

score <- scoringRules::logs_sample(observed, dat)
logsscore <- score[-1] # drop first point
mean(logsscore)

score <- scoringRules::crps_sample(observed, dat)
crpsscore <- score[-1] # drop first point
mean(crpsscore)

```

### 2022-04-13 greta/mcmc.Rmd
10 replicates

@@
 
-sims <- purrr::map_dfr(1:100, \(i) simulate(), .id = "i")
+sims <- purrr::map_dfr(1:10, \(i) simulate(), .id = "i")
 sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
@@
 
+First, a single 'observed' sample from the true:
+
 ```{r}
-
 library(scoringRules)
 
-observed <- sims %>% filter(i == 1) %>% pull(N)
-posterior_sims
+scores <- function(observed, dat) {
+  logsscore <- scoringRules::logs_sample(observed, dat)
+  crpsscore <- scoringRules::crps_sample(observed, dat)
+  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
+
+}
+# ensemble predictions
 dat <- posterior_sims |> pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> select(-t) |> as.matrix()
 
-score <- scoringRules::logs_sample(observed, dat)
-logsscore <- score[-1] # drop first point
-mean(logsscore)
-
-score <- scoringRules::crps_sample(observed, dat)
-crpsscore <- score[-1] # drop first point
-mean(crpsscore)
+# Score just the 1st replicate "true" observations
+observed <- sims |> filter(i == 1) |> pull(N)
+scores(observed, dat)
 
 ```
+```{r}
+# score over all replicates:
+
+rep_scores <- sims |> group_by(i) |> group_map(~ scores(.x$N, dat)) |> bind_rows()
+```
+
+```{r}
+rep_scores |> summarise(across(.fns= base::mean))
+
+library(patchwork)
+ggplot(rep_scores) + geom_histogram(aes(crps)) + 
+(ggplot(rep_scores) + geom_histogram(aes(logs)) )
+
+```

### 2022-04-13 greta/mcmc.Rmd
1-rep, with transistion
100 reps (again)
different test/train sets!

@@
 library(tidyverse)
-set.seed(1234567)
+library(greta)
+library(bayesplot)
+
+set.seed(4242)
+train_reps <- 10
+train_t_max <- 250
+test_t_max <- 250
+test_reps <- 100
 
@@
 }
+```
 
+```{r}
 # View the bistable curve: 
@@
 
-sims <- purrr::map_dfr(1:10, \(i) simulate(), .id = "i")
-sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
+train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
+test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
+
+## show training data
+train |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.2)
 ```
@@
 ```{r}
-gsims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
+gsims <- train |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
 
@@
   purrr::transpose() %>%
-  map_dfr(function(q) simulate() ,.id = "i")
+  map_dfr(function(q) simulate(t_max = test_t_max) ,.id = "i")
 
-bind_rows(sims, posterior_sims, .id = "model")|> 
+bind_rows(test, posterior_sims, .id = "model")|> 
   ggplot(aes(t, N, col=model, group=interaction(model,i))) +
@@
 
+```
+
+```{r}
+true <- tibble(a = 0.75, b = 0.25, sigma = 1e-2) |> gather(variable, value)
+
+posterior_samples  |> gather(variable, value) |> ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
 ```
@@
 # Score just the 1st replicate "true" observations
-observed <- sims |> filter(i == 1) |> pull(N)
+observed <- test |> filter(i == 1) |> pull(N)
 scores(observed, dat)
@@
 
-rep_scores <- sims |> group_by(i) |> group_map(~ scores(.x$N, dat)) |> bind_rows()
+rep_scores <- test |> group_by(i) |> group_map(~ scores(.x$N, dat)) |> bind_rows()
 ```

### 2022-04-14 greta/mcmc.Rmd
scoring 1000 reps

@@
 test_t_max <- 250
-test_reps <- 100
+test_reps <- 1000
 
@@
 ```{r}
+## draw test_reps number of samples
+
 posterior_samples <- 
   bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(100)
+  sample_n(test_reps)
 posterior_sims <- posterior_samples %>%
@@
   ggplot(aes(t, N, col=model, group=interaction(model,i))) +
-           geom_line(alpha=0.3)
+           geom_line(alpha=0.1)
 
@@
 # ensemble predictions
-dat <- posterior_sims |> pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> select(-t) |> as.matrix()
-
-# Score just the 1st replicate "true" observations
-observed <- test |> filter(i == 1) |> pull(N)
-scores(observed, dat)
+dat <- 
+  posterior_sims |> 
+  pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
+  select(-t) |> as.matrix()
 
@@
 # score over all replicates:
-
-rep_scores <- test |> group_by(i) |> group_map(~ scores(.x$N, dat)) |> bind_rows()
+bench::bench_time({
+rep_scores <- 
+  test |> 
+  group_by(i) |> 
+  group_map(~ scores(.x$N, dat)) |> 
+  bind_rows()
+})
 ```

### 2022-04-14 darts/scoring.Rmd
Having problems with this

---
title: "scoring"
output: html_document
date: '2022-04-13'
---

```{r}
data <- read_csv("forecasts/block_rnn_stochastic_10.csv.gz")[-1]
emp_draws <- data %>% filter(true_model==TRUE) %>% select(-true_model)
forecasts <- data %>% filter(true_model==FALSE) %>% select(-true_model)
```

```{r}
library(scoringRules)

scores <- function(observed, dat) {
  logsscore <- scoringRules::logs_sample(observed, dat)
  crpsscore <- scoringRules::crps_sample(observed, dat)
  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
}

# Score just the 1st replicate "true" observations
observed <- emp_draws %>%  filter(iter == 0, ensemble==0) %>%  pull(value)
dat <- forecasts %>%  filter(iter == 0, ensemble==0) %>%  select(ensemble, value) %>% as.matrix()
scores(observed, dat)

```
```{r}
# score over all replicates:
all_scores = list()
for (i in (0:4)) {
  for (e in 0:99){
    dat <- forecasts %>%  filter(iter == i, ensemble == e) %>%  select(ensemble, value) %>% as.matrix()
    rep_scores <- emp_draws %>% filter(iter==i) %>%  group_by(ensemble) %>%  group_map(~ scores(.x$value, dat)) %>%  bind_rows()
    all_scores <- append(all_scores, rep_scores)
  }
  print(i)
}

all_scores %>%  summarise(across(.fns= base::mean))

library(patchwork)
ggplot(all_scores) + geom_histogram(aes(crps)) + 
(ggplot(all_scores) + geom_histogram(aes(logs)) )

```

### 2022-04-14 darts/scoring.Rmd
quick revisions for fast scoring

@@
 title: "scoring"
-output: html_document
+output: github_document
 date: '2022-04-13'
 ---
+
+```{r}
+library(scoringRules)
+library(tidyverse)
+```
 
@@
 }
+```
 
-# Score just the 1st replicate "true" observations
-observed <- emp_draws %>%  filter(iter == 0, ensemble==0) %>%  pull(value)
-dat <- forecasts %>%  filter(iter == 0, ensemble==0) %>%  select(ensemble, value) %>% as.matrix()
-scores(observed, dat)
+```{r}
 
+## Currently it looks like only one tp_model and one ml_model, but 5 iterations:
+## We will need to generalize this code to map over multiple tp_models and ml_models etc
+# forecasts %>% count(tp_model, ml_model, iter)
+
+## pivot forecast into a matrix of size(time) x size(ensembles)
+i <- 1
+dat <- forecasts |> filter(iter == i) |>
+  select(t = time, i = ensemble, N = value) |> 
+ pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
+  select(-t) |> as.matrix()
+
+test <- emp_draws |> filter(iter == i)  |>
+  select(t = time, i = ensemble, N = value)
 ```
+
 ```{r}
-# score over all replicates:
-all_scores = list()
-for (i in (0:4)) {
-  for (e in 0:99){
-    dat <- forecasts %>%  filter(iter == i, ensemble == e) %>%  select(ensemble, value) %>% as.matrix()
-    rep_scores <- emp_draws %>% filter(iter==i) %>%  group_by(ensemble) %>%  group_map(~ scores(.x$value, dat)) %>%  bind_rows()
-    all_scores <- append(all_scores, rep_scores)
-  }
-  print(i)
-}
+bench::bench_time({
+rep_scores <- 
+  test |> 
+  group_by(i) |> 
+  group_map(~ scores(.x$N, dat)) |> 
+  bind_rows()
+})
+```
 
-all_scores %>%  summarise(across(.fns= base::mean))
-
+```{r}
 library(patchwork)

### 2022-04-14 darts/scoring.Rmd
group over models and inter

@@
 data <- read_csv("forecasts/block_rnn_stochastic_10.csv.gz")[-1]
-emp_draws <- data %>% filter(true_model==TRUE) %>% select(-true_model)
-forecasts <- data %>% filter(true_model==FALSE) %>% select(-true_model)
 ```
@@
 
-## Currently it looks like only one tp_model and one ml_model, but 5 iterations:
-## We will need to generalize this code to map over multiple tp_models and ml_models etc
-# forecasts %>% count(tp_model, ml_model, iter)
+score_all <- function(data, groups = NULL) {
 
-## pivot forecast into a matrix of size(time) x size(ensembles)
-i <- 1
-dat <- forecasts |> filter(iter == i) |>
+true <- data |> filter(true_model)
+fcst <- data |> filter(!true_model)
+
+dat <- fcst |>
   select(t = time, i = ensemble, N = value) |> 
@@
 
-test <- emp_draws |> filter(iter == i)  |>
+test <- true  |>
   select(t = time, i = ensemble, N = value)
-```
 
-```{r}
-bench::bench_time({
 rep_scores <- 
@@
   bind_rows()
-})
+
+bind_cols(rep_scores, groups)
+}
+
+rep_scores <- data |> 
+  group_by(iter, ml_model, tp_model) |>
+  group_map(~ score_all(.x, .y) ,.keep = TRUE) |> 
+  bind_rows()
 ```
@@
 ```{r}
-library(patchwork)
-ggplot(all_scores) + geom_histogram(aes(crps)) + 
-(ggplot(all_scores) + geom_histogram(aes(logs)) )
+# crude demo
+ggplot(rep_scores) + geom_density(aes(crps, color=iter, group=iter)) + facet_grid(ml_model~tp_model)
 

### 2022-04-14 darts/scoring.Rmd
add some comments

@@
 
-scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
+# trivial wrapper to score both logs and crps
+scores <- function(observed, predicted) {
+  logsscore <- scoringRules::logs_sample(observed, predicted)
+  crpsscore <- scoringRules::crps_sample(observed, predicted)
+  # drops the first point (initial condition)
   data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
@@
 
-dat <- fcst |>
-  select(t = time, i = ensemble, N = value) |> 
- pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
-  select(-t) |> as.matrix()
+# key trick is to make forecast a t_max x ensembles matrix for scoringRules:
+predicted <- fcst |>
+  select(t = time, i = ensemble, value) |> 
+  pivot_wider(id_cols = "t", names_from="i", values_from = "value") |> 
+  select(-t) |> 
+  as.matrix()
 
-test <- true  |>
-  select(t = time, i = ensemble, N = value)
+# just select and rename relevant columns
+test <- true |>  select(t = time, i = ensemble, value)
 
+# Here we go! score each test timeseries against the entire predicted ensemble!
 rep_scores <- 
@@
   group_by(i) |> 
-  group_map(~ scores(.x$N, dat)) |> 
+  group_map(~ scores(.x$value, predicted)) |> 
   bind_rows()
 
+# add any "grouping variables" on as additional columns for bookkeeping
 bind_cols(rep_scores, groups)
 }
+```
 
+```{r}
+# now actually run the scoring over all iterations, ml_models, tp_models:
+
+bench::bench_time({
 rep_scores <- data |> 
@@
   bind_rows()
+
+})
 ```
@@
 ```{r}
-# crude demo
-ggplot(rep_scores) + geom_density(aes(crps, color=iter, group=iter)) + facet_grid(ml_model~tp_model)
+
+# rough plot, probably not the best visual of this.
+ggplot(rep_scores) + 
+  geom_density(aes(crps, color=iter, group=iter)) + 
+  facet_grid(ml_model~tp_model)
 

### 2022-04-15 darts/scoring.Rmd
Box plot is up

@@
 ```{r}
-data <- read_csv("forecasts/block_rnn_stochastic_10.csv.gz")[-1]
+data <- list.files(path = "./forecasts/",
+               pattern = "*.csv", 
+               full.names = T) %>% 
+        read_csv() 
+data <- data[-1]
 ```
@@
 predicted <- fcst |>
-  select(t = time, i = ensemble, value) |> 
+  select(t = time, i = ensemble, value) |> # Not sure if this should be iter here hmm but this might be addressed later
   pivot_wider(id_cols = "t", names_from="i", values_from = "value") |> 
@@
   test |> 
-  group_by(i) |> 
+  group_by(i) |> # group by i? 
   group_map(~ scores(.x$value, predicted)) |> 
@@
 rep_scores <- data |> 
-  group_by(iter, ml_model, tp_model) |>
+  group_by(iter, ml_model, tp_model, case) |>
   group_map(~ score_all(.x, .y) ,.keep = TRUE) |> 
@@
 ```{r}
+# Logs Plot
+rep_scores %>% 
+  filter(case == 10 |  case == 100 | case == 1000) %>% 
+  #filter(tp_model == "stochastic") %>% 
+  ggplot(aes(ml_model, logs)) + 
+  geom_boxplot(outlier.shape = 1) +
+  #geom_jitter() + 
+  facet_grid(tp_model~case, scales="free")
+```
 
-# rough plot, probably not the best visual of this.
-ggplot(rep_scores) + 
-  geom_density(aes(crps, color=iter, group=iter)) + 
-  facet_grid(ml_model~tp_model)
-
+```{r}
+# CRPS Plot
+rep_scores %>% 
+  filter(case == 10 |  case == 100 | case == 1000) %>% 
+  #filter(tp_model == "stochastic") %>% 
+  ggplot(aes(ml_model, crps)) + 
+  geom_boxplot(outlier.shape = 1) +
+  #geom_jitter() + 
+  facet_grid(tp_model~case, scales="free")
 ```

### 2022-04-16 darts/scoring.Rmd
Spent a little too much time on this :grimacing:

@@
 ```
+
+Trying to get single curves set up here
+```{r}
+plot_single <- function(data, case_, iter_) {
+    # I checked that both "Truth" cases are the same between models but potential error source here -- should rethink csv saving
+    t_data <- data %>% 
+                filter(case == case_ & iter == iter_) %>% 
+                transform(ml_model = ifelse(true_model==TRUE, "Truth", ml_model)) %>% 
+                transform(true_model = ifelse(ml_model == "Truth", FALSE, true_model))
+
+    t_data %>% 
+        filter(case == case_ & iter == iter_ & true_model == FALSE) %>%
+        group_by(time, true_model, ml_model) %>% 
+        mutate(`5%` = quantile(value, c(.05))) %>% 
+        mutate(`95%` = quantile(value, c(.95))) %>% 
+        ggplot() +
+        geom_ribbon(aes(x=time, ymin=`5%`, ymax=`95%`, color=ml_model, fill=ml_model), alpha=0.4)
+}
+
+plot_single(data, "nontipped", 0)
+
+```
+
+```{r}
+plot_single(data, "tipped", 1)
+```

### 2022-04-16 darts/tune.py
Now doing tuning with purely stochastic model

@@ -2,7 +2,7 @@ import numpy as np
 import sys
 sys.path.append("../")
 from darts import TimeSeries
-from darts.models import RNNModel, BlockRNNModel
+from darts.models import RNNModel, BlockRNNModel, TCNModel
 from darts.utils.likelihood_models import LaplaceLikelihood
 from utils import preprocessed_t_series
 import random
@@ -76,8 +76,10 @@ if args.model == "lstm":
     from tune_hyperparams.lstm import hyperparameters
 elif args.model == "block_rnn":
     from tune_hyperparams.block_rnn import hyperparameters
+elif args.model == "tcn":
+    from tune_hyperparams.tcn import hyperparameters
     
-models = {"lstm": RNNModel, "block_rnn": BlockRNNModel}
+models = {"lstm": RNNModel, "block_rnn": BlockRNNModel, "tcn": TCNModel}
 
 for i in range(args.n_trials):
     # Randomly selecting a parameter dictionary

### 2022-04-16 model.py

@@ -5,7 +5,7 @@ class stochastic_tp():
     def __init__(self, N=0.75, t_max=250):
         self.replicates = 0
         self.t_max = t_max
-        self.mu = -5e-4
+        self.mu = 0
         self.sigma = 1e-2
         self.N_init = N
         self.N = N

### 2022-04-16 darts/StatsForecastAutoARIMA.py
Getting this on here for future reference

@@ -1,0 +1,51 @@
+from darts.models import StatsForecastAutoARIMA
+from darts import TimeSeries
+import sys
+sys.path.append("../")
+import matplotlib
+matplotlib.use("Agg")
+import matplotlib.pyplot as plt
+from utils import preprocessed_t_series, truth_dist
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument(
+    "-t",
+    "--tp_model",
+    default="stochastic",
+    type=str,
+    help="Select which model to use for the tipping point (stochastic/saddle)",
+)
+parser.add_argument(
+    "--case",
+    default="tipped",
+    type=str,
+    help="Select which case for the training series",
+)
+args = parser.parse_args()
+
+if args.case.lower() == "tipped" and args.tp_model == "stochastic":
+    train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
+elif args.case.lower() == "nontipped" and args.tp_model == "stochastic":
+    train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
+elif args.case.lower() == "both" and args.tp_model == "stochastic":
+    train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
+    train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
+    train_series = train_series.concatenate(train_series_, axis="sample")
+        
+input_len =25
+output_len=24
+for i in range(3):
+    train_series[:input_len].plot(color="blue", label='truth')
+    t_series, v_series = train_series.split_before(input_len)
+    t_dist = truth_dist("stochastic", t_series, input_len, output_len, n_samples=100)
+    
+    t_max = 250-input_len
+    
+    t_dist.plot(low_quantile=0.025, high_quantile=0.975, color="blue", label="_nolegend_", linestyle="dotted")
+    model = StatsForecastAutoARIMA().fit(t_series)
+    preds = model.predict(t_max, num_samples=100)
+    preds.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
+
+    plt.savefig(f"plots/StatsForecastAutoARIMA_{i}")
+    plt.clf()

### 2022-04-16 darts/StatsForecastAutoARIMA.py
Adding make dir/ line just in case

@@ -46,6 +46,8 @@ for i in range(3):
     model = StatsForecastAutoARIMA().fit(t_series)
     preds = model.predict(t_max, num_samples=100)
     preds.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
-
+    
+    if not os.path.exists("plots/"):
+        os.makedirs("plots/")
     plt.savefig(f"plots/StatsForecastAutoARIMA_{i}")
     plt.clf()

### 2022-04-18 greta/hopf_mcmc.Rmd
hopf

---
output: github_document
---

```{r}
library(tidyverse)
library(greta)
library(bayesplot)

set.seed(4242)
train_reps <- 10
train_t_max <- 100
test_t_max <- 100
test_reps <- 100

np.clip <- function(x, a, b) {
  if(x < a) return(a)
  if(x > b) return(b)
  x
}
```

```{r}

# Cycles start around K >= 21
# some stochastic resonance visible before then
step <- function(N, eta,  r = 0.75, c = 0.25, Ko = 15, delta=0.1, t=1) {

  # K will go linearly from 15 to 25 with delta = 0.1
  # Make delta negative and start with a bigger Ko to run in reverse
  K <- Ko + delta*t

  Nt <- numeric(2)
  Nt[1] <- N[1] * exp(r * (1 - N[1]/K) - c * N[2] + eta[1])
  Nt[2] <- N[1] * exp(r * (1 - N[1]/K) ) * (1 - exp(-c * N[2]+ eta[2]) )
  Nt[1] <- np.clip(Nt[1], 0, 100)
  Nt[2] <- np.clip(Nt[2], 0, 100)

  Nt
}

# simulate
simulate <- function(N_init = c(9,.1),
                     t_max = 100,
                     mu =0, # -5e-4,
                     sigma = 2e-2,
                     r = 0.75, 
                     c = 0.1,
                     Ko = 15,
                     delta = 0.1
                     ) {
  eta <- array(rnorm(2*t_max, mu, sigma), c(t_max,2))
  N   <- array(NA, c(t_max,2))

  N[1,] <- N_init
  for (t in 1:(t_max-1)) {
    N[t+1,] <- step(N[t,], eta[t,], r=r, c=c, Ko=Ko, delta=delta, t=t)
  }
  tibble::tibble(t = 1:t_max, H = N[,1], P = N[,2])
}
```

```{r}
t_max <- train_t_max + test_t_max
sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, delta=-0.1, Ko=30), .id = "i")
train <- sim |> filter(t <= train_t_max)
test <- sim |> filter(t > train_t_max)
# test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")

## show training data
train |> ggplot(aes(H, P, group=i)) + geom_path(alpha=0.4)

sim |> 
  pivot_longer(c(P,H), values_to="density", names_to="species") |>
  ggplot(aes(t, density, col=species, group=interaction(i,species))) + geom_line(alpha=0.2) +
  geom_vline(aes(xintercept = train_t_max))
```

### 2022-04-19 greta/hopf_decrease_mcmc.Rmd
some mcmc work

--- 2022-04-18 greta/hopf_mcmc.Rmd
@@
 t_max <- train_t_max + test_t_max
-sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, delta=-0.1, Ko=30), .id = "i")
+sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, delta=-0.08, Ko=30), .id = "i")
 train <- sim |> filter(t <= train_t_max)

### 2022-04-19 greta/stochastic_mcmc.Rmd
some mcmc work

--- 2022-04-14 greta/mcmc.Rmd
@@
 set.seed(4242)
-train_reps <- 10
+train_reps <- 100
 train_t_max <- 250
 test_t_max <- 250
-test_reps <- 1000
+test_reps <- 100
 
@@
 # View the bistable curve: 
-# curve( (0.75 - x)**2 * (0.25 - x), 0, 1)
+#curve( (0.75 - x)**2 * (0.25 - x), 0.749999, 0.750001)
+#curve(0*x,0,1,add = TRUE)
 
+#curve((0.25 - x)**2 * (0.75 - x), 0, 1)
+#curve(0*x,0,1,add = TRUE)
+```
+
+```{r}
 step <- function(N, eta,  a = 0.75, b = 0.25) {
@@
 simulate <- function(N_init = 0.75,
-                     t_max = 250L,
-                     mu =0, # -5e-4,
+                     t_max = 1250L,
+                       p = list(
                        sigma = 1e-2,
                        a = 0.75, 
-                     b = 0.25) {
-  eta <- rnorm(t_max, mu, sigma)
+                       b = 0.25)) {
+  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
   N   <- numeric(t_max)
@@
   for (t in 1:(t_max-1)) {
-    N[t+1] <- step(N[t], eta[t], a = a, b = b)
+    N[t+1] <- step(N[t], eta[t], a = p$a, b = p$b)
   }
@@
   purrr::transpose() %>%
-  map_dfr(function(q) simulate(t_max = test_t_max) ,.id = "i")
+  map_dfr(function(q) simulate(t_max = test_t_max, p = q) ,.id = "i")
 
-bind_rows(test, posterior_sims, .id = "model")|> 
+bind_rows(
+  mutate(test, model="true"), 
+  mutate(posterior_sims, model="predicted")
+)|> 
   ggplot(aes(t, N, col=model, group=interaction(model,i))) +
@@
 true <- tibble(a = 0.75, b = 0.25, sigma = 1e-2) |> gather(variable, value)
-
-posterior_samples  |> gather(variable, value) |> ggplot() + 
+bind_rows(map(draws, as_tibble)) |>
+  gather(variable, value) |> ggplot() + 
   geom_histogram(aes(value), bins = 30)  +

### 2022-04-19 model.py
Fixed hopf, my bad

@@ -1,5 +1,6 @@
 import numpy as np
 import matplotlib.pyplot as plt
+import copy
 
 class stochastic_tp():
     def __init__(self, N=0.75, t_max=250):
@@ -80,10 +81,10 @@ class saddle_node_tp():
         plt.savefig(file_name)
 
 class hopf_tp():
-    def __init__(self, N=[9, 0.1], t_max=100, delta=0.1, K_init=30):
+    def __init__(self, N=[9, 0.1], t_max=100, delta=0.1, K_init=15):
         self.t_max = t_max
         self.mu = 0
-        self.sigma = 0.02
+        self.sigma = .02
         self.N_init = N
         self.N = N
         self.reps = 0
@@ -96,8 +97,9 @@ class hopf_tp():
     def step(self):
         self.K += self.delta
         self.eta = np.random.normal(self.mu, self.sigma, 2)
-        self.N[0] = self.N[0] * np.exp(self.r * (1 - self.N[0] / self.K) - self.c * self.N[1] + self.eta[0])
-        self.N[1] = self.N[0] * np.exp(self.r * (1 - self.N[0] / self.K)) * (1 - np.exp(-self.c * self.N[1] + self.eta[1]))
+        N = copy.deepcopy(self.N[0])
+        self.N[0] = N * np.exp(self.r * (1 - N / self.K) - self.c * self.N[1] + self.eta[0])
+        self.N[1] = N * np.exp(self.r * (1 - N / self.K)) * (1 - np.exp(-self.c * self.N[1] + self.eta[1]))
         
         self.N = np.clip(self.N, 0 , 100)
         return self.N
@@ -106,16 +108,19 @@ class hopf_tp():
         self.reset()
         return np.array([self.step() for t in range(self.t_max)])
     
-    def collect_samples(self, reps, random_alpha=False):
+    def collect_samples(self, reps):
         self.reps += reps
         self.samples = np.array([self.draw_replicate() for rep in range(reps)])
+        
         return self.samples
     
     def reset(self):
         self.N = self.N_init
         self.K = self.K_init
+        
     
     def plot(self, file_name):
         for idx, sample in enumerate(self.samples):
-            plt.plot(np.linspace(1, self.t_max, self.t_max), sample, alpha=0.1, color="b")
+            plt.plot(np.linspace(1, self.t_max, self.t_max), sample[:,0], alpha=0.1, color="b")
[…]

### 2022-04-19 greta/hopf_mcmc.Rmd

@@
 
+Cycles start around `K >= 21`, though some stochastic resonance visible before then.
+`K` will go linearly from 15 to 25 with `delta = 0.1`.
+Make delta negative and start with a bigger `Ko` to run in reverse.
+
 ```{r}
-
-# Cycles start around K >= 21
-# some stochastic resonance visible before then
-step <- function(N, eta,  r = 0.75, c = 0.25, Ko = 15, delta=0.1, t=1) {
-
-  # K will go linearly from 15 to 25 with delta = 0.1
-  # Make delta negative and start with a bigger Ko to run in reverse
-  K <- Ko + delta*t
-
-  Nt <- numeric(2)
-  Nt[1] <- N[1] * exp(r * (1 - N[1]/K) - c * N[2] + eta[1])
-  Nt[2] <- N[1] * exp(r * (1 - N[1]/K) ) * (1 - exp(-c * N[2]+ eta[2]) )
-  Nt[1] <- np.clip(Nt[1], 0, 100)
-  Nt[2] <- np.clip(Nt[2], 0, 100)
-
-  Nt
+step <- function(H, P, eta,  t=0, 
+                 p = list(r = 0.75, c = 0.25, Ko = 15, delta=0.1)) {
+  K <- p$Ko + p$delta*t
+  N <- numeric(2)
+  N[1] <- H * exp(p$r * (1 - H/K) - p$c * P + eta[1])
+  N[2] <- H * exp(p$r * (1 - H/K) ) * (1 - exp(-p$c * P+ eta[2]) )
+  N[1] <- np.clip(N[1], 0, 100)
+  N[2] <- np.clip(N[2], 0, 100)
+  N
 }
 
-# simulate
-simulate <- function(N_init = c(9,.1),
-                     t_max = 100,
-                     mu =0, # -5e-4,
-                     sigma = 2e-2,
-                     r = 0.75, 
-                     c = 0.1,
-                     Ko = 15,
-                     delta = 0.1
+simulate <- function(t_max = 100, 
+                     p = list(H_init = 9, P_init = 1,
+                              r = 0.75, c = 0.1,  Ko = 15, delta = 0.1,
+                              sigma_H = 2e-2, sigma_P = 2e-2)
                      ) {
-  eta <- array(rnorm(2*t_max, mu, sigma), c(t_max,2))
+  eta <- array(c(rnorm(t_max, 0, p$sigma_H), 
+                 rnorm(t_max, 0, p$sigma_P)),
+               dim=c(t_max,2))
   N   <- array(NA, c(t_max,2))
 
-  N[1,] <- N_init
+  N[1,] <- c(p$H_init, p$P_init)
   for (t in 1:(t_max-1)) {
-    N[t+1,] <- step(N[t,], eta[t,], r=r, c=c, Ko=Ko, delta=delta, t=t)
+    N[t+1,] <- step(N[t,1], N[t,2], eta[t,], t=t, p = p)
   }
@@
 t_max <- train_t_max + test_t_max
-sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, delta=-0.1, Ko=30), .id = "i")
+ p = list(H_init = 9, P_init = 1,
+          r = 0.75, c = 0.1,  Ko = 14, delta = 0.08,
+          sigma_H = 2e-2, sigma_P = 2e-2)
+sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, p = p), .id = "i")
 train <- sim |> filter(t <= train_t_max)
@@
 ```
+
+```{r}
+gsims <- train |> 
+  group_by(i) |> 
+  mutate(H1 = lead(H),
+         P1 = lead(P)) |> 
+  filter(t<max(t)) |> 
+  ungroup()
+
+stopifnot(all(gsims$P1 > 0))
+
+H_t  <- gsims$H 
+H_t1 <- log( gsims$H1 ) 
+P_t  <- gsims$P
+P_t1 <- log( gsims$P1 )
+t <- gsims$t
+```
+
+```{r}
+library(greta)
+r <- uniform(0, 10)
+c <- uniform(0, 10)
+Ko <- uniform(0, 50)
+delta <- uniform(-1,1)
+sigma_H <- uniform(0, 1)
+sigma_P <- uniform(0, 1)
+
+K <- Ko + delta * t
+mu_H <- log( H_t ) +  (r * (1 - H_t/K) - c * P_t) 
+distribution(H_t1) <- normal(mu_H, sigma_H )
+
+mu_P <- log( H_t ) +  (r * (1 - H_t/K) ) + log(1 - exp(-c * P_t) )
+distribution(P_t1) <- normal(mu_P, sigma_P )
+m <- model(r, c, Ko, delta, sigma_H, sigma_P)
+```
+
+```{r}
+mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("hopf_cache"))
+bench::bench_time({                 
+    draws <- mmcmc(m, 
+                 n_samples = 10000, warmup = 5000,
+                 chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+#test_reps <- 100
+## draw test_reps number of samples
+left_off <- train |> 
+  filter(t == train_t_max) |> 
+  select(H,P) |> 
+  rename(H_init = H, P_init = P) |>
+  slice_sample(n = test_reps, replace=TRUE)
+
+posterior_samples <- 
+  bind_rows(map(draws, as_tibble)) %>% 
+  sample_n(test_reps) |>
+  bind_cols(left_off)
+
+posterior_sims <- posterior_samples %>%
+  purrr::transpose() %>%
+  map_dfr(function(q) simulate(t_max = test_t_max,
+                               p = q
+                               ) ,.id = "i") |>
+  mutate(t = t+train_t_max)
+```
+
+```{r}
+bind_rows(mutate(train, type="training"), 
+          mutate(test,type="observed"),  
+          mutate(posterior_sims, type="predicted")) |>
+  rename("host" = H, "parasite" = P) |>
+pivot_longer(c(host, parasite), values_to="density", names_to="species") |>
+  ggplot(aes(t, density, col=type, group=interaction(i,type))) + geom_line(alpha=0.2) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+```
+
+```{r}
+true <- as_tibble(p) |> select(-H_init, -P_init)  |> gather(variable, value)
+
+bind_rows(map(draws, as_tibble)) |>
+  gather(variable, value) |> ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+```
+
+# Scoring
+
+First, a single 'observed' sample from the true:
+
+```{r}
+library(scoringRules)
+
+scores <- function(observed, dat) {
+  logsscore <- scoringRules::logs_sample(observed, dat)
+  crpsscore <- scoringRules::crps_sample(observed, dat)
+  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
+
+}
+# ensemble predictions
+dat <- 
+  posterior_sims |> 
+  pivot_wider(id_cols = "t", names_from="i", values_from = "P") |> 
+  select(-t) |> as.matrix()
+
+```
+```{r}
+# score over all replicates:
+bench::bench_time({
+rep_scores <- 
+  test |> 
+  group_by(i) |> 
+  group_map(~ scores(.x$P, dat)) |> 
+  bind_rows()
+})
+```
+
+```{r}
+rep_scores |> summarise(across(.fns= base::mean))
+
+library(patchwork)
+ggplot(rep_scores) + geom_histogram(aes(crps)) + 
+(ggplot(rep_scores) + geom_histogram(aes(logs)) )
+
+```

### 2022-04-19 model.py
Cleaner fix here

@@ -97,9 +97,9 @@ class hopf_tp():
     def step(self):
         self.K += self.delta
         self.eta = np.random.normal(self.mu, self.sigma, 2)
-        N = copy.deepcopy(self.N[0])
-        self.N[0] = N * np.exp(self.r * (1 - N / self.K) - self.c * self.N[1] + self.eta[0])
-        self.N[1] = N * np.exp(self.r * (1 - N / self.K)) * (1 - np.exp(-self.c * self.N[1] + self.eta[1]))
+        g = self.N[0] * np.exp(self.r * (1 - self.N[0] / K))
+        self.N[0] = g * np.exp(- self.c * self.N[1] + self.eta[0])
+        self.N[1] = g * (1 - np.exp(-self.c * self.N[1] + self.eta[1]))
         
         self.N = np.clip(self.N, 0 , 100)
         return self.N

### 2022-04-20 darts/train_ensemble.py
Intermediate changes here

@@ -9,9 +9,9 @@ import matplotlib.pyplot as plt
 from darts import TimeSeries
 from pandas import DataFrame
 import torch
-from darts.models import RNNModel, BlockRNNModel, TCNModel
+from darts.models import RNNModel, BlockRNNModel, TransformerModel
 from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist, import_hypers, make_df
+from utils import preprocessed_t_series, truth_dist, make_df, get_train_series
 import argparse
 import yaml
 from functools import reduce
@@ -22,12 +22,12 @@ parser.add_argument(
     "--model",
     default="lstm",
     type=str,
-    help="model to train with (lstm or block_rnn)",
+    help="model to train with (lstm/block_rnn/transformer/gru)",
 )
 parser.add_argument(
     "-s",
     "--n_samples",
-    default=1000,
+    default=100,
     type=int,
     help="# of samples to train on",
 )
@@ -55,12 +55,7 @@ parser.add_argument(
     "--tp_model",
     default="stochastic",
     type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle)",
-)
-parser.add_argument(
-    "--hyper_file",
-    default="lstm",
-    help="Select which hyperparameter file to load hypers from",
+    help="Select which model to use for the tipping point (stochastic/saddle/hopf)",
 )
 parser.add_argument(
     "-c",
@@ -69,42 +64,50 @@ parser.add_argument(
     type=str,
     help="Special cases to consider (none, tipped, non_tipped, both)",
 )
+parser.add_argument(
+    "--reverse",
+    action="store_true",
+    help="Flag to use decreasing Hopf model",
+)
 args = parser.parse_args()
 
+if args.model == "lstm": 
+    if args.tp_model == "stochastic":
+        from train_hyperparams.stochastic_lstm import hyperparameters
+    elif args.tp_model == "saddle":
+        from train_hyperparams.saddle_lstm import hyperparameters
+    elif args.tp_model == "hopf":
+        from train_hyperparams.hopf_lstm import hyperparameters
+elif args.model == "gru":
+    if args.tp_model == "stochastic":
+        from train_hyperparams.stochastic_gru import hyperparameters
+    elif args.tp_model == "saddle":
+        from train_hyperparams.saddle_gru import hyperparameters
+    elif args.tp_model == "hopf":
[…]

### 2022-04-20 darts/utils.py

@@ -1,6 +1,6 @@
 import sys
 sys.path.append("../")
-from model import stochastic_tp, saddle_node_tp
+from model import stochastic_tp, saddle_node_tp, hopf_tp
 import numpy as np
 from darts import TimeSeries
 from pandas import RangeIndex, DataFrame
@@ -9,9 +9,31 @@ models = {"stochastic" : stochastic_tp,
           "saddle" : saddle_node_tp,
           }
 
-# Need to change this so that it doesn't intake args
-def preprocessed_t_series(model, n_samples):
-    _data = models[model.lower()]()
+def get_train_series(args):
+    # Selecting case to pick training set
+    if args.case.lower() == "tipped" and args.tp_model == "stochastic":
+        train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
+    elif args.case.lower() == "nontipped" and args.tp_model == "stochastic":
+        train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
+    elif args.case.lower() == "both" and args.tp_model == "stochastic":
+        train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
+        train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
+        train_series = train_series.concatenate(train_series_, axis="sample")
+    elif args.case.lower() == "none":
+        # Generating time series
+        train_series, _ = preprocessed_t_series(args.tp_model, args.n_samples, reverse=args.reverse)
+    
+    return train_series
+
+def preprocessed_t_series(model, n_samples, reverse=False):
+    if model == "hopf":
+        if reverse:
+            _data = hopf_tp(K_init=30, delta=-0.08)
+        else:
+            _data = hopf_tp(K_init=14, delta=0.08)
+    else:
+        _data = models[model.lower()]()
+        
     training_data = _data.collect_samples(n_samples)
 
     # Preprocessing time series to
@@ -21,13 +43,20 @@ def preprocessed_t_series(model, n_samples):
             _ts[j].append(training_data[i, j])
 
     vals = np.array(_ts).reshape(training_data.shape[1], 1, n_samples)
-    return TimeSeries.from_times_and_values(RangeIndex(250),vals), vals.reshape(250, n_samples).T
+    t_max = 250 if model != "hopf" else 100
[…]

### 2022-04-20 greta/hopf_mcmc.Rmd

@@
 
+```{r message=FALSE}
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
+```
+
 ```{r}
-library(tidyverse)
-library(greta)
-library(bayesplot)
-
 set.seed(4242)
-train_reps <- 10
+train_reps <- 1
 train_t_max <- 100
 test_t_max <- 100
-test_reps <- 100
+test_reps <- train_reps # same as train_reps since test cases are continuations of the training rep time series
 
@@
 ```{r}
-#test_reps <- 100
+test_reps <- 100
 ## draw test_reps number of samples
@@
 ```{r}
-bind_rows(mutate(train, type="training"), 
+combined <- 
+  bind_rows(
+    mutate(train, type="training"), 
     mutate(test,type="observed"),  
-          mutate(posterior_sims, type="predicted")) |>
+    mutate(posterior_sims, type="predicted") |> filter(i %in% 1:3)
+    ) |>
   rename("host" = H, "parasite" = P) |>
-pivot_longer(c(host, parasite), values_to="density", names_to="species") |>
-  ggplot(aes(t, density, col=type, group=interaction(i,type))) + geom_line(alpha=0.2) +
+  pivot_longer(c(host, parasite), values_to="density", names_to="species")
+
+combined |> 
+  ggplot(aes(t, density, col=type, group=interaction(i,type))) + 
+  geom_line(alpha=0.5) +
   geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
@@
 
-First, a single 'observed' sample from the true:
-
 ```{r}
@@
 }
+```
+
+We score predictions of each variable seperately.  First the predator (Parasite):
+
+```{r}
 # ensemble predictions
-dat <- 
+P_dat <- 
   posterior_sims |> 
@@
 ```
+
 ```{r}
 # score over all replicates:
-bench::bench_time({
+
 rep_scores <- 
@@
   group_by(i) |> 
-  group_map(~ scores(.x$P, dat)) |> 
+  group_map(~ scores(.x$P, P_dat)) |> 
   bind_rows()
-})
-```
 
-```{r}
 rep_scores |> summarise(across(.fns= base::mean))
 
-library(patchwork)
-ggplot(rep_scores) + geom_histogram(aes(crps)) + 
-(ggplot(rep_scores) + geom_histogram(aes(logs)) )
-
 ```

### 2022-04-21 darts/train_ensemble.py
Hopefully fixed plotting issues now

@@ -11,7 +11,7 @@ from pandas import DataFrame
 import torch
 from darts.models import RNNModel, BlockRNNModel, TransformerModel
 from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist, make_df, get_train_series
+from utils import preprocessed_t_series, truth_dist, make_df, get_train_series, plot
 import argparse
 import yaml
 from functools import reduce
@@ -138,20 +138,27 @@ if args.evaluate:
         np.random.seed(i)
         torch.manual_seed(i)
         
-        train_series, _ = preprocessed_t_series(args.tp_model, 1)
-        train_series[:input_len].plot(color="blue", label='truth')
-        t_series, v_series = train_series.split_before(input_len)
+        train_series = preprocessed_t_series(args.tp_model, 1)
+        
+        if args.tp_model == "hopf":
+            x, y = train_series[:input_len].univariate_component(0), train_series[:input_len].univariate_component(1)
+            x.plot(color="blue", label='truth')
+            y.plot(color="gold", label="truth")
+        else:
+            train_series[:input_len].plot(color="blue", label='truth')
+            
+        t_series = train_series[:input_len]
         t_dist = truth_dist(args.tp_model, t_series, input_len, output_len, n_samples=100, reverse=args.reverse)
         
-        t_max = 250-input_len
+        t_max = 250-input_len if args.tp_model != "hopf" else 200-input_len
+        plot(args.tp_model, t_dist, "blue", "gold")
         
-        t_dist.plot(low_quantile=0.025, high_quantile=0.975, color="blue", label="_nolegend_", linestyle="dotted")
         ensemble_preds = []
         for model in models:
             ensemble_preds.append(model.predict(t_max, t_series, num_samples=100))
         ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
         
-        ensemble_series.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
+        plot(args.tp_model, ensemble_series, "orange", "green")
         df = make_df(ensemble_series, t_dist, args.tp_model, args.model.lower(), args.case, args.n_samples)
         df["iter"] = i
         final_df = final_df.append(df, ignore_index=True)

### 2022-04-21 darts/utils.py

@@ -21,7 +21,7 @@ def get_train_series(args):
         train_series = train_series.concatenate(train_series_, axis="sample")
     elif args.case.lower() == "none":
         # Generating time series
-        train_series, _ = preprocessed_t_series(args.tp_model, args.n_samples, reverse=args.reverse)
+        train_series = preprocessed_t_series(args.tp_model, args.n_samples, reverse=args.reverse)
     
     return train_series
 
@@ -35,35 +35,43 @@ def preprocessed_t_series(model, n_samples, reverse=False):
         _data = models[model.lower()]()
         
     training_data = _data.collect_samples(n_samples)
-
+    # Note training_data.shape[1] is the length of the time series
     # Preprocessing time series to
     _ts = [[] for i in range(training_data.shape[1])]
     for i in range(n_samples):
         for j in range(training_data.shape[1]):
             _ts[j].append(training_data[i, j])
-
-    vals = np.array(_ts).reshape(training_data.shape[1], 1, n_samples)
+    
     t_max = 250 if model != "hopf" else 100
-    return TimeSeries.from_times_and_values(RangeIndex(t_max),vals), vals.reshape(t_max, n_samples).T
+    component_dim = 1 if model!= "hopf" else 2
+    
+    vals = np.array(_ts).reshape(training_data.shape[1], component_dim, n_samples)
+    return TimeSeries.from_times_and_values(RangeIndex(t_max),vals)
 
 def truth_dist(model, t_series, input_len, output_len, n_samples=100, reverse=False):
-    N_init = t_series[-1].values()[0][0]
+    N_init = t_series[-1].values()[0] # previously had another [0]
     t_max = 250-input_len if model != "hopf" else 200-input_len
     
+    t = len(t_series)
     if model == "hopf":
         if reverse:
-            _data = hopf_tp(N_init, t_max, K_init=30, delta=-0.08)
+            delta=-0.08
+            K_init = 30 + delta * t 
+            _data = hopf_tp(N_init, t_max, K_init=K_init, delta=delta)
         else:
-            _data = hopf_tp(N_init, t_max, K_init=14, delta=0.08)
+            delta=0.08
+            K_init = 14 + delta * t
+            _data = hopf_tp(N_init, t_max, K_init=K_init, delta=delta)
     else:
-        _data = models[model.lower()](N_init, t_max)
+        _data = models[model.lower()](N_init[0], t_max)
     
[…]

### 2022-04-22 greta/hopf_decrease_mcmc.Rmd
mcmc :turtle:

@@
 ```
+
+```{r}
+gsims <- train |> 
+  group_by(i) |> 
+  mutate(H1 = lead(H),
+         P1 = lead(P)) |> 
+  filter(t<max(t)) |> 
+  ungroup()
+
+stopifnot(all(gsims$P1 > 0))
+
+H_t  <- gsims$H 
+H_t1 <- log( gsims$H1 ) 
+P_t  <- gsims$P
+P_t1 <- log( gsims$P1 )
+t <- gsims$t
+```
+
+```{r}
+library(greta)
+r <- uniform(0, 10)
+c <- uniform(0, 10)
+Ko <- uniform(0, 50)
+delta <- uniform(-1,1)
+sigma_H <- uniform(0, 1)
+sigma_P <- uniform(0, 1)
+
+K <- Ko + delta * t
+mu_H <- log( H_t ) +  (r * (1 - H_t/K) - c * P_t) 
+distribution(H_t1) <- normal(mu_H, sigma_H )
+
+mu_P <- log( H_t ) +  (r * (1 - H_t/K) ) + log(1 - exp(-c * P_t) )
+distribution(P_t1) <- normal(mu_P, sigma_P )
+m <- model(r, c, Ko, delta, sigma_H, sigma_P)
+```
+
+```{r}
+mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("hopf_decrease_cache"))
+bench::bench_time({                 
+    draws <- mmcmc(m, 
+                 n_samples = 10000, warmup = 5000,
+                 chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+## draw test_reps number of samples
+left_off <- train |> 
+  filter(t == train_t_max) |> 
+  select(H,P) |> 
+  rename(H_init = H, P_init = P) |>
+  slice_sample(n = test_reps, replace=TRUE)
+
+posterior_samples <- 
+  bind_rows(map(draws, as_tibble)) %>% 
+  sample_n(test_reps) |>
+  bind_cols(left_off)
+
+posterior_sims <- posterior_samples %>%
+  purrr::transpose() %>%
+  map_dfr(function(q) simulate(t_max = test_t_max,
+                               p = q
+                               ) ,.id = "i") |>
+  mutate(t = t+train_t_max)
+```
+
+```{r}
+combined <- 
+  bind_rows(
+    mutate(train, type="training"), 
+    mutate(test,type="observed"),  
+    mutate(posterior_sims, type="predicted") |> filter(i %in% 1:3)
+    ) |>
+  rename("host" = H, "parasite" = P) |>
+  pivot_longer(c(host, parasite), values_to="density", names_to="species")
+
+combined |> 
+  ggplot(aes(t, density, col=type, group=interaction(i,type))) + 
+  geom_line(alpha=0.5) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+```
+
+```{r}
+true <- as_tibble(p) |> select(-H_init, -P_init)  |> gather(variable, value)
+
+bind_rows(map(draws, as_tibble)) |>
+  gather(variable, value) |> ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+```
+
+# Scoring
+
+```{r}
+library(scoringRules)
+
+scores <- function(observed, dat) {
+  logsscore <- scoringRules::logs_sample(observed, dat)
+  crpsscore <- scoringRules::crps_sample(observed, dat)
+  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
+
+}
+```
+
+We score predictions of each variable seperately.  First the predator (Parasite):
+
+```{r}
+# ensemble predictions
+P_dat <- 
+  posterior_sims |> 
+  pivot_wider(id_cols = "t", names_from="i", values_from = "P") |> 
+  select(-t) |> as.matrix()
+
+```
+
+```{r}
+# score over all replicates:
+
+rep_scores <- 
+  test |> 
+  group_by(i) |> 
+  group_map(~ scores(.x$P, P_dat)) |> 
+  bind_rows()
+
+rep_scores |> summarise(across(.fns= base::mean))
+
+```

### 2022-04-22 greta/hopf_mcmc.Rmd
mcmc :turtle:

@@
 set.seed(4242)
-train_reps <- 1
+train_reps <- 10
 train_t_max <- 100
 test_t_max <- 100
-test_reps <- train_reps # same as train_reps since test cases are continuations of the training rep time series
+test_reps <- 100
 
@@
 ```{r}
-test_reps <- 100
 ## draw test_reps number of samples

### 2022-04-22 greta/sn_mcmc.Rmd
sn mcmc :turtle:

--- 2022-04-19 greta/stochastic_mcmc.Rmd
@@
 set.seed(4242)
-train_reps <- 100
-train_t_max <- 250
-test_t_max <- 250
+train_reps <- 1
+train_t_max <- 75
+test_t_max <- 75
 test_reps <- 100
@@
 ```{r}
-# View the bistable curve: 
-#curve( (0.75 - x)**2 * (0.25 - x), 0.749999, 0.750001)
-#curve(0*x,0,1,add = TRUE)
+step <- function(N, 
+                 eta,
+                 t,
+                 pars) {
 
-#curve((0.25 - x)**2 * (0.75 - x), 0, 1)
-#curve(0*x,0,1,add = TRUE)
-```
-
-```{r}
-step <- function(N, eta,  a = 0.75, b = 0.25) {
-  N <- N + (a - N)**2 * (b - N) + eta
+        h <- np.clip(pars$h_init + pars$alpha * t, 0, 0.27)
+        N <- N + pars$r * N * (1 - N / pars$K) - 
+          h * (N**2 / (pars$s**2 + N**2)) + eta
         N <- np.clip(N, 0, 1)
+        N
   N
@@
 simulate <- function(N_init = 0.75,
-                     t_max = 1250L,
-                       p = list(
-                       sigma = 1e-2,
-                       a = 0.75, 
-                       b = 0.25)) {
-  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
+                     t_max = 250L,
+                     pars = list(
+                        r = 1,
+                        K = 1,
+                        s = 0.1,
+                        h_init = .15,
+                        alpha = 0.0015,
+                        mu = 0,
+                        sigma = 0.00
+                      )
+                    ) {
+  eta <- rnorm(t_max, pars$mu, pars$sigma) # mu = 0, no drift
   N   <- numeric(t_max)
@@
   for (t in 1:(t_max-1)) {
-    N[t+1] <- step(N[t], eta[t], a = p$a, b = p$b)
+    N[t+1] <- step(N[t], eta[t], t, pars)
   }
@@
 }
+```
 
-train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
-test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
+```{r}
+pars = list(r = 1,
+            K = 1,
+            s = 0.1,
+            h_init = .15,
+            alpha = 0.0015,
+            mu = 0,
+            sigma = 0.02
+          )
+t_max <- train_t_max + test_t_max
+sims <- purrr::map_dfr(1:train_reps, 
+                       \(i) simulate(t_max=t_max, pars = pars), 
+                       .id = "i")
 
-## show training data
-train |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.2)
+train <- sims |> filter(t <= train_t_max)
+test <- sims |> filter(t > train_t_max)
+```
+
+```{r}
+sims <- bind_rows(test, train, .id = "data")
+sims |> 
+  ggplot(aes(t, N, group=interaction(i,data), col=data)) + 
+  geom_line()
 ```
@@
 gsims <- train |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
-
 x_t <- gsims$N
 x_t1 <- gsims$xt1
+t <- gsims$t
 
 library(greta)
-a <- uniform(0, 10)
-b <- uniform(0, 10)
+r <- uniform(0, 10)
+K <- uniform(0, 10)
+s <- uniform(0,10)
+h_init <- uniform(0,10)
+alpha <- uniform(0, 10)
 sigma <- uniform(0, 10)
-mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
-distribution(x_t1) <- normal(mean, sigma )
-m <- model(a, b, sigma)
+h <- h_init + alpha * t
+mean <-  x_t + r*x_t*(1 - x_t / K) - h*(x_t^2 / (s^2 + x_t^2))
+distribution(x_t1) <- normal(mean, sigma)
+m <- model(r, K, s, h_init, alpha, sigma)
 ```
@@
 ```{r}
-mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))
+mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("sn_cache"))
 
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 60000, warmup = 50000, chains = 4, verbose = FALSE)
 })
@@
 library(patchwork)
-ggplot(rep_scores) + geom_histogram(aes(crps)) + 
-(ggplot(rep_scores) + geom_histogram(aes(logs)) )
+ggplot(rep_scores) + geom_histogram(aes(crps), bins=20) + 
+(ggplot(rep_scores) + geom_histogram(aes(logs), bins=20) )
 

### 2022-04-22 greta/hopf_mcmc.Rmd
mcmc ...

@@
 ```{r}
-step <- function(H, P, eta,  t=0, 
-                 p = list(r = 0.75, c = 0.25, Ko = 15, delta=0.1)) {
-  K <- p$Ko + p$delta*t
+step <- function(H, P, eta,  t=0, p) {
+  K <- p$Ko + p$delta*(t+p$t_init)
   N <- numeric(2)
@@
                               r = 0.75, c = 0.1,  Ko = 15, delta = 0.1,
-                              sigma_H = 2e-2, sigma_P = 2e-2)
+                              sigma_H = 2e-2, sigma_P = 2e-2, t_init = 0)
                      ) {
@@
           r = 0.75, c = 0.1,  Ko = 14, delta = 0.08,
-          sigma_H = 2e-2, sigma_P = 2e-2)
+          sigma_H = 2e-2, sigma_P = 2e-2, t_init = 0)
 sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, p = p), .id = "i")

### 2022-04-22 greta/sn_mcmc.Rmd
mcmc ...

@@
 set.seed(4242)
-train_reps <- 1
+train_reps <- 100
 train_t_max <- 75
@@
 
-        h <- np.clip(pars$h_init + pars$alpha * t, 0, 0.27)
+        h <- np.clip(pars$h_init + pars$alpha * (pars$t_init + t), 0, 0.27)
         N <- N + pars$r * N * (1 - N / pars$K) - 
           h * (N**2 / (pars$s**2 + N**2)) + eta
-        N <- np.clip(N, 0, 1)
+        N <- np.clip(N, 0, 100)
         N
@@
                         mu = 0,
-                        sigma = 0.00
+                        sigma = 0.00,
+                        t_init = 0
                       )
@@
   }
-  tibble::tibble(t = 1:t_max, N = N)
+  tibble::tibble(t = pars$t_init + 1:t_max, N = N)
 }
@@
             mu = 0,
-            sigma = 0.02
+            sigma = 0.02,
+            t_init = 0
           )
@@
 K <- uniform(0, 10)
-s <- uniform(0,10)
-h_init <- uniform(0,10)
+s <- uniform(0, 10)
+h_init <- uniform(0, 10)
 alpha <- uniform(0, 10)
@@
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 60000, warmup = 50000, chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 100000, warmup = 80000, chains = 4, verbose = FALSE)
 })
@@
 ## draw test_reps number of samples
+left_off <- train |> 
+  filter(t == train_t_max) |> 
+  select(N) |> 
+  rename(N_init = N) |>
+  slice_sample(n = test_reps, replace=TRUE)
 
 posterior_samples <- 
-  bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(test_reps)
-posterior_sims <- posterior_samples %>%
-  purrr::transpose() %>%
-  map_dfr(function(q) simulate(t_max = test_t_max, p = q) ,.id = "i")
+  bind_rows(map(draws, as_tibble)) |> 
+  sample_n(test_reps) |>
+  mutate(t_init = train_t_max) |>
+  bind_cols(left_off)
+
+posterior_sims <- posterior_samples |>
+  mutate(mu=0) |>
+  purrr::transpose() |>
+  map_dfr(function(q) simulate(t_max = test_t_max, pars = q, N_init = q$N_init) ,.id = "i")
 
 bind_rows(
+  mutate(train, model="historical"),
   mutate(test, model="true"), 
@@
   ggplot(aes(t, N, col=model, group=interaction(model,i))) +
-           geom_line(alpha=0.1)
+           geom_line(alpha=0.2)
 
@@
 ```{r}
-true <- tibble(a = 0.75, b = 0.25, sigma = 1e-2) |> gather(variable, value)
+true <- as_tibble(pars) |> gather(variable, value)
 bind_rows(map(draws, as_tibble)) |>

### 2022-04-22 greta/hopf_mcmc.Rmd
work in progress

@@
 set.seed(4242)
-train_reps <- 10
+train_reps <- 1
 train_t_max <- 100
@@
   }
-  tibble::tibble(t = 1:t_max, H = N[,1], P = N[,2])
+  tibble::tibble(t = p$t_init + 1:t_max, H = N[,1], P = N[,2])
 }
 ```
+
+# Increasing K
 
@@
   pivot_longer(c(P,H), values_to="density", names_to="species") |>
-  ggplot(aes(t, density, col=species, group=interaction(i,species))) + geom_line(alpha=0.2) +
+  ggplot(aes(t, density, col=species, group=interaction(i,species))) + 
+  geom_line() +
   geom_vline(aes(xintercept = train_t_max))
@@
 library(greta)
-r <- uniform(0, 10)
-c <- uniform(0, 10)
+r <- lognormal(0, 1)
+c <- lognormal(0, 1)
 Ko <- uniform(0, 50)
 delta <- uniform(-1,1)
-sigma_H <- uniform(0, 1)
-sigma_P <- uniform(0, 1)
+sigma_H <- lognormal(0, 1)
+sigma_P <- lognormal(0, 1)
 
@@
 ```{r}
+true <- as_tibble(p) |> select(-H_init, -P_init)  |> gather(variable, value)
+
+bind_rows(map(draws, as_tibble)) |>
+  gather(variable, value) |> ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+```
+
+```{r}
 ## draw test_reps number of samples
@@
   sample_n(test_reps) |>
+  mutate(t_init = train_t_max) |>
   bind_cols(left_off)
@@
                                p = q
-                               ) ,.id = "i") |>
-  mutate(t = t+train_t_max)
+                               ) ,.id = "i")
 ```
@@
     mutate(test,type="observed"),  
-    mutate(posterior_sims, type="predicted") |> filter(i %in% 1:3)
+    mutate(posterior_sims, type="predicted") 
     ) |>
@@
   pivot_longer(c(host, parasite), values_to="density", names_to="species")
+```
 
+```{r}
 combined |> 
+  group_by(t,type,species) |> 
+  summarise(mean = mean(density), sd = sd(density), .groups = "drop") |> 
+ ggplot(aes(t, col=type)) + 
+  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_line(aes(y=mean)) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+```
+
+```{r}
+combined |> 
+      filter(i %in% 1:train_reps) |> 
   ggplot(aes(t, density, col=type, group=interaction(i,type))) + 
@@
   geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+```
+
+# Decreasing K
+
+```{r}
+t_max <- train_t_max + test_t_max
+ p = list(H_init = 9, P_init = 1,
+          r = 0.75, c = 0.1,  Ko = 30, delta = -0.08,
+          sigma_H = 2e-2, sigma_P = 2e-2, t_init = 0)
+sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, p = p), .id = "i")
+train <- sim |> filter(t <= train_t_max)
+test <- sim |> filter(t > train_t_max)
+# test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
+
+## show training data
+train |> ggplot(aes(H, P, group=i)) + geom_path(alpha=0.4)
+
+sim |> 
+  pivot_longer(c(P,H), values_to="density", names_to="species") |>
+  ggplot(aes(t, density, col=species, group=interaction(i,species))) + geom_line(alpha=1) +
+  geom_vline(aes(xintercept = train_t_max))
+```
+
+```{r}
+gsims <- train |> 
+  group_by(i) |> 
+  mutate(H1 = lead(H),
+         P1 = lead(P)) |> 
+  filter(t<max(t)) |> 
+  ungroup()
+
+stopifnot(all(gsims$P1 > 0))
+
+H_t  <- gsims$H 
+H_t1 <- log( gsims$H1 ) 
+P_t  <- gsims$P
+P_t1 <- log( gsims$P1 )
+t <- gsims$t
+```
+
+```{r}
+library(greta)
+r <- lognormal(0, 1)
+c <- lognormal(0, 1)
+Ko <- uniform(0, 50)
+delta <- uniform(-1,1)
+sigma_H <- lognormal(0, 1)
+sigma_P <- lognormal(0, 1)
+
+K <- Ko + delta * t
+mu_H <- log( H_t ) +  (r * (1 - H_t/K) - c * P_t) 
+distribution(H_t1) <- normal(mu_H, sigma_H )
+
+mu_P <- log( H_t ) +  (r * (1 - H_t/K) ) + log(1 - exp(-c * P_t) )
+distribution(P_t1) <- normal(mu_P, sigma_P )
+m <- model(r, c, Ko, delta, sigma_H, sigma_P)
+```
+
+```{r}
+mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("hopf_decreasing_cache"))
+bench::bench_time({                 
+    draws <- mmcmc(m, 
+                 n_samples = 10000, warmup = 5000,
+                 chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
 ```
@@
   facet_wrap(~variable, scales = "free")
+```
+
+```{r}
+## draw test_reps number of samples
+left_off <- train |> 
+  filter(t == train_t_max) |> 
+  select(H,P) |> 
+  rename(H_init = H, P_init = P) |>
+  slice_sample(n = test_reps, replace=TRUE)
+
+posterior_samples <- 
+  bind_rows(map(draws, as_tibble)) %>% 
+  sample_n(test_reps) |>
+  mutate(t_init = train_t_max) |>
+  bind_cols(left_off)
+
+posterior_sims <- posterior_samples %>%
+  purrr::transpose() %>%
+  map_dfr(function(q) simulate(t_max = test_t_max,
+                               p = q
+                               ) ,.id = "i")
+```
+
+```{r}
+combined <- 
+  bind_rows(
+    mutate(train, type="training"), 
+    mutate(test,type="observed"),  
+    mutate(posterior_sims, type="predicted") 
+    ) |>
+  rename("host" = H, "parasite" = P) |>
+  pivot_longer(c(host, parasite), values_to="density", names_to="species")
+```
+
+```{r}
+combined |> 
+  group_by(t,type,species) |> 
+  summarise(mean = mean(density), sd = sd(density), .groups = "drop") |> 
+ ggplot(aes(t, col=type)) + 
+  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_line(aes(y=mean)) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+```
+
+```{r}
+combined |> 
+      filter(i %in% 1:train_reps) |> 
+  ggplot(aes(t, density, col=type, group=interaction(i,type))) + 
+  geom_line(alpha=0.5) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
 ```

### 2022-04-23 greta/mcmc.Rmd
generalizing functions a bit

--- 2022-04-13 greta/mcmc.Rmd
@@
 library(tidyverse)
-set.seed(1234567)
-
-np.clip <- function(x, a, b) {
-  if(x < a) return(a)
-  if(x > b) return(b)
-  x
-}
-
-# View the bistable curve: 
-# curve( (0.75 - x)**2 * (0.25 - x), 0, 1)
-
-step <- function(N, eta,  a = 0.75, b = 0.25) {
-  N <- N + (a - N)**2 * (b - N) + eta
-  N <- np.clip(N, 0, 1)
-  N
-}
-
-# simulate
-simulate <- function(N_init = 0.75,
-                     t_max = 250L,
-                     mu =0, # -5e-4,
-                     sigma = 1e-2,
-                     a = 0.75, 
-                     b = 0.25) {
-  eta <- rnorm(t_max, mu, sigma)
-  N   <- numeric(t_max)
-  N[1] <- N_init
-  for (t in 1:(t_max-1)) {
-    N[t+1] <- step(N[t], eta[t], a = a, b = b)
-  }
-  tibble::tibble(t = 1:t_max, N = N)
-}
-
-sims <- purrr::map_dfr(1:10, \(i) simulate(), .id = "i")
-sims |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.06)
+library(greta)
+library(bayesplot)
+source("R/saddle_node.R")
 ```
@@
 ```{r}
-gsims <- sims |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
+set.seed(4242)
+pars = list(r = 1,
+            K = 1,
+            s = 0.1,
+            h_init = .15,
+            alpha = 0.0015,
+            mu = 0,
+            sigma = 0.02,
+            t_init = 0
+          )
+t_max <- train_t_max + test_t_max
 
-x_t <- gsims$N
-x_t1 <- gsims$xt1
-
-library(greta)
-a <- uniform(0, 10)
-b <- uniform(0, 10)
-sigma <- uniform(0, 10)
-mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
-distribution(x_t1) <- normal(mean, sigma )
-m <- model(a, b, sigma)
+simulate <- simulate_sn
 ```
@@
 ```{r}
-mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))
+train_reps <- 100
+train_t_max <- 75
+test_t_max <- 75
+test_reps <- 100
+
+sims <- purrr::map_dfr(1:train_reps, 
+                       \(i) simulate(t_max=t_max, pars = pars), 
+                       .id = "i")
+train <- sims |> filter(t <= train_t_max)
+test <- sims |> filter(t > train_t_max)
+```
+
+```{r}
+m <- greta_model_sn(train)
 
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 100000, warmup = 80000, chains = 4, verbose = FALSE)
 })
@@
 bayesplot::mcmc_trace(draws)
+plot_posteriors(draws, pars)
 ```
@@
 ```{r}
-posterior_samples <- 
-  bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(100)
-posterior_sims <- posterior_samples %>%
-  purrr::transpose() %>%
-  map_dfr(function(q) simulate() ,.id = "i")
+fc <- forecast(draws, train, test, simulate)
 
-bind_rows(sims, posterior_sims, .id = "model")|> 
-  ggplot(aes(t, N, col=model, group=interaction(model,i))) +
-           geom_line(alpha=0.3)
+fc |> ggplot(aes(t, N, col=model, group=interaction(model,i))) +
+           geom_line(alpha=0.2)
 
@@
 ```{r}
-library(scoringRules)
 
-scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
-  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
-
-}
 # ensemble predictions
-dat <- posterior_sims |> pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> select(-t) |> as.matrix()
-
-# Score just the 1st replicate "true" observations
-observed <- sims |> filter(i == 1) |> pull(N)
-scores(observed, dat)
+dat <- 
+  posterior_sims |> 
+  pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
+  select(-t) |> as.matrix()
 
@@
 # score over all replicates:
-
-rep_scores <- sims |> group_by(i) |> group_map(~ scores(.x$N, dat)) |> bind_rows()
+bench::bench_time({
+rep_scores <- 
+  test |> 
+  group_by(i) |> 
+  group_map(~ scores(.x$N, dat)) |> 
+  bind_rows()
+})
 ```

### 2022-04-23 greta/R/saddle_node.R

@@ -1,0 +1,71 @@
+library(tidyverse)
+library(greta)
+library(bayesplot)
+library(scoringRules)
+
+
+
+np.clip <- function(x, a, b) {
+  if(x < a) return(a)
+  if(x > b) return(b)
+  x
+}
+
+step_sn <- function(N, 
+                 eta,
+                 t,
+                 pars) {
+  h <- np.clip(pars$h_init + pars$alpha * (pars$t_init + t), 0, 0.27)
+  N <- N + pars$r * N * (1 - N / pars$K) - 
+    h * (N**2 / (pars$s**2 + N**2)) + eta
+  N <- np.clip(N, 0, 100)
+  N
+}
+# simulate
+simulate_sn <- function(N_init = 0.75,
+                     t_max = 250L,
+                     pars = list(
+                       r = 1,
+                       K = 1,
+                       s = 0.1,
+                       h_init = .15,
+                       alpha = 0.0015,
+                       mu = 0,
+                       sigma = 0.00,
+                       t_init = 0
+                     )
+) {
+
+  eta <- rnorm(t_max, pars$mu, pars$sigma) # mu = 0, no drift
+  N   <- numeric(t_max)
+  N[1] <- N_init
+  for (t in 1:(t_max-1)) {
+    N[t+1] <- step_sn(N[t], eta[t], t, pars)
+  }
+  tibble::tibble(t = pars$t_init + 1:t_max, N = N)
+}
+
+
+greta_model_sn <- function(train) {
+  gsims <- train |> 
+    group_by(i) |> 
+    mutate(xt1 = lead(N)) |>
+    filter(t<max(t))
+  x_t <- gsims$N
+  x_t1 <- gsims$xt1
+  t <- gsims$t
+  
+  r <- uniform(0, 10)
+  K <- uniform(0, 10)
+  s <- uniform(0, 10)
+  h_init <- uniform(0, 10)
+  alpha <- uniform(0, 10)
+  sigma <- uniform(0, 10)
+  h <- h_init + alpha * t
+  mean <-  x_t + r*x_t*(1 - x_t / K) - h*(x_t^2 / (s^2 + x_t^2))
+  distribution(x_t1) <- normal(mean, sigma)
+  m <- model(r, K, s, h_init, alpha, sigma)
+  
+}
+
+

### 2022-04-23 greta/R/utils.R

@@ -1,0 +1,59 @@
+library(tidyverse)
+library(memoise)
+library(greta)
+library(scoringRules)
+
+mmcmc <- memoise::memoise(greta::mcmc, cache = memoise::cache_filesystem("cache"))
+
+plot_posteriors <- function(draws, pars) {
+  true <- as_tibble(pars) |> gather(variable, value)
+  bind_rows(map(draws, as_tibble)) |>
+    gather(variable, value) |> ggplot() + 
+    geom_histogram(aes(value), bins = 30)  +
+    geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+    facet_wrap(~variable, scales = "free")
+}
+
+
+forecast <- function(draws, train, test, simulate) {
+  train_t_max <- max(train$t)
+  
+  # get initial conditions for forecast from final state distribution in training samples
+  left_off <- train |> 
+    filter(t == train_t_max) |> 
+    select(N) |> 
+    rename(N_init = N) |>
+    slice_sample(n = test_reps, replace=TRUE)
+  
+  # sample from posteriors
+  posterior_samples <- 
+    bind_rows(map(draws, as_tibble)) |> 
+    sample_n(test_reps) |>
+    mutate(t_init = train_t_max) |>
+    bind_cols(left_off)
+  
+  # run simulate() with each row of parameters
+  posterior_sims <- 
+    posterior_samples |>
+    mutate(mu=0) |>
+    purrr::transpose() |>
+    map_dfr(function(q) 
+      simulate(t_max = test_t_max, pars = q, N_init = q$N_init),
+      .id = "i")
+  
+  # Combine as single data.frame
+  bind_rows(
+    mutate(train, model="historical"),
+    mutate(test, model="true"), 
+    mutate(posterior_sims, model="predicted"))
+}
+
+
+
+
+scores <- function(observed, dat) {
+  logsscore <- scoringRules::logs_sample(observed, dat)
+  crpsscore <- scoringRules::crps_sample(observed, dat)
+  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
+  
+}

### 2022-04-24 greta/hopf_mcmc.Rmd
refactor to shared functions

--- 2022-04-22 greta/hopf_mcmc.Rmd
@@
 ---
+
+Cycles start around `K >= 21`, though some stochastic resonance visible before then.
+`K` will go linearly from 15 to 25 with `delta = 0.1`.
+Make delta negative and start with a bigger `Ko` to run in reverse.
 
@@
 library(bayesplot, quietly = TRUE)
+source("R/utils.R")
+source("R/hopf.R")
 ```
@@
 
-np.clip <- function(x, a, b) {
-  if(x < a) return(a)
-  if(x > b) return(b)
-  x
-}
+simulate <- simulate_hopf
 ```
 
-Cycles start around `K >= 21`, though some stochastic resonance visible before then.
-`K` will go linearly from 15 to 25 with `delta = 0.1`.
-Make delta negative and start with a bigger `Ko` to run in reverse.
+# Increasing K
 
 ```{r}
-step <- function(H, P, eta,  t=0, p) {
-  K <- p$Ko + p$delta*(t+p$t_init)
-  N <- numeric(2)
-  N[1] <- H * exp(p$r * (1 - H/K) - p$c * P + eta[1])
-  N[2] <- H * exp(p$r * (1 - H/K) ) * (1 - exp(-p$c * P+ eta[2]) )
-  N[1] <- np.clip(N[1], 0, 100)
-  N[2] <- np.clip(N[2], 0, 100)
-  N
-}
+t_max <- train_t_max + test_t_max
+ p = list(H_init = 9, 
+          P_init = 1,
+          r = 0.75, 
+          c = 0.1,
+          Ko = 14,
+          delta = 0.08,
+          sigma_H = 2e-2,
+          sigma_P = 2e-2,
+          t_init = 0)
 
-simulate <- function(t_max = 100, 
-                     p = list(H_init = 9, P_init = 1,
-                              r = 0.75, c = 0.1,  Ko = 15, delta = 0.1,
-                              sigma_H = 2e-2, sigma_P = 2e-2, t_init = 0)
-                     ) {
-  eta <- array(c(rnorm(t_max, 0, p$sigma_H), 
-                 rnorm(t_max, 0, p$sigma_P)),
-               dim=c(t_max,2))
-  N   <- array(NA, c(t_max,2))
+sim <- purrr::map_dfr(1:test_reps, 
+                      \(i) simulate(t_max=t_max, p = p),
+                      .id = "i")
+train <- sim |> filter(t <= train_t_max, i <= train_reps)
+test <- sim |> filter(t > train_t_max)
 
-  N[1,] <- c(p$H_init, p$P_init)
-  for (t in 1:(t_max-1)) {
-    N[t+1,] <- step(N[t,1], N[t,2], eta[t,], t=t, p = p)
-  }
-  tibble::tibble(t = 1:t_max, H = N[,1], P = N[,2])
-}
 ```
@@
 ```{r}
-t_max <- train_t_max + test_t_max
- p = list(H_init = 9, P_init = 1,
-          r = 0.75, c = 0.1,  Ko = 14, delta = 0.08,
-          sigma_H = 2e-2, sigma_P = 2e-2, t_init = 0)
-sim <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=t_max, p = p), .id = "i")
-train <- sim |> filter(t <= train_t_max)
-test <- sim |> filter(t > train_t_max)
-# test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
-
-## show training data
-train |> ggplot(aes(H, P, group=i)) + geom_path(alpha=0.4)
-
-sim |> 
-  pivot_longer(c(P,H), values_to="density", names_to="species") |>
-  ggplot(aes(t, density, col=species, group=interaction(i,species))) + geom_line(alpha=0.2) +
-  geom_vline(aes(xintercept = train_t_max))
-```
-
-```{r}
-gsims <- train |> 
-  group_by(i) |> 
-  mutate(H1 = lead(H),
-         P1 = lead(P)) |> 
-  filter(t<max(t)) |> 
-  ungroup()
-
-stopifnot(all(gsims$P1 > 0))
-
-H_t  <- gsims$H 
-H_t1 <- log( gsims$H1 ) 
-P_t  <- gsims$P
-P_t1 <- log( gsims$P1 )
-t <- gsims$t
-```
-
-```{r}
-library(greta)
-r <- uniform(0, 10)
-c <- uniform(0, 10)
-Ko <- uniform(0, 50)
-delta <- uniform(-1,1)
-sigma_H <- uniform(0, 1)
-sigma_P <- uniform(0, 1)
-
-K <- Ko + delta * t
-mu_H <- log( H_t ) +  (r * (1 - H_t/K) - c * P_t) 
-distribution(H_t1) <- normal(mu_H, sigma_H )
-
-mu_P <- log( H_t ) +  (r * (1 - H_t/K) ) + log(1 - exp(-c * P_t) )
-distribution(P_t1) <- normal(mu_P, sigma_P )
-m <- model(r, c, Ko, delta, sigma_H, sigma_P)
-```
-
-```{r}
-mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("hopf_cache"))
-bench::bench_time({                 
+m <- greta_model_hopf(train)
 draws <- mmcmc(m, 
@@
              chains = 4, verbose = FALSE)
-})
+
 ```
+## MCMC evaluation
 
@@
 bayesplot::mcmc_trace(draws)
-```
-
-```{r}
-## draw test_reps number of samples
-left_off <- train |> 
-  filter(t == train_t_max) |> 
-  select(H,P) |> 
-  rename(H_init = H, P_init = P) |>
-  slice_sample(n = test_reps, replace=TRUE)
-
-posterior_samples <- 
-  bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(test_reps) |>
-  bind_cols(left_off)
-
-posterior_sims <- posterior_samples %>%
-  purrr::transpose() %>%
-  map_dfr(function(q) simulate(t_max = test_t_max,
-                               p = q
-                               ) ,.id = "i") |>
-  mutate(t = t+train_t_max)
+plot_posteriors(draws,p)
 ```
@@
 combined <- 
-  bind_rows(
-    mutate(train, type="training"), 
-    mutate(test,type="observed"),  
-    mutate(posterior_sims, type="predicted") |> filter(i %in% 1:3)
-    ) |>
-  rename("host" = H, "parasite" = P) |>
-  pivot_longer(c(host, parasite), values_to="density", names_to="species")
+  compare_forecast(draws, train, test, simulate, vars = c("H", "P")) |> 
+  pivot_longer(c(H, P), values_to="value", names_to="variable")
 
-combined |> 
-  ggplot(aes(t, density, col=type, group=interaction(i,type))) + 
-  geom_line(alpha=0.5) +
-  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~species, ncol=1)
+write_csv(combined, "data/hopf_increasing.csv.gz")
 ```
@@
 ```{r}
-true <- as_tibble(p) |> select(-H_init, -P_init)  |> gather(variable, value)
+combined |> 
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+ ggplot(aes(t, col=type)) + 
+  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_line(aes(y=mean)) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+```
 
-bind_rows(map(draws, as_tibble)) |>
-  gather(variable, value) |> ggplot() + 
-  geom_histogram(aes(value), bins = 30)  +
-  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
-  facet_wrap(~variable, scales = "free")
+```{r}
+single <- 
+  combined |> 
+  filter(i %in% 1:train_reps) |> 
+  ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
+    geom_line() +
+    geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+```
+
+```{r}
+scores <- bind_rows(
+  rep_scores(combined, "P"),
+  rep_scores(combined, "H") |> 
+  mutate(scenario="hopf_increasing", model="MCMC") 
+)
+```
+
+# Decreasing K
+
+```{r}
+t_max <- train_t_max + test_t_max
+p = p
+p$delta <- -0.08
+p$Ko <- 30
+
+sim <- purrr::map_dfr(1:test_reps, 
+                      \(i) simulate(t_max=t_max, p = p),
+                      .id = "i")
+train <- sim |> filter(t <= train_t_max, i <= train_reps)
+test <- sim |> filter(t > train_t_max)
+```
+
+```{r}
+m2 <- greta_model_hopf(train)
+draws2 <- mmcmc(m2, 
+               n_samples = 10000, warmup = 5000,
+               chains = 4, verbose = FALSE)
+
+```
+
+## MCMC evaluation
+
+```{r}
+bayesplot::mcmc_trace(draws2)
+plot_posteriors(draws2, p)
+```
+
+```{r}
+combined <- 
+  compare_forecast(draws2, train, test, simulate, vars = c("H", "P")) |> 
+  pivot_longer(c(H, P), values_to="value", names_to="variable")
+write_csv(combined, "data/hopf_decreasing.csv.gz")
+
+```
+
+```{r}
+combined |> 
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+   ggplot(aes(t, col=type)) + 
+    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+    geom_line(aes(y=mean)) +
+    geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+```
+
+```{r}
+single <- combined |> 
+  filter(i %in% 1:train_reps) |> 
+  ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
+  geom_line() +
+  geom_vline(aes(xintercept = train_t_max)) +
+  facet_wrap(~variable, ncol=1)
 ```
@@
 ```{r}
-library(scoringRules)
 
-scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
-  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
-
-}
-```
-
-We score predictions of each variable seperately.  First the predator (Parasite):
-
-```{r}
-# ensemble predictions
-P_dat <- 
-  posterior_sims |> 
-  pivot_wider(id_cols = "t", names_from="i", values_from = "P") |> 
-  select(-t) |> as.matrix()
+scores <- bind_rows(
+  rep_scores(combined, "P"),
+  rep_scores(combined, "H") |> 
+  mutate(scenario="hopf_decreasing", model="MCMC") 
+) |> bind_rows(scores)
 
@@
 ```{r}
-# score over all replicates:
-
-rep_scores <- 
-  test |> 
-  group_by(i) |> 
-  group_map(~ scores(.x$P, P_dat)) |> 
-  bind_rows()
-
-rep_scores |> summarise(across(.fns= base::mean))
-
+scores |> ggplot(aes(variable, crps)) + geom_boxplot() ~ facet_wrap(~scenario)
 ```

### 2022-04-24 greta/stochastic_mcmc.Rmd
refactor to shared functions

@@
 set.seed(4242)
-train_reps <- 100
+train_reps <- 1
 train_t_max <- 250
@@
 
-np.clip <- function(x, a, b) {
-  if(x < a) return(a)
-  if(x > b) return(b)
-  x
-}
 ```
@@
 ```{r}
-# View the bistable curve: 
-#curve( (0.75 - x)**2 * (0.25 - x), 0.749999, 0.750001)
-#curve(0*x,0,1,add = TRUE)
 
-#curve((0.25 - x)**2 * (0.75 - x), 0, 1)
-#curve(0*x,0,1,add = TRUE)
 ```
@@
 ```{r}
-step <- function(N, eta,  a = 0.75, b = 0.25) {
-  N <- N + (a - N)**2 * (b - N) + eta
-  N <- np.clip(N, 0, 1)
-  N
-}
-
-# simulate
-simulate <- function(N_init = 0.75,
-                     t_max = 1250L,
-                       p = list(
-                       sigma = 1e-2,
-                       a = 0.75, 
-                       b = 0.25)) {
-  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
-  N   <- numeric(t_max)
-  N[1] <- N_init
-  for (t in 1:(t_max-1)) {
-    N[t+1] <- step(N[t], eta[t], a = p$a, b = p$b)
-  }
-  tibble::tibble(t = 1:t_max, N = N)
-}
 
@@
 ```{r}
-gsims <- train |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
 
-x_t <- gsims$N
-x_t1 <- gsims$xt1
-
-library(greta)
-a <- uniform(0, 10)
-b <- uniform(0, 10)
-sigma <- uniform(0, 10)
-mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
-distribution(x_t1) <- normal(mean, sigma )
-m <- model(a, b, sigma)
+m <- greta_model_stoch(train)
 ```
@@
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 60000, warmup = 50000, chains = 4, verbose = FALSE)
 })
@@
 bayesplot::mcmc_trace(draws)
+plot_posteriors(draws, p)
 ```

### 2022-04-24 greta/R/hopf.R

@@ -1,0 +1,63 @@
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
+
+step_hopf <- function(H, P, eta,  t=0, p) {
+  K <- p$Ko + p$delta*(t+p$t_init)
+  N <- numeric(2)
+  N[1] <- H * exp(p$r * (1 - H/K) - p$c * P + eta[1])
+  N[2] <- H * exp(p$r * (1 - H/K) ) * (1 - exp(-p$c * P+ eta[2]) )
+  N[1] <- np.clip(N[1], 0, 100)
+  N[2] <- np.clip(N[2], 0, 100)
+  N
+}
+
+simulate_hopf <- function(t_max = 100, p) {
+  eta <- array(c(rnorm(t_max, 0, p$sigma_H), 
+                 rnorm(t_max, 0, p$sigma_P)),
+               dim=c(t_max,2))
+  N   <- array(NA, c(t_max,2))
+  
+  N[1,] <- c(p$H_init, p$P_init)
+  for (t in 1:(t_max-1)) {
+    N[t+1,] <- step_hopf(N[t,1], N[t,2], eta[t,], t=t, p = p)
+  }
+  tibble::tibble(t = p$t_init + 1:t_max, H = N[,1], P = N[,2])
+}
+
+
+greta_model_hopf <- function(train) {
+  gsims <- train |> 
+    group_by(i) |> 
+    mutate(H1 = lead(H),
+           P1 = lead(P)) |> 
+    filter(t<max(t)) |> 
+    ungroup()
+  stopifnot(all(gsims$P1 > 0))
+  
+  H_t  <- gsims$H 
+  H_t1 <- log( gsims$H1 ) 
+  P_t  <- gsims$P
+  P_t1 <- log( gsims$P1 )
+  t <- gsims$t
+  
+  
+  
+  r <- lognormal(0, 1)
+  c <- lognormal(0, 1)
+  Ko <- uniform(0, 50)
+  delta <- uniform(-1,1)
+  sigma_H <- lognormal(0, 1)
+  sigma_P <- lognormal(0, 1)
+  
+  K <- Ko + delta * t
+  mu_H <- log( H_t ) +  (r * (1 - H_t/K) - c * P_t) 
+  distribution(H_t1) <- normal(mu_H, sigma_H )
+  
+  
[…]

### 2022-04-24 greta/R/saddle_node.R

@@ -1,48 +1,39 @@
-library(tidyverse)
-library(greta)
-library(bayesplot)
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
 library(scoringRules)
-
-
-
-np.clip <- function(x, a, b) {
-  if(x < a) return(a)
-  if(x > b) return(b)
-  x
-}
 
 step_sn <- function(N, 
                  eta,
                  t,
-                 pars) {
-  h <- np.clip(pars$h_init + pars$alpha * (pars$t_init + t), 0, 0.27)
-  N <- N + pars$r * N * (1 - N / pars$K) - 
-    h * (N**2 / (pars$s**2 + N**2)) + eta
+                 p) {
+  h <- np.clip(p$ho + p$alpha * (p$t_init + t), 0, 0.27)
+  N <- N + p$r * N * (1 - N / p$K) - 
+    h * (N**2 / (p$s**2 + N**2)) + eta
   N <- np.clip(N, 0, 100)
   N
 }
 # simulate
-simulate_sn <- function(N_init = 0.75,
-                     t_max = 250L,
-                     pars = list(
+simulate_sn <- function(t_max = 250L,
+                     p = list(
                        r = 1,
                        K = 1,
                        s = 0.1,
-                       h_init = .15,
+                       ho = .15,
                        alpha = 0.0015,
-                       mu = 0,
                        sigma = 0.00,
-                       t_init = 0
+                       t_init = 0,
+                       N_init = 0.75
                      )
 ) {
 
-  eta <- rnorm(t_max, pars$mu, pars$sigma) # mu = 0, no drift
+  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
   N   <- numeric(t_max)
-  N[1] <- N_init
+  N[1] <- p$N_init
   for (t in 1:(t_max-1)) {
-    N[t+1] <- step_sn(N[t], eta[t], t, pars)
+    N[t+1] <- step_sn(N[t], eta[t], t, p)
   }
-  tibble::tibble(t = pars$t_init + 1:t_max, N = N)
+  tibble::tibble(t = p$t_init + 1:t_max, N = N)
 }
 
 
@@ -58,14 +49,14 @@ greta_model_sn <- function(train) {
   r <- uniform(0, 10)
[…]

### 2022-04-24 greta/R/stochastic.R

@@ -1,0 +1,50 @@
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
+
+# View the bistable curve: 
+#curve( (0.75 - x)**2 * (0.25 - x), 0.749999, 0.750001)
+#curve(0*x,0,1,add = TRUE)
+
+#curve((0.25 - x)**2 * (0.75 - x), 0, 1)
+#curve(0*x,0,1,add = TRUE)
+
+step_stoch <- function(N, eta,  a = 0.75, b = 0.25) {
+  N <- N + (a - N)**2 * (b - N) + eta
+  N <- np.clip(N, 0, 1)
+  N
+}
+
+# simulate
+simulate_stoch <- function(N_init = 0.75,
+                     t_max = 250L,
+                     p = list(
+                       sigma = 1e-2,
+                       a = 0.75, 
+                       b = 0.25)) {
+  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
+  N   <- numeric(t_max)
+  N[1] <- N_init
+  for (t in 1:(t_max-1)) {
+    N[t+1] <- step_stoch(N[t], eta[t], a = p$a, b = p$b)
+  }
+  tibble::tibble(t = 1:t_max, N = N)
+}
+
+
+greta_model_stoch <- function(train) {
+  gsims <- train |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
+  
+  
+  x_t <- gsims$N
+  x_t1 <- gsims$xt1
+  
+  library(greta)
+  a <- uniform(0, 10)
+  b <- uniform(0, 1)
+  sigma <- uniform(0, 10)
+  mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
+  distribution(x_t1) <- normal(mean, sigma)
+  m <- model(a, b, sigma)
+  m
+}

### 2022-04-24 greta/R/utils.R

@@ -1,12 +1,24 @@
 library(tidyverse)
+library(tidyselect)
 library(memoise)
 library(greta)
 library(scoringRules)
 
+
+
+np.clip <- function(x, a, b) {
+  if(x < a) return(a)
+  if(x > b) return(b)
+  x
+}
+
 mmcmc <- memoise::memoise(greta::mcmc, cache = memoise::cache_filesystem("cache"))
 
 plot_posteriors <- function(draws, pars) {
-  true <- as_tibble(pars) |> gather(variable, value)
+  true <- as_tibble(pars) |> 
+    select(-contains("init")) |> 
+    gather(variable, value)
+  
   bind_rows(map(draws, as_tibble)) |>
     gather(variable, value) |> ggplot() + 
     geom_histogram(aes(value), bins = 30)  +
@@ -15,37 +27,47 @@ plot_posteriors <- function(draws, pars) {
 }
 
 
-forecast <- function(draws, train, test, simulate) {
+
+get_inits <- function(train, vars = "N") {
   train_t_max <- max(train$t)
   
-  # get initial conditions for forecast from final state distribution in training samples
+  # initial conditions for forecast using final state distribution in training 
   left_off <- train |> 
     filter(t == train_t_max) |> 
-    select(N) |> 
-    rename(N_init = N) |>
-    slice_sample(n = test_reps, replace=TRUE)
-  
-  # sample from posteriors
-  posterior_samples <- 
+    select(all_of(vars)) |> 
+    slice_sample(n = test_reps, replace=TRUE) 
+  names(left_off) <- paste0(names(left_off), "_init")
+  left_off |>
+    mutate(t_init = train_t_max)
+}
+
+sample_posteriors <- function(draws, inits = NULL, test_reps = 100) {  
     bind_rows(map(draws, as_tibble)) |> 
     sample_n(test_reps) |>
-    mutate(t_init = train_t_max) |>
-    bind_cols(left_off)
-  
+    bind_cols(inits)
+}
+
+forecast_dist <- function(posterior_samples, simulate) {
   # run simulate() with each row of parameters
-  posterior_sims <- 
+
     posterior_samples |>
-    mutate(mu=0) |>
     purrr::transpose() |>
     map_dfr(function(q) 
-      simulate(t_max = test_t_max, pars = q, N_init = q$N_init),
+      simulate(t_max = test_t_max, p = q),
       .id = "i")
-  
-  # Combine as single data.frame
+}
+
+
[…]

### 2022-04-24 greta/hopf_mcmc.Rmd
stoch

@@
 
-```{r message=FALSE}
+```{r setup, message=FALSE}
 library(tidyverse, quietly = TRUE)
@@
 ```{r}
-combined <- 
-  compare_forecast(draws, train, test, simulate, vars = c("H", "P")) |> 
-  pivot_longer(c(H, P), values_to="value", names_to="variable")
+combined <-  compare_forecast(draws, train, test, simulate, vars = c("H", "P"))
 
@@
 ```{r}
-scores <- bind_rows(
+scores <- 
+  bind_rows(
     rep_scores(combined, "P"),
-  rep_scores(combined, "H") |> 
+    rep_scores(combined, "H") 
+  ) |> 
   mutate(scenario="hopf_increasing", model="MCMC")
-)
 ```
@@
 ```{r}
-combined <- 
-  compare_forecast(draws2, train, test, simulate, vars = c("H", "P")) |> 
-  pivot_longer(c(H, P), values_to="value", names_to="variable")
+combined <- compare_forecast(draws2, train, test, simulate, vars = c("H", "P"))
+
 write_csv(combined, "data/hopf_decreasing.csv.gz")
@@
 ```{r}
+scores <- 
+  bind_rows(
+    rep_scores(combined, "P"),
+    rep_scores(combined, "H") 
+  ) |> 
+  mutate(scenario="hopf_decreasing", 
+         model="MCMC", 
+         reps = train_reps) |>
+  bind_rows(scores)
 
-scores <- bind_rows(
-  rep_scores(combined, "P"),
-  rep_scores(combined, "H") |> 
-  mutate(scenario="hopf_decreasing", model="MCMC") 
-) |> bind_rows(scores)
+write_csv(scores, "data/scores_hopf.csv.gz")
 

### 2022-04-24 greta/stochastic_mcmc.Rmd
stoch

@@
 
+```{r setup, message=FALSE}
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
+source("R/utils.R")
+source("R/hopf.R")
+```
+
 ```{r}
-library(tidyverse)
-library(greta)
-library(bayesplot)
-
 set.seed(4242)
@@
 test_reps <- 100
-
+simulate <- simulate_stoch
 ```
@@
 ```{r}
-
+train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
+test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
 ```
@@
 ```{r}
-
-train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
-test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
-
-## show training data
-train |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.2)
+m <- greta_model_stoch(train)
+bench::bench_time({                 
+  draws <- mmcmc(m, n_samples = 60000, warmup = 50000,
+                 chains = 4, verbose = FALSE)
+})
 ```
@@
 ```{r}
-
-m <- greta_model_stoch(train)
+## draw test_reps number of samples
+combined <- compare_forecast(draws, train, test, simulate, vars = "N") 
+write_csv(combined, "data/stochastic.csv.gz")
 ```
@@
 ```{r}
-mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))
+scores <-
+  rep_scores(combined, "N") |> 
+  mutate(scenario="stochastic", 
+         model="MCMC", 
+         reps = train_reps)
 
-bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 60000, warmup = 50000, chains = 4, verbose = FALSE)
-})
+write_csv(scores, "data/scores_stochastic.csv.gz")
+
 ```
@@
 ```{r}
-## draw test_reps number of samples
-
-posterior_samples <- 
-  bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(test_reps)
-posterior_sims <- posterior_samples %>%
-  purrr::transpose() %>%
-  map_dfr(function(q) simulate(t_max = test_t_max, p = q) ,.id = "i")
-
-bind_rows(
-  mutate(test, model="true"), 
-  mutate(posterior_sims, model="predicted")
-)|> 
-  ggplot(aes(t, N, col=model, group=interaction(model,i))) +
-           geom_line(alpha=0.1)
-
+combined |> 
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+ ggplot(aes(t, col=type)) + 
+  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_line(aes(y=mean)) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
 ```
@@
 ```{r}
-true <- tibble(a = 0.75, b = 0.25, sigma = 1e-2) |> gather(variable, value)
-bind_rows(map(draws, as_tibble)) |>
-  gather(variable, value) |> ggplot() + 
-  geom_histogram(aes(value), bins = 30)  +
-  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
-  facet_wrap(~variable, scales = "free")
+single <- 
+  combined |> 
+  filter(i %in% 1:train_reps) |> 
+  ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
+    geom_line() +
+    geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
 ```
-
-# Scoring
-
-First, a single 'observed' sample from the true:
-
-```{r}
-library(scoringRules)
-
-scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
-  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
-
-}
-# ensemble predictions
-dat <- 
-  posterior_sims |> 
-  pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
-  select(-t) |> as.matrix()
-
-```
-```{r}
-# score over all replicates:
-bench::bench_time({
-rep_scores <- 
-  test |> 
-  group_by(i) |> 
-  group_map(~ scores(.x$N, dat)) |> 
-  bind_rows()
-})
-```
-
-```{r}
-rep_scores |> summarise(across(.fns= base::mean))
-
-library(patchwork)
-ggplot(rep_scores) + geom_histogram(aes(crps), bins=20) + 
-(ggplot(rep_scores) + geom_histogram(aes(logs), bins=20) )
-
-```

### 2022-04-24 greta/R/stochastic.R

@@ -16,15 +16,15 @@ step_stoch <- function(N, eta,  a = 0.75, b = 0.25) {
 }
 
 # simulate
-simulate_stoch <- function(N_init = 0.75,
-                     t_max = 250L,
-                     p = list(
-                       sigma = 1e-2,
-                       a = 0.75, 
-                       b = 0.25)) {
+simulate_stoch <- function(t_max = 250L,
+                           p = list(
+                             sigma = 1e-2,
+                             a = 0.75, 
+                             b = 0.25,
+                             N_init = 0.75)) {
   eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
   N   <- numeric(t_max)
-  N[1] <- N_init
+  N[1] <- p$N_init
   for (t in 1:(t_max-1)) {
     N[t+1] <- step_stoch(N[t], eta[t], a = p$a, b = p$b)
   }

### 2022-04-24 greta/R/utils.R

@@ -67,19 +67,19 @@ compare_forecast <- function(draws, train, test, simulate, vars, test_reps = 100
   bind_rows(
     mutate(train, type="historical"),
     mutate(test, type="true"), 
-    mutate(posterior_sims, type="predicted"))
+    mutate(posterior_sims, type="predicted")) |>
+    pivot_longer(vars, values_to="value", names_to="variable")
 }
 
 
 
 
-scores <- function(observed, dat) {
+compute_scores <- function(observed, dat) {
   logsscore <- scoringRules::logs_sample(observed, dat)
   crpsscore <- scoringRules::crps_sample(observed, dat)
   data.frame(logs = logsscore[-1], crps =  crpsscore[-1])
   
 }
-
 
 rep_scores <- function(combined, var) {
   obs <- 
@@ -94,7 +94,7 @@ rep_scores <- function(combined, var) {
   combined |> 
     filter(type == "true", variable == var) |> 
     group_by(i) |> 
-    group_modify(~ scores(.x$value, obs)) |>
+    group_modify(~ compute_scores(.x$value, obs)) |>
     mutate(variable = var)
   
 }

### 2022-04-24 greta/hopf_mcmc.Rmd
set test_t_max

@@
 ```{r}
-combined <-  compare_forecast(draws, train, test, simulate, vars = c("H", "P"))
+combined <-  compare_forecast(draws, train, test, simulate, vars = c("H", "P"),
+                              test_reps, test_t_max)
 
@@
 ```{r}
-combined <- compare_forecast(draws2, train, test, simulate, vars = c("H", "P"))
+combined <- compare_forecast(draws2, train, test, simulate, vars = c("H", "P"),
+                              test_reps, test_t_max)
 

### 2022-04-24 greta/sn_mcmc.Rmd
set test_t_max

--- 2022-04-19 greta/stochastic_mcmc.Rmd
@@
 
+```{r setup, message=FALSE}
+library(tidyverse, quietly = TRUE)
+library(greta, quietly = TRUE)
+library(bayesplot, quietly = TRUE)
+source("R/utils.R")
+source("R/hopf.R")
+```
+
 ```{r}
-library(tidyverse)
-library(greta)
-library(bayesplot)
-
 set.seed(4242)
-train_reps <- 100
+train_reps <- 1
 train_t_max <- 250
@@
 test_reps <- 100
-
-np.clip <- function(x, a, b) {
-  if(x < a) return(a)
-  if(x > b) return(b)
-  x
-}
+simulate <- simulate_sn
 ```
@@
 ```{r}
-# View the bistable curve: 
-#curve( (0.75 - x)**2 * (0.25 - x), 0.749999, 0.750001)
-#curve(0*x,0,1,add = TRUE)
+pars = list(r = 1,
+            K = 1,
+            s = 0.1,
+            h_init = .15,
+            alpha = 0.0015,
+            mu = 0,
+            sigma = 0.02,
+            t_init = 0
+          )
+t_max <- train_t_max + test_t_max
 
-#curve((0.25 - x)**2 * (0.75 - x), 0, 1)
-#curve(0*x,0,1,add = TRUE)
+sim <- purrr::map_dfr(1:test_reps, 
+                      \(i) simulate(t_max=t_max, p = p),
+                      .id = "i")
+train <- sim |> filter(t <= train_t_max, i <= train_reps)
+test <- sim |> filter(t > train_t_max)
 ```
@@
 ```{r}
-step <- function(N, eta,  a = 0.75, b = 0.25) {
-  N <- N + (a - N)**2 * (b - N) + eta
-  N <- np.clip(N, 0, 1)
-  N
-}
-
-# simulate
-simulate <- function(N_init = 0.75,
-                     t_max = 1250L,
-                       p = list(
-                       sigma = 1e-2,
-                       a = 0.75, 
-                       b = 0.25)) {
-  eta <- rnorm(t_max, 0, p$sigma) # mu = 0, no drift
-  N   <- numeric(t_max)
-  N[1] <- N_init
-  for (t in 1:(t_max-1)) {
-    N[t+1] <- step(N[t], eta[t], a = p$a, b = p$b)
-  }
-  tibble::tibble(t = 1:t_max, N = N)
-}
-
-train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
-test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
-
-## show training data
-train |> ggplot(aes(t, N, group=i)) + geom_line(alpha=0.2)
-```
-
-```{r}
-gsims <- train |> group_by(i) |> mutate(xt1 = lead(N)) |> filter(t<max(t))
-
-x_t <- gsims$N
-x_t1 <- gsims$xt1
-
-library(greta)
-a <- uniform(0, 10)
-b <- uniform(0, 10)
-sigma <- uniform(0, 10)
-mean <-  x_t + (a - x_t) ^ 2 * (b - x_t)
-distribution(x_t1) <- normal(mean, sigma )
-m <- model(a, b, sigma)
-```
-
-```{r}
-mmcmc <- memoise::memoise(mcmc, cache = memoise::cache_filesystem("mcmc_cache"))
-
+m <- greta_model_stoch(train)
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 10000, warmup = 5000, chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 60000, warmup = 50000,
+                 chains = 4, verbose = FALSE)
 })
@@
 ```{r}
-bayesplot::mcmc_trace(draws)
+## draw test_reps number of samples
+combined <- compare_forecast(draws, train, test, simulate, vars = "N",
+                              test_reps, test_t_max) 
+write_csv(combined, "data/stochastic.csv.gz")
 ```
@@
 ```{r}
-## draw test_reps number of samples
+scores <-
+  rep_scores(combined, "N") |> 
+  mutate(scenario="stochastic", 
+         model="MCMC", 
+         reps = train_reps)
 
-posterior_samples <- 
-  bind_rows(map(draws, as_tibble)) %>% 
-  sample_n(test_reps)
-posterior_sims <- posterior_samples %>%
-  purrr::transpose() %>%
-  map_dfr(function(q) simulate(t_max = test_t_max, p = q) ,.id = "i")
-
-bind_rows(
-  mutate(test, model="true"), 
-  mutate(posterior_sims, model="predicted")
-)|> 
-  ggplot(aes(t, N, col=model, group=interaction(model,i))) +
-           geom_line(alpha=0.1)
+write_csv(scores, "data/scores_stochastic.csv.gz")
 
@@
 ```{r}
-true <- tibble(a = 0.75, b = 0.25, sigma = 1e-2) |> gather(variable, value)
-bind_rows(map(draws, as_tibble)) |>
-  gather(variable, value) |> ggplot() + 
-  geom_histogram(aes(value), bins = 30)  +
-  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
-  facet_wrap(~variable, scales = "free")
-```
-
-# Scoring
-
-First, a single 'observed' sample from the true:
-
-```{r}
-library(scoringRules)
-
-scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
-  data.frame(logs = mean(logsscore[-1]), crps =  mean(crpsscore[-1]))
-
-}
-# ensemble predictions
-dat <- 
-  posterior_sims |> 
-  pivot_wider(id_cols = "t", names_from="i", values_from = "N") |> 
-  select(-t) |> as.matrix()
-
-```
-```{r}
-# score over all replicates:
-bench::bench_time({
-rep_scores <- 
-  test |> 
-  group_by(i) |> 
-  group_map(~ scores(.x$N, dat)) |> 
-  bind_rows()
-})
+bayesplot::mcmc_trace(draws)
+plot_posteriors(draws, p)
 ```
@@
 ```{r}
-rep_scores |> summarise(across(.fns= base::mean))
+combined |> 
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+ ggplot(aes(t, col=type)) + 
+  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_line(aes(y=mean)) +
+  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+```
 
-library(patchwork)
-ggplot(rep_scores) + geom_histogram(aes(crps)) + 
-(ggplot(rep_scores) + geom_histogram(aes(logs)) )
-
+```{r}
+single <- 
+  combined |> 
+  filter(i %in% 1:train_reps) |> 
+  ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
+    geom_line() +
+    geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
 ```

### 2022-04-24 greta/stochastic_mcmc.Rmd
set test_t_max

@@
 ## draw test_reps number of samples
-combined <- compare_forecast(draws, train, test, simulate, vars = "N") 
+combined <- compare_forecast(draws, train, test, simulate, vars = "N",
+                              test_reps, test_t_max) 
 write_csv(combined, "data/stochastic.csv.gz")

### 2022-04-24 greta/R/utils.R

@@ -47,22 +47,21 @@ sample_posteriors <- function(draws, inits = NULL, test_reps = 100) {
     bind_cols(inits)
 }
 
-forecast_dist <- function(posterior_samples, simulate) {
+forecast_dist <- function(posterior_samples, simulate, test_t_max) {
   # run simulate() with each row of parameters
-
     posterior_samples |>
-    purrr::transpose() |>
-    map_dfr(function(q) 
-      simulate(t_max = test_t_max, p = q),
-      .id = "i")
+      purrr::transpose() |> 
+      map_dfr(\(q) simulate(t_max = test_t_max, p = q),
+                            .id = "i")
 }
 
 
-compare_forecast <- function(draws, train, test, simulate, vars, test_reps = 100) {
+compare_forecast <- function(draws, train, test, simulate, vars,
+                             test_reps = 100, test_t_max = 100) {
   inits <- get_inits(train, vars)
   posterior_sims <-
     sample_posteriors(draws, inits, test_reps=test_reps) |>
-    forecast_dist(simulate)
+    forecast_dist(simulate, test_t_max)
 
   bind_rows(
     mutate(train, type="historical"),

### 2022-04-24 greta/sn_mcmc.Rmd
:broom:

@@
 source("R/utils.R")
-source("R/hopf.R")
+source("R/saddle_node.R")
 ```
@@
 ```{r}
-pars = list(r = 1,
+p <- list(r = 1,
           K = 1,
           s = 0.1,
-            h_init = .15,
+          ho = .15,
           alpha = 0.0015,
@@
           sigma = 0.02,
-            t_init = 0
+          t_init = 0,
+          N_init = 0.75
         )
@@
 ```{r}
-m <- greta_model_stoch(train)
+m <- greta_model_sn(train)
 bench::bench_time({                 

### 2022-04-24 greta/stochastic_mcmc.Rmd
:broom:

@@
 source("R/utils.R")
-source("R/hopf.R")
+source("R/stochastic.R")
 ```

### 2022-04-26 greta/R/utils.R
fix stochastic case

@@ -12,7 +12,8 @@ np.clip <- function(x, a, b) {
   x
 }
 
-mmcmc <- memoise::memoise(greta::mcmc, cache = memoise::cache_filesystem("cache"))
+mmcmc <- memoise::memoise(greta::mcmc,
+                          cache = memoise::cache_filesystem("cache"))
 
 plot_posteriors <- function(draws, pars) {
   true <- as_tibble(pars) |> 
@@ -56,9 +57,15 @@ forecast_dist <- function(posterior_samples, simulate, test_t_max) {
 }
 
 
-compare_forecast <- function(draws, train, test, simulate, vars,
-                             test_reps = 100, test_t_max = 100) {
-  inits <- get_inits(train, vars)
+compare_forecast <- function(draws,
+                             train,
+                             test,
+                             simulate,
+                             vars,
+                             test_reps = 100,
+                             test_t_max = 100,
+                             inits = get_inits(train, vars)) {
+
   posterior_sims <-
     sample_posteriors(draws, inits, test_reps=test_reps) |>
     forecast_dist(simulate, test_t_max)

### 2022-04-26 greta/R/utils.R
bugfix

@@ -28,7 +28,7 @@ plot_posteriors <- function(draws, pars) {
 
 
 
-get_inits <- function(train, vars = "N") {
+get_inits <- function(train, vars = "N", test_reps = 100) {
   train_t_max <- max(train$t)
   
   # initial conditions for forecast using final state distribution in training 
@@ -58,7 +58,7 @@ forecast_dist <- function(posterior_samples, simulate, test_t_max) {
 
 compare_forecast <- function(draws, train, test, simulate, vars,
                              test_reps = 100, test_t_max = 100) {
-  inits <- get_inits(train, vars)
+  inits <- get_inits(train, vars, test_reps = 100)
   posterior_sims <-
     sample_posteriors(draws, inits, test_reps=test_reps) |>
     forecast_dist(simulate, test_t_max)

### 2022-04-26 greta/sn_mcmc_with.Rmd

---
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse, quietly = TRUE)
library(greta, quietly = TRUE)
library(bayesplot, quietly = TRUE)
source("R/utils.R")
source("R/saddle_node.R")
```

```{r}
set.seed(4242)
train_reps <- 100
train_t_max <- 250
test_t_max <- 250
test_reps <- 100
simulate <- simulate_sn
```

```{r}
p <- list(r = 1,
          K = 1,
          s = 0.1,
          ho = .15,
          alpha = 0.0015/4,
          sigma = 0.02,
          t_init = 0,
          N_init = 0.75
        )
train <- purrr::map_dfr(1:train_reps, 
                      \(i) simulate(t_max=train_t_max, p = p),
                      .id = "i")
test <- purrr::map_dfr(1:test_reps, 
                      \(i) simulate(t_max=test_t_max, p = p),
                      .id = "i")

```

```{r}
m <- greta_model_sn(train)
bench::bench_time({                 
  draws <- mmcmc(m, n_samples = 700000, warmup = 600000,
                 chains = 6, verbose = FALSE)
})
```

```{r}
## draw test_reps number of samples
combined <- compare_forecast(draws, train, test, simulate, vars = "N",
                              test_reps, test_t_max)
write_csv(combined, "data/stochastic.csv.gz")
```

```{r}
scores <-
  rep_scores(combined, "N") |> 
  mutate(scenario="stochastic", 
         model="MCMC", 
         reps = train_reps)

write_csv(scores, "data/scores_stochastic.csv.gz")

```

```{r}
bayesplot::mcmc_trace(draws)
plot_posteriors(draws, p)
```

```{r}
combined |> 
  group_by(t,type,variable) |> 
  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
 ggplot(aes(t, col=type)) + 
  geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
  geom_line(aes(y=mean)) +
  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
```

```{r}
  combined |> 
  filter(i %in% 1:train_reps) |> 
  ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
    geom_line() +
    geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
```

```{r}
n <- test_reps*10
prior_sims <- tibble(
    r = runif(n, 0, 2),
    K = runif(n, 0, 2), # runif(n, 0, 2),
    s = runif(n, 0, 0.2),
    ho = runif(n, 0, 0.2), # rlnorm(n, 0,1),
    alpha = runif(n, 0, 0.001),
    sigma = runif(n, 0, 0.1) 
    ) |>
  bind_cols(get_inits(train, "N", n)) |> 
  forecast_dist(simulate_sn, test_t_max) |>
  mutate(type = "prior") |>
  pivot_longer("N", values_to="value", names_to="variable")
prior_sims |>
  ggplot(aes(t, value, group=i)) + geom_line(alpha=0.01)
```

### 2022-04-26 greta/stochastic_mcmc.Rmd

@@
 set.seed(4242)
-train_reps <- 1
+train_reps <- 10
 train_t_max <- 250
@@
 ```{r}
-train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max), .id = "i")
-test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max), .id = "i")
+p = list(sigma = 1e-2,
+         a = 0.75, 
+         b = 0.25,
+         N_init = 0.75)
+train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max, p), .id = "i")
+test <- purrr::map_dfr(1:test_reps, \(i) simulate(t_max=test_t_max, p), .id = "i")
 ```
@@
 ## draw test_reps number of samples
+inits <- data.frame(N_init=p$N_init, t_init = 0)
 combined <- compare_forecast(draws, train, test, simulate, vars = "N",
-                              test_reps, test_t_max) 
+                              test_reps, test_t_max, inits = inits
+                             ) 
 write_csv(combined, "data/stochastic.csv.gz")
@@
 ```{r}
+ribbon <- 
 combined |> 
@@
  ggplot(aes(t, col=type)) + 
-  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
   geom_line(aes(y=mean)) +
   geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+#ribbon
 ```
@@
 ```{r}
-single <- 
 combined |> 
-  filter(i %in% 1:train_reps) |> 
+  filter(i %in% sample(i, 30)) |> 
   ggplot(aes(t, value, col=type, group=interaction(i,type))) + 
-    geom_line() +
+    geom_line(alpha=.30) +
     geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+
 ```

### 2022-04-26 greta/sn_mcmc_with.Rmd
more saddle-node runs:
There are now 2 SN cases: "saddlenode-with" which trains on data with transitions (and forecasts thus start at state 0), and "saddlenode", which attempts the 'true forecast'.

@@
 set.seed(4242)
-train_reps <- 100
+train_reps <- 10
 train_t_max <- 250
@@
           ho = .15,
-          alpha = 0.0015/4,
+          alpha = 0.0015/2,
           sigma = 0.02,
@@
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 700000, warmup = 600000,
+  draws <- mmcmc(m, n_samples = 10000, warmup = 9000,
                  chains = 6, verbose = FALSE)
@@
 ```{r}
-## draw test_reps number of samples
+## compare to simulations starting at the initial conditions 
+# instead of continuing forward from training data
+inits <- data.frame(N_init=p$N_init, t_init = 0)
 combined <- compare_forecast(draws, train, test, simulate, vars = "N",
-                              test_reps, test_t_max)
-write_csv(combined, "data/stochastic.csv.gz")
+                              test_reps, test_t_max, inits = inits
+                             )
+
+write_csv(combined, "data/saddlenode_with.csv.gz")
 ```
@@
   rep_scores(combined, "N") |> 
-  mutate(scenario="stochastic", 
+  mutate(scenario="saddle-node", 
          model="MCMC", 
@@
 
-write_csv(scores, "data/scores_stochastic.csv.gz")
+write_csv(scores, "data/scores_saddlenode_with.csv.gz")
 
@@
   geom_line(aes(y=mean)) +
-  geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
+  facet_wrap(~variable, ncol=1)
 ```

### 2022-04-26 manuscript/manuscript.Rmd
more saddle-node runs:
There are now 2 SN cases: "saddlenode-with" which trains on data with transitions (and forecasts thus start at state 0), and "saddlenode", which attempts the 'true forecast'.

@@
 abstract: |
-  Lorem Ipsum whatever
+  1. Meaningful representations of uncertainty are an essential component of effective ecological forecasting.
+  2. Machine learning
+  3. We examine scenarios in which stochasticity and non-linearity can lead to extreme uncertainty under common ecological models.  
+  4. We show that ..
 
@@
 
+Uncertainty [@Schindler2015]
+
 # Materials and Methods

### 2022-04-26 greta/sn_mcmc.Rmd
rerun sn

@@
 set.seed(4242)
-train_reps <- 1
+train_reps <- 100
 train_t_max <- 250
@@
           ho = .15,
-          alpha = 0.0015,
-          mu = 0,
+          alpha = 0.0015/4,
           sigma = 0.02,
@@
 test <- sim |> filter(t > train_t_max)
+
+#sim |> ggplot(aes(t, N, group=i))+geom_line(alpha = 0.1) + geom_vline((aes(xintercept=train_t_max))) 
 ```
@@
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 60000, warmup = 50000,
-                 chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 700000, warmup = 600000,
+                 chains = 6, verbose = FALSE)
 })
@@
                               test_reps, test_t_max)
-write_csv(combined, "data/stochastic.csv.gz")
+write_csv(combined, "data/saddlenode.csv.gz")
 ```
@@
 
-write_csv(scores, "data/scores_stochastic.csv.gz")
+write_csv(scores, "data/scores_saddlenode.csv.gz")
 
@@
  ggplot(aes(t, col=type)) + 
-  geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
   geom_line(aes(y=mean)) +
@@
 ```{r}
-single <- 
   combined |> 
@@
 ```
+
+```{r}
+n <- test_reps*10
+prior_sims <- tibble(
+    r = runif(n, 0, 2),
+    K = runif(n, 0, 2), # runif(n, 0, 2),
+    s = runif(n, 0, 0.2),
+    ho = runif(n, 0, 0.2), # rlnorm(n, 0,1),
+    alpha = runif(n, 0, 0.001),
+    sigma = runif(n, 0, 0.1) 
+    ) |>
+  bind_cols(get_inits(train, "N", n)) |> 
+  forecast_dist(simulate_sn, test_t_max) |>
+  mutate(type = "prior") |>
+  pivot_longer("N", values_to="value", names_to="variable")
+prior_sims |>
+  ggplot(aes(t, value, group=i)) + geom_line(alpha=0.01)
+```

### 2022-04-26 darts/train_ensemble.py
Removing python plotting and phew xarrays

@@ -1,4 +1,3 @@
-import pandas as pd
 import numpy as np
 import os
 import sys
@@ -13,13 +12,14 @@ from darts.models import RNNModel, BlockRNNModel, TransformerModel
 from darts.utils.likelihood_models import LaplaceLikelihood
 from utils import preprocessed_t_series, truth_dist, make_df, get_train_series, plot
 import argparse
-import yaml
 from functools import reduce
+from darts.dataprocessing.transformers import Scaler
 
+scaler = Scaler()
 parser = argparse.ArgumentParser()
 parser.add_argument(
-    "-m",
-    "--model",
+    "-f",
+    "--forecasting_model",
     default="lstm",
     type=str,
     help="model to train with (lstm/block_rnn/transformer/gru)",
@@ -51,11 +51,10 @@ parser.add_argument(
     help="Seed selection",
 )
 parser.add_argument(
-    "-t",
-    "--tp_model",
+    "--sim_model",
     default="stochastic",
     type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle/hopf)",
+    help="Select which dynamics to use for the tipping point (stochastic/saddle/hopf)",
 )
 parser.add_argument(
     "-c",
@@ -65,42 +64,54 @@ parser.add_argument(
     help="Special cases to consider (none, tipped, non_tipped, both)",
 )
 parser.add_argument(
-    "--reverse",
+    "--decrease",
     action="store_true",
     help="Flag to use decreasing Hopf model",
 )
 args = parser.parse_args()
 
-if args.model == "lstm": 
-    if args.tp_model == "stochastic":
+if args.forecasting_model == "lstm": 
+    if args.sim_model == "stochastic":
         from train_hyperparams.stochastic_lstm import hyperparameters
-    elif args.tp_model == "saddle":
+    elif args.sim_model == "saddle":
         from train_hyperparams.saddle_lstm import hyperparameters
-    elif args.tp_model == "hopf":
-        from train_hyperparams.hopf_lstm import hyperparameters
-elif args.model == "gru":
-    if args.tp_model == "stochastic":
+    elif args.sim_model == "hopf":
+        if args.decrease:
+            from train_hyperparams.hopf_decrease_lstm import hyperparameters
+        else:
+            from train_hyperparams.hopf_lstm import hyperparameters
+elif args.forecasting_model == "gru":
+    if args.sim_model == "stochastic":
         from train_hyperparams.stochastic_gru import hyperparameters
-    elif args.tp_model == "saddle":
[…]

### 2022-04-26 darts/utils.py

@@ -2,6 +2,7 @@ import sys
 sys.path.append("../")
 from model import stochastic_tp, saddle_node_tp, hopf_tp
 import numpy as np
+import pandas as pd
 from darts import TimeSeries
 from pandas import RangeIndex, DataFrame
 
@@ -11,17 +12,17 @@ models = {"stochastic" : stochastic_tp,
 
 def get_train_series(args):
     # Selecting case to pick training set
-    if args.case.lower() == "tipped" and args.tp_model == "stochastic":
+    if args.case.lower() == "tipped" and args.sim_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-    elif args.case.lower() == "nontipped" and args.tp_model == "stochastic":
+    elif args.case.lower() == "nontipped" and args.sim_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-    elif args.case.lower() == "both" and args.tp_model == "stochastic":
+    elif args.case.lower() == "both" and args.sim_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
         train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
         train_series = train_series.concatenate(train_series_, axis="sample")
     elif args.case.lower() == "none":
         # Generating time series
-        train_series = preprocessed_t_series(args.tp_model, args.n_samples, reverse=args.reverse)
+        train_series = preprocessed_t_series(args.sim_model, args.n_samples, reverse=args.decrease)
     
     return train_series
 
@@ -43,16 +44,23 @@ def preprocessed_t_series(model, n_samples, reverse=False):
             _ts[j].append(training_data[i, j])
     
     t_max = 250 if model != "hopf" else 100
-    component_dim = 1 if model!= "hopf" else 2
     
-    vals = np.array(_ts).reshape(training_data.shape[1], component_dim, n_samples)
+    if model == "hopf":
+        _ts = np.array(_ts)
+        _vals = []
+        for t_slice in _ts:
+            _vals.append(t_slice.T)
+        vals = np.array(_vals)
+    else:
+        vals = np.array(_ts).reshape(training_data.shape[1], 1, n_samples)
+        
     return TimeSeries.from_times_and_values(RangeIndex(t_max),vals)
 
[…]

### 2022-04-27 darts/StatsForecastAutoARIMA.py
Cleaning up a bit

@@ -1,4 +1,4 @@
-from darts.models import StatsForecastAutoARIMA
+from darts.models import StatsForecastAutoARIMA, KalmanForecaster, NaiveDrift
 from darts import TimeSeries
 import sys
 sys.path.append("../")
@@ -7,6 +7,7 @@ matplotlib.use("Agg")
 import matplotlib.pyplot as plt
 from utils import preprocessed_t_series, truth_dist
 import argparse
+import os
 
 parser = argparse.ArgumentParser()
 parser.add_argument(
@@ -22,6 +23,13 @@ parser.add_argument(
     type=str,
     help="Select which case for the training series",
 )
+parser.add_argument(
+    "-m",
+    "--model",
+    default="arima",
+    type=str,
+    help="Select which model to train",
+)
 args = parser.parse_args()
 
 if args.case.lower() == "tipped" and args.tp_model == "stochastic":
@@ -32,10 +40,16 @@ elif args.case.lower() == "both" and args.tp_model == "stochastic":
     train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
     train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
     train_series = train_series.concatenate(train_series_, axis="sample")
-        
+
+_model = {"arima" : StatsForecastAutoARIMA, "naive" : NaiveDrift}[args.model]
 input_len =25
 output_len=24
-for i in range(3):
+
+if not os.path.exists("plots/"):
+        os.makedirs("plots/")
+if not os.path.exists(f"plots/{args.model}"):
+        os.makedirs(f"plots/{args.model}")
+for i in range(1):
     train_series[:input_len].plot(color="blue", label='truth')
     t_series, v_series = train_series.split_before(input_len)
     t_dist = truth_dist("stochastic", t_series, input_len, output_len, n_samples=100)
@@ -43,11 +57,9 @@ for i in range(3):
     t_max = 250-input_len
     
     t_dist.plot(low_quantile=0.025, high_quantile=0.975, color="blue", label="_nolegend_", linestyle="dotted")
-    model = StatsForecastAutoARIMA().fit(t_series)
+    model = _model().fit(t_series)
     preds = model.predict(t_max, num_samples=100)
     preds.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
     
[…]

### 2022-04-27 darts/exponential_smoothing.py

@@ -1,46 +1,0 @@
-from darts.models import ExponentialSmoothing
-from darts.utils.utils import ModelMode
-import sys
-sys.path.append("../")
-import matplotlib
-matplotlib.use("Agg")
-import matplotlib.pyplot as plt
-from utils import preprocessed_t_series, truth_dist
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-t",
-    "--tp_model",
-    default="stochastic",
-    type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle)",
-)
-parser.add_argument(
-    "--random_alpha",
-    action="store_true",
-    help="Select whether to use a random alpha or not on saddle node tp",
-)
-args = parser.parse_args()
-
-train_series = preprocessed_t_series(args, 100)
-
-model = ExponentialSmoothing(trend=ModelMode.NONE, seasonal=None, seasonal_periods=0)
-
-input_len =25
-output_len=24
-for i in range(3):
-    train_series = preprocessed_t_series(args, 1)
-    train_series[:input_len].plot(color="blue", label='truth')
-    t_series, v_series = train_series.split_before(input_len)
-    t_dist = truth_dist("stochastic", t_series, input_len, output_len, n_samples=100, random_alpha=args.random_alpha)
-    
-    t_max = 250-input_len
-    
-    t_dist.plot(low_quantile=0.025, high_quantile=0.975, color="blue", label="_nolegend_", linestyle="dotted")
-    preds = model.historical_forecasts(train_series, num_samples=1000, start=3, forecast_horizon=24, retrain=True)
-    preds.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
-    train_series.plot(color="blue", label="_nolegend_")
-
-    plt.savefig(f"plots/es_test_{i}")
-    plt.clf()

### 2022-04-27 darts/kalman_filter.py

@@ -1,45 +1,0 @@
-from darts.models import KalmanFilter
-import sys
-sys.path.append("../")
-import matplotlib
-matplotlib.use("Agg")
-import matplotlib.pyplot as plt
-from utils import preprocessed_t_series, truth_dist
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-t",
-    "--tp_model",
-    default="stochastic",
-    type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle)",
-)
-parser.add_argument(
-    "--random_alpha",
-    action="store_true",
-    help="Select whether to use a random alpha or not on saddle node tp",
-)
-args = parser.parse_args()
-
-train_series = preprocessed_t_series(args, 1)
-model = KalmanFilter(dim_x=1)
-model.fit(train_series)
-
-input_len =25
-output_len=24
-for i in range(3):
-    train_series = preprocessed_t_series(args, 1)
-    train_series[:input_len].plot(color="blue", label='truth')
-    t_series, v_series = train_series.split_before(input_len)
-    t_dist = truth_dist("stochastic", t_series, input_len, output_len, n_samples=100, random_alpha=args.random_alpha)
-    
-    t_max = 250-input_len
-    
-    t_dist.plot(low_quantile=0.025, high_quantile=0.975, color="blue", label="_nolegend_", linestyle="dotted")
-    preds = model.filter(train_series, num_samples=100)
-    preds.plot(low_quantile=0.025, high_quantile=0.975, linestyle="dotted", color="orange", label='prediction')
-    train_series.plot(color="blue", label="_nolegend_")
-
-    plt.savefig(f"plots/kalman_filter_{i}")
-    plt.clf()

### 2022-04-27 darts/scoring.py

@@ -1,32 +1,0 @@
-import properscoring as ps
-import pandas as pd
-import numpy as np
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-f",
-    "--file",
-    default="block_rnn_saddle_1000.csv.gz",
-    type=str,
-    help="file from forecasts to score",
-)
-args = parser.parse_args()
-
-df = pd.read_csv(f"forecasts/{args.file}", index_col=0)
-scores = []
-for i in range(5):
-    iter_df = df[df["iter"] == i]
-    emp_df = iter_df[iter_df["true_model"] == True]
-    forecast_df = iter_df[iter_df["true_model"] == False]
-    iter_scores = []
-    
-    for t in emp_df["time"].unique():
-        observation = emp_df[emp_df["time"] == t]["value"].to_numpy()
-        forecasts = forecast_df[forecast_df["time"] == t]["value"].to_numpy()
-        for forecast in forecasts:
-            score = ps.crps_ensemble(forecast, observation)
-            iter_scores.append(score)
-    scores.append(iter_scores)
-scores = np.array(scores).flatten().mean()
-print(score)

### 2022-04-27 darts/train.py

@@ -1,111 +1,0 @@
-import pandas as pd
-import numpy as np
-import os
-import sys
-sys.path.append("../")
-import matplotlib
-matplotlib.use("Agg")
-import matplotlib.pyplot as plt
-from darts import TimeSeries
-import torch
-from darts.models import RNNModel, TCNModel, TransformerModel
-from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist
-import argparse
-import yaml
-
-parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-m",
-    "--model",
-    default="lstm",
-    type=str,
-    help="model to train with (lstm, tcn or transformer)",
-)
-parser.add_argument(
-    "-s",
-    "--n_samples",
-    default=1000,
-    type=int,
-    help="# of samples to train on",
-)
-parser.add_argument(
-    "-o",
-    "--output_file_name",
-    default="trash",
-    type=str,
-    help="File name of plots",
-)
-parser.add_argument(
-    "-p",
-    "--plot",
-    action="store_true",
-    help="Flag whether to plot",
-)
-parser.add_argument(
-    "--seed",
-    default=42,
-    type=int,
-    help="Seed selection",
-)
-parser.add_argument(
-    "-t",
-    "--tp_model",
-    default="stochastic",
-    type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle)",
-)
-parser.add_argument(
-    "--random_alpha",
-    action="store_true",
-    help="Select whether to use a random alpha or not on saddle node tp",
-)
-args = parser.parse_args()
-np.random.seed(args.seed)
-
-if args.model == "lstm":
-    from train_hyperparams.lstm import hyperparameters
-elif args.model == "tcn":
-    from train_hyperparams.tcn import hyperparameters
-elif args.model == "transformer":
-    from train_hyperparams.transformer import hyperparameters
-
-hyperparameters["random_state"] = args.seed
-
-model = {"lstm" : RNNModel, "tcn" : TCNModel, "transformer" : TransformerModel}[args.model.lower()]
-# Generating time series
-train_series, _ = preprocessed_t_series(args.tp_model, 1, args.random_alpha)
-
-my_model = model(
-    likelihood=LaplaceLikelihood(),
-    **hyperparameters
-)
-
-my_model.fit(
-    train_series,
-    verbose=True,
-)
-
-# Generating time series
-if args.plot:
[…]

### 2022-04-27 darts/train_ensemble.py
Small changes

@@ -45,12 +45,6 @@ parser.add_argument(
     help="Flag whether to plot and save a csv of samples",
 )
 parser.add_argument(
-    "--seed",
-    default=42,
-    type=int,
-    help="Seed selection",
-)
-parser.add_argument(
     "--sim_model",
     default="stochastic",
     type=str,
@@ -121,6 +115,7 @@ for i in range(42, 47):
     train_series = get_train_series(args)
     if args.sim_model == "hopf":
         train_series = scaler.fit_transform(train_series)
+    hyperparameters["model_name"] = f"{args.forecasting_model}_{args.sim_model}_{args.n_samples}_{args.case}_{args.decrease}_{i}"
     my_model = model(
         likelihood=LaplaceLikelihood(),
         **hyperparameters

### 2022-04-27 darts/utils.py

@@ -32,6 +32,8 @@ def preprocessed_t_series(model, n_samples, reverse=False):
             _data = hopf_tp(K_init=30, delta=-0.08)
         else:
             _data = hopf_tp(K_init=14, delta=0.08)
+    elif model == "stochastic":
+        _data = saddle_node_tp(N=0.55, alpha=0, h=0.26)
     else:
         _data = models[model.lower()]()
         

### 2022-04-27 darts/StatsForecastAutoARIMA.py
Getting arima script updated

@@ -5,17 +5,17 @@ sys.path.append("../")
 import matplotlib
 matplotlib.use("Agg")
 import matplotlib.pyplot as plt
-from utils import preprocessed_t_series, truth_dist
+from utils import preprocessed_t_series, truth_dist, make_df
 import argparse
 import os
 
 parser = argparse.ArgumentParser()
 parser.add_argument(
-    "-t",
-    "--tp_model",
+    "-s",
+    "--sim_model",
     default="stochastic",
     type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle)",
+    help="Select which model to use for the tipping point (stochastic/saddle/hopf)",
 )
 parser.add_argument(
     "--case",
@@ -24,42 +24,44 @@ parser.add_argument(
     help="Select which case for the training series",
 )
 parser.add_argument(
-    "-m",
-    "--model",
+    "-f",
+    "--forecasting_model",
     default="arima",
     type=str,
     help="Select which model to train",
 )
+parser.add_argument(
+    "-o",
+    "--output_file_name",
+    default="trash",
+    type=str,
+    help="Select which filename to use",
+)
 args = parser.parse_args()
 
-if args.case.lower() == "tipped" and args.tp_model == "stochastic":
+if args.case.lower() == "tipped" and args.sim_model == "stochastic":
     train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-elif args.case.lower() == "nontipped" and args.tp_model == "stochastic":
+elif args.case.lower() == "nontipped" and args.sim_model == "stochastic":
     train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-elif args.case.lower() == "both" and args.tp_model == "stochastic":
+elif args.case.lower() == "both" and args.sim_model == "stochastic":
     train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
     train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
     train_series = train_series.concatenate(train_series_, axis="sample")
 
-_model = {"arima" : StatsForecastAutoARIMA, "naive" : NaiveDrift}[args.model]
+_model = {"arima" : StatsForecastAutoARIMA,}[args.forecasting_model]
 input_len =25
 output_len=24
+if not os.path.exists("forecasts/"):
+    os.makedirs("forecasts/")
 
[…]

### 2022-04-27 darts/find_set.py

@@ -6,7 +6,9 @@ import argparse
 flag_non = False
 flag_tip = False
 while True:
-    x, y = preprocessed_t_series("stochastic", 1)
+    x = preprocessed_t_series("stochastic", 1)
+    y = x.all_values()
+    
     if count_tipped(y) == 0:
         x.to_csv("stochastic_non_tipped.csv")
         flag_non = True

### 2022-04-27 darts/utils.py

@@ -6,7 +6,7 @@ import pandas as pd
 from darts import TimeSeries
 from pandas import RangeIndex, DataFrame
 
-models = {"stochastic" : stochastic_tp, 
+models = { 
           "saddle" : saddle_node_tp,
           }
 
@@ -72,6 +72,8 @@ def truth_dist(model, t_series, input_len, output_len, n_draws=100, reverse=Fals
             delta=0.08
             K_init = 14 + delta * t
             _data = hopf_tp(N_init, t_max, K_init=K_init, delta=delta)
+    elif model == "stochastic":
+        _data = saddle_node_tp(N=N_init, t_max=t_max, alpha=0, h=0.26)
     else:
         _data = models[model.lower()](N_init[0], t_max)
     
@@ -99,16 +101,16 @@ def truth_dist(model, t_series, input_len, output_len, n_draws=100, reverse=Fals
         vals = np.array(_ts).reshape(training_data.shape[1], 1, n_draws)
         
     stop = 250 if model != "hopf" else 200
-    
+    import pdb; pdb.set_trace()
     return TimeSeries.from_times_and_values(RangeIndex(start=start_t, stop=stop), vals)
   
 def count_tipped(vals):
-    n_vals = len(vals)
+    n_samples = vals.shape[2]
     count = 0
-    for i in range(n_vals):
-        if vals[i, -1] < 0.6:
+    for i in range(n_samples):
+        if vals[-1, 0, i] < 0.3:
             count += 1
-    return count / n_vals
+    return count / n_samples
 
 def import_hypers(tp_model, ml_model):
     if ml_model == "lstm": 
@@ -193,14 +195,6 @@ def make_df(prediction_series, truth_series, historical_series, tp_model, ml_mod
     
     return df
     
-def plot(tp_model, series, color1="blue", color2="yellow", prediction=False):
-    label = "_nolegend_" if not prediction else "prediction"
-    if tp_model == "hopf":
-        x, y = series.univariate_component(0), series.univariate_component(1)
-        x.plot(low_quantile=0.025, high_quantile=0.975, color=color1, label=label, linestyle="dotted")
-        y.plot(low_quantile=0.025, high_quantile=0.975, color=color2, label=label, linestyle="dotted")
-    else:
-        series.plot(low_quantile=0.025, high_quantile=0.975, color=color1, label=label, linestyle="dotted")
     
 def reshape_xarray(_ts):
     _vals = []

### 2022-04-27 darts/find_set.py
More small changes

@@ -2,20 +2,32 @@ import numpy as np
 from utils import preprocessed_t_series, count_tipped
 import argparse
 
+parser = argparse.ArgumentParser()
+parser.add_argument(
+    "-s",
+    "--sim_model",
+    default="stochastic",
+    type=str,
+    help="Select which model to use for the tipping point (stochastic/saddle/hopf)",
+)
+args = parser.parse_args()
 
-flag_non = False
-flag_tip = False
-while True:
-    x = preprocessed_t_series("stochastic", 1)
-    y = x.all_values()
+if args.sim_model == "stochastic":
+    flag_non = False
+    flag_tip = False
+    while True:
+        series = preprocessed_t_series(args.sim_model, 1)
+        vals = series.all_values()
+        
+        if count_tipped(vals) == 0:
+            series.to_csv(f"{args.sim_model}_nontipped.csv")
+            flag_non = True
     
-    if count_tipped(y) == 0:
-        x.to_csv("stochastic_non_tipped.csv")
-        flag_non = True
-
-    if count_tipped(y) == 1:
-        x.to_csv("stochastic_tipped.csv")
-        flag_tip = True
+        if count_tipped(vals) == 1:
+            series.to_csv(f"{args.sim_model}_tipped.csv")
+            flag_tip = True
+        
+        if flag_non and flag_tip:
+            break
     
-    if flag_non and flag_tip:
-        break
+    

### 2022-04-27 darts/train_ensemble.py

@@ -168,5 +168,3 @@ if args.evaluate:
         final_df = final_df.append(df, ignore_index=True)
         
     final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)
-
-

### 2022-04-27 darts/utils.py

@@ -26,7 +26,7 @@ def get_train_series(args):
     
     return train_series
 
-def preprocessed_t_series(model, n_samples, reverse=False):
+def preprocessed_t_series(model, n_samples, reverse=False, saddle_tip=False):
     if model == "hopf":
         if reverse:
             _data = hopf_tp(K_init=30, delta=-0.08)
@@ -34,8 +34,12 @@ def preprocessed_t_series(model, n_samples, reverse=False):
             _data = hopf_tp(K_init=14, delta=0.08)
     elif model == "stochastic":
         _data = saddle_node_tp(N=0.55, alpha=0, h=0.26)
-    else:
-        _data = models[model.lower()]()
+    elif modle == "saddle":
+        if saddle_tip:
+            _data = saddle_node_tp(alpha=.0015/2)
+        else:
+            _data = saddle_node_tp(alpha=.0015/4)
+
         
     training_data = _data.collect_samples(n_samples)
     # Note training_data.shape[1] is the length of the time series
@@ -104,11 +108,11 @@ def truth_dist(model, t_series, input_len, output_len, n_draws=100, reverse=Fals
     import pdb; pdb.set_trace()
     return TimeSeries.from_times_and_values(RangeIndex(start=start_t, stop=stop), vals)
   
-def count_tipped(vals):
+def count_tipped(vals, threshold=0.3):
     n_samples = vals.shape[2]
     count = 0
     for i in range(n_samples):
-        if vals[-1, 0, i] < 0.3:
+        if vals[-1, 0, i] < threshold:
             count += 1
     return count / n_samples
 

### 2022-04-27 greta/stochastic_mcmc.Rmd

@@
 source("R/utils.R")
-source("R/stochastic.R")
+source("R/saddle_node.R")
 ```
@@
 set.seed(4242)
-train_reps <- 10
+train_reps <- 1
 train_t_max <- 250
@@
 test_reps <- 100
-simulate <- simulate_stoch
+simulate <- simulate_sn
 ```
@@
 ```{r}
-p = list(sigma = 1e-2,
-         a = 0.75, 
-         b = 0.25,
-         N_init = 0.75)
+p <- list(r = 1,
+          K = 1,
+          s = 0.1,
+          ho = .26,
+          alpha = 0.0,
+          sigma = 0.02,
+          t_init = 0,
+          N_init = 0.55
+        )
 train <- purrr::map_dfr(1:train_reps, \(i) simulate(t_max=train_t_max, p), .id = "i")
@@
 ```{r}
-m <- greta_model_stoch(train)
+ggplot(train, aes(t, N, group=i)) + geom_line(alpha=0.1)
+```
+
+```{r}
+m <- greta_model_sn(train, 
+                    r     = uniform(0, 2),
+                    K     = uniform(0, 2),
+                    s     = uniform(0.0, 1),
+                    ho    = uniform(0, 1),
+                    alpha = uniform(0,0.00001),
+                    sigma = uniform(0, 0.05))
 bench::bench_time({                 
-  draws <- mmcmc(m, n_samples = 60000, warmup = 50000,
-                 chains = 4, verbose = FALSE)
+  draws <- mmcmc(m, n_samples = 90000, warmup = 70000,
+                 chains = 8, verbose = FALSE)
 })
@@
   geom_vline(aes(xintercept = train_t_max)) + facet_wrap(~variable, ncol=1)
-#ribbon
+ribbon
 ```

### 2022-04-27 darts/train_ensemble.py
Initial checks looked good here

@@ -10,7 +10,7 @@ from pandas import DataFrame
 import torch
 from darts.models import RNNModel, BlockRNNModel, TransformerModel
 from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist, make_df, get_train_series, plot
+from utils import preprocessed_t_series, truth_dist, make_df, get_train_series
 import argparse
 from functools import reduce
 from darts.dataprocessing.transformers import Scaler
@@ -27,7 +27,7 @@ parser.add_argument(
 parser.add_argument(
     "-s",
     "--n_samples",
-    default=100,
+    default=10,
     type=int,
     help="# of samples to train on",
 )
@@ -49,6 +49,12 @@ parser.add_argument(
     default="stochastic",
     type=str,
     help="Select which dynamics to use for the tipping point (stochastic/saddle/hopf)",
+)
+parser.add_argument(
+    "--tipped",
+    default=True,
+    type=bool,
+    help="Whether to use the tipped or nontipped training set",
 )
 parser.add_argument(
     "-c",
@@ -111,11 +117,11 @@ models = []
 for i in range(42, 47):
     hyperparameters["random_state"] = i
     np.random.seed(i)
-    
+    import pdb; pdb.set_trace()
     train_series = get_train_series(args)
     if args.sim_model == "hopf":
         train_series = scaler.fit_transform(train_series)
-    hyperparameters["model_name"] = f"{args.forecasting_model}_{args.sim_model}_{args.n_samples}_{args.case}_{args.decrease}_{i}"
+    hyperparameters["model_name"] = f"{args.forecasting_model}_{args.sim_model}_{args.n_samples}_{args.case}_{args.decrease}_{args.tipped}_{i}"
     my_model = model(
         likelihood=LaplaceLikelihood(),
         **hyperparameters

### 2022-04-27 darts/utils.py

@@ -12,21 +12,19 @@ models = {
 
 def get_train_series(args):
     # Selecting case to pick training set
-    if args.case.lower() == "tipped" and args.sim_model == "stochastic":
+    if args.tipped and args.sim_model == "stochastic":
         train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-    elif args.case.lower() == "nontipped" and args.sim_model == "stochastic":
-        train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-    elif args.case.lower() == "both" and args.sim_model == "stochastic":
-        train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-        train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-        train_series = train_series.concatenate(train_series_, axis="sample")
-    elif args.case.lower() == "none":
+    elif not args.tipped and args.sim_model == "stochastic":
+        train_series = TimeSeries.from_csv("stochastic_nontipped.csv", time_col="time")
+    elif args.sim_model == "saddle":
+        train_series =  preprocessed_t_series(args.sim_model, args.n_samples, tipped=args.tipped)
+    else: # Hopf
         # Generating time series
         train_series = preprocessed_t_series(args.sim_model, args.n_samples, reverse=args.decrease)
     
     return train_series
 
-def preprocessed_t_series(model, n_samples, reverse=False, saddle_tip=False):
+def preprocessed_t_series(model, n_samples, reverse=False, tipped=False):
     if model == "hopf":
         if reverse:
             _data = hopf_tp(K_init=30, delta=-0.08)
@@ -34,8 +32,8 @@ def preprocessed_t_series(model, n_samples, reverse=False, saddle_tip=False):
             _data = hopf_tp(K_init=14, delta=0.08)
     elif model == "stochastic":
         _data = saddle_node_tp(N=0.55, alpha=0, h=0.26)
-    elif modle == "saddle":
-        if saddle_tip:
+    elif model == "saddle":
+        if tipped:
             _data = saddle_node_tp(alpha=.0015/2)
         else:
             _data = saddle_node_tp(alpha=.0015/4)

### 2022-04-28 manuscript/manuscript.Rmd
some text :pen:

@@
 ---
+
+```{r setup}
+library(tidyverse)
+library(patchwork)
+```
+
 # Introduction
 
-Uncertainty [@Schindler2015]
+Uncertainty important to forecasting [@Deitze2018] and  decision-making [@Schindler2015]
+
+Machine learning is increasingly effective at forecasting, but methods often focus on creation or assessment only of point predictions [@]. 
+Because uncertainty can play such an intrinsic role in ecological dynamics [@Boettiger2018; @Coulson2004; @Oveskienen], it is essential that machine learning applications to ecological forecasting are able to address probablistic forecasts adequately.
+We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
 
 # Materials and Methods
+
+<!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
+We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
+Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
+While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [].
+These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
+May's model exhibits alternative stable states. 
+In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis.
+The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.
+
+Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
+Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
+This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date.
+No-analog scenarios [@Williams]
+
+```{r fig.cap="Forecast scenarios. A. a Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation."}
+hopf <- read_csv("../greta/data/hopf_increasing.csv.gz")
+panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
+  rename(species = variable) |> 
+  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
+  ggplot(aes(t, value, group=interaction(species, i, type), col=species)) + 
+  geom_line(show.legend = FALSE, alpha = 0.7) + theme_bw()+  scale_color_grey() +
+  ggtitle("A. Hopf bifurcation")
+
+sn <- read_csv("../greta/data/saddlenode.csv.gz")
+panelB <- sn |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
+  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
+  geom_path(alpha=0.5, show.legend = FALSE) +
+  scale_color_grey(end=0.6) + theme_bw() + 
+  ggtitle("B. Saddle-node bifurcation")
+
+stoch <- read_csv("../greta/data/stochastic.csv.gz")
+panelC <- stoch |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
+  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
+  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
+  ggtitle("C. Stochastic transition")
+
+panelA / panelB / panelC
+
+```
+
+## Scenario 1: Hopf Bifurcation
+
+The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, orignally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@Nicholson1935; @Nicholson1954a; @Nicholson1954b; @Hastings].
+We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,
+
+\begin{equation}
+\label{hopf}
+\end{equation}
+
+Following @Dakos2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, $K = K_0 + \delta t$, which drives a Hopf bifurcation as $K$ becomes sufficiently large.
+In a Hopf bifurcation, a stable node starts an oscillatory pattern, which continues to grow in amplitude as the bifurcation parameter continues to increase.
+This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
+If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
+However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@Boettiger2018].
+
+## Scenario 2: The saddle-node bifurcation
+
+A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
+May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@May1979].
+As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
+Meanwhile, the resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time).
+
+## Scenario 3: The stochastic transition
+
+Hardest of all events to predict are those transitions which are predominately driven by a random component. 
+An example of such a transition event is possible to observe in May's consumer-resource model, in which a stochastic term occasionally results in a transition between alternative stable states.
+In such cases, no forecast can precisely predict when a transition will occur, but it is nonetheless possible to deduce the correct distribution of waiting times knowing the correct model. 
+In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g.
+@vanKampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorsfallacy].
+
+## Method group 1: Markov Chain Monte Carlo
+
+As a reference case, we consider the forecast produced by an MCMC estimate of model parameters, _given the true model_.
+This represents an idealized case where the nature of the underlying process is known precisely.
+Uncertainty comes from parameter estimates and intrinsic stochasticity specified in the model, but does not reflect any uncertainty in our knowledge of the model structure. 
+Alternative model structures, even when capable of producing the same nonlinear phenomena (i.e. the same bifurcations) will give very different forecasts. 
+Even alternative prior distributions of the parameters will generally yield alternate forecasts, as likelihood ridges are common to nonlinear models.
+Thus, this case represents a theoretical upper bound for the performance of forecasts by techniques which do not make such strong assumptions about the underlying processes.
+
+## Method group 2: Statistical models (ARIMA)
+
+## Method group 3: Machine Learning models
 

### 2022-04-28 darts/evaluate.py
Getting close to finishing evals

@@ -10,7 +10,7 @@ from pandas import DataFrame
 import torch
 from darts.models import RNNModel, BlockRNNModel, TransformerModel
 from darts.utils.likelihood_models import LaplaceLikelihood
-from utils import preprocessed_t_series, truth_dist, make_df, get_train_series, plot
+from utils import preprocessed_t_series, truth_dist, make_df, get_train_series
 import argparse
 from functools import reduce
 from darts.dataprocessing.transformers import Scaler
@@ -29,7 +29,7 @@ parser.add_argument(
     "--n_samples",
     default=100,
     type=int,
-    help="# of samples to train on",
+    help="# of samples model was trained on",
 )
 parser.add_argument(
     "-o",
@@ -39,49 +39,37 @@ parser.add_argument(
     help="File name of plots",
 )
 parser.add_argument(
-    "-e",
-    "--evaluate",
-    action="store_true",
-    help="Flag whether to plot and save a csv of samples",
-)
-parser.add_argument(
-    "--seed",
-    default=42,
-    type=int,
-    help="Seed selection",
-)
-parser.add_argument(
     "--sim_model",
     default="stochastic",
     type=str,
     help="Select which dynamics to use for the tipping point (stochastic/saddle/hopf)",
 )
 parser.add_argument(
-    "-c",
-    "--case",
-    default="none",
-    type=str,
-    help="Special cases to consider (none, tipped, non_tipped, both)",
+    "--tipped",
+    action="store_true",
+    help="Use model trained on tipped or nontipped data",
 )
 parser.add_argument(
     "--decrease",
     action="store_true",
-    help="Flag to use decreasing Hopf model",
+    help="Use model trained on decreasing or increasing hopf model",
 )
 args = parser.parse_args()
 
 
-model = {"lstm" : RNNModel, "gru" : RNNModel, "block_rnn" : BlockRNNModel, "transformer" : TransformerModel}[args.forecasting_model.lower()]
-
-model_name = f"{args.forecasting_model}_{args.sim_model}_{args.n_samples}_{args.case}_{args.decrease}_{args.i}"
-model = model.load_model(f"darts_logs/{model_name}")
+_model = {"lstm" : RNNModel, "gru" : RNNModel, "block_rnn" : BlockRNNModel, "transformer" : TransformerModel}[args.forecasting_model.lower()]
+models = []
+for i in range(42, 47):
[…]

### 2022-04-28 darts/evaluate.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+python evaluate.py -f lstm --sim_model hopf -s 10 --seed 42 -o block_hopf_10_decrease

### 2022-04-28 darts/find_set.py

@@ -1,3 +1,4 @@
+
 import numpy as np
 from utils import preprocessed_t_series, count_tipped
 import argparse
@@ -29,5 +30,3 @@ if args.sim_model == "stochastic":
         
         if flag_non and flag_tip:
             break
-    
-    

### 2022-04-28 darts/utils.py

@@ -103,7 +103,7 @@ def truth_dist(model, t_series, input_len, output_len, n_draws=100, reverse=Fals
         vals = np.array(_ts).reshape(training_data.shape[1], 1, n_draws)
         
     stop = 250 if model != "hopf" else 200
-    import pdb; pdb.set_trace()
+    
     return TimeSeries.from_times_and_values(RangeIndex(start=start_t, stop=stop), vals)
   
 def count_tipped(vals, threshold=0.3):

### 2022-04-28 manuscript/manuscript.Rmd
:pen:

@@
 
-\begin{equation}
+\begin{align}
+H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K}\right) - c P_t + \eta_{H,t}
+P_{t+1} &= H_t \exp \left( r \left(1  \tfrac{H}{K}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right)
 \label{hopf}
-\end{equation}
+\end{align}
 
@@
 As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
-Meanwhile, the resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time).
+The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
+For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range ecological and socio-ecological systems [@Scheffer2001].
+
+\begin{align}
+\end{align}
+
+If the environment slowly alters one of the parameters (say, the encounter efficiency, `a`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
+Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state.
+Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
+Due both this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have become a nearly ubiquitous [@Scheffer2009]
+
+Saddle-node b
 

### 2022-04-28 darts/evaluate.py
progress on evaluation after debugging

@@ -15,6 +15,9 @@ import argparse
 from functools import reduce
 from darts.dataprocessing.transformers import Scaler
 
+
+# NB Cannot use Hopf model here as we don't have the data scaler,
+# so have to retrain these ones to evaluate.
 scaler = Scaler()
 parser = argparse.ArgumentParser()
 parser.add_argument(
@@ -27,9 +30,9 @@ parser.add_argument(
 parser.add_argument(
     "-s",
     "--n_samples",
-    default=100,
+    default=1,
     type=int,
-    help="# of samples model was trained on",
+    help="Ignore this",
 )
 parser.add_argument(
     "-o",
@@ -77,7 +80,7 @@ for i in range(5):
     torch.manual_seed(i)
     n_draws = 100
     
-    train_series = preprocessed_t_series(args.sim_model, 1)
+    train_series = get_train_series(args)
     
     if args.sim_model == "hopf":
         t_series = train_series[-input_len:]
@@ -99,7 +102,12 @@ for i in range(5):
         ensemble_preds.append(_preds)
     ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
     
-    df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), args.case, args.n_samples, i)
+    if args.sim_model == "stochastic" or args.sim_model == "saddle":
+        case = "tipped" if args.tipped else "nontipped"
+    elif args.sim_model == "hopf":
+        case = "decrease" if args.decrease else "increase"
+    
+    df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
     final_df = final_df.append(df, ignore_index=True)
         
     final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)

### 2022-04-28 darts/evaluate.sh

@@ -1,2 +1,23 @@
 #!/bin/bash
-python evaluate.py -f lstm --sim_model hopf -s 10 --seed 42 -o block_hopf_10_decrease
+# Need to evaluate hopf from training atm! this is due to data scaler not being portable 
+# (I could write a function here to save training set but time is short rn)
+
+python evaluate.py -f lstm --sim_model saddle -s 1 -o lstm_saddle_1
+python evaluate.py -f block_rnn --sim_model saddle -s 1 -o block_saddle_1
+python evaluate.py -f gru --sim_model saddle -s 1 -o gru_saddle_1
+python evaluate.py -f transformer --sim_model saddle -s 1 -o transformer_saddle_1
+
+python evaluate.py -f lstm --sim_model saddle -s 10 -o lstm_saddle_10
+python evaluate.py -f block_rnn --sim_model saddle -s 10 -o block_saddle_10
+python evaluate.py -f gru --sim_model saddle -s 10 -o gru_saddle_10
+python evaluate.py -f transformer --sim_model saddle -s 10 -o transformer_saddle_10
+
+python evaluate.py -f lstm --sim_model stochastic -s 1 -o lstm_stochastic_1 --tipped
+python evaluate.py -f block_rnn --sim_model stochastic -s 1 -o block_stochastic_1 --tipped
+python evaluate.py -f gru --sim_model stochastic -s 1 -o gru_stochastic_1 --tipped
+python evaluate.py -f transformer --sim_model stochastic -s 1 -o transformer_stochastic_1 --tipped
+
+python evaluate.py -f lstm --sim_model stochastic -s 10 -o lstm_stochastic_10 --tipped
+python evaluate.py -f block_rnn --sim_model stochastic -s 10 -o block_stochastic_10 --tipped
+python evaluate.py -f gru --sim_model stochastic -s 10 -o gru_stochastic_10 --tipped
+python evaluate.py -f transformer --sim_model stochastic -s 10 -o transformer_stochastic_10 --tipped

### 2022-04-29 darts/evaluate.py
debugging fix hopefully

@@ -67,10 +67,9 @@ for i in range(42, 47):
     model = _model.load_model(f"darts_logs/{model_name}/_model.pth.tar")
     models.append(model)
 
-
-
 if not os.path.exists("forecasts/"):
-        os.makedirs("forecasts/")
+    os.makedirs("forecasts/")
+    
 input_len = 25
 output_len = 24
 final_df = DataFrame()
@@ -80,32 +79,23 @@ for i in range(5):
     torch.manual_seed(i)
     n_draws = 100
     
-    train_series = get_train_series(args)
+    train_series = preprocessed_t_series(args.sim_model, 1)
     
-    if args.sim_model == "hopf":
-        t_series = train_series[-input_len:]
-    else:
-        t_series = train_series[:input_len]
+    t_series = train_series[:input_len] # Will want to revisit this
     
     # truth_dist 
-    start_t = input_len if args.sim_model != "hopf" else 100
+    start_t = input_len
     t_dist = truth_dist(args.sim_model, t_series, input_len, output_len, n_draws=n_draws, reverse=args.decrease, start_t=start_t)
     
-    t_max = 250-input_len if args.sim_model != "hopf" else 100
+    t_max = 250-input_len
     
     ensemble_preds = []
     for model in models:
-        if args.sim_model == "hopf":
-            _preds = scaler.inverse_transform(model.predict(t_max, t_series, num_samples=n_draws))
-        else: 
-            _preds = model.predict(t_max, t_series, num_samples=n_draws)
+        _preds = model.predict(t_max, t_series, num_samples=n_draws)
         ensemble_preds.append(_preds)
     ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
     
-    if args.sim_model == "stochastic" or args.sim_model == "saddle":
-        case = "tipped" if args.tipped else "nontipped"
-    elif args.sim_model == "hopf":
-        case = "decrease" if args.decrease else "increase"
+    case = "tipped" if args.tipped else "nontipped"
     
     df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
     final_df = final_df.append(df, ignore_index=True)

### 2022-04-29 greta/sn_mcmc.Rmd
refreshed sn
fix typo in metadata

@@
   rep_scores(combined, "N") |> 
-  mutate(scenario="stochastic", 
+  mutate(scenario="saddle_node", 
          model="MCMC", 

### 2022-04-29 manuscript/manuscript.Rmd
:pen:

@@
 Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
-While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [].
+While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
 These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
@@
 \begin{align}
-H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K}\right) - c P_t + \eta_{H,t}
-P_{t+1} &= H_t \exp \left( r \left(1  \tfrac{H}{K}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right)
+H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) - c P_t + \eta_{H,t} \\
+P_{t+1} &= H_t \exp \left( r \left(1  \tfrac{H}{K_t}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right) \\
+K_{t+1} &= K_t + \delta
 \label{hopf}
@@
 
-Following @Dakos2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, $K = K_0 + \delta t$, which drives a Hopf bifurcation as $K$ becomes sufficiently large.
-In a Hopf bifurcation, a stable node starts an oscillatory pattern, which continues to grow in amplitude as the bifurcation parameter continues to increase.
+Following @Dakos2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
+In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
 This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
@@
 \begin{align}
+N_{t+1} &= N_t + r N_t \left(1 - \frac{N_t}{K} \right) - \frac{h_t N_t^2}{s^2 + N_t^2} + \eta_t \\
+h_{t+1} &= h_t + \alpha \\
+\eta_t &\sim \mathcal{N}(0, \sigma)
 \end{align}
 
-If the environment slowly alters one of the parameters (say, the encounter efficiency, `a`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
+If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
 Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state.
 Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
-Due both this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have become a nearly ubiquitous [@Scheffer2009]
-
-Saddle-node b
+Due both this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
+These dynamics have long been identified as an acute challenge to forecasting, with effort focused primarily on identifying any 'early warning signs' that a catastrophic bifurcation might be occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
 
@@
 
-Hardest of all events to predict are those transitions which are predominately driven by a random component. 
+Perhaps the most difficult of all events to predict are those in which large transitions are predominately driven by a random component. 
 An example of such a transition event is possible to observe in May's consumer-resource model, in which a stochastic term occasionally results in a transition between alternative stable states.
@@
 @vanKampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorsfallacy].
+To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.
+
+In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.
+Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
+Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
+we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
+Based on this data, our forecaster seeks to identify the distribution of expected transition times for analgous systems starting from the same initial condition.
+This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [].
+(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.
+In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
 
@@
 
+<!-- Key distinction here from MCMC above is not on how model parameters are estimated, but rather on approaches that aren't making strong mechanistic assumptions about process, but are merely phenomonological type models
+-->
+
 ## Method group 3: Machine Learning models
+
+<!--
+Basic typology of machine learning methods, how we distinguish "ML" from statistical methods (i.e.
+bias-variance tradeoff; no criteria to prove that the algorithm produces an unbiased estimate of a given summary statistic)
+
+Stuff about how most machine learning techniques provide only point forecasts
+-->
+
+## Forecast skill: strictly proper scores
+
+To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @Gneiting2007's property of a strictly proper score.
+This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
+In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e.
+have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
+It is worth noting that this property applies specifically to probablistic forecasts, and is not true of all metrics often used to compare forecasts.
+Concerns about over-fitting arise in most types of model estimation, are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@].   
+This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.
+
+While all strictly proper scores have this property, this does not guarantee that all such metrics will share the same relative ranking between forecasts.
+We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
+Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS
 
@@
 
+By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
+Note that observations are constrained to a smaller region of state space (i.e. a smaller range of population sizes) than will be realized in future. 
+This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.
+
+```{r}
+hopf |> 
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+  rename(species = variable) |> 
+  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
+  ggplot(aes(t, col=type)) + 
+    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+    geom_line(aes(y=mean)) +
+    #geom_vline(aes(xintercept = 100)) + 
+    facet_wrap(~species, ncol=1) +
+    theme_bw() + 
+    ggtitle("A. Hopf bifurcation")
+```
+
+```{r}
+panelB <- sn |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
+  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
+  geom_path(alpha=0.5, show.legend = FALSE) +
+  scale_color_grey(end=0.6) + theme_bw() + 
+  ggtitle("B. Saddle-node bifurcation")
+
+panelC <- stoch |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
+  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
+  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
+  ggtitle("C. Stochastic transition")
+
+panelA / panelB / panelC
+```
+
+```{r}
+mcmc_scores <- bind_rows(
+  read_csv("../greta/data/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
+read_csv("../greta/data/scores_stochastic.csv.gz"),
+read_csv("../greta/data/scores_hopf.csv.gz") |> 
+  filter(scenario == "hopf_increasing") |>
+  mutate(scenario = forcats::fct_recode(scenario, hopf = "hopf_increasing"),
+         reps = 1)
+) |> rename(method=model)
+mcmc_scores |> 
+  pivot_longer(c(logs, crps), names_to = "metric", values_to = "score") |> 
+  ggplot(aes(method, score, col=variable)) + geom_boxplot() + 
+  facet_grid(metric~scenario, scales="free_y")
+```
+
 # Discussion
+
+Ecological systems have long been acknowledged as complex, due not only to the immense span of dimension and scale such processes involve, but also the frequency of emergent and non-linear phenomena such as stochastic resonance, including bifurcations, tipping points, and hysteresis examined here.
+Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@Clark2001; @Dietze2018].
+What methods will be most reliable in the face of such uncertainty?
+
+reflects a fundamental challenge of ecological forecasting.
+
+non-linearity,
+We have compared the ability of several cutting edge machine learning approaches for forecasting against statistical and process-based models.
+
+- Ecological forecasting invariably is hard
 

### 2022-05-02 darts/evaluate.py
More additions from this wknd

@@ -32,7 +32,7 @@ parser.add_argument(
     "--n_samples",
     default=1,
     type=int,
-    help="Ignore this",
+    help="Number of samples that the model was trained on",
 )
 parser.add_argument(
     "-o",
@@ -64,7 +64,7 @@ _model = {"lstm" : RNNModel, "gru" : RNNModel, "block_rnn" : BlockRNNModel, "tra
 models = []
 for i in range(42, 47):
     model_name = f"{args.forecasting_model}_{args.sim_model}_{args.n_samples}_{args.decrease}_{args.tipped}_{i}"
-    model = _model.load_model(f"darts_logs/{model_name}/_model.pth.tar")
+    model = _model.load_from_checkpoint(f"{model_name}", file_name="last-epoch=499.ckpt")
     models.append(model)
 
 if not os.path.exists("forecasts/"):
@@ -74,14 +74,23 @@ input_len = 25
 output_len = 24
 final_df = DataFrame()
 
+if not os.path.exists("forecasts/"):
+    os.makedirs("forecasts/")
+input_len = 25
+output_len = 24
+final_df = DataFrame()
 for i in range(5):
     np.random.seed(i)
     torch.manual_seed(i)
     n_draws = 100
     
-    train_series = preprocessed_t_series(args.sim_model, 1)
+    args.n_samples = 1
+    train_series = get_train_series(args)
     
-    t_series = train_series[:input_len] # Will want to revisit this
+    if args.sim_model == "stochastic":
+        t_series = train_series[:input_len]
+    elif args.sim_model == "saddle":
+        t_series = train_series[-input_len:]
     
     # truth_dist 
     start_t = input_len
@@ -96,8 +105,8 @@ for i in range(5):
     ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
     
     case = "tipped" if args.tipped else "nontipped"
-    
+        
     df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
     final_df = final_df.append(df, ignore_index=True)
-        
-    final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)
+    
+final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)

### 2022-05-02 darts/evaluate.sh

@@ -2,22 +2,22 @@
 # Need to evaluate hopf from training atm! this is due to data scaler not being portable 
 # (I could write a function here to save training set but time is short rn)
 
-python evaluate.py -f lstm --sim_model saddle -s 1 -o lstm_saddle_1
-python evaluate.py -f block_rnn --sim_model saddle -s 1 -o block_saddle_1
-python evaluate.py -f gru --sim_model saddle -s 1 -o gru_saddle_1
-python evaluate.py -f transformer --sim_model saddle -s 1 -o transformer_saddle_1
+python evaluate.py -f lstm --sim_model saddle -s 1 -o lstm_saddle_1_nontipped
+python evaluate.py -f block_rnn --sim_model saddle -s 1 -o block_saddle_1_nontipped
+python evaluate.py -f gru --sim_model saddle -s 1 -o gru_saddle_1_nontipped
+python evaluate.py -f transformer --sim_model saddle -s 1 -o transformer_saddle_1_nontipped
 
-python evaluate.py -f lstm --sim_model saddle -s 10 -o lstm_saddle_10
-python evaluate.py -f block_rnn --sim_model saddle -s 10 -o block_saddle_10
-python evaluate.py -f gru --sim_model saddle -s 10 -o gru_saddle_10
-python evaluate.py -f transformer --sim_model saddle -s 10 -o transformer_saddle_10
+python evaluate.py -f lstm --sim_model saddle -s 10 -o lstm_saddle_10_nontipped
+python evaluate.py -f block_rnn --sim_model saddle -s 10 -o block_saddle_10_nontipped
+python evaluate.py -f gru --sim_model saddle -s 10 -o gru_saddle_10_nontipped
+python evaluate.py -f transformer --sim_model saddle -s 10 -o transformer_saddle_10_nontipped
 
-python evaluate.py -f lstm --sim_model stochastic -s 1 -o lstm_stochastic_1 --tipped
-python evaluate.py -f block_rnn --sim_model stochastic -s 1 -o block_stochastic_1 --tipped
-python evaluate.py -f gru --sim_model stochastic -s 1 -o gru_stochastic_1 --tipped
-python evaluate.py -f transformer --sim_model stochastic -s 1 -o transformer_stochastic_1 --tipped
+python evaluate.py -f lstm --sim_model stochastic -s 1 -o lstm_stochastic_1_tipped --tipped
[…]

### 2022-05-02 darts/hopf_eval.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+python train_ensemble.py -f lstm -s 1 -o lstm_hopf_1_increase -e --sim_model hopf
+python train_ensemble.py -f block_rnn -s 1 -o block_hopf_1_increase -e --sim_model hopf
+python train_ensemble.py -f gru -s 1 -o gru_hopf_1_increase  -e --sim_model hopf
+python train_ensemble.py -f transformer -s 1 -o tranformer_hopf_1_increase  -e --sim_model hopf
+
+python train_ensemble.py -f lstm -s 10 -o lstm_hopf_10_increase  -e --sim_model hopf
+python train_ensemble.py -f block_rnn -s 10 -o block_hopf_10_increase -e --sim_model hopf
+python train_ensemble.py -f gru -s 10 -o gru_hopf_10_increase -e --sim_model hopf
+python train_ensemble.py -f transformer -s 10 -o tranformer_hopf_10_increase -e --sim_model hopf

### 2022-05-02 darts/train_ensemble.py

@@ -140,7 +140,8 @@ if args.evaluate:
         torch.manual_seed(i)
         n_draws = 100
         
-        train_series = preprocessed_t_series(args.sim_model, 1)
+        args.n_samples = 1
+        train_series = get_train_series(args)
         
         if args.sim_model == "hopf":
             t_series = train_series[-input_len:]

### 2022-05-03 manuscript/manuscript.Rmd
Mostly reading not a lot of writing today
Adding more notes

@@
 
+We present ARIMA forecasts as model-free statistical analogs to the forecasts made using Bayesian inference with MCMC.
+not making mechanistic assumptions. Worst-case scenario, presumption here is that future will be similar to the past
+
 ## Method group 3: Machine Learning models
+
+Ml models tend to focus on point forecasts, not sequence to sequence forecasts. 
+Majority of techniques here come from language models.
+Need concept of memory here, either via attention (transformer) or memory (recurrent networks)
+Recent work has put forward that encoder-decoder structure should excel here
+uncertainty is represented by learning parameters for a distribution, in this case we use a poisson distribution. 
+So we estimate uncertainty intervals by making draws from a fitted distribution at each time point.
+In practice lots of problems with instability, we get around this by taking ensemble approach -> learning multiple models over different seeds, using collective forecast
+otherwise can come across models that struggle
 

### 2022-05-04 manuscript/manuscript.Rmd
Slow going

--- 2022-04-29 manuscript/manuscript.Rmd
@@
 
+We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
+In contrast to the ideal case MCMC-based forecasts, ARIMA-based forecasts should serve as a lower limit for the performance of non-mechanistic models.
+Since ARIMA models make the assumption that the future will resemble the past via their auto-regressive and moving average components, ARIMA models are not well-suited for tipping point problems.
+Uncertainty intervals here are estimated directly from parameters of the model.
+
 ## Method group 3: Machine Learning models
+
+Lastly, we examine how deep learning algorithms perform.
+Over the past decade, deep learning has become popular for a broad range of challenging predictive tasks.
+However, there have been a relatively small number of studies investigating how deep learning methods perform with sequence to sequence time series forecasting [@kao, @lyu, @Du].
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important to improving forecasting performance. 
+To investigate these assertions, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
+
+Many of the common deep learning algorithms are not designed to estimate uncertainty intervals, so it is necessary to consider how deep learning methods can be used for uncertainty estimation. 
+With the deep learning models that we present, we take the same approach wherein we set the output of each machine learning model to be the parameters of a poisson distribution.
+To estimate the uncertainty interval at any point, we collect samples from the distribution described by the parameters outputted by the model.
+
+A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds.
+We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds.
+To avoid this instability concern, we ensemble output samples across 5 different initialzation seeds.
+In this way, we can regularize for instability for seed initialization.
 

### 2022-05-04 darts/StatsForecastAutoARIMA.py
Fixed hopf bug hopefully

@@ -5,63 +5,88 @@ sys.path.append("../")
 import matplotlib
 matplotlib.use("Agg")
 import matplotlib.pyplot as plt
-from utils import preprocessed_t_series, truth_dist, make_df
+from utils import get_train_series, truth_dist, make_df
 import argparse
 import os
 
+scaler = Scaler()
 parser = argparse.ArgumentParser()
-parser.add_argument(
-    "-s",
-    "--sim_model",
-    default="stochastic",
-    type=str,
-    help="Select which model to use for the tipping point (stochastic/saddle/hopf)",
-)
-parser.add_argument(
-    "--case",
-    default="tipped",
-    type=str,
-    help="Select which case for the training series",
-)
 parser.add_argument(
     "-f",
     "--forecasting_model",
     default="arima",
     type=str,
-    help="Select which model to train",
+    help="model to train with (arima)",
+)
+parser.add_argument(
+    "-s",
+    "--n_samples",
+    default=1,
+    type=int,
+    help="Number of samples that the model was trained on",
 )
 parser.add_argument(
     "-o",
     "--output_file_name",
     default="trash",
     type=str,
-    help="Select which filename to use",
+    help="File name of plots",
+)
+parser.add_argument(
+    "--sim_model",
+    default="stochastic",
+    type=str,
+    help="Select which dynamics to use for the tipping point (stochastic/saddle/hopf)",
+)
+parser.add_argument(
+    "--tipped",
+    action="store_true",
+    help="Use model trained on tipped or nontipped data",
+)
+parser.add_argument(
+    "--decrease",
+    action="store_true",
+    help="Use model trained on decreasing or increasing hopf model",
 )
 args = parser.parse_args()
 
-if args.case.lower() == "tipped" and args.sim_model == "stochastic":
-    train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-elif args.case.lower() == "nontipped" and args.sim_model == "stochastic":
-    train_series = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
-elif args.case.lower() == "both" and args.sim_model == "stochastic":
-    train_series = TimeSeries.from_csv("stochastic_tipped.csv", time_col="time")
-    train_series_ = TimeSeries.from_csv("stochastic_non_tipped.csv", time_col="time")
[…]

### 2022-05-04 darts/train_ensemble.py

@@ -149,7 +149,7 @@ if args.evaluate:
             t_series = train_series[:input_len]
         
         # truth_dist 
-        start_t = input_len if args.sim_model != "hopf" else 100
+        start_t = input_len
         t_dist = truth_dist(args.sim_model, t_series, input_len, output_len, n_draws=n_draws, reverse=args.decrease, start_t=start_t)
         
         t_max = 250-input_len if args.sim_model != "hopf" else 100

### 2022-05-04 manuscript/manuscript.Rmd
use hive partition structure, example plotting

--- 2022-04-29 manuscript/manuscript.Rmd
@@
 ```{r fig.cap="Forecast scenarios. A. a Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation."}
-hopf <- read_csv("../greta/data/hopf_increasing.csv.gz")
+hopf <- read_csv("../greta/forecasts/hopf_increasing.csv.gz")
 panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
@@
 
-sn <- read_csv("../greta/data/saddlenode.csv.gz")
+sn <- read_csv("../greta/forecasts/saddlenode.csv.gz")
 panelB <- sn |> filter(type == "historical", i == 1) |>
@@
 
-stoch <- read_csv("../greta/data/stochastic.csv.gz")
+stoch <- read_csv("../greta/forecasts/stochastic.csv.gz")
 panelC <- stoch |> filter(type == "historical", i == 1) |>
@@
 ```{r}
-panelB <- sn |> filter(type == "historical", i == 1) |>
-  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
-  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
-  geom_path(alpha=0.5, show.legend = FALSE) +
-  scale_color_grey(end=0.6) + theme_bw() + 
-  ggtitle("B. Saddle-node bifurcation")
+panelB <- sn |>
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+  rename(species = variable) |>
+  ggplot(aes(t, col=type)) + 
+    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+    geom_line(aes(y=mean)) +
+    #geom_vline(aes(xintercept = 100)) + 
+    facet_wrap(~species, ncol=1) +
+    theme_bw()
 
-panelC <- stoch |> filter(type == "historical", i == 1) |>
-  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
-  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
-  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
+panelC <- stoch |>
+  group_by(t,type,variable) |> 
+  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
+  rename(species = variable) |>
+  ggplot(aes(t, col=type)) + 
+    geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
+    geom_line(aes(y=mean)) +
+    #geom_vline(aes(xintercept = 100)) + 
+    facet_wrap(~species, ncol=1) +
+    theme_bw() + 
   ggtitle("C. Stochastic transition")
@@
 mcmc_scores <- bind_rows(
-  read_csv("../greta/data/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
-read_csv("../greta/data/scores_stochastic.csv.gz"),
-read_csv("../greta/data/scores_hopf.csv.gz") |> 
+  read_csv("../greta/forecasts/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
+read_csv("../greta/forecasts/scores_stochastic.csv.gz"),
+read_csv("../greta/forecasts/scores_hopf.csv.gz") |> 
   filter(scenario == "hopf_increasing") |>

### 2022-05-05 darts/StatsForecastAutoARIMA.py
Hopf arima up now

@@ -5,11 +5,14 @@ sys.path.append("../")
 import matplotlib
 matplotlib.use("Agg")
 import matplotlib.pyplot as plt
+from pandas import DataFrame
+import numpy as np
+import torch
 from utils import get_train_series, truth_dist, make_df
 import argparse
+from functools import reduce
 import os
 
-scaler = Scaler()
 parser = argparse.ArgumentParser()
 parser.add_argument(
     "-f",
@@ -76,13 +79,22 @@ for i in range(5):
     start_t = input_len
     t_dist = truth_dist(args.sim_model, t_series, input_len, output_len, n_draws=n_draws, reverse=args.decrease, start_t=start_t)
     
-    t_max = 250-input_len if args.sim_model != hopf else 100
+    t_max = 250-input_len if args.sim_model != "hopf" else 100
     
     ensemble_preds = []
-    for model in models:
-        _preds = model.predict(t_max, t_series, num_samples=n_draws)
+    if args.sim_model != "hopf":
+        model = _model().fit(series=t_series)
+        _preds = model.predict(t_max, num_samples=n_draws)
         ensemble_preds.append(_preds)
-    ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
+        ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="sample"), ensemble_preds)
+    else:
+        h_t_series, p_t_series = t_series.univariate_component(0), t_series.univariate_component(1)
+        for series in [h_t_series, p_t_series]:
+            model = _model().fit(series=series)
+            _preds = model.predict(t_max, num_samples=n_draws)
+            ensemble_preds.append(_preds)
+      
+        ensemble_series = reduce(lambda a, b: a.concatenate(b, axis="component"), ensemble_preds)
     
     case = "tipped" if args.tipped else "nontipped"
         

### 2022-05-05 darts/arima_eval.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+python StatsForecastAutoARIMA.py -s 1 -o arima_hopf_1_increase --sim_model hopf
+python StatsForecastAutoARIMA.py -s 1 -o arima_stochastic_1_tipped --sim_model stochastic --tipped
+python StatsForecastAutoARIMA.py -s 1 -o arima_saddle_1_nontipped --sim_model saddle

### 2022-05-05 darts/StatsForecastAutoARIMA.py
Updating some paths

@@ -101,4 +101,4 @@ for i in range(5):
     df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
     final_df = final_df.append(df, ignore_index=True)
     
-final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)
+final_df.to_csv(f"forecasts/reps={args.n_samples}/{args.output_file_name}.csv.gz", index=False)

### 2022-05-05 darts/evaluate.py

@@ -109,4 +109,4 @@ for i in range(5):
     df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
     final_df = final_df.append(df, ignore_index=True)
     
-final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)
+final_df.to_csv(f"forecasts/reps={args.n_samples}/{args.output_file_name}.csv.gz", index=False)

### 2022-05-05 darts/train_ensemble.py

@@ -131,6 +131,8 @@ for i in range(42, 47):
 if args.evaluate:
     if not os.path.exists("forecasts/"):
             os.makedirs("forecasts/")
+    if not os.path.exists(f"forecasts/reps={args.n_samples}"):
+            os.makedirs(f"forecasts/reps={args.n_samples}")
     input_len = hyperparameters["input_chunk_length"]
     output_len = hyperparameters["output_chunk_length"]
     final_df = DataFrame()
@@ -171,4 +173,4 @@ if args.evaluate:
         df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
         final_df = final_df.append(df, ignore_index=True)
         
-    final_df.to_csv(f"forecasts/{args.output_file_name}.csv.gz", index=False)
+    final_df.to_csv(f"forecasts/reps={args.n_samples}/{args.output_file_name}.csv.gz", index=False)

### 2022-05-05 manuscript/manuscript.Rmd
Premptive commit

@@
 
-Lastly, we examine how deep learning algorithms perform.
-Over the past decade, deep learning has become popular for a broad range of challenging predictive tasks.
+Over the past decade, deep learning has become popular for a broad range of challenging predictive tasks;
 However, there have been a relatively small number of studies investigating how deep learning methods perform with sequence to sequence time series forecasting [@kao, @lyu, @Du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important to improving forecasting performance. 
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important to optimizing forecasting performance. 
 To investigate these assertions, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.

### 2022-05-05 darts/train_ensemble.py
Patched hopf bug

@@ -159,7 +159,8 @@ if args.evaluate:
         ensemble_preds = []
         for model in models:
             if args.sim_model == "hopf":
-                _preds = scaler.inverse_transform(model.predict(t_max, t_series, num_samples=n_draws))
+                input_series = scaler.transform(t_series)
+                _preds = scaler.inverse_transform(model.predict(t_max, input_series, num_samples=n_draws))
             else: 
                 _preds = model.predict(t_max, t_series, num_samples=n_draws)
             ensemble_preds.append(_preds)

### 2022-05-05 darts/train_ensemble.py
Hopf bug patch

@@ -61,6 +61,7 @@ parser.add_argument(
     help="Flag to use decreasing Hopf model",
 )
 args = parser.parse_args()
+final_dest = f"forecasts/reps={args.n_samples}/{args.output_file_name}.csv.gz"
 
 if args.forecasting_model == "lstm": 
     if args.sim_model == "stochastic":
@@ -173,5 +174,5 @@ if args.evaluate:
             
         df = make_df(ensemble_series, t_dist, t_series, args.sim_model, args.forecasting_model.lower(), case, args.n_samples, i)
         final_df = final_df.append(df, ignore_index=True)
-        
-    final_df.to_csv(f"forecasts/reps={args.n_samples}/{args.output_file_name}.csv.gz", index=False)
+    
+    final_df.to_csv(final_dest, index=False)

### 2022-05-05 darts-plots.R
edits

@@ -91,57 +91,3 @@ combined |> filter(simulation=="stochastic", group == 1) |>
 
 #####################
 
-
-
-
-compute_scores <- function(observed, dat) {
-  logsscore <- scoringRules::logs_sample(observed, dat)
-  crpsscore <- scoringRules::crps_sample(observed, dat)
-  data.frame(logs = logsscore, crps =  crpsscore)
-  
-}
-
-rep_scores <- function(predicted, sim) {
-  obs <- 
-    predicted |> 
-    pivot_wider(id_cols = "t", 
-                names_from="iter",
-                values_from = "value") |> 
-    select(-t) |> 
-    as.matrix()
-
-  scores <- sim |> 
-    group_by(iter) |> 
-    group_modify(~ compute_scores(.x$value, obs), .keep = TRUE) 
-  
-  ## crude way to restore metadata
-  scores <- sim |> 
-    select(t, variable, simulation) |> 
-    bind_cols(scores)
-  scores
-}
-
-
-score_it <- function(scenario, model, var, darts, reference) {
-  predicted <- darts |> 
-    filter(simulation=={{scenario}}, 
-           forecasting_model == {{model}}, 
-           variable == {{var}},
-           group==1, 
-           reps==1,
-           type == "predicted",
-           ) |>
-    collect() 
-  sim <- reference |> 
-    filter(simulation=={{scenario}},
-           variable=={{var}},
-           type == "true") |>
-    rename(iter = i) |> 
-    collect()
-  scores <- rep_scores(predicted, sim)
-
-}
-scores <- score_it(scenario = "hopf", "gru", "H", darts, reference) 
-
-scores |> 
-  ggplot(aes(t, crps, group=iter)) + geom_point()

### 2022-05-05 manuscript/manuscript.Rmd
edits
file path

@@
 ```{r fig.cap="Forecast scenarios. A. a Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation."}
-hopf <- read_csv("../greta/forecasts/hopf_increasing.csv.gz")
+hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
 panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
@@
 
-sn <- read_csv("../greta/forecasts/saddlenode.csv.gz")
+sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
 panelB <- sn |> filter(type == "historical", i == 1) |>
@@
 
-stoch <- read_csv("../greta/forecasts/stochastic.csv.gz")
+stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
 panelC <- stoch |> filter(type == "historical", i == 1) |>
@@
 mcmc_scores <- bind_rows(
-  read_csv("../greta/forecasts/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
-read_csv("../greta/forecasts/scores_stochastic.csv.gz"),
-read_csv("../greta/forecasts/scores_hopf.csv.gz") |> 
+  read_csv("../greta/scores/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
+read_csv("../greta/scores/scores_stochastic.csv.gz"),
+read_csv("../greta/scores/scores_hopf.csv.gz") |> 
   filter(scenario == "hopf_increasing") |>

### 2022-05-06 manuscript/manuscript.Rmd
fix eq
Some edits here

@@
 \begin{align}
-H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) - c P_t + \eta_{H,t} \\
+H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) - c P_t + \eta_{H,t} \right) \\
 P_{t+1} &= H_t \exp \left( r \left(1  \tfrac{H}{K_t}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right) \\
@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-In contrast to the ideal case MCMC-based forecasts, ARIMA-based forecasts should serve as a lower limit for the performance of non-mechanistic models.
-Since ARIMA models make the assumption that the future will resemble the past via their auto-regressive and moving average components, ARIMA models are not well-suited for tipping point problems.
-Uncertainty intervals here are estimated directly from parameters of the model.
+Since ARIMA models make the assumption that the future will resemble the past via their auto-regressive and moving average components [@cite], ARIMA models are not well-suited for tipping point problems.
+Thus, ARIMA-based forecasts should serve as a lower threshold of acceptability for the performance of non-mechanistic models.
+In contrast to inference with MCMC, uncertainty is estimated directly from the learned parameters [@cite].
 
@@
 
-Over the past decade, deep learning has become popular for a broad range of challenging predictive tasks;
-However, there have been a relatively small number of studies investigating how deep learning methods perform with sequence to sequence time series forecasting [@kao, @lyu, @Du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important to optimizing forecasting performance. 
-To investigate these assertions, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
+Over the past decade, deep learning has become very popular for a broad range of challenging predictive tasks [@cite];
+However, there have been a relatively small number of studies investigating how deep learning methods perform on sequence to sequence time series forecasting problems [@kao, @lyu, @Du].
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important for improving forecasting performance. 
+To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
 
-Many of the common deep learning algorithms are not designed to estimate uncertainty intervals, so it is necessary to consider how deep learning methods can be used for uncertainty estimation. 
-With the deep learning models that we present, we take the same approach wherein we set the output of each machine learning model to be the parameters of a poisson distribution.
-To estimate the uncertainty interval at any point, we collect samples from the distribution described by the parameters outputted by the model.
+Machine learning models are often used to make point forecasts, yet for their application to complex time series problems like tipping points, it will be necessary to use methods that estimate uncertainty.
+For all the machine learning models in this study, we use the same method for uncertainty estimation.
+Each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
+To estimate uncertainty intervals, we simply draw samples from the learned distribution at each time step and report the 95% confidence interval.
 
-A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds.
-We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds.
-To avoid this instability concern, we ensemble output samples across 5 different initialzation seeds.
-In this way, we can regularize for instability for seed initialization.
+A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@cite].
+We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
+A proposed reason for this instability is that the machine learning models often get stuck on local optima of loss surfaces [@modelstability].
+To regularize against the common initialization instability problem, each machine learning forecast in [@figures] is an ensemble forecast over 5 different initialization seeds.
+
+Bias Variance tradeoff.
 
@@
 have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
-It is worth noting that this property applies specifically to probablistic forecasts, and is not true of all metrics often used to compare forecasts.
+It is worth noting that this property applies specifically to probabilistic forecasts, and is not true of all metrics often used to compare forecasts.
 Concerns about over-fitting arise in most types of model estimation, are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@].   

### 2022-05-09 manuscript/manuscript.Rmd
Looking a little better here

@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-Since ARIMA models make the assumption that the future will resemble the past via their auto-regressive and moving average components [@cite], ARIMA models are not well-suited for tipping point problems.
+Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], ARIMA models are not well-suited for tipping point problems.
 Thus, ARIMA-based forecasts should serve as a lower threshold of acceptability for the performance of non-mechanistic models.
-In contrast to inference with MCMC, uncertainty is estimated directly from the learned parameters [@cite].
+In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@cite].
+Since ARIMA is a very commonly encountered forecasting method, we will refer readers to [@cite] for further discussion.
 
@@
 
-Over the past decade, deep learning has become very popular for a broad range of challenging predictive tasks [@cite];
-However, there have been a relatively small number of studies investigating how deep learning methods perform on sequence to sequence time series forecasting problems [@kao, @lyu, @Du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is important for improving forecasting performance. 
+Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @Du].
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance.
+Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
+Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
 To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
+We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.
 
-Machine learning models are often used to make point forecasts, yet for their application to complex time series problems like tipping points, it will be necessary to use methods that estimate uncertainty.
-For all the machine learning models in this study, we use the same method for uncertainty estimation.
-Each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
-To estimate uncertainty intervals, we simply draw samples from the learned distribution at each time step and report the 95% confidence interval.
+Machine learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use methods that estimate uncertainty. 
+For all the machine learning models in this study, we use the same method for uncertainty estimation:
+each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
+To estimate uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.
 
@@
 We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
-A proposed reason for this instability is that the machine learning models often get stuck on local optima of loss surfaces [@modelstability].
-To regularize against the common initialization instability problem, each machine learning forecast in [@figures] is an ensemble forecast over 5 different initialization seeds.
-
-Bias Variance tradeoff.
+One proposed reason for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
+Another likely reason for this instability is overfitting to the training set.
+Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
+A common method to reduce overfitting is to perform K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
+To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 models that were trained with different initialization seeds.
+We found this simple ensemble method to be an effective way to improve the generalizability for each ML model.
 

### 2022-05-09 darts-plots.R
tweaking plots

@@ -68,23 +68,25 @@ combined |> filter(simulation=="hopf", reps == 1, group == 1) |>
 
 
 
-combined |> filter(simulation=="saddle", group == 1) |> 
+combined |> filter(simulation=="saddle", group == 1, reps == 1) |> 
   ggplot(aes(t, col=type, fill=type)) + 
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = saddle_sims) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_sims) +
   geom_line(aes(y=mean)) +
   geom_line(aes(y=mean), data = saddle_sims) +
-  facet_grid(forecasting_model ~ reps, scales = "free") +
+  facet_wrap(~forecasting_model, scales = "free") +
+  #facet_grid(forecasting_model ~ reps, scales = "free") +
   theme_bw() + 
   ggtitle("B. Saddle-Node bifurcation")
 
-combined |> filter(simulation=="stochastic", group == 1) |> 
+combined |> filter(simulation=="stochastic", group == 1, reps==1) |> 
   ggplot(aes(t, col=type, fill=type)) + 
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = stochastic_sims) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_sims) +
   geom_line(aes(y=mean)) +
   geom_line(aes(y=mean), data = stochastic_sims) +
-  facet_grid(forecasting_model ~ reps, scales = "free") +
+  facet_wrap(~forecasting_model, scales = "free") +
+  #facet_grid(forecasting_model ~ reps, scales = "free") +
   theme_bw() + 
   ggtitle("C. Stochastic transition")
 

### 2022-05-09 scores-plots.R
processed scores & score plots

@@ -1,0 +1,156 @@
+library(arrow)
+library(tidyverse)
+
+bound <- function(x, percentile = 0.975) {
+  dplyr::nth(x, percentile * length(x), order_by = x)
+}
+
+reference <- arrow::open_dataset("greta/forecasts/", format="csv") |> 
+  filter(type != "predicted") |> 
+  collect() |> rename(iter = i) |>
+  mutate(variable = forcats::fct_recode(variable, 
+                                        X = "N")) 
+
+
+darts <- arrow::open_dataset("darts/forecasts/", format="csv")|> filter(type == "predicted")
+greta <- arrow::open_dataset("greta/forecasts/", format="csv") |>
+  mutate(forecasting_model = "MCMC", reps=1, group=1) |> 
+  filter(type == "predicted") |> rename(iter = i) |> collect() |>
+  mutate(variable = forcats::fct_recode(variable,  X = "N")) 
+
+
+
+compute_scores <- function(observed, dat) {
+  logsscore <- scoringRules::logs_sample(observed, dat)
+  crpsscore <- scoringRules::crps_sample(observed, dat)
+  data.frame(logs = logsscore, crps =  crpsscore)
+  
+}
+
+rep_scores <- function(predicted, sim) {
+
+  ## ensure corresponding time interval in both
+  match_sim <- sim |> filter(t %in% unique(predicted$t)) 
+  match_predicted <- predicted |> filter(t %in% unique(match_sim$t)) 
+  
+  
+    
+  # T x iter
+  obs <- 
+    match_predicted |> 
+    pivot_wider(id_cols = "t", 
+                names_from="iter",
+                values_from = "value") |> 
+    select(-t) |> 
+    as.matrix()
+
+
+  scores <- match_sim |>
+    group_by(iter) |> 
+    group_modify(~ compute_scores(.x$value, obs), .keep = TRUE) 
+  
+  ## crude way to restore metadata
+  scores <- match_sim |> 
+    select(t, variable, simulation) |> 
+    bind_cols(scores)
+  scores
+}
+
+score_it <- function(scenario, model, var, darts, reference) {
+  predicted <- darts |> 
+    filter(simulation=={{scenario}}, 
+           forecasting_model =={{model}}, 
+           variable == {{var}},
+           group==1, 
+           reps==1,
+           type == "predicted",
+    ) |>
+    collect() 
[…]

### 2022-05-09 manuscript/manuscript.Rmd

@@
 
-Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @Du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance.
+Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @du].
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance -- see @kao, @lyu and @du for more details on these concepts.
 Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
@@
 One proposed reason for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
-Another likely reason for this instability is overfitting to the training set.
+Another likely reason is that machine learning models commonly overfit the training data.
 Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
-A common method to reduce overfitting is to perform K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
-To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 models that were trained with different initialization seeds.
-We found this simple ensemble method to be an effective way to improve the generalizability for each ML model.
+One frequently cited method to reduce overfitting is K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
+To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
+We found this simple ensemble method to be an effective way to improve generalizability.
 

### 2022-05-10 manuscript/manuscript.Rmd
Adding some inline comments

@@
 While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
+<!-- MFL: Using 'tactical' might throw readers off that aren't familiar with tact vs strat distinction -->
 These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
@@
 The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
-For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range ecological and socio-ecological systems [@Scheffer2001].
+For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@Scheffer2001].
 
@@
 Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
-Due both this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-These dynamics have long been identified as an acute challenge to forecasting, with effort focused primarily on identifying any 'early warning signs' that a catastrophic bifurcation might be occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
+These dynamics have long been identified as an acute challenge to forecasting, with effort focused primarily on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+<!-- MFL: I think we can make the first half of this sentence clearer.
+I think some readers might get lost here.
+-->
 Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
@@
 we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
-Based on this data, our forecaster seeks to identify the distribution of expected transition times for analgous systems starting from the same initial condition.
+Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
 This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [].
@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], ARIMA models are not well-suited for tipping point problems.
-Thus, ARIMA-based forecasts should serve as a lower threshold of acceptability for the performance of non-mechanistic models.
+Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], these models are not well-suited for problems with complex bifurcation dyanmics.
+Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
 In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@cite].
@@
 Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance -- see @kao, @lyu and @du for more details on these concepts.
+Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance -- see @kao, @lyu and @du for more context on memory, attention and encoder-decoders.
 Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
@@
 
-Machine learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use methods that estimate uncertainty. 
+Machine learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use probablisitic forecasts. 
 For all the machine learning models in this study, we use the same method for uncertainty estimation:
 each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
-To estimate uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.
+To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.
 
@@
 We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
-One proposed reason for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
-Another likely reason is that machine learning models commonly overfit the training data.
+One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
+Another likely cause is that machine learning models commonly overfit the training data [@ying].
 Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
-One frequently cited method to reduce overfitting is K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
+One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
 To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
@@
 have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
+<!-- MFL: I think we can take out the parenthetical here and assume the reader knows what we mean by overfitting-->
 It is worth noting that this property applies specifically to probabilistic forecasts, and is not true of all metrics often used to compare forecasts.
-Concerns about over-fitting arise in most types of model estimation, are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@].   
+<!-- MFL: hmm I am not completely following this point here -->
+Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@].
+<!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
 This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.

### 2022-05-10 manuscript/manuscript.Rmd
Changing structure on ml section

@@
 
-Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @du].
-Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance -- see @kao, @lyu and @du for more context on memory, attention and encoder-decoders.
-Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
-Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
-To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
-We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.
-
-Machine learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use probablisitic forecasts. 
-For all the machine learning models in this study, we use the same method for uncertainty estimation:
+Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
+Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use probablisitic forecasts. 
+For all the deep learning models in this study, we use the same method for uncertainty estimation:
 each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
@@
 Another likely cause is that machine learning models commonly overfit the training data [@ying].
-Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
+Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
+With so many parameters, deep learning models tend to have high variance and thus overfit the training data [@ying].
 One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
 To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
-We found this simple ensemble method to be an effective way to improve generalizability.
+We found this simple ensemble method to be an effective way to improve generalizability in our limited data regime.
+
+Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao, @lyu, @du].
+Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
+Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
+To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
+We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.
 

### 2022-05-10 manuscript/manuscript.Rmd
Little update on a ml paragraph

@@
 For all the deep learning models in this study, we use the same method for uncertainty estimation:
-each model is trained to learn the parameters of a poisson distribution that maximizes the log likelihood. 
+each model is trained to learn the parameters of a Poisson distribution that maximizes the log likelihood. 
 To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.
@@
 Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
-With so many parameters, deep learning models tend to have high variance and thus overfit the training data [@ying].
+With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequnce of the bias-variance trade-off common across statistics and machine learning [@ying].
 One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
@@
 Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao, @lyu, @du].
+Herein we will provide some background on what these machine learning methods are and their benefits.
+
 Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].

### 2022-05-10 manuscript/manuscript.Rmd
Tiny one sentence edit

@@
 For all the deep learning models in this study, we use the same method for uncertainty estimation:
-each model is trained to learn the parameters of a Poisson distribution that maximizes the log likelihood. 
+each model is trained to learn the parameters of a Poisson distribution to maximize the log likelihood. 
 To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.

### 2022-05-10 manuscript/manuscript.Rmd
Another little change here

@@
 Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
-With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequnce of the bias-variance trade-off common across statistics and machine learning [@ying].
+With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@ying].
 One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.

### 2022-05-10 manuscript/manuscript.Rmd
Transformer paragraph up

@@
 To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
-We found this simple ensemble method to be an effective way to improve generalizability in our limited data regime.
+We found this simple ensemble method to be an effective way to improve generalizability in the limited data regime.
 
@@
 
-Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
+### Recurrent Neural Networks
+
+Recurrent neural networks are the predominant memory-based ML method.
+Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps.
+By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
+However, a standard recurrent neural network, i.e.
+a feed forward neural network that provides self-feedback, is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
+Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
+These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung for more details.
+While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung], so we investigate both methods.
+
+### Transformers
+
+The Transformer is a state of the art ML architecture which uses an attention mechanism to model long and short term dependencies on sequence to sequence tasks [@vaswani].
+Specifically, Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
+Because of self-attention, Transformers do not need to process data sequentially like recurrent neural networks do, and thus can be parallelized, offering significant computational advantages [@vaswani].
+The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate these models here.
+
+### Encoder-Decoder
+
 Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
-To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
-We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.
 

### 2022-05-11 scores-plots.R
:pen:

@@ -121,12 +121,13 @@ scores <- arrow::open_dataset("scores", format="csv") |> collect()
 # shift logs score to positive?
 shift <- scores |> filter(!is.infinite(logs)) |> mutate(logs = logs - min(logs) +.01)
 
-crps <- scores |> 
+scores |> 
   ggplot(aes(forecasting_model, crps)) + 
   geom_boxplot() +
-  facet_wrap(~simulation, scales = "free", ncol=1) + theme_bw()
+  facet_wrap(~simulation, scales = "free", ncol=1) +
+  theme_bw()
 
-logs<- shift |>
+shift |>
   ggplot(aes(forecasting_model, logs)) + 
   geom_boxplot() +
   facet_wrap(~simulation, scales = "free", ncol=1) + 
@@ -134,7 +135,7 @@ logs<- shift |>
 
 
 # scores over time
-over_time <- scores |> collect() |> 
+over_time <- shift |> 
   group_by(t, simulation,forecasting_model ) |> 
   summarise(logs = mean(logs), 
             crps = mean(crps)) 

### 2022-05-11 manuscript/manuscript.Rmd

@@
 Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
-Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use probablisitic forecasts. 
+Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will often be necessary to use multi-step, probablisitic forecasts. 
 For all the deep learning models in this study, we use the same method for uncertainty estimation:
@@
 Recurrent neural networks are the predominant memory-based ML method.
-Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps.
+Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky].
 By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
-However, a standard recurrent neural network, i.e.
-a feed forward neural network that provides self-feedback, is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
+However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
 Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
@@
 
-The Transformer is a state of the art ML architecture which uses an attention mechanism to model long and short term dependencies on sequence to sequence tasks [@vaswani].
-Specifically, Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
-Because of self-attention, Transformers do not need to process data sequentially like recurrent neural networks do, and thus can be parallelized, offering significant computational advantages [@vaswani].
-The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate these models here.
+The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani].
+Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
+For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
+Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani].
+The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate Transformers in this study.
 
@@
 
-Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
+Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@learningphrase, @cite, @cite].
+Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
+It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever].
+Of the models that we present, Block RNNs and Transformers are the only models that use encoder-decoders.
 

### 2022-05-11 manuscript/manuscript.Rmd
plots

@@
 ```{r}
-hopf |> 
-  group_by(t,type,variable) |> 
-  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
-  rename(species = variable) |> 
-  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
-  ggplot(aes(t, col=type)) + 
-    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+
+forecasts <- read_parquet("data/forecasts.parquet")
+observations <- read_parquet("data/observations.parquet")
+
+hopf_obs <- observations |> filter(simulation=="hopf")
+stochastic_obs <- observations |> filter(simulation=="stochastic")
+saddle_obs <- observations |> filter(simulation=="saddle")
+```
+
+```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
+forecasts |> filter(simulation=="hopf", reps == 1, group == 1) |> 
+  ggplot(aes(t, col=type, fill=type)) + 
+  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
+  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_sims) +
   geom_line(aes(y=mean)) +
-    #geom_vline(aes(xintercept = 100)) + 
-    facet_wrap(~species, ncol=1) +
+  geom_line(aes(y=mean), data = hopf_obs) +
+  facet_grid(forecasting_model ~ variable, scales = "free") +
   theme_bw() + 
-    ggtitle("A. Hopf bifurcation")
+  ggtitle("Hopf bifurcation")
 ```
@@
 ```{r}
-panelB <- sn |>
-  group_by(t,type,variable) |> 
-  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
-  rename(species = variable) |>
-  ggplot(aes(t, col=type)) + 
-    geom_ribbon(aes(ymin = mean-2*sd, ymax = mean+2*sd, fill=type), alpha=0.5) +
+forecasts |> filter(simulation=="saddle", group == 1, reps == 1) |> 
+  ggplot(aes(t, col=type, fill=type)) + 
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_sims) +
   geom_line(aes(y=mean)) +
-    #geom_vline(aes(xintercept = 100)) + 
-    facet_wrap(~species, ncol=1) +
-    theme_bw()
-
-panelC <- stoch |>
-  group_by(t,type,variable) |> 
-  summarise(mean = mean(value), sd = sd(value), .groups = "drop") |> 
-  rename(species = variable) |>
-  ggplot(aes(t, col=type)) + 
-    geom_ribbon(aes(ymin = pmax(mean-2*sd,0), ymax = mean+2*sd, fill=type), alpha=0.5) +
-    geom_line(aes(y=mean)) +
-    #geom_vline(aes(xintercept = 100)) + 
-    facet_wrap(~species, ncol=1) +
+  geom_line(aes(y=mean), data = saddle_obs) +
+  facet_wrap(~forecasting_model, scales = "free") +
+  #facet_grid(forecasting_model ~ reps, scales = "free") +
   theme_bw() + 
-  ggtitle("C. Stochastic transition")
-
-panelA / panelB / panelC
+  ggtitle("Saddle-Node bifurcation")
 ```
@@
 ```{r}
-mcmc_scores <- bind_rows(
-  read_csv("../greta/scores/scores_saddlenode.csv.gz") |> mutate(scenario="saddlenode"),
-read_csv("../greta/scores/scores_stochastic.csv.gz"),
-read_csv("../greta/scores/scores_hopf.csv.gz") |> 
-  filter(scenario == "hopf_increasing") |>
-  mutate(scenario = forcats::fct_recode(scenario, hopf = "hopf_increasing"),
-         reps = 1)
-) |> rename(method=model)
-mcmc_scores |> 
-  pivot_longer(c(logs, crps), names_to = "metric", values_to = "score") |> 
-  ggplot(aes(method, score, col=variable)) + geom_boxplot() + 
-  facet_grid(metric~scenario, scales="free_y")
+forecasts |> filter(simulation=="stochastic", group == 1, reps==1) |> 
+  ggplot(aes(t, col=type, fill=type)) + 
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_sims) +
+  geom_line(aes(y=mean)) +
+  geom_line(aes(y=mean), data = stochastic_obs) +
+  facet_wrap(~forecasting_model, scales = "free") +
+  #facet_grid(forecasting_model ~ reps, scales = "free") +
+  theme_bw() + 
+  ggtitle("Stochastic transition")
+```
+
+```{r}
+scores <- arrow::open_dataset("../scores", format="csv") |> collect()
+
+# shift logs score to strictly positive, non-infinite values so we can use a log-scale
+shift <- scores |> 
+  filter(!is.infinite(logs)) |> 
+  mutate(logs = logs - min(logs) +.01) |>
+  pivot_longer(c(logs, crps), 
+               names_to = "metric", 
+               values_to = "score")
+```
+
+```{r}
+shift |> 
+  ggplot(aes(forecasting_model, score, color=forecasting_model, fill=forecasting_model)) + 
+  geom_violin() +
+  facet_grid(metric ~ simulation, scales = "free") + 
+  scale_y_log10() +
+  theme_bw() + 
+  coord_flip()
+
+  ggplot(aes(simulation, score, color=forecasting_model, fill=forecasting_model)) + 
+  geom_violin() +
+  facet_wrap(~ metric, scales = "free", ncol=1) + 
+  scale_y_log10() + 
+  theme_bw()
+```
+
+```{r}
+shift |>
+  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
+  geom_jitter(alpha = 0.01, show.legend = FALSE) +
+  facet_grid(metric ~ simulation, scales = "free") + 
+  scale_y_log10() +
+  theme_bw() + 
+  coord_flip()
+
+```
+
+```{r}
+# scores over time, averaged over replicates
+over_time <- shift |> 
+  group_by(t, simulation,forecasting_model, metric ) |> 
+  summarise(score = mean(score))
+
+over_time |> 
+  ggplot(aes(t, score, col = forecasting_model)) + 
+  geom_point(size=1) +
+  facet_grid(metric ~ simulation,
+             scales="free") + 
+  theme_bw() + scale_y_log10()
+
 ```

### 2022-05-12 manuscript/manuscript.Rmd
tweaking plots, some writing
added a bit explicitly about frameworks

@@
 Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
-While 'tactical' population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
-<!-- MFL: Using 'tactical' might throw readers off that aren't familiar with tact vs strat distinction -->
+While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
 These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
@@
 This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date.
+
 No-analog scenarios [@Williams]
+
+We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
+After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting.
+Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
+Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
+While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
+Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@scoringRules].
+We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
+Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.
 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-These dynamics have long been identified as an acute challenge to forecasting, with effort focused primarily on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-<!-- MFL: I think we can make the first half of this sentence clearer.
-I think some readers might get lost here.
--->
-Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001]. 
+Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
 
@@
 have a higher likelihood than the true process, it is not possible for these models to overfit the data against which they are scored.
-<!-- MFL: I think we can take out the parenthetical here and assume the reader knows what we mean by overfitting-->
-It is worth noting that this property applies specifically to probabilistic forecasts, and is not true of all metrics often used to compare forecasts.
-<!-- MFL: hmm I am not completely following this point here -->
-Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@].
+It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts.
+Not all common metrics metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
+<!-- MFL: hmm I am not completely following this point here.
+CB: added some examples of non-strictly-proper scoring metrics.
+For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
+Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@Mehta2019].
 <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
@@
 
-While all strictly proper scores have this property, this does not guarantee that all such metrics will share the same relative ranking between forecasts.
+Not even all strictly proper scores will agree on the same relative ranking between forecasts.
 We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
-Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS
+Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
 
 # Results
+
+We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifucation, saddle-node bifurcation, stochastic transition).
+Using model-based simulations allows us to examine performance against an ensemble of replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
 
@@
 shift |>
-  ggplot(aes(forecasting_model, score, color=forecasting_model, fill=forecasting_model)) + 
-  geom_violin() +
-  facet_grid(metric ~ simulation, scales = "free") + 
-  scale_y_log10() +
-  theme_bw() + 
-  coord_flip()
-
-  ggplot(aes(simulation, score, color=forecasting_model, fill=forecasting_model)) + 
-  geom_violin() +
-  facet_wrap(~ metric, scales = "free", ncol=1) + 
+  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
+  geom_jitter(alpha = 0.006, show.legend = FALSE) + 
+  coord_flip() +
+  facet_grid(simulation ~ metric, scales = "free") + 
   scale_y_log10() +
   theme_bw()
-```
-
-```{r}
-shift |>
-  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
-  geom_jitter(alpha = 0.01, show.legend = FALSE) +
-  facet_grid(metric ~ simulation, scales = "free") + 
-  scale_y_log10() +
-  theme_bw() + 
-  coord_flip()
 
@@
   group_by(t, simulation,forecasting_model, metric ) |> 
-  summarise(score = mean(score))
+  summarise(sd = sd(score),
+            score = mean(score),
+            .groups = "drop") |>
+  group_by(simulation,forecasting_model, metric) |> 
+  mutate(t = t - min(t)) |> ungroup()
 
-over_time |> 
+p1 <- over_time |> filter(metric == "crps") |>
   ggplot(aes(t, score, col = forecasting_model)) + 
-  geom_point(size=1) +
-  facet_grid(metric ~ simulation,
-             scales="free") + 
+  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
+                  shape=".", alpha=0.4, show.legend=FALSE) +
+  geom_line(show.legend=FALSE) +
+  facet_wrap(~simulation, scales="free", ncol=1) + 
+  theme_bw()
+
+p2 <- over_time |> filter(metric == "logs") |>
+  ggplot(aes(t, score, col = forecasting_model)) + 
+  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
+                  shape=".", alpha=0.4, show.legend=FALSE) +
+  geom_line(show.legend=FALSE) +
+  facet_wrap(~simulation, scales="free", ncol=1) + 
+  scale_y_log10() +
+  theme_bw()
+
+p1 + p2
+```
+```{r}
+over_time |> filter(metric == "logs") |>
+  ggplot(aes(t, score, col = forecasting_model)) + 
+  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
+                  shape=".", alpha=0.4, show.legend=FALSE) +
+  geom_line(show.legend=FALSE) +
+ # facet_wrap(~simulation, scales="free", ncol=1) + 
+  facet_grid(simulation ~ forecasting_model, scales="free") + 
   theme_bw() + scale_y_log10()
-
 ```

### 2022-05-12 darts-plots.R

@@ -9,7 +9,6 @@ reference <- arrow::open_dataset("greta/forecasts/", format="csv") |>
   filter(type != "predicted") |> 
   collect()
 
-
 sims <- reference |>
   group_by(simulation, variable, type, t) |> 
   summarise(mean = mean(value), 
@@ -22,10 +21,12 @@ sims <- reference |>
                                         X = "N")) 
 
 
-darts <- arrow::open_dataset("darts/forecasts/", format="csv")
-darts <- darts |> filter(type == "predicted")
-greta <- arrow::open_dataset("greta/forecasts/", format="csv") |> mutate(forecasting_model = "MCMC", reps=1, group=1)
-greta <- greta |>  filter(type == "predicted") |> rename(iter = i)
+darts <- arrow::open_dataset("darts/forecasts/", format="csv") |>
+  filter(type == "predicted")
+greta <- arrow::open_dataset("greta/forecasts/", format="csv") |>
+  mutate(forecasting_model = "MCMC", reps=1, group=1) |>
+  filter(type == "predicted") |> 
+  rename(iter = i)
 
   
   
@@ -50,41 +51,44 @@ combined <- bind_rows(darts_df, greta_df) |>
                                         host = "H", 
                                         parasitoid = "P", 
                                         X = "N")) 
+arrow::write_parquet(combined, "manuscript/data/forecasts.parquet")
+arrow::write_parquet(sims, "manuscript/data/observations.parquet")
 
-hopf_sims <- sims |> filter(simulation=="hopf")
-stochastic_sims <- sims |> filter(simulation=="stochastic")
-saddle_sims <- sims |> filter(simulation=="saddle")
 
-combined |> filter(simulation=="hopf", reps == 1, group == 1) |> 
+forecasts <- read_parquet("manuscript/data/forecasts.parquet")
+observations <- read_parquet("manuscript/data/observations.parquet")
+
+hopf_obs <- observations |> filter(simulation=="hopf")
+stochastic_obs <- observations |> filter(simulation=="stochastic")
+saddle_obs <- observations |> filter(simulation=="saddle")
+
+forecasts |> filter(simulation=="hopf", reps == 1, group == 1) |> 
   ggplot(aes(t, col=type, fill=type)) + 
[…]

### 2022-05-12 manuscript/manuscript.Rmd
missed one

@@
   geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_sims) +
+  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_obs) +
   geom_line(aes(y=mean)) +

### 2022-05-12 manuscript/manuscript.Rmd
missed a few

@@
   geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_sims) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_obs) +
   geom_line(aes(y=mean)) +
@@
   geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_sims) +
+  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_obs) +
   geom_line(aes(y=mean)) +

### 2022-05-12 manuscript/manuscript.Rmd
ensemble model scores + updated plots

@@
 
-```{r setup}
+```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
+knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
 library(tidyverse)
 library(patchwork)
+library(arrow)
+
+bound <- function(x, percentile = 0.975) {
+  dplyr::nth(x, percentile * length(x), order_by = x)
+}
 ```
@@
 
-forecasts <- read_parquet("data/forecasts.parquet")
+forecasts <- read_parquet("data/forecasts.parquet") |> filter(forecasting_model != "ml_ensemble")
 observations <- read_parquet("data/observations.parquet")
 
-hopf_obs <- observations |> filter(simulation=="hopf")
-stochastic_obs <- observations |> filter(simulation=="stochastic")
-saddle_obs <- observations |> filter(simulation=="saddle")
+hopf_obs <- observations |> mutate(variable = as.character(variable)) |> filter(iter %in% 1:15, simulation == "hopf", variable !="X")
+saddle_obs <- observations |> filter(iter %in% 1:15, simulation == "saddle")
+stochastic_obs <- observations |> filter(iter %in% 1:15, simulation == "stochastic")
+
 ```
@@
 ```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
-forecasts |> filter(simulation=="hopf", reps == 1, group == 1) |> 
-  ggplot(aes(t, col=type, fill=type)) + 
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha=0.5, data = hopf_obs) +
-  geom_line(aes(y=mean)) +
-  geom_line(aes(y=mean), data = hopf_obs) +
+forecasts  |> 
+  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = hopf_obs) +
+  geom_line(alpha=0.1) +
+  geom_line(data = filter(hopf_obs, type=="historical")) +
   facet_grid(forecasting_model ~ variable, scales = "free") +
   theme_bw()  + 
+  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
   ggtitle("Hopf bifurcation")
@@
 ```{r}
-forecasts |> filter(simulation=="saddle", group == 1, reps == 1) |> 
-  ggplot(aes(t, col=type, fill=type)) + 
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = saddle_obs) +
-  geom_line(aes(y=mean)) +
-  geom_line(aes(y=mean), data = saddle_obs) +
+
+forecasts |> 
+  filter(iter %in% 1:15, simulation == "saddle") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = saddle_obs) +
+  geom_line(alpha=0.1) +
+  geom_line(data = filter(saddle_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  #facet_grid(forecasting_model ~ reps, scales = "free") +
-  theme_bw() + 
+  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
   ggtitle("Saddle-Node bifurcation")
@@
 ```{r}
-forecasts |> filter(simulation=="stochastic", group == 1, reps==1) |> 
-  ggplot(aes(t, col=type, fill=type)) + 
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5) +
-  geom_ribbon(aes(ymin = pmax(lower,0), ymax = upper), alpha=0.5, data = stochastic_obs) +
-  geom_line(aes(y=mean)) +
-  geom_line(aes(y=mean), data = stochastic_obs) +
+
+forecasts |>
+  filter(iter %in% 1:15, simulation == "stochastic") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
+  geom_line(alpha=0.2) +
+  geom_line(data = filter(stochastic_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  #facet_grid(forecasting_model ~ reps, scales = "free") +
-  theme_bw() + 
+  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
   ggtitle("Stochastic transition")
@@
   ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
-  geom_jitter(alpha = 0.006, show.legend = FALSE) + 
+  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
   coord_flip() +
@@
 ```{r}
-# scores over time, averaged over replicates
-over_time <- shift |> 
-  group_by(t, simulation,forecasting_model, metric ) |> 
-  summarise(sd = sd(score),
-            score = mean(score),
-            .groups = "drop") |>
+shift |> 
   group_by(simulation,forecasting_model, metric) |> 
-  mutate(t = t - min(t)) |> ungroup()
+  mutate(t = t - min(t)) |> ungroup() |> 
+  filter(metric == "crps", t<100) |> 
+  ggplot(aes(t, score, col = forecasting_model, fill=forecasting_model)) + 
+  geom_line(aes(group=iter), alpha = 0.05, show.legend=FALSE) +
+  facet_grid(simulation ~ forecasting_model, scales="free") + 
+  theme_bw() 
+```
 
-p1 <- over_time |> filter(metric == "crps") |>
-  ggplot(aes(t, score, col = forecasting_model)) + 
-  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
-                  shape=".", alpha=0.4, show.legend=FALSE) +
-  geom_line(show.legend=FALSE) +
-  facet_wrap(~simulation, scales="free", ncol=1) + 
-  theme_bw()
-
-p2 <- over_time |> filter(metric == "logs") |>
-  ggplot(aes(t, score, col = forecasting_model)) + 
-  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
-                  shape=".", alpha=0.4, show.legend=FALSE) +
-  geom_line(show.legend=FALSE) +
-  facet_wrap(~simulation, scales="free", ncol=1) + 
-  scale_y_log10() +
-  theme_bw()
-
-p1 + p2
-```
 ```{r}
-over_time |> filter(metric == "logs") |>
-  ggplot(aes(t, score, col = forecasting_model)) + 
-  geom_pointrange(aes(y = score, ymin = score - sd, ymax = score + sd),
-                  shape=".", alpha=0.4, show.legend=FALSE) +
-  geom_line(show.legend=FALSE) +
- # facet_wrap(~simulation, scales="free", ncol=1) + 
+shift |> 
+  group_by(simulation,forecasting_model, metric) |> 
+  mutate(t = t - min(t)) |> ungroup() |> 
+  filter(metric == "logs", t<100) |> 
+  ggplot(aes(t, score, col = forecasting_model, fill=forecasting_model)) + 
+  geom_line(aes(group=iter), alpha = 0.05, show.legend=FALSE) +
   facet_grid(simulation ~ forecasting_model, scales="free") + 

### 2022-05-12 darts-plots.R

@@ -1,99 +1,81 @@
 library(arrow)
 library(tidyverse)
-
+library(forcats)
 bound <- function(x, percentile = 0.975) {
   dplyr::nth(x, percentile * length(x), order_by = x)
 }
 
 reference <- arrow::open_dataset("greta/forecasts/", format="csv") |> 
   filter(type != "predicted") |> 
-  collect()
-
-sims <- reference |>
-  group_by(simulation, variable, type, t) |> 
-  summarise(mean = mean(value), 
-            upper = bound(value, 0.975), 
-            lower = bound(value, 0.025), 
-            .groups = "drop") |>
-  mutate(variable = forcats::fct_recode(variable, 
+  collect() |>
+  mutate(variable = as.character(fct_recode(variable, 
                                         host = "H", 
                                         parasitoid = "P", 
-                                        X = "N")) 
+                                        X = "N")),
+         type = as.character(fct_recode(type, observed = "true"))) |>
+  rename(iter = i)
 
 
+# group indicates the identifier for training data (0:5)
+# reps indicates how many replicate training data sets were used (1 or 10)
+# here we focus only on the case of a single training series.
+# note that the MCMC data only consider rep=1, group=1 case anyway.
 darts <- arrow::open_dataset("darts/forecasts/", format="csv") |>
-  filter(type == "predicted")
+  filter(type == "predicted") |> 
+  filter(group == 1, reps == 1)
 greta <- arrow::open_dataset("greta/forecasts/", format="csv") |>
   mutate(forecasting_model = "MCMC", reps=1, group=1) |>
   filter(type == "predicted") |> 
   rename(iter = i)
 
-  
-  
-## Compute summary stats in arrow before reading into R
-darts_df <- darts |>  
-  collect() |> 
-  group_by(simulation, variable, type, t, reps, group, forecasting_model) |> 
-  summarise(mean = mean(value), 
-            upper = bound(value, 0.975), 
-            lower = bound(value, 0.025), 
-            .groups = "drop")
-greta_df <- greta |>
-  collect() |> 
-  group_by(simulation, variable, type, t, reps, group, forecasting_model) |> 
-  summarise(mean = mean(value), 
[…]

### 2022-05-12 scores-plots.R

@@ -12,12 +12,21 @@ reference <- arrow::open_dataset("greta/forecasts/", format="csv") |>
                                         X = "N")) 
 
 
-darts <- arrow::open_dataset("darts/forecasts/", format="csv")|> filter(type == "predicted")
+darts <- arrow::open_dataset("darts/forecasts/", format="csv")|>
+  filter(type == "predicted")
+ensemble <- darts |> 
+  mutate(iter = paste0(iter,"-", forecasting_model), 
+                       forecasting_model="ml_ensemble") |>
+  collect()
+
+## eek big
+#darts <- bind_rows(collect(darts), collect(ensemble))
+
+
 greta <- arrow::open_dataset("greta/forecasts/", format="csv") |>
   mutate(forecasting_model = "MCMC", reps=1, group=1) |> 
   filter(type == "predicted") |> rename(iter = i) |> collect() |>
   mutate(variable = forcats::fct_recode(variable,  X = "N")) 
-
 
 
 compute_scores <- function(observed, dat) {
@@ -81,17 +90,6 @@ cases <- darts |>
   select(simulation, forecasting_model, variable) |>
   distinct() |> 
   collect()
-
-
-
-x <- cases[1,]
-s <- score_it(scenario="stochastic",
-         model = "block_rnn",
-         var ="X", darts, reference)
-
-
-
-
 for (i in 1:nrow(cases)) {
       scenario <- cases$simulation[[i]]
       model <- cases$forecasting_model[[i]]
@@ -113,45 +111,66 @@ for (i in 1:nrow(cases)) {
   write_csv(df, paste0("scores/", paste(scenario, model, var,sep="-"), ".csv.gz"))
 }
 
+cases <- ensemble |> 
+  select(simulation, forecasting_model, variable) |>
+  distinct() |> 
+  collect()
+for (i in 1:nrow(cases)) {
+  scenario <- cases$simulation[[i]]
+  model <- cases$forecasting_model[[i]]
+  var <- cases$variable[[i]]
+  df <- score_it(scenario, model, var, ensemble, reference)
+  write_csv(df, paste0("scores/", paste(scenario, model, var,sep="-"), ".csv.gz"))
+}
 
 
 
-scores <- arrow::open_dataset("scores", format="csv") |> collect()
 
-# shift logs score to positive?
[…]

### 2022-05-12 manuscript/manuscript.Rmd
bullet pts for discussion

@@
 
-reflects a fundamental challenge of ecological forecasting.
+- overall forecasting bifurcations is difficult, even in the idealized case of ample measurement data and knowing the structural model exactly.
+
+- ML techniques are mostly bracketed by ARIMA and MCMC as expected (better than arima, worse than estimates using true model), though details vary (can do better than MCMC etc).
+
+- MCMC is most relevant to the saddle-node, which is an especially hard problem for the model-free methods.  block rnn does surprisingly well here all things considered.
+
+- the stochastic context is in some sense the easiest for  ML/statistical methods, as it is the only case where training data includes the phenomenon of the transition.
+
+- MCMC does not always out-perform ML techniques, particularly in CRPS terms. Log likelihood-based skill score
+
+- Forecast skill assessment can be challenging.
+it is essential to pay attention to skill over time.
+Forecast error is not a smooth function of horizon, nor is it always increasing with longer horizons.
+
+- Examining means and summary statistics of variation (confidence intervals) can be limiting too, obscuring aspects such as bimodally distributed scores.
 

### 2022-05-13 manuscript/manuscript.Rmd
Some updates

@@
 Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@Clark2001; @Dietze2018].
-What methods will be most reliable in the face of such uncertainty?
+This motivates the question, "What methods will be most reliable in the face of unobserved conditions?"
+
+In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
+We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
+When machine learning models are able to observe transition events, as in the stochastic case, they generally outperformed MCMC-based forecasting but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+Yet, there are nuances to these assessments of performance since forecasting skill varies over time as we show in @figureX.
+For instance, while most of the ML methods outperform MCMC in the stochastic transition example when forecasting skill is averaged over all time points, MCMC produces more replicates with relatively small errors than the ML models.
+
+There are a number of questions that we have left unanswered in our study that we hope will be addressed in future work.
+In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so general conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
+Neither can we claim that the ML methods employed will translate well to all transition event forecasting  problems in reality, since working with real data will introduce additional difficulties.
+For instance, to train a neural network on more than one real time series, we would likely have to employ a specialized algorithm that would allow for generalization over varying time and state scales.
+
+Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
+This paper is not intended to give a comprehensive answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve.
+Instead, this paper intends to serve as an initial exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
 
@@
 non-linearity,
-We have compared the ability of several cutting edge machine learning approaches for forecasting against statistical and process-based models.
 

### 2022-05-13 manuscript/manuscript.Rmd
Tiny edits

@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], these models are not well-suited for problems with complex bifurcation dyanmics.
+Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], these models are not well-suited for problems with complex bifurcation dynamics.
 Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
@@
 This paper is not intended to give a comprehensive answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve.
-Instead, this paper intends to serve as an initial exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
+Instead, this paper aims to serve as a first exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
 

### 2022-05-13 manuscript/manuscript.Rmd
add means

--- 2022-05-12 manuscript/manuscript.Rmd
@@
   ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
+  #geom_boxplot(col="black", outlier.size=0, outlier.shape = ".", outlier.alpha=0) +
   geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
+  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
   coord_flip() +

### 2022-05-13 manuscript/manuscript.Rmd
intro text

@@
 ---
-title: Estimating Uncertainty for Tipping Points with Deep Learning
+title: "Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning"
 authors:
@@
 
-Uncertainty important to forecasting [@Deitze2018] and decision-making [@Schindler2015]
-
-Machine learning is increasingly effective at forecasting, but methods often focus on creation or assessment only of point predictions [@]. 
-Because uncertainty can play such an intrinsic role in ecological dynamics [@Boettiger2018; @Coulson2004; @Oveskienen], it is essential that machine learning applications to ecological forecasting are able to address probablistic forecasts adequately.
+Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@Deitze2018; @Schindler2015].
+Meanwhile, rapid advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@something]. 
+Will these emerging methods improve the capacity for forecasts in ecological systems as well?
+Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @Coulson2004; @Oveskienen2010].
+These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are simultaneously more important and more difficult to predict than more gradual changes.
+Here, we examine several of the best-known examples of critical transitions in ecological systems.
 We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.

### 2022-05-13 manuscript/manuscript.Rmd
a few refs

@@
 We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
-Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
-While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
-These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
+Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
+Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018].
+These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
 May's model exhibits alternative stable states. 
@@
 Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
-This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analogue in the historical data to date.
-
-No-analog scenarios [@Williams]
+This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data to date.
+Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].
 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001]. 
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
 Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2012].
 
@@
 In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g.
-@vanKampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorsfallacy].
+@vankampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorfallacy].
 To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.
@@
 Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
-This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [].
+This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
 (Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.

### 2022-05-13 manuscript/manuscript.Rmd
Most references are up now

--- 2022-05-13 manuscript/manuscript.Rmd
@@
 
-Uncertainty important to forecasting [@Deitze2018] and decision-making [@Schindler2015]
+Uncertainty important to forecasting [@dietze_iterative_2018] and decision-making [@schindler_portfolio_2015]
 
 Machine learning is increasingly effective at forecasting, but methods often focus on creation or assessment only of point predictions [@]. 
-Because uncertainty can play such an intrinsic role in ecological dynamics [@Boettiger2018; @Coulson2004; @Oveskienen], it is essential that machine learning applications to ecological forecasting are able to address probablistic forecasts adequately.
+Because uncertainty can play such an intrinsic role in ecological dynamics [@Boettiger2018; @hallett_why_2004; @Oveskienen], it is essential that machine learning applications to ecological forecasting are able to address probablistic forecasts adequately.
 We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
@@
 We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
-Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
+Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@@nicholson_balance_1935]. 
 While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
-These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@Folke2009; @].
+These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@folke_principles_2009].
 May's model exhibits alternative stable states. 
@@
 While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
-Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@scoringRules].
+Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@gneiting_strictly_2007].
 We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
@@
 
-The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, orignally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@Nicholson1935; @Nicholson1954a; @Nicholson1954b; @Hastings].
+The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
 We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,
@@
 
-Following @Dakos2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
+Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
 In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
@@
 If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
-However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@Boettiger2018].
+However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
 
@@
 A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
-May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@May1979].
+May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
 As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
 The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
-For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@Scheffer2001].
+For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].
 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001]. 
-Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2014].
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001]. 
+Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@hadly_end_2015].
 
@@
 In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g.
-@vanKampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorsfallacy].
+@kampen_stochastic_1992], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@boettiger_early_2012].
 To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.
@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], these models are not well-suited for problems with complex bifurcation dyanmics.
+Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@hyndman_forecasting_2018], these models are not well-suited for problems with complex bifurcation dynamics.
 Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
-In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@cite].
-Since ARIMA is a very commonly encountered forecasting method, we will refer readers to [@cite] for further discussion.
+In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@hyndman_forecasting_2018].
+Since ARIMA is a very commonly encountered forecasting method, we will refer readers to @hyndman_forecasting_2018 for further discussion.
 
@@
 
-Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
+Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis_statistical_2018].
 Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will often be necessary to use multi-step, probablisitic forecasts. 
@@
 
-A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@cite].
+A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
 We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
-One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
-Another likely cause is that machine learning models commonly overfit the training data [@ying].
+One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
+Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
 Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
-With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@ying].
+With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
 One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
@@
 
-Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao, @lyu, @du].
+Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020].
 Herein we will provide some background on what these machine learning methods are and their benefits.
@@
 
-Recurrent neural networks are the predominant memory-based ML method.
-Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky].
-By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
-However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
-Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
-These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung for more details.
-While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung], so we investigate both methods.
+Recurrent neural networks are the predominant memory-based deep learning method.
+Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky_fundamentals_2020].
+By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus learn temporal dependencies.
+However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@pascanu_difficulty_2013], so there have been specialized neural network architectures designed to avoid these gradient problems.
+Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
+These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
+While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung_empirical_2014], so we investigate both methods.
 
@@
 
-The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani].
-Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
+The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani_attention_2017].
+Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani_attention_2017].
 For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
-Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani].
-The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate Transformers in this study.
+Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
+The Transformer is likely to be a foundational method for future AI research [@bommasani_opportunities_2021], so we considered it critical to investigate Transformers in this study.
 
@@
 
-Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@learningphrase, @cite, @cite].
+Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@cho_learning_2014].
 Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
-It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever].
+It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever_sequence_2014].
 Of the models that we present, Block RNNs and Transformers are the only models that use encoder-decoders.
@@
 
-To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @Gneiting2007's property of a strictly proper score.
+To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @gneiting_strictly_2007's property of a strictly proper score.
 This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
@@
 For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
-Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@Mehta2019].
+Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019].
 <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
@@
 Not even all strictly proper scores will agree on the same relative ranking between forecasts.
-We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
+We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
 Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
@@
   ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
-  #geom_boxplot(col="black", outlier.size=0, outlier.shape = ".", outlier.alpha=0) +
   geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
-  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
   coord_flip() +
@@
 Ecological systems have long been acknowledged as complex, due not only to the immense span of dimension and scale such processes involve, but also the frequency of emergent and non-linear phenomena such as stochastic resonance, including bifurcations, tipping points, and hysteresis examined here.
-Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@Clark2001; @Dietze2018].
+Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@clark_ecological_2001; @dietze_iterative_2018].
 This motivates the question, "What methods will be most reliable in the face of unobserved conditions?"

### 2022-05-13 manuscript/manuscript.Rmd

@@
 
-Uncertainty important to forecasting [@dietze_iterative_2018] and decision-making [@schindler_portfolio_2015]
+Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
+Meanwhile, rapid advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@something]. 
+Will these emerging methods improve the capacity for forecasts in ecological systems as well?
+Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
+These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are simultaneously more important and more difficult to predict than more gradual changes.
+Here, we examine several of the best-known examples of critical transitions in ecological systems.
 
-Machine learning is increasingly effective at forecasting, but methods often focus on creation or assessment only of point predictions [@]. 
-Because uncertainty can play such an intrinsic role in ecological dynamics [@Boettiger2018; @hallett_why_2004; @Oveskienen], it is essential that machine learning applications to ecological forecasting are able to address probablistic forecasts adequately.
 We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
@@
 We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology:
-Robert May's consumer-resource model [@May1979], and the Nicholson-Bailey parasatoid-host model [@@nicholson_balance_1935]. 
-While population dynamics models used to inform decision making in fields such as fisheries are often much more complex, these simple models nevertheless exhibit rich nonlinear dynamics thought to be typical in ecological systems [@].
-These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points [] which has had important theoretical and practical management outcomes [@folke_principles_2009].
+
+Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@nicholson_balance_1935]. 
+Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018].
+These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
+
 May's model exhibits alternative stable states. 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001]. 
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001; @Folke2004]. 
 Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dei2011; @Dei2012] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@hadly_end_2015].
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
 

### 2022-05-13 manuscript/manuscript.Rmd
Deleting some new lines

@@
 Here, we examine several of the best-known examples of critical transitions in ecological systems.
-
 We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
@@
 We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
-
 Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@nicholson_balance_1935]. 
@@
 These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
-
 May's model exhibits alternative stable states. 

### 2022-05-13 manuscript/manuscript.Rmd
refs

--- 2022-05-13 manuscript/manuscript.Rmd
@@
 
-Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
+Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@Deitze2018; @Schindler2015].
 Meanwhile, rapid advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@something]. 
 Will these emerging methods improve the capacity for forecasts in ecological systems as well?
-Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
+Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @Coulson2004; @Oveskienen2010].
 These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are simultaneously more important and more difficult to predict than more gradual changes.
 Here, we examine several of the best-known examples of critical transitions in ecological systems.
-
 We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
@@
 We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
-
-Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@nicholson_balance_1935]. 
+Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@Nicholson1935]. 
 Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018].
 These textbook models have been well studied and form the basis of half a century of research in ecology, including much recent work on topics such as resilience and tipping points which has had important theoretical and practical management outcomes [@Folke2004; @Fischer2009; @Polasky2011].
-
 May's model exhibits alternative stable states. 
@@
 While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
-Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@gneiting_strictly_2007].
+Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability (Logs) scores for forecast ensembles [@scoringRules].
+Following popular conventions, we express both skill scores in error-orientation, that is, larger values indicate worse skill (higher degree of error).
 We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
@@
 
-The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
+The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, orignally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@Nicholson1935; @Nicholson1954a; @Nicholson1954b; @Hastings].
 We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,
@@
 
-Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
+Following @Dakos2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
 In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
@@
 If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
-However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
+However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@Boettiger2018].
+Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.
 
@@
 A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
-May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
+May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@May1979].
 As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
 The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
-For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].
+For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@Scheffer2001].
 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001; @Folke2004]. 
-Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
+Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@Carpenter2011], and postulated as a model for global change [@Barnosky2012].
+Examples use the following parameters: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = 0.15$, $\alpha = 0.000375$, $\sigma = 0.02$, $N_0 = 0.75$.
 
@@
 In the case of small noise, transitions are Poisson distributed, such that the distribution of waiting times is roughly exponential [e.g.
-@kampen_stochastic_1992], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@boettiger_early_2012].
+@vankampen], though post-hoc the trajectories of such transitions can be mistaken for saddle-node transitions [@prosectorfallacy].
 To consider such cases, we will again use May's alternative stable state model, though this time leaving all parameters fixed.
@@
 In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
+Model definition is the same as May's model for the saddle node with fixed parameter $h$, values: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = h = 0.26$, $\alpha = 0$, $\sigma = 0.02$, $N_0 = 0.55$.
 
@@
 
-<!-- Key distinction here from MCMC above is not on how model parameters are estimated, but rather on approaches that aren't making strong mechanistic assumptions about process, but are merely phenomonological type models
+<!-- Key distinction here from MCMC above is not on how model parameters are estimated, but rather on approaches that aren't making strong mechanistic assumptions about process, but are merely phenomenological type models
 -->
@@
 We present forecasts produced by ARIMA models as the model-free analogs to the forecasts made using parameter estimation with MCMC.
-Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@hyndman_forecasting_2018], these models are not well-suited for problems with complex bifurcation dynamics.
+Since ARIMA models make the assumption that the future will resemble the past via ARIMA's auto-regressive and moving average components [@cite], these models are not well-suited for problems with complex bifurcation dyanmics.
 Thus, ARIMA-based forecasts should be treated as a lower bound for the performance of non-mechanistic models.
-In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@hyndman_forecasting_2018].
-Since ARIMA is a very commonly encountered forecasting method, we will refer readers to @hyndman_forecasting_2018 for further discussion.
+In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@cite].
+Since ARIMA is a very commonly encountered forecasting method, we will refer readers to [@cite] for further discussion.
 
@@
 
-Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis_statistical_2018].
+Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
 Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will often be necessary to use multi-step, probablisitic forecasts. 
@@
 
-A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
+A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@cite].
 We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
-One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
-Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
+One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
+Another likely cause is that machine learning models commonly overfit the training data [@ying].
 Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
-With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
+With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@ying].
 One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
@@
 
-Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020].
+Recently, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance on time series data [@kao, @lyu, @du].
 Herein we will provide some background on what these machine learning methods are and their benefits.
@@
 
-Recurrent neural networks are the predominant memory-based deep learning method.
-Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky_fundamentals_2020].
-By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus learn temporal dependencies.
-However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@pascanu_difficulty_2013], so there have been specialized neural network architectures designed to avoid these gradient problems.
-Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
-These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
-While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung_empirical_2014], so we investigate both methods.
+Recurrent neural networks are the predominant memory-based ML method.
+Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky].
+By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
+However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
+Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
+These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung for more details.
+While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung], so we investigate both methods.
 
@@
 
-The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani_attention_2017].
-Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani_attention_2017].
+The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani].
+Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
 For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
-Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
-The Transformer is likely to be a foundational method for future AI research [@bommasani_opportunities_2021], so we considered it critical to investigate Transformers in this study.
+Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani].
+The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate Transformers in this study.
 
@@
 
-Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@cho_learning_2014].
+Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@learningphrase, @cite, @cite].
 Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
-It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever_sequence_2014].
+It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever].
 Of the models that we present, Block RNNs and Transformers are the only models that use encoder-decoders.
@@
 
-To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @gneiting_strictly_2007's property of a strictly proper score.
+To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @Gneiting2007's property of a strictly proper score.
 This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
@@
 For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
-Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019].
+Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@Mehta2019].
 <!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
@@
 Not even all strictly proper scores will agree on the same relative ranking between forecasts.
-We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
+We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @scoringRules; @Gneiting2014].
 Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
@@
 
-We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifucation, saddle-node bifurcation, stochastic transition).
-Using model-based simulations allows us to examine performance against an ensemble of replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
-
+We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
+In addition to these cases, we also consider an "ensemble model", generated from drawing from the distribution of all models except the MCMC model. 
+Such ensemble techniques can better reflect uncertainty than relying on any single method [@something_about_ensembles].
+For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
+Using model-based simulations allows us to examine performance against an multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
 By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
-Note that observations are constrained to a smaller region of state space (i.e. a smaller range of population sizes) than will be realized in future. 
+Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e.
+very small population sizes).
 This no-analog aspect of forecasting bifurcation dynamics means that even with many sample points in the training data _and_ perfect knowledge of the true model structure, posterior distributions of parameter values are still influenced by the choice of priors.
-
-```{r}
-
-forecasts <- read_parquet("data/forecasts.parquet") |> filter(forecasting_model != "ml_ensemble")
-observations <- read_parquet("data/observations.parquet")
-
-hopf_obs <- observations |> mutate(variable = as.character(variable)) |> filter(iter %in% 1:15, simulation == "hopf", variable !="X")
-saddle_obs <- observations |> filter(iter %in% 1:15, simulation == "saddle")
-stochastic_obs <- observations |> filter(iter %in% 1:15, simulation == "stochastic")
-
-```
-
-```{r fig.cap="Forecasts of the Hopf bifurcation under each model"}
-forecasts  |> 
-  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
-  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
-  geom_line(alpha=0.1, data = hopf_obs) +
-  geom_line(alpha=0.1) +
-  geom_line(data = filter(hopf_obs, type=="historical")) +
-  facet_grid(forecasting_model ~ variable, scales = "free") +
-  theme_bw()  + 
-  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
-  ggtitle("Hopf bifurcation")
-```
-
-```{r}
-
-forecasts |> 
-  filter(iter %in% 1:15, simulation == "saddle") |> 
-  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
-  geom_line(alpha=0.1, data = saddle_obs) +
-  geom_line(alpha=0.1) +
-  geom_line(data = filter(saddle_obs, type=="historical")) +
-  facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
-  ggtitle("Saddle-Node bifurcation")
-```
-
-```{r}
-
-forecasts |>
-  filter(iter %in% 1:15, simulation == "stochastic") |> 
-  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
-  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
-  geom_line(alpha=0.2) +
-  geom_line(data = filter(stochastic_obs, type=="historical")) +
-  facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
-  ggtitle("Stochastic transition")
-```
 
@@
 ```{r}
+#| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
+#| Smaller scores are better (indicating smaller errors).
+#| Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
+#| simulations of the given scenario.  "
 shift |> 
   ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
-  #geom_boxplot(col="black", outlier.size=0, outlier.shape = ".", outlier.alpha=0) +
   geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
@@
   coord_flip() +
+  labs(y="") +
   facet_grid(simulation ~ metric, scales = "free") + 
@@
   theme_bw()
+```
 
+Overall forecasting skill scores are summarized in figure 2.
+
+```{r}
+forecasts <- read_parquet("data/forecasts.parquet") |> filter(forecasting_model != "ml_ensemble")
+observations <- read_parquet("data/observations.parquet")
+
+hopf_obs <- observations |> mutate(variable = as.character(variable)) |> filter(iter %in% 1:15, simulation == "hopf", variable !="X")
+saddle_obs <- observations |> filter(iter %in% 1:15, simulation == "saddle")
+stochastic_obs <- observations |> filter(iter %in% 1:15, simulation == "stochastic")
+
+```
+
+```{r }
+#| fig2
+#| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
+#| The bifurcation occurs soon after forecasting period begins,
+#| leading to progressively larger and larger oscillations.
+#| Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
+#| Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations.
+#| Only last 25 time points of training data are shown."
+
+forecasts  |> 
+  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = hopf_obs) +
+  geom_line(alpha=0.1) +
+  geom_line(data = filter(hopf_obs, type=="historical")) +
+  facet_grid(forecasting_model ~ variable, scales = "free") +
+  theme_bw()  + 
+  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
+  ggtitle("Hopf bifurcation")
+```
+
+Forecasts of the Hopf bifurcation (Figure 2) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
+All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
+Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
+Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
+However, like the ML models, it fails to predict the growing amplitude of those oscillations.
+Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
+
+Those overestimates mean that the initial skill scores of the MCMC model are distinctly worse than those of any ML model during the early phase of the forecasts (Figures 6-7). 
+This error shows up particularly in the CRPS scores (Figure 6), which have a higher penalty for over-estimating the spread than the negative log probabilities (Figure 7).
+While the ML models fail to predict the core dynamical phenomenon of growing oscillations,
+
+<!-- would be interesting to revisit this with lower stochasticity such that the cyclic behavior is more crisp, and the quasi-cycles in the training data are smaller-->
+
+```{r }
+#| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model.
+#| Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult."
+
+forecasts |> 
+  filter(iter %in% 1:15, simulation == "saddle") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = saddle_obs) +
+  geom_line(alpha=0.1) +
+  geom_line(data = filter(saddle_obs, type=="historical")) +
+  facet_wrap(~forecasting_model, scales = "free") +
+  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
+  ggtitle("Saddle-Node bifurcation")
+```
+
+The saddle node bifurcation
+
+```{r}
+#| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model.
+#| In contrast to the other challenges, this case considers the prediction of replicate systems starting from the same initial condition, rather than forecasting the future evolution of the model after the stochastic transition has already occurred."
+forecasts |>
+  filter(iter %in% 1:15, simulation == "stochastic") |> 
+  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
+  geom_line(alpha=0.2) +
+  geom_line(data = filter(stochastic_obs, type=="historical")) +
+  facet_wrap(~forecasting_model, scales = "free") +
+  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
+  ggtitle("Stochastic transition")
 ```
@@
 Ecological systems have long been acknowledged as complex, due not only to the immense span of dimension and scale such processes involve, but also the frequency of emergent and non-linear phenomena such as stochastic resonance, including bifurcations, tipping points, and hysteresis examined here.
-Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@clark_ecological_2001; @dietze_iterative_2018].
+Calls for increased forecasting efforts from ecologists frequently reference the role of changing climate and other anthropogenic change, which raise the challenge of prediction in no-analog environments, anticipating ecosystem responses to conditions that have not been previously observed [@Clark2001; @Dietze2018].
 This motivates the question, "What methods will be most reliable in the face of unobserved conditions?"
@@
 
+Our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
+Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018].
+Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
+Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
+Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
+We leave these cases to future exploration.
+
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
 This paper is not intended to give a comprehensive answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve.
-Instead, this paper aims to serve as a first exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
+Instead, this paper intends to serve as an initial exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
 
@@
 
+- Forecasts with 'assimilation' i.e. ability to re-estimate after seeing some of the data, rather than looking at a single forecast with a long horizon.
+
 non-linearity,

### 2022-05-13 manuscript/manuscript.Rmd

--- 2022-05-13 manuscript/manuscript.Rmd
@@
 While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
-Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@gneiting_strictly_2007].
+
+Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability (Logs) scores for forecast ensembles [@gneiting_strictly_2007].
+Following popular conventions, we express both skill scores in error-orientation, that is, larger values indicate worse skill (higher degree of error).
+
 We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
@@
 However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
+Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.
 
@@
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
-Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001; @Folke2004]. 
-Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
+
+Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
+Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
+Examples use the following parameters: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = 0.15$, $\alpha = 0.000375$, $\sigma = 0.02$, $N_0 = 0.75$.
 
@@
 
+Our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
+Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018].
+Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
+Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
+Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
+We leave these cases to future exploration.
+
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.

### 2022-05-13 manuscript/manuscript.Rmd
results, fig captions

@@
 Not even all strictly proper scores will agree on the same relative ranking between forecasts.
-We will focus on two of the most common such skill metrics, CRPS score and logs score [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
+We will focus on two of the most common such skill metrics, CRPS score and log probability score (negative log likelihood) [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
 Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
+Note that while the minus log-likelihood can be negative for sufficiently high probability densities, we use a fixed scalar shift of logs score to ensure the log skill score is strictly positive, which facilitates visualization without impacting relative rankings.
 
@@
 Such ensemble techniques can better reflect uncertainty than relying on any single method [@something_about_ensembles].
-For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
+For simplicity, we consider the un-weighted case, where each model is represented equally in the ensemble.
 Using model-based simulations allows us to examine performance against an multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
@@
 scores <- arrow::open_dataset("../scores", format="csv") |> collect()
-
 # shift logs score to strictly positive, non-infinite values so we can use a log-scale
@@
 #| Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
-#| simulations of the given scenario.  "
+#| simulations of the given scenario."
 shift |> 
@@
   coord_flip() +
-  labs(y="") +
   facet_grid(simulation ~ metric, scales = "free") + 
   scale_y_log10() +
-  theme_bw()
+  theme_bw() +
+  theme(axis.title.y = element_blank())
 ```
 
-Overall forecasting skill scores are summarized in figure 2.
+Overall forecasting skill scores for each model across all three scenarios are summarized in figure 2.
+Average scores (black lines) hide wide variation in forecast skill.
+Generally, ML performance tends to be bracketed between MCMC (essentially the theoretical optimum), and the statistical ARIMA model, though sometimes performing worse than ARIMA or better than MCMC.
+Under scenarios with alternative stable states (saddle and stochastic), the distribution of scores is often bimodal for ML models, though not MCMC. 
+The ML ensemble model often performs as well as the best ML model on average.
+Note that a wide prediction of uncertainty does not mean a wide range in the score skill -- for instance the ensemble model which has the widest array of outcomes often has the a relatively tight distribution of score, especially in logs skill.
+This reflects the relative contributions of accuracy and uncertainty as components in the forecasts.
+Overall, ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
 
@@
 ```{r }
-#| fig2
+#| fig3
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
@@
 
-Forecasts of the Hopf bifurcation (Figure 2) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
+Forecasts of the Hopf bifurcation (Figure 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
 All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
@@
 Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
-
-Those overestimates mean that the initial skill scores of the MCMC model are distinctly worse than those of any ML model during the early phase of the forecasts (Figures 6-7). 
-This error shows up particularly in the CRPS scores (Figure 6), which have a higher penalty for over-estimating the spread than the negative log probabilities (Figure 7).
-While the ML models fail to predict the core dynamical phenomenon of growing oscillations,
+Despite this, overall all methods score comparably in CRPS score (Fig 2), with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
+A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.
 
@@
 
-The saddle node bifurcation
+The saddle node bifurcation proves even more difficult for most methods (Figure 4).
+Only the MCMC model anticipates the sharp transition to an alternative state, though this behavior is more baked into the method from the start which takes the true model structure as given. 
+Even accurate estimation of the MCMC requires slightly informative priors, though still broad enough to reflect a wide range of possible outcomes. 
+Two ML methods -- Block RNNs and transformers -- resemble a naive prediction extrapolating the last observed state, failing even to reflect the slow downward trend of the training data.  
+LSTM indicates greater uncertainty, while GRU shows very large variability which spans the alternative stable state range. 
+With additional tuning, better performance may be possible for these ML models. 
+The selected ARIMA model reflects wide uncertainty that is nevertheless not broad enough to span the alternative stable state.
+Consequentially, the MCMC estimate easily outperforms the ML models (Fig 2).
+The distribution of scores over
 
@@
 ```
+Machine learning methods do markedly better on the stochastic transition scenario than in the two bifurcation scenarios (Fig 5).
+This occurs because the training data includes the transition phenomenon of interest.
+All ML models accurately capture the dynamics of a sharp transition between alternative stable states -- a dynamic the statistical ARIMA model entirely fails to reflect.
+Stochastic transition events should be approximately exponentially distributed, as seen in the wide range of waiting times for transitions to occur in replicates of the true 'observed' process (Fig 5).
+Transformer and Block RNN distribution times are much more concentrated, while again GRU and especially the LSTM do a better job reflecting the uncertainty in range of transition times.
 
 ```{r}
+#| fig6
+#| fig.cap="CRPS scores over time for each scenario.
+#| Each line represents the scores against a replicate of time series observations from 'true model'.
+#| Patterns show forecast skill generally get worse over time -
+#| gradually in the case of Hopf bifurcation or suddenly in repsonse to the saddle-node bifurcation.
+#| The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window.
+#| In the stochastic transition context, this creates two branches, where high values indicate periods of
+#| time when the forecast predicts the wrong state and lower branch indicates the correct state."
+
 shift |> 
@@
   facet_grid(simulation ~ forecasting_model, scales="free") + 
-  theme_bw() 
+  theme_bw()  + ggtitle("CRPS scores over time")
 ```
@@
 ```{r}
+#| fig.cap="Logs skill score over time. 
+#| Forecasts which underestimate uncertainty do substantially worse in logs score than in CRPS score.
+#| Comparing this panel to those in Figure 6 highlights scenarios that most often underestimate uncertainty.
+#| Generally, MCMC performs better relative to other models under this metric than it does under CRPS,
+#| reflecting the bias-variance tradeoff of machine learning."
+
 shift |> 
@@
   facet_grid(simulation ~ forecasting_model, scales="free") + 
-  theme_bw() + scale_y_log10()
+  theme_bw() + scale_y_log10() + ggtitle("Logs skill score over time")
 ```
@@
 
-non-linearity,
-
-- Ecological forecasting invariably is hard
-
 \pagebreak

### 2022-05-13 manuscript/manuscript.Rmd

@@
 
-Forecasting plays an important and rapidly growing role in both testing fundamental understanding of ecological processes and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
-Meanwhile, rapid advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@something]. 
+Forecasting plays an important and rapidly growing role in both testing our fundamental understanding of ecological processes, and informing ecological applications and conservation decision-making [@dietze_iterative_2018; @schindler_portfolio_2015].
+Meanwhile, recent advances in machine learning have rapidly improved the prevalence and accuracy of short term forecasts in many fields [@kao_exploring_2020; @lyu_lstm_2020; @du_multivariate_2020]. 
 Will these emerging methods improve the capacity for forecasts in ecological systems as well?
 Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
-These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are simultaneously more important and more difficult to predict than more gradual changes.
+These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
 Here, we examine several of the best-known examples of critical transitions in ecological systems.
@@
 <!-- some more verbose bits about bifurcation models probably belongs in discussion instead -->
-We will focus the analysis of several different forecasting scenarios based around two classic models in population ecology: 
-Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasatoid-host model [@nicholson_balance_1935]. 
+We will focus the analysis on several different forecasting scenarios based around two classic models in population ecology: 
+Robert May's consumer-resource model [@May1977], and the Nicholson-Bailey parasitoid-host model [@nicholson_balance_1935]. 
 Though these models may appear simple when measured against high-dimensional and parameter rich models found in some management contexts such as fisheries, they can exhibit rich nonlinear dynamics and provide greater capacity to generalize [@Levins1966; @Getz2018].
@@
 
-Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is particularly important question for ecological systems and global environmental change problems.
+Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is a particularly important question for ecological systems and global environmental change problems.
 Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
@@
 
-We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
-After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting.
-Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
-Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
-While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
+We provide fully reproducible coded examples in R and Python for performing, scoring, and visualizing each of the forecasts considered here.
+After significant time spent considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting.
+Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
+Our model-based MCMC forecasts use the `greta` framework, an R library that uses Python-based `tensorflow` probability to achieve better performance.
+While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
 
@@
 
-We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
+We expect greater convergence between methods available in R and Python in the future, as already illustrated in the example of `greta`.
 Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.
@@
 A yet more difficult forecasting scenario is created by the saddle-node bifurcation.
-May's consumer-resource model is a one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
-As in the Nicholson-Bailey model, in the absence of that predation the resource population density grows under a density-dependent pattern described by a logistic function.
+May's consumer-resource model is an one-dimensional model describing the growth of a 'resource' population (e.g. herbivore) which is grazed by a consumer [@may_population_1979].
+As in the Nicholson-Bailey model, in the absence of that predation, the resource population density grows under a density-dependent pattern described by a logistic function.
 The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
@@
 Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
-Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
+Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on its historical values, we consider an alternative framing of the task:
 we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
@@
 In contrast to inference with MCMC, uncertainty with ARIMA models is estimated directly from the learned parameters [@hyndman_forecasting_2018].
-Since ARIMA is a very commonly encountered forecasting method, we will refer readers to @hyndman_forecasting_2018 for further discussion.
+Since ARIMA is a commonly encountered method, we will refer readers to @hyndman_forecasting_2018 for further discussion.
 
@@
 Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
-Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized. 
+Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized [@dar_farewell_2021]. 
 With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
-One frequently cited method to reduce overfitting is K-fold cross validation, but this approach cannot be effectively employed when there is one or few time series in the training set.
+One frequently used method to reduce overfitting is K-fold cross validation [@raschka_model_2020], but this approach cannot be effectively employed when there is one or few time series in the training set.
 To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
@@
 Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani_attention_2017].
-For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
-Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
+For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent neural network would process all the words as one phrase.
+Because of self-attention, Transformers do not need to process data sequentially and thus can be parallelized, offering significant computational advantages [@vaswani_attention_2017].
 The Transformer is likely to be a foundational method for future AI research [@bommasani_opportunities_2021], so we considered it critical to investigate Transformers in this study.
@@
 We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
-In addition to these cases, we also consider an "ensemble model", generated from drawing from the distribution of all models except the MCMC model. 
-Such ensemble techniques can better reflect uncertainty than relying on any single method [@something_about_ensembles].
-For simplicity, we consider the un-weighted case, where each model is represented equally in the ensemble.
-Using model-based simulations allows us to examine performance against an multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
+In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
+Such ensemble techniques can better reflect uncertainty than relying on any single method [@gneiting_weather_2005].
+For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
+Using model-based simulations allows us to examine performance against a multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
 By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
@@
 Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
-However, like the ML models, it fails to predict the growing amplitude of those oscillations.
+However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
 Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.

### 2022-05-14 manuscript/manuscript.Rmd
Small edits

@@
 Here, we examine several of the best-known examples of critical transitions in ecological systems.
-We evaluate the ability of some of the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
+We evaluate the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
 
@@
 
-The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parsatiod, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
+The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parasitoid, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
 We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,
@@
 Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
-In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
+In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
 This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
@@
 
-We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, stochastic transition).
+We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic transition).
 In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
@@
 For simplicity, we consider the unweighted case, where each model is represented equally in the ensemble.
-Using model-based simulations allows us to examine performance against a multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance.
+Using model-based simulations allows us to examine performance against multiple (n=100) replicates of the "true" process, which further helps identify differences that may occur solely due to chance. 
 By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.

### 2022-05-14 manuscript/manuscript.Rmd
More edits all around here

@@
 Assessing the accuracy of forecasting methods in the face of such bifurcation dynamics is a particularly important question for ecological systems and global environmental change problems.
-Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters (i.e. the 'environment' in which those dynamics take place.)
-This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data to date.
+Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters.
+This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data.
 Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].
@@
 Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
-Our model-based MCMC forecasts use the `greta` framework, an R library that uses Python-based `tensorflow` probability to achieve better performance.
-While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)
+Our model-based MCMC forecasts use the `greta` framework, a R library that uses Python-based `tensorflow` probability to achieve better performance.
+While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores.
 
-Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability (Logs) scores for forecast ensembles [@gneiting_strictly_2007].
+Our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability (Logs) scores for forecast ensembles [@gneiting_strictly_2007].
 Following popular conventions, we express both skill scores in error-orientation, that is, larger values indicate worse skill (higher degree of error).
@@
 The resource population is also grazed by a consumer at a rate given by a Holling type III s-curve (typically used to model handling time). 
-For a certain range of parameter choices, this model supports alternative stable state dynamics, and this model has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].
+For a certain range of parameter choices, this model supports alternative stable state dynamics, and has been identified and employed in explaining alternative stable state dynamics in a broad range of ecological and socio-ecological systems [@scheffer_catastrophic_2001].
 
@@
 
-As a reference case, we consider the forecast produced by an MCMC estimate of model parameters, _given the true model_.
+As a reference case, we consider forecasts produced by MCMC estimates of model parameters, _given the true model_.
 This represents an idealized case where the nature of the underlying process is known precisely.
@@
 each model is trained to learn the parameters of a Poisson distribution to maximize the log likelihood. 
-To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.
+To compute uncertainty intervals, we draw samples from the learned distribution at each time step.
 
 A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
-We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
+We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds.
 One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
@@
 One frequently used method to reduce overfitting is K-fold cross validation [@raschka_model_2020], but this approach cannot be effectively employed when there is one or few time series in the training set.
-To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
-We found this simple ensemble method to be an effective way to improve generalizability in the limited data regime.
+To remedy the instability problem, we use an ensemble-based method, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
+We found this simple ensemble technique to be an effective way to improve generalizability in the limited data regime.
 
@@
 Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
-These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
+These methods avoid gradient problems by regulating the self-feedback via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
 While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung_empirical_2014], so we investigate both methods.
@@
 
-Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@cho_learning_2014].
+Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@aitken_understanding_2021].
 Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
-It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever_sequence_2014].
-Of the models that we present, Block RNNs and Transformers are the only models that use encoder-decoders.
+It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task much easier [@sutskever_sequence_2014].
+Note that it is possible to use any type of neural network as the encoder and the decoder, but it is most common to use recurrent neural networks or networks with attention mechanisms [@aitken_understanding_2021].
+Of the models that we present, Block RNNs and Transformers have encoder-decoder structures.
 
@@
 
-To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy @gneiting_strictly_2007's property of a strictly proper score.
+To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
 This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
@@
 
-We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic transition).
+We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and Transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic transition).
 In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
@@
 The ML ensemble model often performs as well as the best ML model on average.
-Note that a wide prediction of uncertainty does not mean a wide range in the score skill -- for instance the ensemble model which has the widest array of outcomes often has the a relatively tight distribution of score, especially in logs skill.
+Note that a wide prediction of uncertainty does not mean a wide range in the score skill -- for instance the ensemble model which has the widest array of outcomes often has a relatively tight distribution of score, especially in logs skill.
 This reflects the relative contributions of accuracy and uncertainty as components in the forecasts.
@@
 Even accurate estimation of the MCMC requires slightly informative priors, though still broad enough to reflect a wide range of possible outcomes. 
-Two ML methods -- Block RNNs and transformers -- resemble a naive prediction extrapolating the last observed state, failing even to reflect the slow downward trend of the training data.  
+Two ML methods -- Block RNNs and Transformers -- resemble a naive prediction extrapolating the last observed state, failing even to reflect the slow downward trend of the training data.  
 LSTM indicates greater uncertainty, while GRU shows very large variability which spans the alternative stable state range. 
@@
 ```
+
 Machine learning methods do markedly better on the stochastic transition scenario than in the two bifurcation scenarios (Fig 5).
@@
 In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
-We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-When machine learning models are able to observe transition events, as in the stochastic case, they generally outperformed MCMC-based forecasting but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
-Yet, there are nuances to these assessments of performance since forecasting skill varies over time as we show in @figureX.
-For instance, while most of the ML methods outperform MCMC in the stochastic transition example when forecasting skill is averaged over all time points, MCMC produces more replicates with relatively small errors than the ML models.
+We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
+When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
 
-There are a number of questions that we have left unanswered in our study that we hope will be addressed in future work.
-In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so general conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
-Neither can we claim that the ML methods employed will translate well to all transition event forecasting  problems in reality, since working with real data will introduce additional difficulties.
-For instance, to train a neural network on more than one real time series, we would likely have to employ a specialized algorithm that would allow for generalization over varying time and state scales.
+There are a number of questions that we have left unanswered that we hope will be addressed in future work.
+In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
+Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties.
+Some potential issues include varying state and time scales across time series, non-concurrent sampling, sparse data, and more.
 
-Our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
+Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
 Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018].
@@
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
-This paper is not intended to give a comprehensive answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve.
-Instead, this paper aims to serve as a first exploration on whether deep learning methods should be considered as viable tools for the extremely challenging category of no-analog bifurcation forecasting problems.
+This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be as an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
 

### 2022-05-14 manuscript/manuscript.Rmd
More touching up

@@
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts, like the pronounced bimodality of ML forecasts for the stochastic transition case.
 
@@
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
-This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be as an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
+This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
+Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.
 

### 2022-05-16 manuscript/manuscript.Rmd
:pen:

@@
 #| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
-#| Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult."
+#| Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "
 
@@
 Transformer and Block RNN distribution times are much more concentrated, while again GRU and especially the LSTM do a better job reflecting the uncertainty in range of transition times.
+
+Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
+In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
+In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
+while MCMC scores are worst in mid-range.
+Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
+A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation.
+CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state.
+Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g.
+by the MCMC predictions.
+The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
+after the catastrophic transition) are poor.
+In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.
 
@@
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
-Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts, like the pronounced bimodality of ML forecasts for the stochastic transition case.
+All though most ML-based forecasting applications focus on point predictions, we have emphasized examples which can provide a meaningful estimate of uncertainty.
+When machine learning models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
+Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts.
+For instance, forecast skill varies with the length of the forecast horizon, and not in a monotonic fashion. 
+This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict then transient transitions. 
+Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
+It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than less accurate on average but correctly reflect a greater degree of uncertainty.
+The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.
+
+The success of these ML models on the stochastic transition case is particularly notable.
+Using only a single previous replicate
+
+Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in these critical transitions.
+This is not surprising, as the training data provide no prior example of such behavior (e.g. growing oscillations or a sudden shift).
+Nevertheless, this should be an important reminder of a central difficulty in ecological forecasting.
+Note that in such scenarios, near-term forecasts [@dietze_iterative_2018] may be very accurate right up to the transition event before becoming widely wrong. 
+Nor can the possibility of such non-linear behavior be easily dismissed in ecological models -- the examples considered here have been bedrock of ecological modeling and management practices for over half a century [@Folke2004], and if anything are only too simple, representing a small slice of possible dynamical behavior of more complicated models.
+
+Identifiability
 

### 2022-05-16 manuscript/manuscript.Rmd
:pen:

@@
 
-Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
+Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a super-critical Hopf bifurcation as $K$ becomes sufficiently large.
 In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
+In this model, the Hopf bifurcation is dubbed 'super-critical' as it creates a stable limit cycle instead of an unstable one.
 This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
@@
 ```{r}
-forecasts <- read_parquet("data/forecasts.parquet") |> filter(forecasting_model != "ml_ensemble")
+forecasts <- read_parquet("data/forecasts.parquet") |>
+  filter(forecasting_model != "ml_ensemble")
 observations <- read_parquet("data/observations.parquet")
 
-hopf_obs <- observations |> mutate(variable = as.character(variable)) |> filter(iter %in% 1:15, simulation == "hopf", variable !="X")
+hopf_obs <- observations |> 
+  mutate(variable = as.character(variable)) |>
+  filter(iter %in% 1:15, simulation == "hopf", variable !="X")
 saddle_obs <- observations |> filter(iter %in% 1:15, simulation == "saddle")
@@
 The success of these ML models on the stochastic transition case is particularly notable.
-Using only a single previous replicate
+All methods are given only a single previous replicate of a stochastic transition (Fig 5, red line) on which to base there estimates.
+This is typical of ecological scenarios where data is so often limited.
+While even one observation of a transition is more than the methods have in our other forecasts, this still presents a significant challenge to model estimation.
+Unlike the MCMC case, the ML models have no prior expectation of a model structure that contains sharp transitions -- we might have expected these models to perform little differently than the ARIMA model.
+Given this single replicate, all four ML models successfully capture the phenomenological pattern of a sharp shift between two stable states -- behavior which the structurally simpler family of ARIMA models cannot express.
+This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately models with ML models compared to classical statistical models.
+In this way, the ML models can be seen to impose even fewer assumptions on the phenomenological behavior of the system than the ARIMA model.
+In contrast, the MCMC performance benefits from very strong process-based assumptions, which happen to match the 'true' model in this case and thus provide a comparison of the theoretical optimal performance.
 
-Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in these critical transitions.
+Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in the critical transition scenarios (Hopf and saddle node).
 This is not surprising, as the training data provide no prior example of such behavior (e.g. growing oscillations or a sudden shift).
@@
 
-Identifiability
+It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of super-critical Hopf or saddle node bifurcations. 
+This question is not as easy to answer as it may seem, because of the difficulty in defining what real-world forecasting scenario that may correspond to.
+The scenarios we have considered are true, pure forecasts: the training data comes from a single realization of a specific generative process, and the task is to predict the future states of that system before they occur.
+Would it be possible to train an predictive algorithm on 'analogous' examples of critical transitions?
+For instance, could data from other lakes which may have experienced a critical transition such as eutrophication event in the past be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
+Perhaps, but it depends what we mean by an 'analogous' system.
+Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different.
+
+Identifiability MCMC theoretical limits
 

### 2022-05-16 manuscript/manuscript.Rmd
a little bit more in the discussion

@@
 It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of super-critical Hopf or saddle node bifurcations. 
-This question is not as easy to answer as it may seem, because of the difficulty in defining what real-world forecasting scenario that may correspond to.
+This question is not as easy to answer as it may seem, because of the difficulty in defining the corresponding forecasting scenario.
 The scenarios we have considered are true, pure forecasts: the training data comes from a single realization of a specific generative process, and the task is to predict the future states of that system before they occur.
@@
 Perhaps, but it depends what we mean by an 'analogous' system.
-Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different.
+Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different, even after accounting for re-scaling or non-dimensionalization of the model [@Hastings1996].
+Rarely do ecologists have access to completely controlled replicates for fitting or training models.
+The ability for ML models to successfully generalize from training in such cases remains an open problem and a promising subject of further investigation.
 
-Identifiability MCMC theoretical limits
+Even the MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
+Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
+This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
+It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
+In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
+Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
+Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
+Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
+Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
+<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->
 
@@
 
-- overall forecasting bifurcations is difficult, even in the idealized case of ample measurement data and knowing the structural model exactly.
-
-- ML techniques are mostly bracketed by ARIMA and MCMC as expected (better than arima, worse than estimates using true model), though details vary (can do better than MCMC etc).
-
-- MCMC is most relevant to the saddle-node, which is an especially hard problem for the model-free methods.  block rnn does surprisingly well here all things considered.
-
-- the stochastic context is in some sense the easiest for  ML/statistical methods, as it is the only case where training data includes the phenomenon of the transition.
-
-- MCMC does not always out-perform ML techniques, particularly in CRPS terms. Log likelihood-based skill score
-
-- Forecast skill assessment can be challenging.
-it is essential to pay attention to skill over time.
-Forecast error is not a smooth function of horizon, nor is it always increasing with longer horizons.
-
-- Examining means and summary statistics of variation (confidence intervals) can be limiting too, obscuring aspects such as bimodally distributed scores.
-
-- Forecasts with 'assimilation' i.e. ability to re-estimate after seeing some of the data, rather than looking at a single forecast with a long horizon.
-
 \pagebreak

### 2022-05-16 manuscript/manuscript.Rmd
:pen:

@@
 
+The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
+Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
+This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
+It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
+In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
+Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
+Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
+Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
+Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
+<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->
+
 Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in the critical transition scenarios (Hopf and saddle node).
@@
 
-Even the MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
-Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
-This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
-It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
-In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
-Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
-Because the available data come from only a small region of the dynamical state space, they are consistent with many possible parameterizations of the same model structure -- which creates likelihood ridges and non-identifiability of specific parameter values.
-Using more simplified versions of the dynamical processes in question, such as the canonical form of a bifurcation, can mitigate this issue in some cases.
-Even when such non-identifiability issues cannot be avoided entirely, they can usually be diagnosed by examining the degree of mixing in MCMC sampling and comparing posterior to prior distributions.
-<!-- While similar convergence issues may arise in the estimation of ML models, they may be harder to diagnose. -->
-
 There are a number of questions that we have left unanswered that we hope will be addressed in future work.
@@
 Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
-We leave these cases to future exploration.
+We leave these cases to future exploration rather than attempting to explore all such variations in a single narrative. 
+We have focused on these illustrative examples accompanied by an extensive appendix of reproducible code implemented using well documented and intuitive frameworks.
+We encourage the reader to use these appendices as an entry point into further exploration.
 

### 2022-05-16 manuscript/manuscript.Rmd
Added a small paragraph in the discussion on ML models

--- 2022-05-16 manuscript/manuscript.Rmd
@@
 
-In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
+In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting critical transition events.
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-All though most ML-based forecasting applications focus on point predictions, we have emphasized examples which can provide a meaningful estimate of uncertainty.
-When machine learning models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+Although most ML-based forecasting applications focus on point predictions, we have emphasized examples which can provide a meaningful estimate of uncertainty.
+When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
 An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
-Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts.
-For instance, forecast skill varies with the length of the forecast horizon, and not in a monotonic fashion. 
-This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict then transient transitions. 
+Yet, examining summary statistics, CRPS and log probability scores obscures finer detailed components of the forecasts.
+For instance, forecast skill varies with the length of the forecast horizon in a non-monotonic fashion. 
+This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict than transient transitions. 
 Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
 It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than less accurate on average but correctly reflect a greater degree of uncertainty.
+<!-- MFL: "that are less accurate"?-->
 The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.
@@
 The success of these ML models on the stochastic transition case is particularly notable.
-All methods are given only a single previous replicate of a stochastic transition (Fig 5, red line) on which to base there estimates.
+All methods are given only a single previous replicate of a stochastic transition (Fig 5, red line) on which to base their estimates.
 This is typical of ecological scenarios where data is so often limited.
@@
 Given this single replicate, all four ML models successfully capture the phenomenological pattern of a sharp shift between two stable states -- behavior which the structurally simpler family of ARIMA models cannot express.
-This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately models with ML models compared to classical statistical models.
+This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately modeled with ML models compared to classical statistical models.
 In this way, the ML models can be seen to impose even fewer assumptions on the phenomenological behavior of the system than the ARIMA model.
 In contrast, the MCMC performance benefits from very strong process-based assumptions, which happen to match the 'true' model in this case and thus provide a comparison of the theoretical optimal performance.
+
+When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
+Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning subfield [@aitken_understanding_2021].
+These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
+Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often difficult to provide an explanation for why certain methods over-perform or fail to meet expectations.
 
@@
 The scenarios we have considered are true, pure forecasts: the training data comes from a single realization of a specific generative process, and the task is to predict the future states of that system before they occur.
-Would it be possible to train an predictive algorithm on 'analogous' examples of critical transitions?
+Would it be possible to train a predictive algorithm on 'analogous' examples of critical transitions?
 For instance, could data from other lakes which may have experienced a critical transition such as eutrophication event in the past be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
@@
 Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties.
-Some potential issues include varying state and time scales across time series, non-concurrent sampling, sparse data, and more.
+<!-- MFL: I think we could cut this paragraph out completely now -->
 

### 2022-05-16 manuscript/manuscript.Rmd

--- 2022-05-16 manuscript/manuscript.Rmd
@@
 Given this single replicate, all four ML models successfully capture the phenomenological pattern of a sharp shift between two stable states -- behavior which the structurally simpler family of ARIMA models cannot express.
-This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately models with ML models compared to classical statistical models.
+This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately modeled with ML models compared to classical statistical models.
 In this way, the ML models can be seen to impose even fewer assumptions on the phenomenological behavior of the system than the ARIMA model.
@@
 
+When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
+Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning subfield [@aitken_understanding_2021].
+These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
+Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often difficult to provide an explanation for why certain methods over-perform or fail to meet expectations.
+
 Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in the critical transition scenarios (Hopf and saddle node).

### 2022-05-17 manuscript/manuscript.Rmd

@@
 Will these emerging methods improve the capacity for forecasts in ecological systems as well?
-Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Oveskienen2010].
+Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Ovaskainen2010].
 These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
@@
 Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
-Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
+Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
 Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
@@
 Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
-This is a much stronger assumption than that made by the ML models, though it can potentially be justfied on the basis of a mechanistic understanding of the processes involved.
+This is a much stronger assumption than that made by the ML models, though it can potentially be justified on the basis of a mechanistic understanding of the processes involved.
 It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
@@
 When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
-Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning subfield [@aitken_understanding_2021].
+Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning sub-field [@aitken_understanding_2021].
 These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
-Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often difficult to provide an explanation for why certain methods over-perform or fail to meet expectations.
+Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often impossible to provide an explanation for why certain methods over-perform or fail to meet expectations.
 
@@
 Furthermore, our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
-Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018].
+Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@dietze_iterative_2018].
 Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.

### 2022-05-18 manuscript/manuscript.Rmd
Adding an abstract

@@
 ---
-title: "Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning"
+title: "Limits to Ecological Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning"
 authors:
@@
 abstract: |
-  1. Meaningful representations of uncertainty are an essential component of effective ecological forecasting.
-  2. Machine learning
-  3. We examine scenarios in which stochasticity and non-linearity can lead to extreme uncertainty under common ecological models.  
-  4. We show that ..
+  1.
+In the current age of a rapidly changing environment, it is becoming increasingly important to study critical transitions and how to best anticipate them.
+Critical transitions pose extremely challenging forecasting problems, which necessitate informative uncertainty estimation rather than point forecasts.
+In this study, we apply some of the most cutting edge deep learning methods for probabilistic time series forecasting to several classic ecological models that examine critical transitions.
+  2.
+Our analysis focuses on three different simulated examples of critical transitions: a Hopf bifurcation, a saddle-node bifurcation and a stochastic transition.
+For each scenario, we compare the forecasts from four deep learning models, Long-short Term Memory networks, Gated Recurrent Unit networks, Block Recurrent neural networks and Transformers, to forecasts from an ARIMA model and a MCMC estimated model that is given the true transition dynamics.
+  3.
+We found that the deep learning models were able to perform comparably to the idealized MCMC model on the stochastic transition case, and generally in between the MCMC and ARIMA models on the Hopf and saddle-node bifurcation examples.
+  4.
+Our results establish that deep learning models are a promising class of forecasting methods for critical transition prediction problems, and that deep learning methods warrant further exploration on this challenging class of forecasting problems.
 

### 2022-05-18 manuscript/manuscript.Rmd
Ready for submission

@@
 We found that the deep learning models were able to perform comparably to the idealized MCMC model on the stochastic transition case, and generally in between the MCMC and ARIMA models on the Hopf and saddle-node bifurcation examples.
-  4.
-Our results establish that deep learning models are a promising class of forecasting methods for critical transition prediction problems, and that deep learning methods warrant further exploration on this challenging class of forecasting problems.
+  4. Our results establish that deep learning methods warrant further exploration on the challenging class of critical transition forecasting problems.
 
 keywords:
-  - Time Series
+  - Artificial Intelligence
   - Forecasting
   - Machine Learning
-  - Artificial Intelligence
+  - Time Series
   - Tipping points
@@
   - \usepackage{amssymb}
-#  - \usepackage{lineno}
-#  - \usepackage{endfloat}
-#  - \linenumbers
+  - \usepackage{lineno}
+  - \usepackage{setspace}
+  - \linenumbers
+  - \doublespacing
 
@@
 
-\pagebreak
+# Conflicts of Interest {-}
+
+The authors declare there are no conflicts of interest.
+
+# Authors' Contributions {-}
+
+M.L. and C.B. developed the code and wrote the manuscript.
+
+# Data Availability {-}
+
+We have created a github repository <https://github.com/boettiger-lab/mee_tipping_point_forecasting.> that contains the code used to produce the figures herein.
+
+\newpage
 

### 2022-05-18 manuscript/manuscript.Rmd
Now rendering figure captions

@@
 ---
-title: "Limits to Ecological Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning"
+title: "Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning"
 authors:
@@
 
-```{r }
-#| fig3
+```{r}
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
@@
 ```{r}
-#| fig6
 #| fig.cap="CRPS scores over time for each scenario.

### 2022-08-05 manuscript/manuscript.Rmd
Caption spacing fixes
NSF required acknowledgement
Smaller changes here

@@
 Ecological dynamics are notoriously complex, with uncertainty and non-linearity playing critical roles [@Boettiger2018; @hallett_why_2004; @Ovaskainen2010].
-These challenges are nowhere more evident then in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
+These challenges are nowhere more evident than in _critical transitions_, sudden shifts in the states or patterns of ecosystem dynamics that are more important and more difficult to predict than gradual changes.
 Here, we examine several of the best-known examples of critical transitions in ecological systems.
@@
 In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis.
-The Nicholson-Bailey model is a two species model which contains a Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.
+The Nicholson-Bailey model is a two species model which contains a super-critical Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.
 
@@
 
-```{r fig.cap="Forecast scenarios. A. a Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation."}
+```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation. C. The stochastic transition."}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
@@
 H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) - c P_t + \eta_{H,t} \right) \\
-P_{t+1} &= H_t \exp \left( r \left(1  \tfrac{H}{K_t}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right) \\
+P_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right) \\
 K_{t+1} &= K_t + \delta
@@
 
-Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a super-critical Hopf bifurcation as $K$ becomes sufficiently large.
+Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a super-critical Hopf bifurcation as $K$ becomes sufficiently large.
 In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
@@
 Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state.
-Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
+Unlike the supercritical Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
 Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
@@
 Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
-Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
+Tipping points resulting from saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].
 Examples use the following parameters: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = 0.15$, $\alpha = 0.000375$, $\sigma = 0.02$, $N_0 = 0.75$.
@@
 Note that it is possible to use any type of neural network as the encoder and the decoder, but it is most common to use recurrent neural networks or networks with attention mechanisms [@aitken_understanding_2021].
-Of the models that we present, Block RNNs and Transformers have encoder-decoder structures.
+Of the models that we present, Block RNNs and Transformers have encoder-decoder-based architectures.
 
@@
 To compare forecasts, we focus exclusively on metrics of forecast skill which satisfy the property from @gneiting_strictly_2007 of a strictly proper score.
-This ensures the very desirable behavior that which no probabilistic forecast $Q(x,t)$ can have a score as high as that of the true process $P(x,t)$ on average. 
+This ensures the very desirable behavior that no probabilistic forecast $Q(x,t)$ can have a score as high as the score of the true process $P(x,t)$ on average. 
 In other words, while it is possible for any of the models considered to _overfit_ the data against which they are trained, i.e.
@@
 It is worth noting that this property applies specifically to probabilistic forecasts and not point forecasts.
-Not all common metrics metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
+Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
 <!-- MFL: hmm I am not completely following this point here.
@@
 By taking the true model structure as given, MCMC methods can be used to determine a theoretical limit of forecasting skill.
-Note that in both bifurcation scenarios, future dynamics will see states never previously observed in the historical data that was used to train each of the methods (i.e.
+Note that in both bifurcation scenarios, future dynamics will visit states never previously observed in the historical data that was used to train each of the methods (e.g.
 very small population sizes).
@@
 Consequentially, the MCMC estimate easily outperforms the ML models (Fig 2).
-The distribution of scores over
 
@@
 #|  Each line represents the scores against a replicate of time series observations from 'true model'. 
-#| Patterns show forecast skill generally get worse over time - 
-#| gradually in the case of Hopf bifurcation or suddenly in repsonse to the saddle-node bifurcation. 
+#|  Patterns show forecast skill generally get worse over time -- 
+#| gradually in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
 #|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
@@
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-Although most ML-based forecasting applications focus on point predictions, we have emphasized examples which can provide a meaningful estimate of uncertainty.
-When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
+Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide a meaningful estimate of uncertainty.
+When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and saddle-node examples.
 An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
@@
 Unlike the MCMC case, the ML models have no prior expectation of a model structure that contains sharp transitions -- we might have expected these models to perform little differently than the ARIMA model.
-Given this single replicate, all four ML models successfully capture the phenomenological pattern of a sharp shift between two stable states -- behavior which the structurally simpler family of ARIMA models cannot express.
+Given this single replicate, all four ML models successfully capture the phenomenological pattern of a sharp shift between two stable states -- this is behavior that the structurally simpler family of ARIMA models cannot express.
 This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately modeled with ML models compared to classical statistical models.
@@
 
+# Acknowledgements
+
+This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
+
 # Conflicts of Interest {-}

### 2022-08-05 manuscript/manuscript.Rmd

--- 2022-05-18 manuscript/manuscript.Rmd
@@
 We evaluate the most promising machine learning methods for probabilistic forecasts relative to traditional statistical and mechanistic approaches applied to several classic models in ecology.
+
+“In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states at some time in the future.
+It is important to distinguish this objective from the extensive previous literature on “early warning signs” of critical transitions [eg as reviewed in @Scheffer2015] which has sought to answer only a categorical question: is the system approaching a critical transition?
+More recent work by @Bury2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
+These are important results with considerable success, but do not answer the questions of when a shift might occur or how large that transition will be.
+In light of increasing calls for the importance of such quantitative, probabilistic forecasts [@Deitz2018], we focus on evaluating the potential and limits to prediction in critical transition scenarios.”
 
@@
 Model definition is the same as May's model for the saddle node with fixed parameter $h$, values: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = h = 0.26$, $\alpha = 0$, $\sigma = 0.02$, $N_0 = 0.55$.
+
+**Selecting timescales**
+In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Figure 1.
+For the Hopf bifurcation, forecast begins at t=100 and extends to t =200, for the saddle node, forecast begins at t=250 and extends to t = 500, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t = 250.
+While much attention is often paid to the number of data points in training or testing data, it is essential to realize that these are only meaningful relative to the specific process in question.  
+Thus, in each case we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario.  
+For instance, if the stochastic scenario was restricted to a much shorter timescale used in the hopf case, few replicate simulations would experience a transition at all.  
+If it were made much longer, most of the timeseries would be spent post-transition. 
+Likewise, if the forecast for the Hopf scenario was extended much further into the future under the current parameterization, the system experiences a homoclinic bifurcation at which the population collapses to 0.
+Using different length timescales allows us to consider the three different forecasting tasks illustrated in Figure 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
+These three critical transitions are fundamentally different processes, there is no perfect apples-to-apples parameterization for each that allows the transition to unfold in a way that gives precisely the same time windows.
+
+time at which to start the forecast in each scenario relative to the timing of the bifurcation event (in the case of Hopf and Saddle Node), and relative to the expected distribution of transitions in the stochastic case.
+The rate at which the dynamical process unfolds in each case depends on the parameter values of the model in question.
+At much
 

### 2022-08-05 manuscript/manuscript.Rmd
Tidying new intro paragraph

@@
 
-“In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states at some time in the future.
-It is important to distinguish this objective from the extensive previous literature on “early warning signs” of critical transitions [eg as reviewed in @Scheffer2015] which has sought to answer only a categorical question: is the system approaching a critical transition?
-More recent work by @Bury2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
+In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states.
+It is important to distinguish this objective from the extensive previous literature on “early warning signs” of critical transitions, as reviewed in @scheffer_early-warning_2009, which has sought to answer only a categorical question: is the system approaching a critical transition?
+More recent work by @bury_2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
 These are important results with considerable success, but do not answer the questions of when a shift might occur or how large that transition will be.
-In light of increasing calls for the importance of such quantitative, probabilistic forecasts [@Deitz2018], we focus on evaluating the potential and limits to prediction in critical transition scenarios.”
+In light of increasing calls for the importance of such quantitative, probabilistic forecasts [@dietze_iterative_2018], we focus on evaluating the potential and limits to prediction in critical transition scenarios.
 

### 2022-08-08 manuscript/manuscript.Rmd
Added paragraph on general ml process

@@
 Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis_statistical_2018].
-Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will often be necessary to use multi-step, probablisitic forecasts. 
-For all the deep learning models in this study, we use the same method for uncertainty estimation:
-each model is trained to learn the parameters of a Poisson distribution to maximize the log likelihood. 
-To compute uncertainty intervals, we draw samples from the learned distribution at each time step.
+Deep learning models are often used to make point forecasts, but for their application to ecological time series, it will often be necessary to use multi-step, probabilistic forecasts. 
+For all the deep learning models in this study, we use the same general process.
+Each machine learning model is trained on one time series.
+For the Hopf and saddle node cases, these time series consist of the period leading up to the bifurcation.
+A critical transition is, however, included in the training set for the stochastic transition case.
+Each model is trained to learn the parameters of a Laplace distribution for every time step in the forecast horizon.
+To produce a forecast, we input a time series into a model, then we draw samples from the distributions that were learned during training.
 
@@
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide a meaningful estimate of uncertainty.
+Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide informative estimates of uncertainty.
 When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and saddle-node examples.
@@
 Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
-It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than less accurate on average but correctly reflect a greater degree of uncertainty.
-<!-- MFL: "that are less accurate"?-->
+It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than the scores of forecasts that are less accurate on average but correctly reflect a greater degree of uncertainty.
 The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.
@@
 This provides a clear illustration of the much broader array of phenomenological behaviors that can be accurately modeled with ML models compared to classical statistical models.
-In this way, the ML models can be seen to impose even fewer assumptions on the phenomenological behavior of the system than the ARIMA model.
+In this way, the ML models can be seen as imposing even fewer assumptions on the phenomenological behavior of the system than the ARIMA model.
 In contrast, the MCMC performance benefits from very strong process-based assumptions, which happen to match the 'true' model in this case and thus provide a comparison of the theoretical optimal performance.

### 2022-08-08 manuscript/manuscript.Rmd
Little edit on general ml paragraph
Some more touch ups

@@
 For all the deep learning models in this study, we use the same general process.
-Each machine learning model is trained on one time series.
+Each machine learning model is trained on one time series drawn from the three scenarios described previously.
 For the Hopf and saddle node cases, these time series consist of the period leading up to the bifurcation.
@@
 Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
-This is a much stronger assumption than that made by the ML models, though it can potentially be justified on the basis of a mechanistic understanding of the processes involved.
+This is a much stronger assumption than that made by the ML models, though this assumption can potentially be justified on the basis of a mechanistic understanding of the processes involved.
 It is important not to confound the MCMC example here with the use of MCMC in process based models of real systems.
 In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
+<!-- ML: maybe "In the real world, having access to the _true_ model is never the case:"-->
 Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
@@
 In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
-Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties.
-<!-- MFL: I think we could cut this paragraph out completely now -->
+Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data or observation error.
 

### 2022-08-09 manuscript/manuscript.Rmd
Little adjustments

@@
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
-This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
+This paper is not intended to give a complete answer to whether machine learning models are the the methods that are best suited for tipping point forecasting problems as this will take numerous studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
 Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.

### 2022-08-09 manuscript/manuscript.Rmd
Tiny grammar edit

@@
 In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
-Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data or observation error.
+Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data or observation errors.
 

### 2022-08-09 manuscript/manuscript.Rmd
Last commit before signing off

--- 2022-08-08 manuscript/manuscript.Rmd
@@
 In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
-Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data or observation error.
+Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.
 
@@
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
-This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
+This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
+<!-- I think here would perhaps be a good place to flesh out "promising" tag, stuck on how to do so-->
 Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.

### 2022-08-10 manuscript/manuscript.Rmd

@@
 We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
-Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide informative estimates of uncertainty.
+Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide estimates of uncertainty.
 When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and saddle-node examples.
@@
 Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
-It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than the scores of forecasts that are less accurate on average but correctly reflect a greater degree of uncertainty.
+It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than scores from forecasts that are less accurate on average but correctly reflect a greater degree of uncertainty.
 The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.
@@
 When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
-Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning sub-field [@aitken_understanding_2021].
+Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the field of sequence-to-sequence deep learning [@aitken_understanding_2021].
 These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
@@
 Would it be possible to train a predictive algorithm on 'analogous' examples of critical transitions?
-For instance, could data from other lakes which may have experienced a critical transition such as eutrophication event in the past be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
-Perhaps, but it depends what we mean by an 'analogous' system.
+For instance, could data from other lakes, which may have experienced a critical transition such as an eutrophication event in the past, be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
+Perhaps, but it depends on what we mean by an 'analogous' system.
 Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different, even after accounting for re-scaling or non-dimensionalization of the model [@Hastings1996].
@@
 There are a number of questions that we have left unanswered that we hope will be addressed in future work.
-In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
+In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for critical transition forecasting problems.
 Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.
@@
 Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
-This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
+This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
 <!-- I think here would perhaps be a good place to flesh out "promising" tag, stuck on how to do so-->
-Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.
+Since we have shown that ML models can make relatively good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are worthy of further investigation regarding their application to critical transition forecasting problems.
 

### 2022-08-11 manuscript/manuscript.Rmd
updated figures

@@
 
-```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation. C. The stochastic transition."}
+```{r fig.cap="Forecast scenarios.
+A.
+The Hopf bifurcation: a stable node develops into a limit cycle which gradually grows larger in this predator-prey model.
+B.
+The saddle-node bifurcation in a single species.
+C.
+The stochastic transition in a single species.
+Plots show historical data used to train the algorithm in purple, and ensemble replicate simulations of the the future dynamics in yellow.
+Note how the characteristic time for the critical transition varies across the transitions.", fig.height=6}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
-panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
+panelA <- hopf |> 
+  mutate(type = as.factor(type)) |>
+  filter(type %in% c("historical", "true"), i < 10) |>
   rename(species = variable) |> 
   mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
-  ggplot(aes(t, value, group=interaction(species, i, type), col=species)) + 
-  geom_line(show.legend = FALSE, alpha = 0.7) + theme_bw()+  scale_color_grey() +
-  ggtitle("A. Hopf bifurcation")
+  bind_rows(tibble(i=1,t=1,type="predicted", variable="H", value=NA)) |> #placeholder
+  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
+  ggplot(aes(t, value, group=interaction(species, i, type), col=type, lty=species)) + 
+  geom_line(show.legend = TRUE, alpha = 0.7) +
+  theme_bw() +  
+  scale_color_viridis_d(guide="none") +
+  ggtitle("A. Hopf bifurcation")+
+  theme(legend.position="top")
 
 sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
-panelB <- sn |> filter(type == "historical", i == 1) |>
-  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
+panelB <- sn |>
+  mutate(type = as.factor(type)) |>
+  filter(type == "historical", i == 1) |>
+  bind_rows( filter(sn, type=="true")) |>
+  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
+  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
   ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
   geom_path(alpha=0.5, show.legend = FALSE) +
-  scale_color_grey(end=0.6) + theme_bw() + 
+  scale_color_viridis_d() +
+  theme_bw() + 
   ggtitle("B. Saddle-node bifurcation")
@@
 stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
-panelC <- stoch |> filter(type == "historical", i == 1) |>
-  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
+panelC <- stoch |>
+  filter(type == "historical", i == 1) |>
+  bind_rows(filter(stoch, type=="true", i < 50)) |> 
+  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
+  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
   ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
-  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
-  ggtitle("C. Stochastic transition")
+  geom_path(alpha=0.5) + 
+  scale_color_viridis_d() +
+#  scale_color_grey(end=0.6) + 
+  theme_bw() + 
+  ggtitle("C. Stochastic transition") +
+  theme(legend.position="bottom")
 
@@
   stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
+  scale_x_discrete(limits = rev) +
   coord_flip() + 
@@
 forecasts <- read_parquet("data/forecasts.parquet") |>
-  filter(forecasting_model != "ml_ensemble")
-observations <- read_parquet("data/observations.parquet")
+  filter(forecasting_model != "ml_ensemble") |> 
+  bind_rows(tibble(type= c("future ensemble", "historical"))) |>
+  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))
+
+observations <- read_parquet("data/observations.parquet") |> 
+  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))
 
@@
   filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
+  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
+  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |>
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1) +
   geom_line(alpha=0.1, data = hopf_obs) +
-  geom_line(alpha=0.1) +
   geom_line(data = filter(hopf_obs, type=="historical")) +
@@
   theme_bw()  + 
+  scale_color_viridis_d() +
   coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
@@
   filter(iter %in% 1:15, simulation == "saddle") |> 
+  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
+  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1) +
   geom_line(alpha=0.1, data = saddle_obs) +
-  geom_line(alpha=0.1) +
   geom_line(data = filter(saddle_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
+  theme_bw()  + 
+  scale_color_viridis_d() +
+  coord_cartesian(ylim = c(0,1)) + 
   ggtitle("Saddle-Node bifurcation")
@@
   filter(iter %in% 1:15, simulation == "stochastic") |> 
+  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
+  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
-  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
   geom_line(alpha=0.2) +
+  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="future ensemble")) +
   geom_line(data = filter(stochastic_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
+  theme_bw()  + 
+  scale_color_viridis_d() +
+  coord_cartesian(ylim = c(0,0.8)) + 
   ggtitle("Stochastic transition")
@@
 by the MCMC predictions.
-The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
+The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
 after the catastrophic transition) are poor.
@@
 #|  Each line represents the scores against a replicate of time series observations from 'true model'. 
-#|  Patterns show forecast skill generally get worse over time -- 
-#| gradually in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
+#|  Patterns show forecast skill generally get worse over time -- gradually 
+#| in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
 #|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 

### 2022-08-11 manuscript/manuscript.Rmd
caption

@@
 Plots show historical data used to train the algorithm in purple, and ensemble replicate simulations of the the future dynamics in yellow.
-Note how the characteristic time for the critical transition varies across the transitions.", fig.height=6}
+Note how the characteristic time for the critical transition varies across the transitions.
+Compare to forecasts shown in figures 3-5 (forecasted timeseries always shown in blue).", fig.height=6}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")

### 2022-08-11 manuscript/manuscript.Rmd
tweak

@@
 The stochastic transition in a single species.
-Plots show historical data used to train the algorithm in purple, and ensemble replicate simulations of the the future dynamics in yellow.
+Plots show historical data used to train the algorithm in purple, and replicate simulations of the true dynamics ('future ensemble') in yellow.
 Note how the characteristic time for the critical transition varies across the transitions.
-Compare to forecasts shown in figures 3-5 (forecasted timeseries always shown in blue).", fig.height=6}
+We will examine various models which will each produce probablistic forecast distributions (blue, Fig 3-5) seeking to match the true future ensemble (yellow) as closely as possible.", fig.height=6}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")

### 2022-08-11 manuscript/manuscript.Rmd
png lower res fig 2

@@
 
-```{r fig.cap="Forecast scenarios.
+```{r figure1, fig.cap="Forecast scenarios.
 A.
@@
 
-```{r}
+```{r figure2, dev="png", dpi=300, dev.args = list(type = "cairo-png"), out.width="400px", out.height="300px"}
 #| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
@@
 
-```{r}
+```{r figure3}
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
@@
 
-```{r }
+```{r figure4}
 #| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
@@
 
-```{r}
+```{r figure5}
 #| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
@@
 
-```{r}
+```{r figure6}
 #| fig.cap="CRPS scores over time for each scenario.
@@
 
-```{r}
+```{r figure7}
 #| fig.cap="Logs skill score over time. 

### 2022-08-11 manuscript/manuscript.Rmd
super-critical -> supercritical

@@
 In this one-dimensional model, transitions between these states can occur due to intrinsic stochasticity, external forcing, or the gradual environmental change that results in a catastrophic saddle-node bifurcation and generates hysteresis.
-The Nicholson-Bailey model is a two species model which contains a super-critical Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.
+The Nicholson-Bailey model is a two species model which contains a supercritical Hopf bifurcation, a non-catastrophic bifurcation which either creates or destroys a limit cycle -- a stable oscillatory pattern.
 
@@
 
-Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a super-critical Hopf bifurcation as $K$ becomes sufficiently large.
+Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a supercritical Hopf bifurcation as $K$ becomes sufficiently large.
 In a Hopf bifurcation, a stable node starts an oscillatory pattern which grows in amplitude as the bifurcation parameter continues to increase.
-In this model, the Hopf bifurcation is dubbed 'super-critical' as it creates a stable limit cycle instead of an unstable one.
+In this model, the Hopf bifurcation is dubbed 'supercritical' as it creates a stable limit cycle instead of an unstable one.
 This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
@@
 
-It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of super-critical Hopf or saddle node bifurcations. 
+It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of supercritical Hopf or saddle node bifurcations. 
 This question is not as easy to answer as it may seem, because of the difficulty in defining the corresponding forecasting scenario.

### 2022-08-11 manuscript/manuscript.Rmd
define H_t, P_t
textual tweaks

@@
 
-In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states.
+In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states, as frequently called for in ecological research [@clark2001; @dietze_iterative_2018].
+Such forecasting tasks may arise whenever a manager is interested in knowing the future states of a system: such as setting future catch quotas for a fishery or adjusting eradication effort for an invasive species.
 It is important to distinguish this objective from the extensive previous literature on “early warning signs” of critical transitions, as reviewed in @scheffer_early-warning_2009, which has sought to answer only a categorical question: is the system approaching a critical transition?
-More recent work by @bury_2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
-These are important results with considerable success, but do not answer the questions of when a shift might occur or how large that transition will be.
-In light of increasing calls for the importance of such quantitative, probabilistic forecasts [@dietze_iterative_2018], we focus on evaluating the potential and limits to prediction in critical transition scenarios.
+Recent work such as @bury_2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
+These are important results that with considerable promise [@lapeyrolerie2021], but which nevertheless address a very different question using very different methods. 
+Early warning signals only predict "a big change may be coming soon" -- they do not try to forecast when or how big.
+As we shall see, there is good reason to focus on that more modest, qualitative objective when faced with systems that might produce critical transitions.
+Here, we examine the more ambitious questions of forecasting when and how much change: or more precisely, of making probabilistic forecasts of all future states over a given time horizon.
 
@@
 Note how the characteristic time for the critical transition varies across the transitions.
-We will examine various models which will each produce probablistic forecast distributions (blue, Fig 3-5) seeking to match the true future ensemble (yellow) as closely as possible.", fig.height=6}
+We will examine forecasts of various models (Fig 2) which will each produce probablistic forecast distributions (blue, Fig 3-5) seeking to match the true future ensemble (yellow) as closely as possible.", fig.height=6}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
@@
 
+Where $H_t$ is the population density of the host species at time $t$, (in arbitrary units) and $P_t$ the population density of the parasitoid.
+The time step is defined by the generation time of the parasitoid, which is about two weeks in the case of Nicholson's blowflies [@nicholson_outline_1954].
 Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at a linear rate, which drives a supercritical Hopf bifurcation as $K$ becomes sufficiently large.
@@
 we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
-Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
+Based on this data, our forecaster seeks to identify the distribution of expected transition times for analogous systems starting from the same initial condition. 
 This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
@@
 **Selecting timescales**
-In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Figure 1. 
+In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Fig 1. 
 For the Hopf bifurcation, forecast begins at t=100 and extends to t =200, for the saddle node, forecast begins at t=250 and extends to t = 500, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t = 250.
@@
 Thus, in each case we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario. 
-For instance, if the stochastic scenario was restricted to a much shorter timescale used in the hopf case, few replicate simulations would experience a transition at all. 
+For instance, if the stochastic scenario was restricted to a much shorter timescale used in the Hopf case, few replicate simulations would experience a transition at all. 
 If it were made much longer, most of the timeseries would be spent post-transition. 
 Likewise, if the forecast for the Hopf scenario was extended much further into the future under the current parameterization, the system experiences a homoclinic bifurcation at which the population collapses to 0.
-Using different length timescales allows us to consider the three different forecasting tasks illustrated in Figure 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
+Using different length timescales allows us to consider the three different forecasting tasks illustrated in Fig 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
 These three critical transitions are fundamentally different processes, there is no perfect apples-to-apples parameterization for each that allows the transition to unfold in a way that gives precisely the same time windows.
-
-time at which to start the forecast in each scenario relative to the timing of the bifurcation event (in the case of Hopf and Saddle Node), and relative to the expected distribution of transitions in the stochastic case.
-The rate at which the dynamical process unfolds in each case depends on the parameter values of the model in question.
-At much
 
@@
 
-Overall forecasting skill scores for each model across all three scenarios are summarized in figure 2.
+Overall forecasting skill scores for each model across all three scenarios are summarized in Fig 2.
 Average scores (black lines) hide wide variation in forecast skill.
@@
 This reflects the relative contributions of accuracy and uncertainty as components in the forecasts.
-Overall, ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
+Most ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
+Th get a deeper understanding of these general patterns, we now turn to examine each of the forecast distributions themselves in comparison to the future ensemble produced by the true generative process model, Fig 3-5.
 
@@
 
-```{r figure3}
+```{r figure3, fig.height=7}
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
@@
 
-Forecasts of the Hopf bifurcation (Figure 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
+Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
 All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
@@
 
-The saddle node bifurcation proves even more difficult for most methods (Figure 4).
+The saddle node bifurcation proves even more difficult for most methods (Fig 4).
 Only the MCMC model anticipates the sharp transition to an alternative state, though this behavior is more baked into the method from the start which takes the true model structure as given. 
@@
 by the MCMC predictions.
-The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
+The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
 after the catastrophic transition) are poor.
@@
 #|  Forecasts which underestimate uncertainty do substantially worse in logs score than in CRPS score. 
-#|  Comparing this panel to those in Figure 6 highlights scenarios that most often underestimate uncertainty. 
+#|  Comparing this panel to those in Fig 6 highlights scenarios that most often underestimate uncertainty. 
 #|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
@@
 This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
-<!-- I think here would perhaps be a good place to flesh out "promising" tag, stuck on how to do so-->
-Since we have shown that ML models can make relatively good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are worthy of further investigation regarding their application to critical transition forecasting problems.
+Given the difficulty of forecasting never-before-observed behavior, as illustrated by the Hopf and saddle-node bifurcation scenarios, there is good reason for research to focus more on the kind of qualitative predictions long emphasized in the literature on early warning signals and resilience [@scheffer2015].
+Recently, ML techniques developed for classification rather than the ML methods used in regression and forecasting models considered here have demonstrated a more nuanced ability to reliably detect different classes of critical transitions in time-series data [@bury2021; @lapeyrolerie2021].
+Rather than seeking to provide managers with quantitative, probabilistic forecasts reflecting a broad uncertainty in possible outcomes, this literature has sought to emphasize only a more qualitative form of prediction, such as establishing whether a system is either "resilient" or "approaching a critical transition."
+Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@folke2009; @polasky2011].
+Such approaches could be valuable in concert with probabilistic forecasts considered here, providing a possible mechanism to identify when the probabilistic forecast is least reliable.
 

### 2022-08-14 manuscript/manuscript.Rmd
Added explicit definitions of scores

--- 2022-08-11 manuscript/manuscript.Rmd
@@
   - \usepackage{setspace}
+  - \usepackage{bbm}
   - \linenumbers
@@
 
-```{r figure1, fig.cap="Forecast scenarios.
-A.
-The Hopf bifurcation: a stable node develops into a limit cycle which gradually grows larger in this predator-prey model.
-B.
-The saddle-node bifurcation in a single species.
-C.
-The stochastic transition in a single species.
-Plots show historical data used to train the algorithm in purple, and replicate simulations of the true dynamics ('future ensemble') in yellow.
-Note how the characteristic time for the critical transition varies across the transitions.
-We will examine various models which will each produce probablistic forecast distributions (blue, Fig 3-5) seeking to match the true future ensemble (yellow) as closely as possible.", fig.height=6}
+```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation. C. The stochastic transition."}
 hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
-panelA <- hopf |> 
-  mutate(type = as.factor(type)) |>
-  filter(type %in% c("historical", "true"), i < 10) |>
+panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
   rename(species = variable) |> 
   mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
-  bind_rows(tibble(i=1,t=1,type="predicted", variable="H", value=NA)) |> #placeholder
-  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
-  ggplot(aes(t, value, group=interaction(species, i, type), col=type, lty=species)) + 
-  geom_line(show.legend = TRUE, alpha = 0.7) +
-  theme_bw() +  
-  scale_color_viridis_d(guide="none") +
-  ggtitle("A. Hopf bifurcation")+
-  theme(legend.position="top")
+  ggplot(aes(t, value, group=interaction(species, i, type), col=species)) + 
+  geom_line(show.legend = FALSE, alpha = 0.7) + theme_bw()+  scale_color_grey() +
+  ggtitle("A. Hopf bifurcation")
 
 sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
-panelB <- sn |>
-  mutate(type = as.factor(type)) |>
-  filter(type == "historical", i == 1) |>
-  bind_rows( filter(sn, type=="true")) |>
-  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
-  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
+panelB <- sn |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
   ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
   geom_path(alpha=0.5, show.legend = FALSE) +
-  scale_color_viridis_d() +
-  theme_bw() + 
+  scale_color_grey(end=0.6) + theme_bw() + 
   ggtitle("B. Saddle-node bifurcation")
@@
 stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
-panelC <- stoch |>
-  filter(type == "historical", i == 1) |>
-  bind_rows(filter(stoch, type=="true", i < 50)) |> 
-  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
-  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
+panelC <- stoch |> filter(type == "historical", i == 1) |>
+  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
   ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
-  geom_path(alpha=0.5) + 
-  scale_color_viridis_d() +
-#  scale_color_grey(end=0.6) + 
-  theme_bw() + 
-  ggtitle("C. Stochastic transition") +
-  theme(legend.position="bottom")
+  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
+  ggtitle("C. Stochastic transition")
 
@@
 Not all common metrics often used to compare forecasts are strictly proper -- such as the average root-mean-square error or the average absolute error.
-<!-- MFL: hmm I am not completely following this point here.
-CB: added some examples of non-strictly-proper scoring metrics.
-For instance, {darts} has a whole 'metrics' class, which does not include any strictly proper metrics (because ML researchers don't care about statistical theorems?) -->
 Concerns about over-fitting arise in most types of model estimation and are a particularly acute concern to machine learning methods due to the bias-variance trade-off [@mehta_high-bias_2019].
-<!-- MFL: I don't think I nailed the overfitting part in ML section but we probably want to reference this here-->
 This makes the use of strictly proper scoring especially relevant in assessing machine learning predictions.
@@
 We will focus on two of the most common such skill metrics, CRPS score and log probability score (negative log likelihood) [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
+We have defined these two scores explicitly in Equations 7-8, where $F$ and $f$ are the respective CDF's and PDF's from a probabilistic forecast, and $y$ refers to an observation.
 Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
 Note that while the minus log-likelihood can be negative for sufficiently high probability densities, we use a fixed scalar shift of logs score to ensure the log skill score is strictly positive, which facilitates visualization without impacting relative rankings.
+
+\begin{align} 
+\text{LogS}(F, y) &= -\log{f(y)} \\
+\text{CRPS}(F, y) &= -\int (F(z) - \mathbbm{1}\{y \leq z\})^2 \, dz
+\end{align}
 
@@
 
-```{r figure2, dev="png", dpi=300, dev.args = list(type = "cairo-png"), out.width="400px", out.height="300px"}
+```{r}
 #| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
@@
   stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
-  scale_x_discrete(limits = rev) +
   coord_flip() +
@@
 forecasts <- read_parquet("data/forecasts.parquet") |>
-  filter(forecasting_model != "ml_ensemble") |> 
-  bind_rows(tibble(type= c("future ensemble", "historical"))) |>
-  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))
-
-observations <- read_parquet("data/observations.parquet") |> 
-  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))
+  filter(forecasting_model != "ml_ensemble")
+observations <- read_parquet("data/observations.parquet")
 
@@
 
-```{r figure3}
+```{r}
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
@@
   filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
-  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
-  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |>
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = hopf_obs) +
   geom_line(alpha=0.1) +
-  geom_line(alpha=0.1, data = hopf_obs) +
   geom_line(data = filter(hopf_obs, type=="historical")) +
@@
   theme_bw()  + 
-  scale_color_viridis_d() +
   coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
@@
 
-```{r figure4}
+```{r }
 #| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
@@
   filter(iter %in% 1:15, simulation == "saddle") |> 
-  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
-  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.1, data = saddle_obs) +
   geom_line(alpha=0.1) +
-  geom_line(alpha=0.1, data = saddle_obs) +
   geom_line(data = filter(saddle_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + 
-  scale_color_viridis_d() +
-  coord_cartesian(ylim = c(0,1)) + 
+  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
   ggtitle("Saddle-Node bifurcation")
@@
 
-```{r figure5}
+```{r}
 #| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
@@
   filter(iter %in% 1:15, simulation == "stochastic") |> 
-  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
-  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
   ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
+  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
   geom_line(alpha=0.2) +
-  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="future ensemble")) +
   geom_line(data = filter(stochastic_obs, type=="historical")) +
   facet_wrap(~forecasting_model, scales = "free") +
-  theme_bw()  + 
-  scale_color_viridis_d() +
-  coord_cartesian(ylim = c(0,0.8)) + 
+  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
   ggtitle("Stochastic transition")
@@
 by the MCMC predictions.
-The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
+The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
 after the catastrophic transition) are poor.
@@
 
-```{r figure6}
+```{r}
 #| fig.cap="CRPS scores over time for each scenario.
 #|  Each line represents the scores against a replicate of time series observations from 'true model'. 
-#|  Patterns show forecast skill generally get worse over time -- gradually 
-#| in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
+#|  Patterns show forecast skill generally get worse over time -- 
+#| gradually in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
 #|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
@@
 
-```{r figure7}
+```{r}
 #| fig.cap="Logs skill score over time. 

### 2022-08-14 manuscript/manuscript.Rmd

--- 2022-08-11 manuscript/manuscript.Rmd
@@
 Rather than seeking to provide managers with quantitative, probabilistic forecasts reflecting a broad uncertainty in possible outcomes, this literature has sought to emphasize only a more qualitative form of prediction, such as establishing whether a system is either "resilient" or "approaching a critical transition."
-Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@folke2009; @polasky2011].
+Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@fischer2009; @polasky2011].
 Such approaches could be valuable in concert with probabilistic forecasts considered here, providing a possible mechanism to identify when the probabilistic forecast is least reliable.

### 2022-08-14 manuscript/manuscript.tex

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Limits to Ecological Forecasting: Estimating Uncertainty for
Critical Transitions with Deep Learning}

\author{
    Marcus Lapeyrolerie
   \\
    Department of Environmental Science, Policy, and Management \\
    University of California, Berkeley \\
  Berkeley, California \\
  \texttt{\href{mailto:mlapeyro@berkeley.edu}{\nolinkurl{mlapeyro@berkeley.edu}}} \\
   \And
    Carl Boettiger
   \\
    Department of Environmental Science, Policy, and Management \\
    University of California, Berkeley \\
  Berkeley, California \\
  \texttt{\href{mailto:cboettig@berkeley.edu}{\nolinkurl{cboettig@berkeley.edu}}
(corresponding author)} \\
  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{setspace}
\linenumbers
\doublespacing
\begin{document}
\maketitle

\begin{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In the current age of a rapidly changing environment, it is becoming
  increasingly important to study critical transitions and how to best
  anticipate them. Critical transitions pose extremely challenging
  forecasting problems, which necessitate informative uncertainty
  estimation rather than point forecasts. In this study, we apply some
  of the most cutting edge deep learning methods for probabilistic time
  series forecasting to several classic ecological models that examine
  critical transitions.
\item
  Our analysis focuses on three different simulated examples of critical
  transitions: a Hopf bifurcation, a saddle-node bifurcation and a
  stochastic transition. For each scenario, we compare the forecasts
  from four deep learning models, Long-short Term Memory networks, Gated
  Recurrent Unit networks, Block Recurrent neural networks and
  Transformers, to forecasts from an ARIMA model and a MCMC estimated
  model that is given the true transition dynamics.
\item
  We found that the deep learning models were able to perform comparably
  to the idealized MCMC model on the stochastic transition case, and
  generally in between the MCMC and ARIMA models on the Hopf and
  saddle-node bifurcation examples.
\item
  Our results establish that deep learning methods warrant further
  exploration on the challenging class of critical transition
  forecasting problems.
\end{enumerate}
\end{abstract}

\keywords{
    Artificial Intelligence
   \and
    Forecasting
   \and
    Machine Learning
   \and
    Time Series
   \and
    Tipping points
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Forecasting plays an important and rapidly growing role in both testing
our fundamental understanding of ecological processes, and informing
ecological applications and conservation decision-making (Dietze et al.
2018; Schindler, Armstrong, and Reed 2015). Meanwhile, recent advances
in machine learning have rapidly improved the prevalence and accuracy of
short term forecasts in many fields (Kao et al. 2020; Lyu et al. 2020;
Du et al. 2020). Will these emerging methods improve the capacity for
forecasts in ecological systems as well? Ecological dynamics are
notoriously complex, with uncertainty and non-linearity playing critical
roles (Boettiger 2018a; Hallett et al. 2004; Ovaskainen and Meerson
2010). These challenges are nowhere more evident than in \emph{critical
transitions}, sudden shifts in the states or patterns of ecosystem
dynamics that are more important and more difficult to predict than
gradual changes. Here, we examine several of the best-known examples of
critical transitions in ecological systems. We evaluate the most
promising machine learning methods for probabilistic forecasts relative
to traditional statistical and mechanistic approaches applied to several
classic models in ecology.

In this paper, we focus on the task of producing quantitative,
probabilistic forecasts reflecting the possible distribution of future
states, as frequently called for in ecological research
(\textbf{clark2001?}; Dietze et al. 2018). Such forecasting tasks may
arise whenever a manager is interested in knowing the future states of a
system: such as setting future catch quotas for a fishery or adjusting
eradication effort for an invasive species. It is important to
distinguish this objective from the extensive previous literature on
``early warning signs'' of critical transitions, as reviewed in Scheffer
et al. (2009), which has sought to answer only a categorical question:
is the system approaching a critical transition? Recent work such as
Bury et al. (2021) has introduced ML methods to consider classification
of this transition in four possible categories (Hopf, saddle-node,
transcritical, or no bifurcation) rather than two (bifurcation or not).
These are important results that with considerable promise
(\textbf{lapeyrolerie2021?}), but which nevertheless address a very
different question using very different methods. Early warning signals
only predict ``a big change may be coming soon'' -- they do not try to
forecast when or how big. As we shall see, there is good reason to focus
on that more modest, qualitative objective when faced with systems that
might produce critical transitions. Here, we examine the more ambitious
questions of forecasting when and how much change: or more precisely, of
making probabilistic forecasts of all future states over a given time
horizon.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

We will focus the analysis on several different forecasting scenarios
based around two classic models in population ecology: Robert May's
consumer-resource model (Robert M. May 1977), and the Nicholson-Bailey
parasitoid-host model (A. J. Nicholson and Bailey 1935). Though these
models may appear simple when measured against high-dimensional and
parameter rich models found in some management contexts such as
fisheries, they can exhibit rich nonlinear dynamics and provide greater
capacity to generalize (Levins 1966; Getz et al. 2018). These textbook
models have been well studied and form the basis of half a century of
research in ecology, including much recent work on topics such as
resilience and tipping points which has had important theoretical and
practical management outcomes (Folke et al. 2004; Fischer et al. 2009;
Polasky et al. 2011). May's model exhibits alternative stable states. In
this one-dimensional model, transitions between these states can occur
due to intrinsic stochasticity, external forcing, or the gradual
environmental change that results in a catastrophic saddle-node
bifurcation and generates hysteresis. The Nicholson-Bailey model is a
two species model which contains a supercritical Hopf bifurcation, a
non-catastrophic bifurcation which either creates or destroys a limit
cycle -- a stable oscillatory pattern.

Assessing the accuracy of forecasting methods in the face of such
bifurcation dynamics is a particularly important question for ecological
systems and global environmental change problems. Bifurcations represent
the kind of non-linear responses complex systems can make as the result
of slowly changing parameters. This can create a particularly
challenging forecasting task when such transitions have not been
previously observed in the same system, requiring the forecast to
anticipate dynamics for which there are no corresponding analog in the
historical data. Forecast skill under such no-analog conditions may be
particularly relevant to ecological forecasting in the context of global
change (Williams and Jackson 2007).

We provide fully reproducible coded examples in R and Python for
performing, scoring, and visualizing each of the forecasts considered
here. After significant time spent considering alternative frameworks,
we have emphasized those which best met our requirements for
performance, ease-of-use, flexibility, and support for the latest
probabilistic machine learning models for forecasting. Most of our
forecasts use the \texttt{darts} framework, a sophisticated and well
documented Python library with support for a wide range of methods. Our
model-based MCMC forecasts use the \texttt{greta} framework, a R library
that uses Python-based \texttt{tensorflow} probability to achieve better
performance. While Python-based frameworks currently have the edge in
performance and access modern ML algorithms, they lag behind in
attention to statistical issues such as the computation of strictly
proper skill scores.

Our examples of scoring and visualization will rely on a collection of R
packages, in particular, \texttt{scoringRules} for the efficient
calculation of Continuous Ranked Probability Score (CRPS) and
logarithmic probability (Logs) scores for forecast ensembles (Gneiting
and Raftery 2007). Following popular conventions, we express both skill
scores in error-orientation, that is, larger values indicate worse skill
(higher degree of error).

We expect greater convergence between methods available in R and Python
in the future, as already illustrated in the example of \texttt{greta}.
Complete code for all examples presented here can be found at
\url{https://github.com/boettiger-lab/mee_tipping_point_forecasting}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure1-1.pdf}
\caption{Forecast scenarios. A. The Hopf bifurcation: a stable node
develops into a limit cycle which gradually grows larger in this
predator-prey model. B. The saddle-node bifurcation in a single species.
C. The stochastic transition in a single species. Plots show historical
data used to train the algorithm in purple, and replicate simulations of
the true dynamics (`future ensemble') in yellow. Note how the
characteristic time for the critical transition varies across the
transitions. We will examine forecasts of various models (Fig 2) which
will each produce probablistic forecast distributions (blue, Fig 3-5)
seeking to match the true future ensemble (yellow) as closely as
possible.}
\end{figure}

\hypertarget{scenario-1-hopf-bifurcation}{%
\subsection{Scenario 1: Hopf
Bifurcation}\label{scenario-1-hopf-bifurcation}}

The Nicholson-Bailey model describes a predator-prey dynamic for the
relationship of a host species and an obligate parasitoid, originally
used to model the population dynamics of blowflies (\emph{Lucilia
cuprina}) (A. J. Nicholson and Bailey 1935; A. Nicholson 1954a, 1954b).
We consider the form which includes density dependence in the host
species, and allow for environmental stochasticity,

\begin{align}
H_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) - c P_t + \eta_{H,t} \right) \\
P_{t+1} &= H_t \exp \left( r \left(1 - \tfrac{H}{K_t}\right) \right)  \left(1 -  \exp \left(- c P_t + \eta_{P,t}\right) \right) \\
K_{t+1} &= K_t + \delta
\label{hopf}
\end{align}

Where \(H_t\) is the population density of the host species at time
\(t\), (in arbitrary units) and \(P_t\) the population density of the
parasitoid. The time step is defined by the generation time of the
parasitoid, which is about two weeks in the case of Nicholson's
blowflies (A. Nicholson 1954a). Following Dakos et al. (2012), we
further allow the carrying capacity of the host, \(K\) to slowly
increase at a linear rate, which drives a supercritical Hopf bifurcation
as \(K\) becomes sufficiently large. In a Hopf bifurcation, a stable
node starts an oscillatory pattern which grows in amplitude as the
bifurcation parameter continues to increase. In this model, the Hopf
bifurcation is dubbed `supercritical' as it creates a stable limit cycle
instead of an unstable one. This example illustrates one of the many
kinds of challenges which nonlinear phenomena pose to forecasting: the
``historical'' data prior to the bifurcation never exhibit the cyclical
dynamics of growing amplitude that will emerge after the bifurcation
occurs. If we had used a purely deterministic model, the dynamics would
be constrained to a single stable point, corresponding to a slowly
changing steady-state population size of host and parasitoid
populations. However, stochasticity in this case acts as a source of
some additional information about the dynamics, as the noise excites
quasi-cycles which are visible in the irregular oscillations that appear
significantly prior to the emergence of true limit cycles which follow
the bifurcation (Boettiger 2018b). Examples use the following
parameters: \(H_0 = 9\), \(P_0 = 1\), \(r = 0.75\), \(c = 0.1\),
\(K_0 = 14\), \(\delta = 0.08\), \(\sigma_H = 0.02\),
\(\sigma_P = 0.02\).

\hypertarget{scenario-2-the-saddle-node-bifurcation}{%
\subsection{Scenario 2: The saddle-node
bifurcation}\label{scenario-2-the-saddle-node-bifurcation}}

A yet more difficult forecasting scenario is created by the saddle-node
bifurcation. May's consumer-resource model is an one-dimensional model
describing the growth of a `resource' population (e.g.~herbivore) which
is grazed by a consumer (Robert M. May and Anderson 1979). As in the
Nicholson-Bailey model, in the absence of that predation, the resource
population density grows under a density-dependent pattern described by
a logistic function. The resource population is also grazed by a
consumer at a rate given by a Holling type III s-curve (typically used
to model handling time). For a certain range of parameter choices, this
model supports alternative stable state dynamics, and has been
identified and employed in explaining alternative stable state dynamics
in a broad range of ecological and socio-ecological systems (Scheffer et
al. 2001b).

\begin{align}
N_{t+1} &= N_t + r N_t \left(1 - \frac{N_t}{K} \right) - \frac{h_t N_t^2}{s^2 + N_t^2} + \eta_t \\
h_{t+1} &= h_t + \alpha \\
\eta_t &\sim \mathcal{N}(0, \sigma)
\end{align}

If the environment slowly alters one of the parameters (say, the
encounter efficiency, \texttt{h\_t}, in our formulation), one of the
stable nodes moves closer and closer to the unstable saddle point,
leading to a bifurcation that destroys the stable state, leaving the
system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a
phenomenon known as hysteresis, where it is not sufficient to restore
the environment to the previous parameter values to recover the previous
state. Unlike the supercritical Hopf bifurcation which exhibits a
continuous transition from a stable node to a small limit cycle that
then grows, the saddle-node transition is a discontinuous or so-called
`catastrophic' bifurcation. Due both to this sudden, catastrophic nature
of the transition and the difficulty in reversing the shift after it has
occurred, saddle-node bifurcations have been the subject of intense
study.

Tipping point dynamics have long been identified as an important but
difficult challenge for forecasting (e.g. Scheffer et al. 2001a; Folke
et al. 2004). Much effort in the ecological literature so far has
focused on identifying any `early warning signs' that a catastrophic
bifurcation might occur at all (Scheffer et al. 2009, 2012) rather than
more ambitious attempts to provide quantitative probabilistic forecasts
of the likely distribution of waiting times before such a transition
occurs. Tipping points resulting from saddle-node bifurcations have been
demonstrated in examples ranging from laboratory microcosms (Dai et al.
2012; Dai, Korolev, and Gore 2015) to whole-ecosystem experiments
(Carpenter et al. 2011), and postulated as a model for global change
(Barnosky et al. 2012). Examples use the following parameters:
\(r = 1\), \(K = 1\), \(s = 0.1\), \(h_0 = 0.15\),
\(\alpha = 0.000375\), \(\sigma = 0.02\), \(N_0 = 0.75\).

\hypertarget{scenario-3-the-stochastic-transition}{%
\subsection{Scenario 3: The stochastic
transition}\label{scenario-3-the-stochastic-transition}}

Perhaps the most difficult of all events to predict are those in which
large transitions are predominately driven by a random component. An
example of such a transition event is possible to observe in May's
consumer-resource model, in which a stochastic term occasionally results
in a transition between alternative stable states. In such cases, no
forecast can precisely predict when a transition will occur, but it is
nonetheless possible to deduce the correct distribution of waiting times
knowing the correct model. In the case of small noise, transitions are
Poisson distributed, such that the distribution of waiting times is
roughly exponential (e.g. Kampen 1992), though post-hoc the trajectories
of such transitions can be mistaken for saddle-node transitions
(Boettiger and Hastings 2012). To consider such cases, we will again use
May's alternative stable state model, though this time leaving all
parameters fixed.

In this context, predicting the probability of a transition in the
future based solely on observations prior to a transition occurring is
essentially impossible without additional information constraining the
model estimate, as such data is equally consistent with infinitely many
models or parameter choices which share the same local linearization
about the stable point. Unlike the saddle-node bifurcation, there is no
slowly warping potential basin which can be detected to inform
estimates. Thus, in this scenario, rather than considering the problem
of predicting the future evolution of a single time series based only on
its historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more
comparable systems which includes a previous stochastic transition
event. Based on this data, our forecaster seeks to identify the
distribution of expected transition times for analogous systems starting
from the same initial condition. This parallels actual practice in which
researchers would draw on previous examples of stochastic transitions in
a system - lake-ecosystem shifts, disease emergence, changing fire
regimes, (Scheffer et al. 2001a; Folke et al. 2004). (Note that such
stochastic transitions between alternative stable states can also create
oscillatory-like dynamics when stochasticity is sufficiently high enough
to drive repeated transitions from one attractor to the other and back
again. In such cases, it might be reasonable to estimate a strictly
forward-looking forecast of a single system, predicting the distribution
of these transitions.) Model definition is the same as May's model for
the saddle node with fixed parameter \(h\), values: \(r = 1\),
\(K = 1\), \(s = 0.1\), \(h_0 = h = 0.26\), \(\alpha = 0\),
\(\sigma = 0.02\), \(N_0 = 0.55\).

\textbf{Selecting timescales} In each scenario, t = 0 is the start time
of the training data, while the length of training data and forecast
horizon (with ensembles sampled from the true distribution) is
illustrated in Fig 1. For the Hopf bifurcation, forecast begins at t=100
and extends to t =200, for the saddle node, forecast begins at t=250 and
extends to t = 500, for the stochastic transition, both training data
and forecasting tasks begin at 0 and extend to t = 250. While much
attention is often paid to the number of data points in training or
testing data, it is essential to realize that these are only meaningful
relative to the specific process in question. Thus, in each case we have
selected these time intervals to focus on the dynamical process in
question, which unfolds at a different rate and tempo in each scenario.
For instance, if the stochastic scenario was restricted to a much
shorter timescale used in the Hopf case, few replicate simulations would
experience a transition at all. If it were made much longer, most of the
timeseries would be spent post-transition. Likewise, if the forecast for
the Hopf scenario was extended much further into the future under the
current parameterization, the system experiences a homoclinic
bifurcation at which the population collapses to 0. Using different
length timescales allows us to consider the three different forecasting
tasks illustrated in Fig 1 that focus around predicting the critical
behavior, rather than predicting long periods of relative stasis. These
three critical transitions are fundamentally different processes, there
is no perfect apples-to-apples parameterization for each that allows the
transition to unfold in a way that gives precisely the same time
windows.

\hypertarget{method-group-1-markov-chain-monte-carlo}{%
\subsection{Method group 1: Markov Chain Monte
Carlo}\label{method-group-1-markov-chain-monte-carlo}}

As a reference case, we consider forecasts produced by MCMC estimates of
model parameters, \emph{given the true model}. This represents an
idealized case where the nature of the underlying process is known
precisely. Uncertainty comes from parameter estimates and intrinsic
stochasticity specified in the model, but does not reflect any
uncertainty in our knowledge of the model structure. Alternative model
structures, even when capable of producing the same nonlinear phenomena
(i.e.~the same bifurcations) will give very different forecasts. Even
alternative prior distributions of the parameters will generally yield
alternate forecasts, as likelihood ridges are common to nonlinear
models. Thus, this case represents a theoretical upper bound for the
performance of forecasts by techniques which do not make such strong
assumptions about the underlying processes.

\hypertarget{method-group-2-statistical-models-arima}{%
\subsection{Method group 2: Statistical models
(ARIMA)}\label{method-group-2-statistical-models-arima}}

We present forecasts produced by ARIMA models as the model-free analogs
to the forecasts made using parameter estimation with MCMC. Since ARIMA
models make the assumption that the future will resemble the past via
ARIMA's auto-regressive and moving average components (Hyndman and
Athanasopoulos 2018), these models are not well-suited for problems with
complex bifurcation dynamics. Thus, ARIMA-based forecasts should be
treated as a lower bound for the performance of non-mechanistic models.
In contrast to inference with MCMC, uncertainty with ARIMA models is
estimated directly from the learned parameters (Hyndman and
Athanasopoulos 2018). Since ARIMA is a commonly encountered method, we
will refer readers to Hyndman and Athanasopoulos (2018) for further
discussion.

\hypertarget{method-group-3-machine-learning-models}{%
\subsection{Method group 3: Machine Learning
models}\label{method-group-3-machine-learning-models}}

Over the past decade, deep learning has become very popular for a broad
range of challenging time series prediction problems (Makridakis,
Spiliotis, and Assimakopoulos 2018). Deep learning models are often used
to make point forecasts, but for their application to ecological time
series, it will often be necessary to use multi-step, probabilistic
forecasts. For all the deep learning models in this study, we use the
same general process. Each machine learning model is trained on one time
series drawn from the three scenarios described previously. For the Hopf
and saddle node cases, these time series consist of the period leading
up to the bifurcation. A critical transition is, however, included in
the training set for the stochastic transition case. Each model is
trained to learn the parameters of a Laplace distribution for every time
step in the forecast horizon. To produce a forecast, we input a time
series into a model, then we draw samples from the distributions that
were learned during training.

A major nuisance with deep learning methods is their instability to
hyperparameters and initialization seeds (Madhyastha and Jain 2019). We
found that for the same set of hyperparameters, we could produce starkly
different forecasts if we trained two models with different
initialization seeds. One explanation for this instability is that
machine learning models often get stuck on the local optima of loss
surfaces (Madhyastha and Jain 2019). Another likely cause is that
machine learning models commonly overfit the training data (Mehta et al.
2019). Across deep learning, overfitting is a fundamental issue, arising
from neural networks being highly overparameterized (Dar, Muthukumar,
and Baraniuk 2021). With so many parameters, deep learning models tend
to have high variance and thus overfit the training data, a consequence
of the bias-variance trade-off common across statistics and machine
learning (Mehta et al. 2019). One frequently used method to reduce
overfitting is K-fold cross validation (Raschka 2020), but this approach
cannot be effectively employed when there is one or few time series in
the training set. To remedy the instability problem, we use an
ensemble-based method, wherein each ML forecast is the union of
forecasts from 5 individual models that were trained with different
initialization seeds. We found this simple ensemble technique to be an
effective way to improve generalizability in the limited data regime.

Recently, it has become established that using memory or attention-based
neural networks, and an encoder-decoder architecture is crucial for
improving forecasting performance on time series data (Kao et al. 2020;
Lyu et al. 2020; Du et al. 2020). Herein we will provide some background
on what these machine learning methods are and their benefits.

\hypertarget{recurrent-neural-networks}{%
\subsubsection{Recurrent Neural
Networks}\label{recurrent-neural-networks}}

Recurrent neural networks are the predominant memory-based deep learning
method. Recurrent neural networks differ from feed-forward neural
networks in that a recurrent neural network provides feedback to itself
between time steps (Sherstinsky 2020). By providing self-feedback,
recurrent neural networks are able to retain information from previous
time steps and thus learn temporal dependencies. However, a standard
recurrent neural network is unwieldy to train because of the vanishing
and exploding gradient problem (Pascanu, Mikolov, and Bengio 2013), so
there have been specialized neural network architectures designed to
avoid these gradient problems. Long Short-term Memory and Gated
Recurrent Units Networks are considered to be the state of the art
recurrent neural networks that address exploding and vanishing gradients
(Chung et al. 2014). These methods avoid gradient problems by regulating
the self-feedback via gates which perform operations on the feedback
signal -- see Chung et al. (2014) for more details. While GRU's and
LSTM's commonly outperform standard RNN's, it is difficult to anticipate
whether GRU's or LSTM's will be best suited for any time series problem
(Chung et al. 2014), so we investigate both methods.

\hypertarget{transformers}{%
\subsubsection{Transformers}\label{transformers}}

The Transformer is a state of the art ML architecture that is able to
model long and short term dependencies on sequence to sequence tasks
(Vaswani et al. 2017). Transformers use a mechanism called
self-attention which interrelates different positions of the input
sequence in order to find an informative representation of the input
sequence (Vaswani et al. 2017). For example, if given a sentence, a
transformer could learn the contextual relationship between a subject
and a direct object, but a recurrent neural network would process all
the words as one phrase. Because of self-attention, Transformers do not
need to process data sequentially and thus can be parallelized, offering
significant computational advantages (Vaswani et al. 2017). The
Transformer is likely to be a foundational method for future AI research
(Bommasani et al. 2021), so we considered it critical to investigate
Transformers in this study.

\hypertarget{encoder-decoder}{%
\subsubsection{Encoder-Decoder}\label{encoder-decoder}}

Encoder-decoder architectures have been shown empirically to excel on
sequence to sequence tasks (Aitken et al. 2021). Encoder-decoders work
by processing the input sequence into a fixed-length vector then
decoding the fixed-length vector to the predicted output sequence. It is
thought that by encoding the input sequence to a vector,
encoder-decoders find informative representations of the input sequence
that make the prediction task much easier (Sutskever, Vinyals, and Le
2014). Note that it is possible to use any type of neural network as the
encoder and the decoder, but it is most common to use recurrent neural
networks or networks with attention mechanisms (Aitken et al. 2021). Of
the models that we present, Block RNNs and Transformers have
encoder-decoder-based architectures.

\hypertarget{forecast-skill-strictly-proper-scores}{%
\subsection{Forecast skill: strictly proper
scores}\label{forecast-skill-strictly-proper-scores}}

To compare forecasts, we focus exclusively on metrics of forecast skill
which satisfy the property from Gneiting and Raftery (2007) of a
strictly proper score. This ensures the very desirable behavior that no
probabilistic forecast \(Q(x,t)\) can have a score as high as the score
of the true process \(P(x,t)\) on average. In other words, while it is
possible for any of the models considered to \emph{overfit} the data
against which they are trained, i.e.~have a higher likelihood than the
true process, it is not possible for these models to overfit the data
against which they are scored. It is worth noting that this property
applies specifically to probabilistic forecasts and not point forecasts.
Not all common metrics often used to compare forecasts are strictly
proper -- such as the average root-mean-square error or the average
absolute error. Concerns about over-fitting arise in most types of model
estimation and are a particularly acute concern to machine learning
methods due to the bias-variance trade-off (Mehta et al. 2019). This
makes the use of strictly proper scoring especially relevant in
assessing machine learning predictions.

Not even all strictly proper scores will agree on the same relative
ranking between forecasts. We will focus on two of the most common such
skill metrics, CRPS score and log probability score (negative log
likelihood) (e.g.~see Gneiting and Raftery 2007; Gneiting and Katzfuss
2014). Of the two, the logs probability score puts a much a greater
penalty on unexpected observations than CRPS, and may be more suitable
when the occurrence of unexpected events incurs a particularly high
cost. Note that while the minus log-likelihood can be negative for
sufficiently high probability densities, we use a fixed scalar shift of
logs score to ensure the log skill score is strictly positive, which
facilitates visualization without impacting relative rankings.

\hypertarget{results}{%
\section{Results}\label{results}}

We examine forecast skill for each of the six forecasting methods (MCMC,
ARIMA, block-RNN, GRU, LSTM, and Transformer) in each of our three
scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic
transition). In addition to these cases, we also consider an ``ensemble
model'', generated by drawing from the distribution of all models except
the MCMC model. Such ensemble techniques can better reflect uncertainty
than relying on any single method (Gneiting and Raftery 2005). For
simplicity, we consider the unweighted case, where each model is
represented equally in the ensemble. Using model-based simulations
allows us to examine performance against multiple (n=100) replicates of
the ``true'' process, which further helps identify differences that may
occur solely due to chance. By taking the true model structure as given,
MCMC methods can be used to determine a theoretical limit of forecasting
skill. Note that in both bifurcation scenarios, future dynamics will
visit states never previously observed in the historical data that was
used to train each of the methods (e.g.~very small population sizes).
This no-analog aspect of forecasting bifurcation dynamics means that
even with many sample points in the training data \emph{and} perfect
knowledge of the true model structure, posterior distributions of
parameter values are still influenced by the choice of priors.

\begin{figure}
\includegraphics[width=400px,height=300px]{manuscript_files/figure-latex/figure2-1} \caption{Overall distribution of skill scores across models, including an ensemble of methods. Smaller scores are better (indicating smaller errors). Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'simulations of the given scenario.}\label{fig:figure2}
\end{figure}

Overall forecasting skill scores for each model across all three
scenarios are summarized in Fig 2. Average scores (black lines) hide
wide variation in forecast skill. Generally, ML performance tends to be
bracketed between MCMC (essentially the theoretical optimum), and the
statistical ARIMA model, though sometimes performing worse than ARIMA or
better than MCMC. Under scenarios with alternative stable states (saddle
and stochastic), the distribution of scores is often bimodal for ML
models, though not MCMC. The ML ensemble model often performs as well as
the best ML model on average. Note that a wide prediction of uncertainty
does not mean a wide range in the score skill -- for instance the
ensemble model which has the widest array of outcomes often has a
relatively tight distribution of score, especially in logs skill. This
reflects the relative contributions of accuracy and uncertainty as
components in the forecasts. Most ML scores are comparable to MCMC skill
except for the scenario of the saddle-node bifurcation, where all other
models are much worse. Th get a deeper understanding of these general
patterns, we now turn to examine each of the forecast distributions
themselves in comparison to the future ensemble produced by the true
generative process model, Fig 3-5.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure3-1.pdf}
\caption{Forecasts of the Hopf bifurcation under each model, compared to
15 realizations of the true model. The bifurcation occurs soon after
forecasting period begins,leading to progressively larger and larger
oscillations. Prior to the bifurcation, pseudo-cycles are visible in the
training data due to stochastic excitations. Following the bifurcation,
stochasticity blurs the oscillatory pattern across replicate
simulations. Only last 25 time points of training data are shown.}
\end{figure}

Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across
the phenomenological models (ARIMA and machine learning models). All
models are trained using 100 time points drawn from the period of time
prior to the onset of the Hopf bifurcation, which leads to a stable
limit cycle that gradually grows in magnitude. Most models predict a
roughly constant mean with a spread roughly equal to that created by the
stochastic oscillations around the stable node as seen in the training
data prior to the bifurcation. Notably, the GRU method picks up the
oscillatory nature of the dynamics, despite the fact that no true
oscillations were yet present in the training data. However, like the
other ML models, it fails to predict the growing amplitude of those
oscillations. Having access to the true model structure, the MCMC model
alone predicts the transition into a pattern of oscillations which grows
over time, though it tends to overestimate the amplitude of those
oscillations initially. Despite this, overall all methods score
comparably in CRPS score (Fig 2), with most ML methods actually
out-performing the MCMC score on average (Fig 5), albeit with much
greater variation in individual scores. A clearer picture can be seen by
looking at these skill scores over time (Fig 6-7), which show that MCMC
is initially performing worse (over-predicting variance) but as
oscillations grow further, it starts outperforming the more stationary
forecasts of the ML models.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure4-1.pdf}
\caption{Forecasts of the saddle-node bifurcation under each model,
compared to 15 realizations of the true model. Training data proceeds
the bifurcation, making accurate prediction without knowledge of the
underlying model very difficult.}
\end{figure}

The saddle node bifurcation proves even more difficult for most methods
(Fig 4). Only the MCMC model anticipates the sharp transition to an
alternative state, though this behavior is more baked into the method
from the start which takes the true model structure as given. Even
accurate estimation of the MCMC requires slightly informative priors,
though still broad enough to reflect a wide range of possible outcomes.
Two ML methods -- Block RNNs and Transformers -- resemble a naive
prediction extrapolating the last observed state, failing even to
reflect the slow downward trend of the training data. LSTM indicates
greater uncertainty, while GRU shows very large variability which spans
the alternative stable state range. With additional tuning, better
performance may be possible for these ML models. The selected ARIMA
model reflects wide uncertainty that is nevertheless not broad enough to
span the alternative stable state. Consequentially, the MCMC estimate
easily outperforms the ML models (Fig 2).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure5-1.pdf}
\caption{Forecasts of the stochastic transition under each model,
compared to 15 realizations of the true model. In contrast to the other
challenges, this case considers the prediction of replicate systems
starting from the same initial condition, rather than forecasting the
future evolution of the model after the stochastic transition has
already occurred.}
\end{figure}

Machine learning methods do markedly better on the stochastic transition
scenario than in the two bifurcation scenarios (Fig 5). This occurs
because the training data includes the transition phenomenon of
interest. All ML models accurately capture the dynamics of a sharp
transition between alternative stable states -- a dynamic the
statistical ARIMA model entirely fails to reflect. Stochastic transition
events should be approximately exponentially distributed, as seen in the
wide range of waiting times for transitions to occur in replicates of
the true `observed' process (Fig 5). Transformer and Block RNN
distribution times are much more concentrated, while again GRU and
especially the LSTM do a better job reflecting the uncertainty in range
of transition times.

Examining patterns in the scores over time (Fig 6-7) provides a more
nuanced understanding of the forecast dynamics than aggregate scores
alone (Fig 2). In the Hopf bifurcation, CRPS scores get worse over time
across all methods, including the MCMC forecasts. In the saddle node
bifurcation and stochastic transition, the same pattern holds somewhat
more dramatically for non-MCMC forecast, while MCMC scores are worst in
mid-range. Comparing CRPS scores to logs score also emphasizes the
relative role of uncertainty: for instance, the MCMC scores for the Hopf
bifurcation get steadily worse under CRPS but not under logs score. A
relatively sharp transition can be seen under both MCMC scores on the
Hopf bifurcation once the magnitude of the oscillations exceeds the
variance created by mere stochasticity: the MCMC model no longer
over-estimates the spread of the data, while the ML models now
underestimate that variation. CRPS scores for stochastic transitions
exhibit a distinct two-branch pattern, with scores for a given replicate
being either very high (poor skill) or very low, reflecting whether the
individual `true' replicate matches the mean state predicted by the
forecast or the other state. Logs skill score may be a better measure in
this context, where correctly capturing the uncertainty in the forecast
means that this bi-modal structure in scores can be avoided entirely,
e.g.~by the MCMC predictions. The forecast-skill-over time plots
illustrate different reasons for the bi-modal distribution in skill seen
for the saddle-node and stochastic transition scenarios in Fig 2
respectively: in the case of the saddle node, the two modes are
distinguished by time-horizon: short term forecasts are relatively
accurate, longer term forecasts (i.e.~after the catastrophic transition)
are poor. In the case of the stochastic transition model, the two modes
are not structured by horizon but by replicate, with some replicates
having transitioned and others still in the original state.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure6-1.pdf}
\caption{CRPS scores over time for each scenario. Each line represents
the scores against a replicate of time series observations from `true
model'. Patterns show forecast skill generally get worse over time --
graduallyin the case of Hopf bifurcation or suddenly in response to the
saddle-node bifurcation. The MCMC shows a sharp improvement in forecast
skill over time as it passes the transition window. In the stochastic
transition context, this creates two branches, where high values
indicate periods oftime when the forecast predicts the wrong state and
lower branch indicates the correct state.}
\end{figure}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/figure7-1.pdf}
\caption{Logs skill score over time. Forecasts which underestimate
uncertainty do substantially worse in logs score than in CRPS score.
Comparing this panel to those in Fig 6 highlights scenarios that most
often underestimate uncertainty. Generally, MCMC performs better
relative to other models under this metric than it does under
CRPS,reflecting the bias-variance tradeoff of machine learning.}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Ecological systems have long been acknowledged as complex, due not only
to the immense span of dimension and scale such processes involve, but
also the frequency of emergent and non-linear phenomena such as
stochastic resonance, including bifurcations, tipping points, and
hysteresis examined here. Calls for increased forecasting efforts from
ecologists frequently reference the role of changing climate and other
anthropogenic change, which raise the challenge of prediction in
no-analog environments, anticipating ecosystem responses to conditions
that have not been previously observed (Clark et al. 2001; Dietze et al.
2018). This motivates the question, ``What methods will be most reliable
in the face of unobserved conditions?''

In this paper, we carry out an initial exploration on how deep learning
methods can perform on predicting critical transition events. We compare
the ability of several cutting edge machine learning approaches against
statistical and process-based models, and show that deep learning
methods are generally able to strike a middle ground between what we
consider as acceptable and ideal case forecasting methods, ARIMA and
MCMC-based parameter estimation respectively. Although most ML-based
forecasting applications focus on point predictions, we have emphasized
examples that can provide estimates of uncertainty. When the ML models
are able to observe transition phenomena, as in the stochastic case,
they performed comparably to MCMC-based forecasting with respect to CRPS
and log probability score but under-performed MCMC when there were no
transition events in the training sets as in the Hopf and saddle-node
examples. An ensemble forecast combining the predictions of all four ML
methods generally scores as well or better than any one of the ML
methods alone. Yet, examining summary statistics, CRPS and log
probability scores obscures finer detailed components of the forecasts.
For instance, forecast skill varies with the length of the forecast
horizon in a non-monotonic fashion. This is the result of multiple
factors: for some dynamics, such as those involving tipping points, the
long term behavior can be easier to predict than transient transitions.
Both predicted uncertainty and forecast skill can be better on longer
horizons than on shorter ones, as in the MCMC predictions of tipping
point dynamics. It is also important to remember that probabilistic
forecast skill scores do not only measure how close observations are to
expectation, but also reflect the predicted uncertainty: therefore,
over-confidence about predictive accuracy can result in worse scores
than scores from forecasts that are less accurate on average but
correctly reflect a greater degree of uncertainty. The ability to better
reflect uncertainty rather than better average predictions explains much
of the performance of the ensemble model.

The success of these ML models on the stochastic transition case is
particularly notable. All methods are given only a single previous
replicate of a stochastic transition (Fig 5, red line) on which to base
their estimates. This is typical of ecological scenarios where data is
so often limited. While even one observation of a transition is more
than the methods have in our other forecasts, this still presents a
significant challenge to model estimation. Unlike the MCMC case, the ML
models have no prior expectation of a model structure that contains
sharp transitions -- we might have expected these models to perform
little differently than the ARIMA model. Given this single replicate,
all four ML models successfully capture the phenomenological pattern of
a sharp shift between two stable states -- this is behavior that the
structurally simpler family of ARIMA models cannot express. This
provides a clear illustration of the much broader array of
phenomenological behaviors that can be accurately modeled with ML models
compared to classical statistical models. In this way, the ML models can
be seen as imposing even fewer assumptions on the phenomenological
behavior of the system than the ARIMA model. In contrast, the MCMC
performance benefits from very strong process-based assumptions, which
happen to match the `true' model in this case and thus provide a
comparison of the theoretical optimal performance.

The MCMC case illustrates some of the hard limits to ecological
forecasting of critical transitions. Our MCMC forecast assumes the
\emph{true} model structure, the data-generating process, is known, and
the forecaster need only infer the posterior distribution of model
parameters. This is a much stronger assumption than that made by the ML
models, though this assumption can potentially be justified on the basis
of a mechanistic understanding of the processes involved. It is
important not to confound the MCMC example here with the use of MCMC in
process based models of real systems. In the real world, this is never
the case: all models are at best approximations of the underlying
processes (Oreskes, Shrader-Frechette, and Belitz 1994). Despite this
advantage, even the MCMC forecasts differ from the distribution of the
true process. Because the available data come from only a small region
of the dynamical state space, they are consistent with many possible
parameterizations of the same model structure -- which creates
likelihood ridges and non-identifiability of specific parameter values.
Using more simplified versions of the dynamical processes in question,
such as the canonical form of a bifurcation, can mitigate this issue in
some cases. Even when such non-identifiability issues cannot be avoided
entirely, they can usually be diagnosed by examining the degree of
mixing in MCMC sampling and comparing posterior to prior distributions.

When examining the performance of the ML models, it is clear that there
is no single method that excels in all scenarios. Neither is there one
class of ML methods that outperforms the others -- a fact we found
surprising given the reported dominance of encoder-decoders in the field
of sequence-to-sequence deep learning (Aitken et al. 2021). These
observations underscore the point that ML is a very empirically-driven
field in which there are few guarantees on performance. Furthermore, due
to the black-box-ness of deep learning and other reasons like
instability to initialization seed, it is often impossible to provide an
explanation for why certain methods over-perform or fail to meet
expectations.

Overall, ML models and the more traditional ARIMA model fail to predict
the qualitative shift in dynamic behavior that occurs in the critical
transition scenarios (Hopf and saddle node). This is not surprising, as
the training data provide no prior example of such behavior
(e.g.~growing oscillations or a sudden shift). Nevertheless, this should
be an important reminder of a central difficulty in ecological
forecasting. Note that in such scenarios, near-term forecasts (Dietze et
al. 2018) may be very accurate right up to the transition event before
becoming widely wrong. Nor can the possibility of such non-linear
behavior be easily dismissed in ecological models -- the examples
considered here have been bedrock of ecological modeling and management
practices for over half a century (Folke et al. 2004), and if anything
are only too simple, representing a small slice of possible dynamical
behavior of more complicated models.

It may be natural to ask whether this performance would be remedied if
the ML models were trained on data which includes prior examples of
supercritical Hopf or saddle node bifurcations. This question is not as
easy to answer as it may seem, because of the difficulty in defining the
corresponding forecasting scenario. The scenarios we have considered are
true, pure forecasts: the training data comes from a single realization
of a specific generative process, and the task is to predict the future
states of that system before they occur. Would it be possible to train a
predictive algorithm on `analogous' examples of critical transitions?
For instance, could data from other lakes, which may have experienced a
critical transition such as an eutrophication event in the past, be used
to train machine learning models to predict such events in some focal
lake in the future (Scheffer et al. 2001a)? Perhaps, but it depends on
what we mean by an `analogous' system. Even if the underlying mechanism
was accurately captured by the same model, say, the saddle-node model of
Robert M. May (1977) we consider here, it is likely that most of the
individual model parameters would be quite different, even after
accounting for re-scaling or non-dimensionalization of the model
(Hastings 1996). Rarely do ecologists have access to completely
controlled replicates for fitting or training models. The ability for ML
models to successfully generalize from training in such cases remains an
open problem and a promising subject of further investigation.

There are a number of questions that we have left unanswered that we
hope will be addressed in future work. In this paper, we have explored a
small number of machine learning and statistical models that can be used
for forecasting, so comprehensive conclusions can not be drawn on
whether statistical or machine learning-based approaches are better
suited for critical transition forecasting problems. Neither can we
claim that the ML methods employed will translate well to all sudden
transition event forecasting problems in reality, since working with
real data will introduce additional difficulties like how to deal with
missing data, sparse data and observation errors.

Furthermore, our analysis has focused on the task of making a single
forecast prior to the occurrence of a critical transition. Forecasting
is ideally a more iterative process of data assimilation, where
forecasts are updated with respect to additional observations, rather
than projecting 100s of time steps into the future (Dietze et al. 2018).
Updating a forecast after a critical transition has already occurred may
be of little use in the context of hysteresis, such as under the saddle
node or stochastic transition -- recognizing the alternative stable
state only after the system is stuck in that basin will often be
considered `too late'. Assimilation may be more applicable to the Hopf
bifurcation, where additional observations of slowly growing
oscillations may lead to more accurate forecasts. Such models may even
accurately predict the homoclinic bifurcation that occurs when the limit
cycle grows too large, eventually hitting a saddle point of zero
population size for the host species. We leave these cases to future
exploration rather than attempting to explore all such variations in a
single narrative. We have focused on these illustrative examples
accompanied by an extensive appendix of reproducible code implemented
using well documented and intuitive frameworks. We encourage the reader
to use these appendices as an entry point into further exploration.

Ecological forecasting is invariably difficult, even in the idealized
cases of ample measurement data and clearly identified structural
models. This paper is not intended to give a complete answer to whether
deep learning is the best suited method for tipping point forecasting
problems as this will take numerous studies to resolve; instead, this
paper aims to be an early exploration on whether deep learning methods
should be considered as viable tools for this extremely challenging
class of prediction problems. Given the difficulty of forecasting
never-before-observed behavior, as illustrated by the Hopf and
saddle-node bifurcation scenarios, there is good reason for research to
focus more on the kind of qualitative predictions long emphasized in the
literature on early warning signals and resilience
(\textbf{scheffer2015?}). Recently, ML techniques developed for
classification rather than the ML methods used in regression and
forecasting models considered here have demonstrated a more nuanced
ability to reliably detect different classes of critical transitions in
time-series data (\textbf{bury2021?}; \textbf{lapeyrolerie2021?}).
Rather than seeking to provide managers with quantitative, probabilistic
forecasts reflecting a broad uncertainty in possible outcomes, this
literature has sought to emphasize only a more qualitative form of
prediction, such as establishing whether a system is either
``resilient'' or ``approaching a critical transition.'' Decision
sciences have long emphasized the importance of reconciling the
qualitative predictions of resilience thinking with quantitative
forecasts of future states (\textbf{folke2009?}; \textbf{polasky2011?}).
Such approaches could be valuable in concert with probabilistic
forecasts considered here, providing a possible mechanism to identify
when the probabilistic forecast is least reliable.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This material is based upon work supported by the National Science
Foundation under Grant No.~DBI-1942280.

\hypertarget{conflicts-of-interest}{%
\section*{Conflicts of Interest}\label{conflicts-of-interest}}
\addcontentsline{toc}{section}{Conflicts of Interest}

The authors declare there are no conflicts of interest.

\hypertarget{authors-contributions}{%
\section*{Authors' Contributions}\label{authors-contributions}}
\addcontentsline{toc}{section}{Authors' Contributions}

M.L. and C.B. developed the code and wrote the manuscript.

\hypertarget{data-availability}{%
\section*{Data Availability}\label{data-availability}}
\addcontentsline{toc}{section}{Data Availability}

We have created a github repository
\url{https://github.com/boettiger-lab/mee_tipping_point_forecasting.}
that contains the code used to produce the figures herein.

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aitken_understanding_2021}{}}%
Aitken, Kyle, Vinay V. Ramasesh, Yuan Cao, and Niru Maheswaranathan.
2021. {``Understanding {How} {Encoder}-{Decoder} {Architectures}
{Attend}.''} arXiv. \url{http://arxiv.org/abs/2110.15253}.

\leavevmode\vadjust pre{\hypertarget{ref-Barnosky2012}{}}%
Barnosky, Anthony D., Elizabeth A. Hadly, Jordi Bascompte, Eric L.
Berlow, James H. Brown, Mikael Fortelius, Wayne M. Getz, et al. 2012.
{``Approaching a State Shift in {Earth}'s Biosphere.''} \emph{Nature}
486 (7401): 52--58. \url{https://doi.org/10.1038/nature11018}.

\leavevmode\vadjust pre{\hypertarget{ref-Boettiger2018}{}}%
Boettiger, Carl. 2018a. {``From Noise to Knowledge: How Randomness
Generates Novel Phenomena and Reveals Information.''} \emph{Ecology
Letters}. \url{https://doi.org/10.1111/ele.13085}.

\leavevmode\vadjust pre{\hypertarget{ref-boettiger_noise_2018}{}}%
---------. 2018b. {``From Noise to Knowledge: How Randomness Generates
Novel Phenomena and Reveals Information.''} Edited by Tim Coulson.
\emph{Ecology Letters} 21 (8): 1255--67.
\url{https://doi.org/10.1111/ele.13085}.

\leavevmode\vadjust pre{\hypertarget{ref-boettiger_early_2012}{}}%
Boettiger, Carl, and Alan Hastings. 2012. {``Early Warning Signals and
the Prosecutor's Fallacy.''} \emph{Proceedings of the Royal Society B:
Biological Sciences} 279 (1748): 4734--39.
\url{https://doi.org/10.1098/rspb.2012.2085}.

\leavevmode\vadjust pre{\hypertarget{ref-bommasani_opportunities_2021}{}}%
Bommasani, Rishi, Drew A. Hudson, Ehsan Adeli, Russ Altman, Simran
Arora, Sydney von Arx, Michael S. Bernstein, et al. 2021. {``On the
{Opportunities} and {Risks} of {Foundation} {Models}.''}
\emph{arXiv:2108.07258 {[}Cs{]}}, August.
\url{http://arxiv.org/abs/2108.07258}.

\leavevmode\vadjust pre{\hypertarget{ref-bury_2021}{}}%
Bury, Thomas M., R. I. Sujith, Induja Pavithran, Marten Scheffer,
Timothy M. Lenton, Madhur Anand, and Chris T. Bauch. 2021. {``Deep
Learning for Early Warning Signals of Tipping Points.''}
\emph{Proceedings of the National Academy of Sciences} 118 (39):
e2106140118. \url{https://doi.org/10.1073/pnas.2106140118}.

\leavevmode\vadjust pre{\hypertarget{ref-carpenter_early_2011}{}}%
Carpenter, S. R., J. J. Cole, M. L. Pace, R. Batt, W. A. Brock, T.
Cline, J. Coloso, et al. 2011. {``Early {Warnings} of {Regime} {Shifts}:
{A} {Whole}-{Ecosystem} {Experiment}.''} \emph{Science} 332 (6033):
1079--82. \url{https://doi.org/10.1126/science.1203672}.

\leavevmode\vadjust pre{\hypertarget{ref-chung_empirical_2014}{}}%
Chung, Junyoung, Caglar Gulcehre, KyungHyun Cho, and Yoshua Bengio.
2014. {``Empirical {Evaluation} of {Gated} {Recurrent} {Neural}
{Networks} on {Sequence} {Modeling}.''} \emph{arXiv:1412.3555 {[}Cs{]}},
December. \url{http://arxiv.org/abs/1412.3555}.

\leavevmode\vadjust pre{\hypertarget{ref-clark_ecological_2001}{}}%
Clark, James S., Steven R. Carpenter, Mary Barber, Scott Collins, Andy
Dobson, Jonathan A. Foley, David M. Lodge, et al. 2001. {``Ecological
{Forecasts}: {An} {Emerging} {Imperative}.''} \emph{Science} 293 (5530):
657--60. \url{https://doi.org/10.1126/science.293.5530.657}.

\leavevmode\vadjust pre{\hypertarget{ref-Dai2015}{}}%
Dai, Lei, Kirill S. Korolev, and Jeff Gore. 2015. {``Relation Between
Stability and Resilience Determines the Performance of Early Warning
Signals Under Different Environmental Drivers.''} \emph{Proceedings of
the National Academy of Sciences}, 201418415.
\url{https://doi.org/10.1073/pnas.1418415112}.

\leavevmode\vadjust pre{\hypertarget{ref-Dai2012}{}}%
Dai, Lei, Daan Vorselen, Kirill S Korolev, and J. Gore. 2012. {``Generic
{Indicators} for {Loss} of {Resilience} {Before} a {Tipping} {Point}
{Leading} to {Population} {Collapse}.''} \emph{Science (New York, N.Y.)}
336 (6085): 1175--77. \url{https://doi.org/10.1126/science.1219805}.

\leavevmode\vadjust pre{\hypertarget{ref-dakos_methods_2012}{}}%
Dakos, Vasilis, Stephen R. Carpenter, William A. Brock, Aaron M.
Ellison, Vishwesha Guttal, Anthony R. Ives, Sonia Kéfi, et al. 2012.
{``Methods for {Detecting} {Early} {Warnings} of {Critical}
{Transitions} in {Time} {Series} {Illustrated} {Using} {Simulated}
{Ecological} {Data}.''} Edited by Bülent Yener. \emph{PLoS ONE} 7 (7):
e41010. \url{https://doi.org/10.1371/journal.pone.0041010}.

\leavevmode\vadjust pre{\hypertarget{ref-dar_farewell_2021}{}}%
Dar, Yehuda, Vidya Muthukumar, and Richard G. Baraniuk. 2021. {``A
{Farewell} to the {Bias}-{Variance} {Tradeoff}? {An} {Overview} of the
{Theory} of {Overparameterized} {Machine} {Learning}.''} arXiv.
\url{http://arxiv.org/abs/2109.02355}.

\leavevmode\vadjust pre{\hypertarget{ref-dietze_iterative_2018}{}}%
Dietze, Michael C., Andrew Fox, Lindsay M. Beck-Johnson, Julio L.
Betancourt, Mevin B. Hooten, Catherine S. Jarnevich, Timothy H. Keitt,
et al. 2018. {``Iterative Near-Term Ecological Forecasting: {Needs},
Opportunities, and Challenges.''} \emph{Proceedings of the National
Academy of Sciences} 115 (7): 1424--32.
\url{https://doi.org/10.1073/pnas.1710231115}.

\leavevmode\vadjust pre{\hypertarget{ref-du_multivariate_2020}{}}%
Du, Shengdong, Tianrui Li, Yan Yang, and Shi-Jinn Horng. 2020.
{``Multivariate Time Series Forecasting via Attention-Based
Encoder--Decoder Framework.''} \emph{Neurocomputing} 388 (May): 269--79.
\url{https://doi.org/10.1016/j.neucom.2019.12.118}.

\leavevmode\vadjust pre{\hypertarget{ref-Fischer2009}{}}%
Fischer, Joern, Garry D Peterson, Toby A. Gardner, Line J Gordon, Ioan
Fazey, Thomas Elmqvist, Adam Felton, Carl Folke, and Stephen Dovers.
2009. {``Integrating Resilience Thinking and Optimisation for
Conservation.''} \emph{Trends in Ecology \& Evolution} 24 (10): 549--54.
\url{https://doi.org/10.1016/j.tree.2009.03.020}.

\leavevmode\vadjust pre{\hypertarget{ref-Folke2004}{}}%
Folke, Carl, Steve Carpenter, Brian Walker, Marten Scheffer, Thomas
Elmqvist, Lance Gunderson, and C. S. Holling. 2004. {``Regime {Shifts},
{Resilience}, and {Biodiversity} in {Ecosystem} {Management}.''}
\emph{Annual Review of Ecology, Evolution, and Systematics} 35 (1):
557--81. \url{https://doi.org/10.1146/annurev.ecolsys.35.021103.105711}.

\leavevmode\vadjust pre{\hypertarget{ref-Getz2018}{}}%
Getz, Wayne M., Charles R. Marshall, Colin J. Carlson, Luca Giuggioli,
Sadie J. Ryan, Stephanie S. Romañach, Carl Boettiger, et al. 2018.
{``Making Ecological Models Adequate.''} Edited by Tim Coulson.
\emph{Ecology Letters} 21 (2): 153--66.
\url{https://doi.org/10.1111/ele.12893}.

\leavevmode\vadjust pre{\hypertarget{ref-gneiting_probabilistic_2014}{}}%
Gneiting, Tilmann, and Matthias Katzfuss. 2014. {``Probabilistic
{Forecasting}.''} \emph{Annual Review of Statistics and Its Application}
1 (1): 125--51.
\url{https://doi.org/10.1146/annurev-statistics-062713-085831}.

\leavevmode\vadjust pre{\hypertarget{ref-gneiting_strictly_2007}{}}%
Gneiting, Tilmann, and Adrian E Raftery. 2007. {``Strictly {Proper}
{Scoring} {Rules}, {Prediction}, and {Estimation}.''} \emph{Journal of
the American Statistical Association} 102 (477): 359--78.
\url{https://doi.org/10.1198/016214506000001437}.

\leavevmode\vadjust pre{\hypertarget{ref-gneiting_weather_2005}{}}%
Gneiting, Tilmann, and Adrian E. Raftery. 2005. {``Weather {Forecasting}
with {Ensemble} {Methods}.''} \emph{Science} 310 (5746): 248--49.
\url{https://doi.org/10.1126/science.1115255}.

\leavevmode\vadjust pre{\hypertarget{ref-hallett_why_2004}{}}%
Hallett, T. B., T. Coulson, J. G. Pilkington, T. H. Clutton-Brock, J. M.
Pemberton, and B. T. Grenfell. 2004. {``Why Large-Scale Climate Indices
Seem to Predict Ecological Processes Better Than Local Weather.''}
\emph{Nature} 430 (6995): 71--75.
\url{https://doi.org/10.1038/nature02708}.

\leavevmode\vadjust pre{\hypertarget{ref-Hastings1996}{}}%
Hastings, Alan. 1996. \emph{Population {Biology}: {Concepts} and
{Models}}. Springer.

\leavevmode\vadjust pre{\hypertarget{ref-hyndman_forecasting_2018}{}}%
Hyndman, R. J., and G. Athanasopoulos. 2018. \emph{Forecasting:
Principles and Practice}. 2nd ed. Melbourne, Australia: OTexts.
\href{https://OTexts.com/fpp2}{OTexts.com/fpp2}.

\leavevmode\vadjust pre{\hypertarget{ref-kampen_stochastic_1992}{}}%
Kampen, N. G. van. 1992. \emph{Stochastic Processes in Physics and
Chemistry}. Rev. and enl. ed. North-{Holland} Personal Library.
Amsterdam ; New York: North-Holland.

\leavevmode\vadjust pre{\hypertarget{ref-kao_exploring_2020}{}}%
Kao, I-Feng, Yanlai Zhou, Li-Chiu Chang, and Fi-John Chang. 2020.
{``Exploring a {Long} {Short}-{Term} {Memory} Based {Encoder}-{Decoder}
Framework for Multi-Step-Ahead Flood Forecasting.''} \emph{Journal of
Hydrology} 583 (April): 124631.
\url{https://doi.org/10.1016/j.jhydrol.2020.124631}.

\leavevmode\vadjust pre{\hypertarget{ref-Levins1966}{}}%
Levins, Richard. 1966. {``The Strategy of Model Building in Population
Biology.''} \emph{American Scientist} 54 (4): 421--31.

\leavevmode\vadjust pre{\hypertarget{ref-lyu_lstm_2020}{}}%
Lyu, Pingyang, Ning Chen, Shanjun Mao, and Mei Li. 2020. {``{LSTM} Based
Encoder-Decoder for Short-Term Predictions of Gas Concentration Using
Multi-Sensor Fusion.''} \emph{Process Safety and Environmental
Protection} 137 (May): 93--105.
\url{https://doi.org/10.1016/j.psep.2020.02.021}.

\leavevmode\vadjust pre{\hypertarget{ref-madhyastha_model_2019}{}}%
Madhyastha, Pranava, and Rishabh Jain. 2019. {``On {Model} {Stability}
as a {Function} of {Random} {Seed}.''} \emph{arXiv:1909.10447 {[}Cs,
Stat{]}}, September. \url{http://arxiv.org/abs/1909.10447}.

\leavevmode\vadjust pre{\hypertarget{ref-makridakis_statistical_2018}{}}%
Makridakis, Spyros, Evangelos Spiliotis, and Vassilios Assimakopoulos.
2018. {``Statistical and {Machine} {Learning} Forecasting Methods:
{Concerns} and Ways Forward.''} Edited by Alejandro Raul Hernandez
Montoya. \emph{PLOS ONE} 13 (3): e0194889.
\url{https://doi.org/10.1371/journal.pone.0194889}.

\leavevmode\vadjust pre{\hypertarget{ref-May1977}{}}%
May, Robert M. 1977. {``Thresholds and Breakpoints in Ecosystems with a
Multiplicity of Stable States.''} \emph{Nature} 269 (5628): 471--77.
\url{https://doi.org/10.1038/269471a0}.

\leavevmode\vadjust pre{\hypertarget{ref-may_population_1979}{}}%
May, Robert M., and Roy M. Anderson. 1979. {``Population Biology of
Infectious Diseases: {Part} {II}.''} \emph{Nature} 280 (5722): 455--61.
\url{https://doi.org/10.1038/280455a0}.

\leavevmode\vadjust pre{\hypertarget{ref-mehta_high-bias_2019}{}}%
Mehta, Pankaj, Marin Bukov, Ching-Hao Wang, Alexandre G. R. Day, Clint
Richardson, Charles K. Fisher, and David J. Schwab. 2019. {``A
High-Bias, Low-Variance Introduction to {Machine} {Learning} for
Physicists.''} \emph{Physics Reports} 810 (May): 1--124.
\url{https://doi.org/10.1016/j.physrep.2019.03.001}.

\leavevmode\vadjust pre{\hypertarget{ref-nicholson_balance_1935}{}}%
Nicholson, A. J., and V. A. Bailey. 1935. {``The {Balance} of {Animal}
{Populations}.---{Part} {I}.''} \emph{Proceedings of the Zoological
Society of London} 105 (3): 551--98.
\url{https://doi.org/10.1111/j.1096-3642.1935.tb01680.x}.

\leavevmode\vadjust pre{\hypertarget{ref-nicholson_outline_1954}{}}%
Nicholson, Aj. 1954a. {``An Outline of the Dynamics of Animal
Populations.''} \emph{Australian Journal of Zoology} 2 (1): 9.
\url{https://doi.org/10.1071/ZO9540009}.

\leavevmode\vadjust pre{\hypertarget{ref-nicholson_compensatory_1954}{}}%
---------. 1954b. {``Compensatory Reactions of Populations to Stresses,
and Their Evolutionary Significance.''} \emph{Australian Journal of
Zoology} 2 (1): 1. \url{https://doi.org/10.1071/ZO9540001}.

\leavevmode\vadjust pre{\hypertarget{ref-Oreskes1994}{}}%
Oreskes, Naomi, Kristin Shrader-Frechette, and Kenneth Belitz. 1994.
{``Verification, {Validation}, and {Confirmation} of {Numerical}
{Models} in the {Earth} {Sciences}.''} \emph{Science} 263 (5147):
641--46. \url{https://doi.org/10.1126/science.263.5147.641}.

\leavevmode\vadjust pre{\hypertarget{ref-Ovaskainen2010}{}}%
Ovaskainen, Otso, and Baruch Meerson. 2010. {``Stochastic Models of
Population Extinction.''} \emph{Trends in Ecology \& Evolution} 25 (11):
643--52. \url{https://doi.org/10.1016/j.tree.2010.07.009}.

\leavevmode\vadjust pre{\hypertarget{ref-pascanu_difficulty_2013}{}}%
Pascanu, Razvan, Tomas Mikolov, and Yoshua Bengio. 2013. {``On the
Difficulty of Training {Recurrent} {Neural} {Networks}.''}
\emph{arXiv:1211.5063 {[}Cs{]}}, February.
\url{http://arxiv.org/abs/1211.5063}.

\leavevmode\vadjust pre{\hypertarget{ref-Polasky2011}{}}%
Polasky, Stephen, Stephen R. Carpenter, Carl Folke, and Bonnie Keeler.
2011. {``Decision-Making Under Great Uncertainty: Environmental
Management in an Era of Global Change.''} \emph{Trends in Ecology \&
Evolution}, May, 1--7. \url{https://doi.org/10.1016/j.tree.2011.04.007}.

\leavevmode\vadjust pre{\hypertarget{ref-raschka_model_2020}{}}%
Raschka, Sebastian. 2020. {``Model {Evaluation}, {Model} {Selection},
and {Algorithm} {Selection} in {Machine} {Learning}.''} arXiv.
\url{http://arxiv.org/abs/1811.12808}.

\leavevmode\vadjust pre{\hypertarget{ref-scheffer_early-warning_2009}{}}%
Scheffer, Marten, Jordi Bascompte, William A. Brock, Victor Brovkin,
Stephen R. Carpenter, Vasilis Dakos, Hermann Held, Egbert H. van Nes,
Max Rietkerk, and George Sugihara. 2009. {``Early-Warning Signals for
Critical Transitions.''} \emph{Nature} 461 (7260): 53--59.
\url{https://doi.org/10.1038/nature08227}.

\leavevmode\vadjust pre{\hypertarget{ref-Scheffer2001}{}}%
Scheffer, Marten, Stephen R. Carpenter, Jonathan A. Foley, Carl Folke,
and Brian H Walker. 2001a. {``Catastrophic Shifts in Ecosystems.''}
\emph{Nature} 413 (6856): 591--96.
\url{https://doi.org/10.1038/35098000}.

\leavevmode\vadjust pre{\hypertarget{ref-scheffer_anticipating_2012}{}}%
Scheffer, Marten, Stephen R. Carpenter, Timothy M. Lenton, Jordi
Bascompte, William Brock, Vasilis Dakos, Johan van de Koppel, et al.
2012. {``Anticipating {Critical} {Transitions}.''} \emph{Science} 338
(6105): 344--48. \url{https://doi.org/10.1126/science.1225244}.

\leavevmode\vadjust pre{\hypertarget{ref-scheffer_catastrophic_2001}{}}%
Scheffer, Marten, Steve Carpenter, Jonathan A. Foley, Carl Folke, and
Brian Walker. 2001b. {``Catastrophic Shifts in Ecosystems.''}
\emph{Nature} 413 (6856): 591--96.
\url{https://doi.org/10.1038/35098000}.

\leavevmode\vadjust pre{\hypertarget{ref-schindler_portfolio_2015}{}}%
Schindler, Daniel E, Jonathan B Armstrong, and Thomas E Reed. 2015.
{``The Portfolio Concept in Ecology and Evolution.''} \emph{Frontiers in
Ecology and the Environment} 13 (5): 257--63.
\url{https://doi.org/10.1890/140275}.

\leavevmode\vadjust pre{\hypertarget{ref-sherstinsky_fundamentals_2020}{}}%
Sherstinsky, Alex. 2020. {``Fundamentals of {Recurrent} {Neural}
{Network} ({RNN}) and {Long} {Short}-{Term} {Memory} ({LSTM})
{Network}.''} \emph{Physica D: Nonlinear Phenomena} 404 (March): 132306.
\url{https://doi.org/10.1016/j.physd.2019.132306}.

\leavevmode\vadjust pre{\hypertarget{ref-sutskever_sequence_2014}{}}%
Sutskever, Ilya, Oriol Vinyals, and Quoc V. Le. 2014. {``Sequence to
{Sequence} {Learning} with {Neural} {Networks}.''} \emph{arXiv:1409.3215
{[}Cs{]}}, December. \url{http://arxiv.org/abs/1409.3215}.

\leavevmode\vadjust pre{\hypertarget{ref-vaswani_attention_2017}{}}%
Vaswani, Ashish, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion
Jones, Aidan N. Gomez, Lukasz Kaiser, and Illia Polosukhin. 2017.
{``Attention {Is} {All} {You} {Need}.''}
\url{https://doi.org/10.48550/ARXIV.1706.03762}.

\leavevmode\vadjust pre{\hypertarget{ref-Williams2007}{}}%
Williams, John W., and Stephen T. Jackson. 2007. {``Novel Climates,
No-Analog Communities, and Ecological Surprises.''} \emph{Frontiers in
Ecology and the Environment} 5 (9): 475--82.
\url{https://doi.org/10.1890/070037}.

\end{CSLReferences}

\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2022-08-14 manuscript/manuscript.Rmd

--- 2022-08-11 manuscript/manuscript.Rmd
@@
   - \usepackage{lineno}
+  - \usepackage{bbm}
   - \usepackage{setspace}
@@
 
-In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states, as frequently called for in ecological research [@clark2001; @dietze_iterative_2018].
+In this paper, we focus on the task of producing quantitative, probabilistic forecasts reflecting the possible distribution of future states, as frequently called for in ecological research [@clark_ecological_2001; @dietze_iterative_2018].
 Such forecasting tasks may arise whenever a manager is interested in knowing the future states of a system: such as setting future catch quotas for a fishery or adjusting eradication effort for an invasive species.
@@
 Recent work such as @bury_2021 has introduced ML methods to consider classification of this transition in four possible categories (Hopf, saddle-node, transcritical, or no bifurcation) rather than two (bifurcation or not).
-These are important results that with considerable promise [@lapeyrolerie2021], but which nevertheless address a very different question using very different methods. 
+These are important results with considerable promise [@lapeyrolerie_teaching_2021], but which nevertheless address a very different question using very different methods. 
 Early warning signals only predict "a big change may be coming soon" -- they do not try to forecast when or how big.
@@
 
-**Selecting timescales**
+## Selecting timescales
 In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Fig 1. 
@@
 We will focus on two of the most common such skill metrics, CRPS score and log probability score (negative log likelihood) [e.g. see @gneiting_strictly_2007; @gneiting_probabilistic_2014].
-Of the two, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
+We define these scores explicitly in Equations 7-8, where $F$ and $f$ respectively correspond to the cumulative distribution function and probability density function of the forecast; and, $y$ denotes an observation.
+Of the two metrics, the logs probability score puts a much a greater penalty on unexpected observations than CRPS, and may be more suitable when the occurrence of unexpected events incurs a particularly high cost.
 Note that while the minus log-likelihood can be negative for sufficiently high probability densities, we use a fixed scalar shift of logs score to ensure the log skill score is strictly positive, which facilitates visualization without impacting relative rankings.
+
+\begin{align}
+\text{LogS}(F, y) &= - \log{f(y)} \\
+\text{CRPS}(F, y) &= \int (F(z) - \mathbbm{1}\{y \leq z\})^2 \, dz
+\end{align}
 
@@
 This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
-Given the difficulty of forecasting never-before-observed behavior, as illustrated by the Hopf and saddle-node bifurcation scenarios, there is good reason for research to focus more on the kind of qualitative predictions long emphasized in the literature on early warning signals and resilience [@scheffer2015].
-Recently, ML techniques developed for classification rather than the ML methods used in regression and forecasting models considered here have demonstrated a more nuanced ability to reliably detect different classes of critical transitions in time-series data [@bury2021; @lapeyrolerie2021].
+Given the difficulty of forecasting never-before-observed behavior, as illustrated by the Hopf and saddle-node bifurcation scenarios, there is good reason for research to focus more on the kind of qualitative predictions long emphasized in the literature on early warning signals and resilience [@scheffer_anticipating_2012].
+Recently, ML techniques developed for classification rather than the ML methods used in regression and forecasting models considered here have demonstrated a more nuanced ability to reliably detect different classes of critical transitions in time-series data [@bury_2021; @lapeyrolerie_teaching_2021].
 Rather than seeking to provide managers with quantitative, probabilistic forecasts reflecting a broad uncertainty in possible outcomes, this literature has sought to emphasize only a more qualitative form of prediction, such as establishing whether a system is either "resilient" or "approaching a critical transition."
-Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@folke2009; @polasky2011].
+Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@Fischer2009; @polasky_decision-making_2011].
 Such approaches could be valuable in concert with probabilistic forecasts considered here, providing a possible mechanism to identify when the probabilistic forecast is least reliable.

### 2022-08-16 manuscript/manuscript.Rmd
One line grammar edit

@@
 Bifurcations represent the kind of non-linear responses complex systems can make as the result of slowly changing parameters.
-This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no corresponding analog in the historical data.
+This can create a particularly challenging forecasting task when such transitions have not been previously observed in the same system, requiring the forecast to anticipate dynamics for which there are no analogs in the historical data.
 Forecast skill under such no-analog conditions may be particularly relevant to ecological forecasting in the context of global change [@Williams2007].

### 2022-08-16 manuscript/manuscript.Rmd
Tidying up some more

@@
 A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
-We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds.
+We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained the same model with different initialization seeds.
 One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
@@
 Most ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
-Th get a deeper understanding of these general patterns, we now turn to examine each of the forecast distributions themselves in comparison to the future ensemble produced by the true generative process model, Fig 3-5.
+To get a deeper understanding of these general patterns, we now turn to examine each of the forecast distributions themselves in comparison to the future ensemble produced by the true generative process model, Fig 3-5.
 
@@
 Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
-Despite this, overall all methods score comparably in CRPS score (Fig 2), with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
+Despite this, overall all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
 A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.
@@
 In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
-In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
-while MCMC scores are worst in mid-range.
+In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worse in mid-range.
 Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
@@
 by the MCMC predictions.
-The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e.
+The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, and longer term forecasts (i.e.
 after the catastrophic transition) are poor.

### 2022-08-16 manuscript/manuscript.Rmd
Reverting one edit

@@
 In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
-In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worse in mid-range.
+In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worst in mid-range.
 Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 

### 2022-09-17 manuscript/manuscript.Rmd
Editing some figure captions

@@
 #| fig.cap="CRPS scores over time for each scenario.
-#|  Each line represents the scores against a replicate of time series observations from 'true model'. 
-#|  Patterns show forecast skill generally get worse over time -- gradually 
-#| in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
-#|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
-#|  In the stochastic transition context, this creates two branches, where high values indicate periods of 
-#| time when the forecast predicts the wrong state and lower branch indicates the correct state."
+#|  Each line represents the CRPS scores against a replicate of time series observations from the 'true model'. 
+#|  The general pattern across these plots is that forecast skill gets worse over time -- gradually 
+#|  in the case of a Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
+#|  In the stochastic transition case,  the scores tend to diverge in two branches, where high values indicate periods of 
+#|  time when the forecast predicts the wrong equilibrium state and the lower branch indicates predicting the correct one."
 
@@
 #|  Forecasts which underestimate uncertainty do substantially worse in logs score than in CRPS score. 
-#|  Comparing this panel to those in Fig 6 highlights scenarios that most often underestimate uncertainty. 
+#|  Comparing this panel to those in Figure 6 highlights scenarios that most often underestimate uncertainty. 
 #|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
-#| reflecting the bias-variance tradeoff of machine learning."
+#|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning."
 

### 2022-09-20 manuscript/manuscript.Rmd
Read through finished

@@
 Most of our forecasts use the `darts` framework, a sophisticated and well documented Python library with support for a wide range of methods.
-Our model-based MCMC forecasts use the `greta` framework, a R library that uses Python-based `tensorflow` probability to achieve better performance.
+Our model-based MCMC forecasts use the `greta` framework, a R library that uses Python-based `TensorFlow Probability` to achieve better performance.
 While Python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores.
@@
 The Nicholson-Bailey model describes a predator-prey dynamic for the relationship of a host species and an obligate parasitoid, originally used to model the population dynamics of blowflies (_Lucilia cuprina_) [@nicholson_balance_1935; @nicholson_outline_1954; @nicholson_compensatory_1954].
-We consider the form which includes density dependence in the host species, and allow for environmental stochasticity,
+We consider the form which includes density dependence in the host species, and we allow for environmental stochasticity,
 
@@
 
-Where $H_t$ is the population density of the host species at time $t$, (in arbitrary units) and $P_t$ the population density of the parasitoid.
+Where $H_t$ is the population density of the host species at time $t$, (in arbitrary units) and $P_t$ is the population density of the parasitoid.
 The time step is defined by the generation time of the parasitoid, which is about two weeks in the case of Nicholson's blowflies [@nicholson_outline_1954].
@@
 ## Selecting timescales
-In each scenario,  t = 0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) is illustrated in Fig 1. 
-For the Hopf bifurcation, forecast begins at t=100 and extends to t =200, for the saddle node, forecast begins at t=250 and extends to t = 500, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t = 250.
+In each scenario,  t=0 is the start time of the training data, while the length of training data and forecast horizon (with ensembles sampled from the true distribution) are illustrated in Fig 1. 
+For the Hopf bifurcation, forecasts begin at t=100 and extend to t=200; for the saddle node, forecasts begin at t=250 and extend to t=500; and, for the stochastic transition, both training data and forecasting tasks begin at 0 and extend to t=250.
 While much attention is often paid to the number of data points in training or testing data, it is essential to realize that these are only meaningful relative to the specific process in question. 
-Thus, in each case we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario. 
-For instance, if the stochastic scenario was restricted to a much shorter timescale used in the Hopf case, few replicate simulations would experience a transition at all. 
-If it were made much longer, most of the timeseries would be spent post-transition. 
-Likewise, if the forecast for the Hopf scenario was extended much further into the future under the current parameterization, the system experiences a homoclinic bifurcation at which the population collapses to 0.
+Thus, in each case, we have selected these time intervals to focus on the dynamical process in question, which unfolds at a different rate and tempo in each scenario.
+For instance, if the stochastic scenario was restricted to the much shorter timescale used in the Hopf case, few replicate simulations would experience a transition at all. 
+If length of the stochastic transition timeseries was made much longer, most of the timeseries would be spent post-transition. 
+Likewise, if the forecast horizon for the Hopf scenario was extended much further into the future under the current parameterization, the system would experience a homoclinic bifurcation at which the population collapses to 0.
 Using different length timescales allows us to consider the three different forecasting tasks illustrated in Fig 1 that focus around predicting the critical behavior, rather than predicting long periods of relative stasis.
@@
 
-Recurrent neural networks are the predominant memory-based deep learning method.
+Recurrent neural networks (RNN's) are the predominant memory-based deep learning method.
 Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky_fundamentals_2020].
@@
 However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@pascanu_difficulty_2013], so there have been specialized neural network architectures designed to avoid these gradient problems.
-Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
+Long Short-term Memory (LSTM) and Gated Recurrent Units (GRU) Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung_empirical_2014].
 These methods avoid gradient problems by regulating the self-feedback via gates which perform operations on the feedback signal -- see @chung_empirical_2014 for more details.
@@
 
-### Encoder-Decoder
+### Encoder-Decoders
 
 Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@aitken_understanding_2021].
-Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
+Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding this fixed-length vector to an output sequence.
 It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task much easier [@sutskever_sequence_2014].
 Note that it is possible to use any type of neural network as the encoder and the decoder, but it is most common to use recurrent neural networks or networks with attention mechanisms [@aitken_understanding_2021].
-Of the models that we present, Block RNNs and Transformers have encoder-decoder-based architectures.
+Of the models that we present, Block RNN's are a direct example of an encoder-decoder-based model since a Block RNN employs a RNN as an encoder and a separate RNN as a decoder.
+Transformers also have an encoder and decoder component.
 
@@
 Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
-Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
+Notably, the GRU model picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
 However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
 Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
-Despite this, overall all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
+Despite this, all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
 A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.
@@
 The saddle node bifurcation proves even more difficult for most methods (Fig 4).
-Only the MCMC model anticipates the sharp transition to an alternative state, though this behavior is more baked into the method from the start which takes the true model structure as given. 
+Only the MCMC model anticipates the sharp transition to an alternative state.
 Even accurate estimation of the MCMC requires slightly informative priors, though still broad enough to reflect a wide range of possible outcomes. 
-Two ML methods -- Block RNNs and Transformers -- resemble a naive prediction extrapolating the last observed state, failing even to reflect the slow downward trend of the training data. 
-LSTM indicates greater uncertainty, while GRU shows very large variability which spans the alternative stable state range. 
+Two ML methods -- Block RNN's and Transformers -- resemble a naive prediction extrapolating the last observed state, failing to reflect the slow downward trend of the training data. 
+LSTM's indicate greater uncertainty, while GRU's show very large variability which spans the alternative stable state range. 
 With additional tuning, better performance may be possible for these ML models. 
@@
 In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
-In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worst in mid-range.
+In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecasts, while MCMC scores are at their worst around the middle of the forecast horizon.
 Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
 A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation.
-CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state.
+CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast.
 Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g.
 by the MCMC predictions.
-The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, and longer term forecasts (i.e.
+The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon; short term forecasts are relatively accurate, and longer term forecasts (i.e.
 after the catastrophic transition) are poor.
@@
 The MCMC case illustrates some of the hard limits to ecological forecasting of critical transitions.
-Our MCMC forecast assumes the _true_ model structure, the data-generating process, is known, and the forecaster need only infer the posterior distribution of model parameters.
+Our MCMC forecast assumes that the data-generating process is known, so the forecaster need only infer the posterior distribution of model parameters.
 This is a much stronger assumption than that made by the ML models, though this assumption can potentially be justified on the basis of a mechanistic understanding of the processes involved.
@@
 In the real world, this is never the case: all models are at best approximations of the underlying processes [@Oreskes1994].
-<!-- ML: maybe "In the real world, having access to the _true_ model is never the case:"-->
 Despite this advantage, even the MCMC forecasts differ from the distribution of the true process.
@@
 In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for critical transition forecasting problems.
-Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.
+Neither can we claim that ML methods will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.
 
@@
 We leave these cases to future exploration rather than attempting to explore all such variations in a single narrative.
-We have focused on these illustrative examples accompanied by an extensive appendix of reproducible code implemented using well documented and intuitive frameworks.
-We encourage the reader to use these appendices as an entry point into further exploration.
 

### 2022-09-20 manuscript/manuscript.Rmd
Fixing some captions

@@
 #| fig.cap="Forecasts of the Hopf bifurcation under each model, compared to 15 realizations of the true model. 
-#|  The bifurcation occurs soon after forecasting period begins, 
-#| leading to progressively larger and larger oscillations. 
+#|  The bifurcation occurs soon after the forecasting period begins, 
+#|  leading to progressively larger oscillations. 
 #|  Prior to the bifurcation, pseudo-cycles are visible in the training data due to stochastic excitations. 
 #|  Following the bifurcation, stochasticity blurs the oscillatory pattern across replicate simulations. 
-#|  Only last 25 time points of training data are shown."
+#|  Only the last 25 time points of the training data are shown."
 
@@
 #|  In the stochastic transition case,  the scores tend to diverge in two branches, where high values indicate periods of 
-#|  time when the forecast predicts the wrong equilibrium state and the lower branch indicates predicting the correct one."
+#|  time when the forecast predicts the wrong equilibrium state and the lower branch indicates predictions of the correct one.
+The time axis in this plot and in Figure 7 refers to the time from the beginning of the forecast window, not the time from the beginning of the time series as in other plots."
 

### 2022-09-20 manuscript/manuscript.Rmd
fix var names in dummy 'predicted' entry

--- 2022-09-20 manuscript/manuscript.Rmd
@@
   mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
-  bind_rows(tibble(i=1,t=1,type="predicted", variable="H", value=NA)) |> #placeholder
+  bind_rows(tibble(i=1,t=1,type="predicted", species="host", value=NA)) |> #placeholder
   mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
   ggplot(aes(t, value, group=interaction(species, i, type), col=type, lty=species)) + 
-  geom_line(show.legend = TRUE, alpha = 0.7) +
+  geom_line(alpha = 0.7) +
   theme_bw() +  

### 2022-09-22 manuscript/manuscript.Rmd
Small edits again

--- 2022-09-20 manuscript/manuscript.Rmd
@@
 #|  Smaller scores are better (indicating smaller errors).
-#|  Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
+#|  Black bars indicate means. The points indicate all individual predictions over time and replicate 'true'
 #|  simulations of the given scenario."

### 2022-09-22 manuscript/manuscript.Rmd
Might've missed something

@@
 
-We have created a github repository <https://github.com/boettiger-lab/mee_tipping_point_forecasting.> that contains the code used to produce the figures herein.
+We have created a github repository <https://github.com/boettiger-lab/mee_tipping_point_forecasting> that contains the code used to produce the figures herein.
 

### 2022-09-22 manuscript/manuscript.Rmd
Making sure not forgetting anything here

@@
 We examine forecast skill for each of the six forecasting methods (MCMC, ARIMA, block-RNN, GRU, LSTM, and Transformer) in each of our three scenarios (Hopf bifurcation, saddle-node bifurcation, and stochastic transition).
-In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model. 
+In addition to these cases, we also consider an "ensemble model", generated by drawing from the distribution of all models except the MCMC model -- throughout our figures, this ensemble model is denoted by "ml_ensemble".
 Such ensemble techniques can better reflect uncertainty than relying on any single method [@gneiting_weather_2005].

### 2022-09-22 manuscript/manuscript.Rmd
More small edits

--- 2022-09-22 manuscript/manuscript.Rmd
@@
 #| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
-#|  Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "
+#|  Training data precedes the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "
 
@@
 #|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
-#|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning."
+#|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning.
+#|  The time axis in this plot and in Figure 6 refers to the time from the beginning of the forecast window, not the time from the beginning of the time series as in other plots."
 

### 2022-09-22 manuscript/manuscript.Rmd
Changing word choice in caption

@@
 #|  time when the forecast predicts the wrong equilibrium state and the lower branch indicates predictions of the correct one.
-The time axis in this plot and in Figure 7 refers to the time from the beginning of the forecast window, not the time from the beginning of the time series as in other plots."
+The time axis in this plot and in Figure 7 refers to the time from the beginning of the forecast horizon, not the time from the beginning of the time series as in other plots."
 
@@
 #|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning.
-#|  The time axis in this plot and in Figure 6 refers to the time from the beginning of the forecast window, not the time from the beginning of the time series as in other plots."
+#|  The time axis in this plot and in Figure 6 refers to the time from the beginning of the forecast horizon, not the time from the beginning of the time series as in other plots."
 

### 2022-09-22 manuscript/manuscript.Rmd

--- 2022-09-22 manuscript/manuscript.Rmd
@@
 #|  Generally, MCMC performs better relative to other models under this metric than it does under CRPS, 
-#|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning."
+#|  reflecting the bias-variance tradeoff taken when using biased estimators in machine learning.
+#|  The time axis in this plot and in Figure 6 refers to the time from the beginning of the forecast horizon, not the time from the beginning of the time series as in other plots."
 
